---
title: "Cell Cell Adjacencies in Banksy Niches"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

Here we will use a neighborhood adjacency analysis to look at the adjacency of cell types within niches. 


```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(dbscan)
library(circlize)
library(Seurat)
library(plotly)
library(scales)
library(htmlwidgets)
library(metap)

# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
```


# Preperations

## Loading the data

The objects are here:

RDS object with Leiden Clusters: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds`

    -   `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_embeddings/total_spatial_banksy_clusters_subset_20240719_092550.rds`


Cell type clusters annotation: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv`  


```{r}
# load the banksy embeddings
banksy_embeddings = readRDS(ifelse(
    workdir == '/xchip/beroukhimlab/',
    paste0(workdir,'coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds'),
    '/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/total_spatial_banksy_clusters_subset_20240719_092550.rds'
))
colData(banksy_embeddings)$sample_id = factor(colData(banksy_embeddings)$sample_id)

# Load the cell type clusters annotation
prelim_annotation_res1 = fread(paste0(
    workdir,'coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv'
))

# specify the cluster names 
# this is mainly to keep track of which resolution we are using
cell_res = 1
cell_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k50_res%.1f',cell_res))
niche_res = 1
niche_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.8_k50_res%.1f',niche_res))

# # filter the banksy object to only include the cells with annotated cell types
# annotated_cell_clusters = sort(unique(prelim_annotation_res1$cluster))
# banksy_embeddings = banksy_embeddings[,colData(banksy_embeddings)[,cell_clust_name] %in% annotated_cell_clusters]
```


## Build the color palette for all niches and cell types downstream

```{r}
set.seed(5555)
cluster_classes = levels(colData(banksy_embeddings)[[niche_clust_name]])
niche_color_manual = structure(
    pals::polychrome()[c(1:length(cluster_classes))],
    names = sample(cluster_classes)
)
# for niche resolution 0.5
# niches_to_show = c(1:11, 13, 14, 16, 17)
# for niche resolution 1
niches_to_show = c(1:20, 22:25)

cell_type_to_test = 'Coarse Cell Type'
cell_type_to_test = 'Fine Cell Type'
cell_type_to_test = 'Fine Cell Type UMAP'

cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))
cell_type_color_manual = structure(
    pals::alphabet()[c(1:length(cell_type_class))],
    names = cell_type_class
)
```


## Setting up Neighborhood Analysis

Here we will look at the niches in the data and annotate them using the cell types of self and neighbors.

```{r}
k_geom = 15
locs = spatialCoords(banksy_embeddings)
knn = dbscan::kNN(x = locs, k = k_geom)
knnDT <- data.table(
    from = rep(seq_len(nrow(knn$id)), k_geom),
    from_id = rep(rownames(knn$id), k_geom),
    to = as.vector(knn$id),
    to_id = rownames(knn$id)[as.vector(knn$id)],
    distance = as.vector(knn$dist)
)
cell_cell_type_dt = data.table(
    cell_id = rownames(colData(banksy_embeddings)),
    cell_type = as.numeric(colData(banksy_embeddings)[,cell_clust_name]),
    niche_type = as.factor(colData(banksy_embeddings)[,niche_clust_name]),
    sample_id = colData(banksy_embeddings)$sample_id,
    histology = colData(banksy_embeddings)$histology
)

```

# Cell Cell Adjacency Analysis

To do this, we compare each niche's cell cell adjacency to all other niches' cell cell adjacency. 

```{r}
adjacencies = knnDT[,.(from_id,to_id,distance)][
    cell_cell_type_dt[,.(cell_id,cell_type_to = cell_type)], on = c('to_id' = 'cell_id'),
][
    cell_cell_type_dt[,.(
        cell_id, niche_type_from = niche_type, 
        cell_type_from = cell_type, sample_id, histology
    )], on = c('from_id' = 'cell_id')
][
    ,total_count := .N, .(niche_type_from, sample_id, histology)
][
    # calculate niche specific cell cell adjacency proportions
    ,.(proportion = .N/total_count), .(
        niche_type_from, cell_type_from, cell_type_to, 
        sample_id, histology
    )
][
    # score each cell cell adjacency across niches using a two tailed p value
    ,c('p_value','z_score') := list(
        # important article about deriving empirical p values
        # https://www.ncbi.nlm.nih.gov/pmc/articles/PMC379178/
        ifelse(
            rank(proportion)/(.N+1) < 0.5, 2*rank(proportion)/(.N+1),
            2*(1 - rank(proportion)/(.N+1))
        ),
        (proportion - mean(proportion))/sd(proportion)
    ),.(cell_type_from, cell_type_to, sample_id, histology)
][
    ,.(
        combined_p_value = ifelse(
            sum((p_value > 0) & (p_value <= 1)) >= 2,
            sumlog(p_value)$p, as.numeric(p_value)
        ),
        mean_z_score = mean(z_score), proportion = mean(proportion)
    ), .(niche_type_from, cell_type_from, cell_type_to)
][, q_val := p.adjust(combined_p_value, 'BY')][
    prelim_annotation_res1[
        ,.(cluster, cell_type_test_from = get(cell_type_to_test))
    ], on = c('cell_type_from' = 'cluster')
][
    prelim_annotation_res1[
        ,.(cluster, cell_type_test_to = get(cell_type_to_test))
    ], on = c('cell_type_to' = 'cluster')
]
hist(adjacencies$combined_p_value, breaks = 100)
hist(adjacencies$q_val, breaks = 100)
```

```{r}
lapply(niches_to_show, function(niche){
    ggplot(adjacencies[
        niche_type_from == niche
    ]) +
        geom_point(aes(
            x = factor(
                cell_type_test_from, , levels = c(
                'GABAergic Neuron 1','GABAergic Neuron 2','GABAergic Neuron 3',
                'Glutamatergic Neuron 1','Glutamatergic Neuron 2','Glutamatergic Neuron 3',
                'Unknown Neuron 1','Unknown Neuron 2' ,'RGC 1','RGC 2','Proliferation 1',
                'OPC 1','Oligodendrocytes 1','Astrocyte 1','Astrocyte 2','Endothelial/VLMC 1',
                'Lymphoid 1','Myeloid 1','Myeloid 2', 'Undetermined 1'
            )
            ), 
            y = factor(
                cell_type_test_to, , levels = c(
                'GABAergic Neuron 1','GABAergic Neuron 2','GABAergic Neuron 3',
                'Glutamatergic Neuron 1','Glutamatergic Neuron 2','Glutamatergic Neuron 3',
                'Unknown Neuron 1','Unknown Neuron 2' ,'RGC 1','RGC 2','Proliferation 1',
                'OPC 1','Oligodendrocytes 1','Astrocyte 1','Astrocyte 2','Endothelial/VLMC 1',
                'Lymphoid 1','Myeloid 1','Myeloid 2', 'Undetermined 1'
            )
            ), 
            fill = pmin(
                quantile(mean_z_score, 0.90, na.rm = TRUE), 
                mean_z_score, na.rm = TRUE
            ), color = ifelse(q_val < 0.05, 'significant','not significant'),
            shape = ifelse(mean_z_score > 0, 'Positive','Negative')
        ), stroke = 1) +
        theme(
            text = element_text(size = 15), legend.position = 'right',
            axis.text.x = element_text(angle = 45, hjust = 1)
        ) + scale_fill_gradient2(
            low = 'blue', mid = 'white', high = 'red', midpoint = 0
        ) + scale_shape_manual(values = c('Positive' = 21,'Negative' = 22)) +
        # scale_size_area(max_size = 10) +
        scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
        labs(
            title = paste0('Cell Type Adjacencies in Niche ',niche),
            x = 'Cell Type From',
            y = 'Cell Type To',
            fill = 'Mean Z Score',
            color = 'Significance', 
            shape = 'Direction'
        ) 
})

# ggplot(adjacencies[
#     niche_type_from %in% niches_to_show
# ]) +
#     geom_point(aes(
#         x = cell_type_test_from, y = cell_type_test_to, #size = count, 
#         fill = pmin(
#             quantile(mean_z_score, 0.90, na.rm = TRUE), 
#             mean_z_score, na.rm = TRUE
#         ), color = ifelse(q_val < 0.05 & mean_z_score > 0, 'significant','not significant'),
#         shape = ifelse(mean_z_score > 0, 'Positive','Negative')
#     ), stroke = 1) +
#     facet_wrap(
#         ~niche_type_from, 
#         ncol = round(sqrt(length(niches_to_show)/2))
#     ) +
#     theme(
#         text = element_text(size = 15), legend.position = 'right',
#         axis.text.x = element_text(angle = 45, hjust = 1)
#     ) + scale_fill_gradient2(
#         low = 'blue', mid = 'white', high = 'red', midpoint = 0
#     ) + scale_shape_manual(values = c('Positive' = 21,'Negative' = 22)) +
#     # scale_size_area(max_size = 10) +
#     scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
#     labs(
#         title = 'Cell Type Adjacencies by Niches',
#         x = 'Cell Type From',
#         y = 'Cell Type To',
#         fill = 'Mean Z Score',
#         color = 'Significance', 
#         shape = 'Direction'
#     ) 
```

## Visualizing the adjacency of niches with network graphs

```{r}
#| fig-height: 12
#| fig-width: 12
library(igraph)

min_q_val = min(adjacencies[q_val != 0]$q_val, na.rm = TRUE)
lapply(sort(unique(adjacencies$niche_type_from)), function(niche){
    links = adjacencies[
        q_val <= 0.05 &  
            niche_type_from == niche & 
            mean_z_score >= 0, .(
            source = as.character(cell_type_test_from),
            target = as.character(cell_type_test_to),
            weight = ifelse(
                q_val == 0, -log10(min_q_val), -log10(q_val)
            )
        )
    ]
    g = graph_from_data_frame(links, directed = TRUE)
    # # identify community in graph
    # communities = cluster_walktrap(g, step = 2)
    # data.table(
    #     vertex = V(g)$name, 
    #     community = communities$membership
    # )[order(community)]
    set.seed(55555)
    plot(
        g, #vertex.color = communities$membership,
        edge.width = E(g)$weight/10, edge.arrow.size = 0.5, 
        vertex.size = 5, vertex.label.cex = 1,
        vertex.label.font=3,
        layout = layout_with_fr(g),
        main = paste0('Niche ',niche)
    )
})
# recip_links = reciprocity[
#     , c('source','target') := 
#         tstrsplit(niche_combo, ' - ')
# ][
#     (reciprocal_sig),.(source, target)
# ]
# g = graph_from_data_frame(recip_links, directed = FALSE)
# plot(g, edge.width = 1, edge.arrow.size = 0.5, vertex.size = 10, vertex.label.cex = 1.5)
```


# Cell Cell Adjacency Analysis by histology


```{r}
adjacencies = knnDT[,.(from_id,to_id,distance)][
    cell_cell_type_dt[,.(cell_id,cell_type_to = cell_type)], on = c('to_id' = 'cell_id'),
][
    cell_cell_type_dt[,.(
        cell_id, niche_type_from = niche_type, 
        cell_type_from = cell_type, sample_id, histology
    )], on = c('from_id' = 'cell_id')
][
    ,total_count := .N, .(niche_type_from, sample_id, histology)
][
    # calculate niche specific cell cell adjacency proportions
    ,.(proportion = .N/total_count), .(
        niche_type_from, cell_type_from, cell_type_to, 
        sample_id, histology
    )
][
    # score each cell cell adjacency across niches using a two tailed p value
    ,c('p_value','z_score') := list(
        # important article about deriving empirical p values
        # https://www.ncbi.nlm.nih.gov/pmc/articles/PMC379178/
        ifelse(
            rank(proportion)/(.N+1) < 0.5, 2*rank(proportion)/(.N+1),
            2*(1 - rank(proportion)/(.N+1))
        ),
        (proportion - mean(proportion))/sd(proportion)
    ),.(cell_type_from, cell_type_to, sample_id, histology)
][
    ,.(
        combined_p_value = ifelse(
            sum((p_value > 0) & (p_value <= 1)) >= 2,
            sumlog(p_value)$p, as.numeric(p_value)
        ),
        mean_z_score = mean(z_score), proportion = mean(proportion)
    ), .(niche_type_from, cell_type_from, cell_type_to, histology)
][, q_val := p.adjust(combined_p_value, 'BY')][
    prelim_annotation_res1[
        ,.(cluster, cell_type_test_from = get(cell_type_to_test))
    ], on = c('cell_type_from' = 'cluster')
][
    prelim_annotation_res1[
        ,.(cluster, cell_type_test_to = get(cell_type_to_test))
    ], on = c('cell_type_to' = 'cluster')
]
hist(adjacencies$combined_p_value, breaks = 100)
hist(adjacencies$q_val, breaks = 100)
```

```{r}
lapply(niches_to_show, function(niche){
    ggplot(adjacencies[
        niche_type_from == niche
    ]) +
        geom_point(aes(
            x = factor(
                cell_type_test_from, , levels = c(
                'GABAergic Neuron 1','GABAergic Neuron 2','GABAergic Neuron 3',
                'Glutamatergic Neuron 1','Glutamatergic Neuron 2','Glutamatergic Neuron 3',
                'Unknown Neuron 1','Unknown Neuron 2' ,'RGC 1','RGC 2','Proliferation 1',
                'OPC 1','Oligodendrocytes 1','Astrocyte 1','Astrocyte 2','Endothelial/VLMC 1',
                'Lymphoid 1','Myeloid 1','Myeloid 2', 'Undetermined 1'
            )
            ), 
            y = factor(
                cell_type_test_to, , levels = c(
                'GABAergic Neuron 1','GABAergic Neuron 2','GABAergic Neuron 3',
                'Glutamatergic Neuron 1','Glutamatergic Neuron 2','Glutamatergic Neuron 3',
                'Unknown Neuron 1','Unknown Neuron 2' ,'RGC 1','RGC 2','Proliferation 1',
                'OPC 1','Oligodendrocytes 1','Astrocyte 1','Astrocyte 2','Endothelial/VLMC 1',
                'Lymphoid 1','Myeloid 1','Myeloid 2', 'Undetermined 1'
            )
            ), 
            fill = pmin(
                quantile(mean_z_score, 0.90, na.rm = TRUE), 
                mean_z_score, na.rm = TRUE
            ), color = ifelse(q_val < 0.05, 'significant','not significant'),
            shape = ifelse(mean_z_score > 0, 'Positive','Negative')
        ), stroke = 1) +
        facet_wrap(
            ~histology, 
            ncol = 2
        ) +
        theme(
            text = element_text(size = 15), legend.position = 'right',
            axis.text.x = element_text(angle = 45, hjust = 1)
        ) + scale_fill_gradient2(
            low = 'blue', mid = 'white', high = 'red', midpoint = 0
        ) + scale_shape_manual(values = c('Positive' = 21,'Negative' = 22)) +
        # scale_size_area(max_size = 10) +
        scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
        labs(
            title = paste0('Cell Type Adjacencies in Niche ',niche),
            x = 'Cell Type From',
            y = 'Cell Type To',
            fill = 'Mean Z Score',
            color = 'Significance', 
            shape = 'Direction'
        ) 
})
```

## Visualizing the adjacency of niches with network graphs

```{r}
#| fig-height: 12
#| fig-width: 12
library(igraph)

niche_histo_combo = expand.grid(
    sort(unique(adjacencies$niche_type_from)), 
    unique(adjacencies$histology)
)
mapply(function(x, y){
    links = adjacencies[
        q_val <= 0.05 &  
            niche_type_from == x & 
            histology == y &
            mean_z_score >= 0, .(
            source = as.character(cell_type_test_from),
            target = as.character(cell_type_test_to),
            weight = ifelse(
                q_val == 0, -log10(min(q_val[q_val > 0])),
                -log10(q_val)
            )
        )
    ]
    g = graph_from_data_frame(links, directed = TRUE)
    # # identify community in graph
    # communities = cluster_walktrap(g, step = 2)
    # data.table(
    #     vertex = V(g)$name, 
    #     community = communities$membership
    # )[order(community)]
    set.seed(55555)
    plot(
        g, #vertex.color = communities$membership,
        edge.width = E(g)$weight/10, edge.arrow.size = 0.5, 
        vertex.size = 5, vertex.label.cex = 1,
        vertex.label.font=3,
        layout = layout_with_fr(g),
        main = paste0('Niche ',x,' in ', y)
    )
}, niche_histo_combo$Var1, niche_histo_combo$Var2)
# recip_links = reciprocity[
#     , c('source','target') := 
#         tstrsplit(niche_combo, ' - ')
# ][
#     (reciprocal_sig),.(source, target)
# ]
# g = graph_from_data_frame(recip_links, directed = FALSE)
# plot(g, edge.width = 1, edge.arrow.size = 0.5, vertex.size = 10, vertex.label.cex = 1.5)
```
