---
title: "BANKSY Niches in Space"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

This doc is solely dedicated to visualization of niches in space

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(dbscan)
library(circlize)
library(Seurat)
library(plotly)
library(scales)
library(htmlwidgets)
library(metap)

# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
```


# Load the data

The objects are here:

RDS object with Leiden Clusters: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds`

    -   `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_embeddings/total_spatial_banksy_clusters_subset_20240719_092550.rds`


Cell type clusters annotation: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv`  


```{r}
# load the banksy embeddings
banksy_embeddings = readRDS(ifelse(
    workdir == '/xchip/beroukhimlab/',
    paste0(workdir,'coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds'),
    '/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/total_spatial_banksy_clusters_subset_20240719_092550.rds'
))
colData(banksy_embeddings)$sample_id = factor(colData(banksy_embeddings)$sample_id)

# Load the cell type clusters annotation
prelim_annotation_res1 = fread(paste0(
    workdir,'coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv'
))

# specify the cluster names 
# this is mainly to keep track of which resolution we are using
cell_res = 1
cell_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k50_res%.1f',cell_res))
niche_res = 1
niche_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.8_k50_res%.1f',niche_res))

# # filter the banksy object to only include the cells with annotated cell types
# annotated_cell_clusters = sort(unique(prelim_annotation_res1$cluster))
# banksy_embeddings = banksy_embeddings[,colData(banksy_embeddings)[,cell_clust_name] %in% annotated_cell_clusters]
```


Build the color palette for all niches and cell types downstream

```{r}
set.seed(5555)
cluster_classes = levels(colData(banksy_embeddings)[[niche_clust_name]])
niche_color_manual = structure(
    pals::polychrome()[c(1:length(cluster_classes))],
    names = sample(cluster_classes)
)
# for niche resolution 0.5
niches_to_show = c(1:11, 13, 14, 16, 17)
# for niche resolution 1
# niches_to_show = c(1:20, 22:25)

cell_type_to_test = 'Coarse Cell Type'
cell_type_to_test = 'Fine Cell Type'
cell_type_to_test = 'Fine Cell Type UMAP'

cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))
cell_type_color_manual = structure(
    pals::alphabet()[c(1:length(cell_type_class))],
    names = cell_type_class
)

```


# Looking at the niches in space all together

```{r}
#| fig-height: 12
#| fig-width: 24
graph_dt = melt(
    as.data.table(cbind(
        spatialCoords(banksy_embeddings),
        colData(banksy_embeddings)[c(
            'sample_id','cell_id', 'histology',
            grep('Harmony.*.res',clusterNames(banksy_embeddings), value = TRUE)
        )]
    )),id.vars = c('sample_id','cell_id','histology','sdimx','sdimy'),
    variable.name = 'Cluster Type',value.name = 'Cluster ID'
)[,`Cluster ID` := as.factor(as.numeric(`Cluster ID`))]

g = ggplot(
    graph_dt[`Cluster Type` == niche_clust_name], 
    aes(x=sdimy, y=sdimx, color=`Cluster ID`)
) +
    facet_wrap(histology ~ sample_id, nrow = 2, scale = 'free_x') +
    # subset cells
    # geom_point(size = 0.3, alpha = 0.5) + 
    # all cells
    geom_point(size = 0.00001) + 
    scale_color_manual(values = niche_color_manual) +
    theme_classic() + 
    theme(
        legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        text = element_text(size = 30),
        strip.text = element_text(size = 20),
        strip.background = element_blank()
    ) + guides(
        color = guide_legend(override.aes = list(size=5, alpha = 1), nrow = 3, byrow = TRUE)
    ) + labs(
        title = 'BANKSY Niches in Space',
        x = '',y = ''
    ) +
    coord_flip()
g
# p = ggplotly(g)
# saveWidget(p, paste0(workdir,'/youyun/plgg/code/niche/niche_identity_spatial.html'), selfcontained = TRUE)
```


# Looking at the niches in space separately

```{r}
#| fig-height: 24
#| fig-width: 24
x = unique(graph_dt$sample_id)[1]
lapply(unique(graph_dt$sample_id), function(x){
    ggplot(
        graph_dt[`Cluster Type` == niche_clust_name & sample_id == x], 
        aes(x=sdimy, y=sdimx, color=`Cluster ID`)
    ) +
        facet_wrap(
            histology ~ sample_id + `Cluster ID`, labeller = label_wrap_gen(), 
            ncol = 6, scale = 'free_x'
        ) +
        # subset cells
        # geom_point(size = 0.3, alpha = 0.5) + 
        # all cells
        geom_point(size = 0.00001) + 
        scale_color_manual(values = niche_color_manual) +
        theme_classic() + 
        theme(
            legend.position = "bottom",
            axis.text.x=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks=element_blank(),
            text = element_text(size = 30),
            strip.text = element_text(size = 20),
            strip.background = element_blank()
        ) + guides(
            color = guide_legend(override.aes = list(size=5, alpha = 1), nrow = 3, byrow = TRUE)
        ) + labs(
            title = paste0('BANKSY Niches in Space: ',x),
            x = '',y = ''
        ) +
        coord_flip()
})
```

```{r}
#| fig-height: 12
#| fig-width: 24
x = unique(graph_dt[`Cluster Type` == niche_clust_name]$`Cluster ID`)[1]
lapply(unique(graph_dt[`Cluster Type` == niche_clust_name]$`Cluster ID`), function(x){
    ggplot(
        graph_dt[`Cluster Type` == niche_clust_name & `Cluster ID` == x], 
        aes(x=sdimy, y=sdimx, color=`Cluster ID`)
    ) +
        facet_wrap(
            histology ~ sample_id, 
            nrow = 2, scale = 'free_x'
        ) +
        # subset cells
        # geom_point(size = 0.3, alpha = 0.5) + 
        # all cells
        geom_point(size = 0.00001) + 
        scale_color_manual(values = niche_color_manual) +
        theme_classic() + 
        theme(
            legend.position = "bottom",
            axis.text.x=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks=element_blank(),
            text = element_text(size = 30),
            strip.text = element_text(size = 20),
            strip.background = element_blank()
        ) + guides(
            color = guide_legend(override.aes = list(size=5, alpha = 1), nrow = 3, byrow = TRUE)
        ) + labs(
            title = paste0('BANKSY Niches in Space: ',x),
            x = '',y = ''
        ) +
        coord_flip()
})
```

