---
title: "Niche Niche Adjacencies in Banksy Niches"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

In this doc, we will be look at the niche niche adjacencies patterns to potentially identify tissue domains within our cohort

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(dbscan)
library(circlize)
library(Seurat)
library(plotly)
library(scales)
library(htmlwidgets)
library(metap)

# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
```


# Preperations

## Loading the data

The objects are here:

RDS object with Leiden Clusters: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds`

    -   `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_embeddings/total_spatial_banksy_clusters_subset_20240719_092550.rds`


Cell type clusters annotation: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv`  


```{r}
# load the banksy embeddings
banksy_embeddings = readRDS(ifelse(
    workdir == '/xchip/beroukhimlab/',
    paste0(workdir,'coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds'),
    '/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/total_spatial_banksy_clusters_subset_20240719_092550.rds'
))
colData(banksy_embeddings)$sample_id = factor(colData(banksy_embeddings)$sample_id)

# Load the cell type clusters annotation
prelim_annotation_res1 = fread(paste0(
    workdir,'coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv'
))

# specify the cluster names 
# this is mainly to keep track of which resolution we are using
cell_res = 1
cell_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k50_res%.1f',cell_res))
niche_res = 1
niche_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.8_k50_res%.1f',niche_res))

# # filter the banksy object to only include the cells with annotated cell types
# annotated_cell_clusters = sort(unique(prelim_annotation_res1$cluster))
# banksy_embeddings = banksy_embeddings[,colData(banksy_embeddings)[,cell_clust_name] %in% annotated_cell_clusters]
```


## Build the color palette for all niches and cell types downstream

```{r}
set.seed(5555)
cluster_classes = levels(colData(banksy_embeddings)[[niche_clust_name]])
niche_color_manual = structure(
    pals::polychrome()[c(1:length(cluster_classes))],
    names = sample(cluster_classes)
)
# for niche resolution 0.5
# niches_to_show = c(1:11, 13, 14, 16, 17)
# for niche resolution 1
niches_to_show = c(1:20, 22:25)

cell_type_to_test = 'Coarse Cell Type'
cell_type_to_test = 'Fine Cell Type'
cell_type_to_test = 'Fine Cell Type UMAP'

cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))
cell_type_color_manual = structure(
    pals::alphabet()[c(1:length(cell_type_class))],
    names = cell_type_class
)
```


## Setting up Neighborhood Analysis

Here we will look at the niches in the data and annotate them using the cell types of self and neighbors.

```{r}
k_geom = 15
locs = spatialCoords(banksy_embeddings)
knn = dbscan::kNN(x = locs, k = k_geom)
knnDT <- data.table(
    from = rep(seq_len(nrow(knn$id)), k_geom),
    from_id = rep(rownames(knn$id), k_geom),
    to = as.vector(knn$id),
    to_id = rownames(knn$id)[as.vector(knn$id)],
    distance = as.vector(knn$dist)
)
cell_cell_type_dt = data.table(
    cell_id = rownames(colData(banksy_embeddings)),
    cell_type = as.numeric(colData(banksy_embeddings)[,cell_clust_name]),
    niche_type = as.factor(colData(banksy_embeddings)[,niche_clust_name]),
    sample_id = colData(banksy_embeddings)$sample_id,
    histology = colData(banksy_embeddings)$histology
)

cell_type_cols = setdiff(colnames(prelim_annotation_res1),'cluster')
niche_self_neighbor_dt = knnDT[,c('from','to') := NULL][
    cell_cell_type_dt[,.(cell_id,cell_type)], on = c('to_id' = 'cell_id'),
    cell_type_to := cell_type
][cell_cell_type_dt, on = c('from_id' = 'cell_id')][
    ,.(
    cell_id = from_id, cell_type, niche_type, sample_id, 
    histology, neighbor_cell_type = cell_type_to
)][
    prelim_annotation_res1, on = c('cell_type' = 'cluster')
][
    prelim_annotation_res1, on = c('neighbor_cell_type' = 'cluster'),
    paste0(cell_type_cols,' Neighbors') := mget(paste0('i.',cell_type_cols))
][order(cell_id)]
niche_self_dt = unique(niche_self_neighbor_dt[
    ,mget(c('cell_id','cell_type','niche_type','sample_id','histology',cell_type_cols))
])
```

# Niche Niche Adjacency Analysis

This is done on the entire cohort together

```{r}
observed_adjacency = knnDT[,.(from_id,to_id,distance)][
    cell_cell_type_dt[,.(cell_id,niche_type_to = niche_type)], on = c('to_id' = 'cell_id'),
][
    cell_cell_type_dt[,.(
        cell_id, niche_type_from = niche_type, sample_id, histology
    )], on = c('from_id' = 'cell_id')
][
    ,.(count = .N), .(niche_type_from, niche_type_to, sample_id)
][, run := 'observed']

set.seed(55555)
permutation_num = 200
permuted_adjacencies = do.call('rbind',lapply(c(1:permutation_num), function(x){
    # for each round of permutation, we permute the niche type labels within a sample
    permuted_cell_type_dt = cell_cell_type_dt[,.(
        cell_id,niche_type = sample(niche_type)
    ), .(sample_id, histology)]
    knnDT[,.(from_id,to_id,distance)][
        permuted_cell_type_dt[,.(cell_id, niche_type_to = niche_type)], 
        on = c('to_id' = 'cell_id'),
    ][
        # for the cells themselves, we do not permute the niche type labels
        permuted_cell_type_dt[,.(
            cell_id, niche_type_from = niche_type, sample_id, histology
        )], on = c('from_id' = 'cell_id')
    ][
        ,.(count = .N), .(niche_type_from, niche_type_to, sample_id)
    ][, run := paste0('perm_',x)]
}))

adjacencies = rbind(observed_adjacency, permuted_adjacencies)[
    niche_type_from %in% niches_to_show & niche_type_to %in% niches_to_show
][
    # to each sample, we calculate the mean and sd of the count of the permuted adjacencies
    # and the z score of the observed adjacency
    ,.(
        mean_count = mean(count[run != 'observed']),
        sd_count = sd(count[run != 'observed']),
        observed_count = count[run == 'observed'],
        z_score = (
            count[run == 'observed'] - 
                mean(count[run != 'observed'])
        )/sd(count[run != 'observed'])
    ),
    .(niche_type_from, niche_type_to, sample_id)
][
    ,.(
        # get the two tailed p values from the z scores
        p_value = 2*pnorm(-abs(z_score), lower.tail = TRUE),
        z_score
    ),.(niche_type_from, niche_type_to, sample_id)
][
    !is.na(p_value),.(
        comb_p_value = ifelse(
            sum((p_value > 0) & (p_value <= 1)) >= 2,
            sumlog(p_value)$p, as.numeric(p_value)
        ), mean_z_score = mean(z_score)
    ), .(niche_type_from, niche_type_to)
][, q_val := p.adjust(comb_p_value, method = 'BY')][
    observed_adjacency[,.(count = sum(count)), .(niche_type_from, niche_type_to)], 
    on = c('niche_type_from','niche_type_to')
]

ggplot(adjacencies[
    niche_type_from %in% niches_to_show & niche_type_to %in% niches_to_show
]) +
    geom_point(aes(
        x = niche_type_from, y = niche_type_to, #size = count, 
        fill = pmin(
            quantile(mean_z_score, 0.90, na.rm = TRUE), 
            mean_z_score, na.rm = TRUE
        ), color = ifelse(q_val < 0.05, 'significant','not significant'),
        shape = ifelse(mean_z_score > 0, 'Positive','Negative')
    ), stroke = 1, size = 10) +
    theme(
        text = element_text(size = 15), legend.position = 'right'
    ) + scale_fill_gradient2(
        low = 'blue', mid = 'white', high = 'red', midpoint = 0
    ) + scale_shape_manual(values = c('Positive' = 21,'Negative' = 22)) +
    # scale_size_area(max_size = 10) +
    scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
    labs(
        title = 'Adjacency of Niches',
        x = 'Niche Type From',
        y = 'Niche Type To',
        size = 'Count',
        fill = 'Mean Z Score',
        color = 'Significance', 
        shape = 'Direction'
    ) 

ggplot(adjacencies[
    niche_type_from %in% niches_to_show & niche_type_to %in% niches_to_show
]) +
    geom_point(aes(
        x = niche_type_from, y = niche_type_to, size = count, 
        fill = pmin(
            quantile(mean_z_score, 0.90, na.rm = TRUE), 
            mean_z_score, na.rm = TRUE
        ), color = ifelse(q_val < 0.05, 'significant','not significant'),
        shape = ifelse(mean_z_score > 0, 'Positive','Negative')
    ), stroke = 1, size = 10) +
    theme(
        text = element_text(size = 15), legend.position = 'right'
    ) + scale_fill_gradient2(
        low = 'blue', mid = 'white', high = 'red', midpoint = 0
    ) + scale_shape_manual(values = c('Positive' = 21,'Negative' = 22)) +
    scale_size_area(max_size = 10) +
    scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
    labs(
        title = 'Adjacency of Niches',
        x = 'Niche Type From',
        y = 'Niche Type To',
        size = 'Count',
        fill = 'Mean Z Score',
        color = 'Significance', 
        shape = 'Direction'
    ) 
```

## Visualizing the adjacency of niches with network graphs

```{r}
library(igraph)

links = adjacencies[
    q_val <= 0.05 &  
        niche_type_from %in% niches_to_show & 
        niche_type_to %in% niches_to_show &
        niche_type_from != niche_type_to & 
        mean_z_score >= 0, .(
        source = as.character(niche_type_from),
        target = as.character(niche_type_to),
        weight = pmin(
            quantile(-log10(q_val), 0.95, na.rm = TRUE), 
            -log10(q_val), na.rm = TRUE
        )
    )
]
g = graph_from_data_frame(links, directed = TRUE)
# identify community in graph
communities = cluster_walktrap(g, step = 2)
data.table(
    vertex = V(g)$name, 
    community = communities$membership
)[order(community)]
set.seed(55555)
plot(
    g, vertex.color = communities$membership,
    edge.width = E(g)$weight/10, edge.arrow.size = 0.5, vertex.size = 10, vertex.label.cex = 1.5
)

# recip_links = reciprocity[
#     , c('source','target') := 
#         tstrsplit(niche_combo, ' - ')
# ][
#     (reciprocal_sig),.(source, target)
# ]
# g = graph_from_data_frame(recip_links, directed = FALSE)
# plot(g, edge.width = 1, edge.arrow.size = 0.5, vertex.size = 10, vertex.label.cex = 1.5)
```

# Niche Niche Adjacencies within histology

```{r}
observed_adjacency = knnDT[,.(from_id,to_id,distance)][
    cell_cell_type_dt[,.(cell_id,niche_type_to = niche_type)], on = c('to_id' = 'cell_id'),
][
    cell_cell_type_dt[,.(
        cell_id, niche_type_from = niche_type, sample_id, histology
    )], on = c('from_id' = 'cell_id')
][
    ,.(count = .N), .(niche_type_from, niche_type_to, sample_id, histology)
][, run := 'observed']

set.seed(55555)
permutation_num = 10
permuted_adjacencies = do.call('rbind',lapply(c(1:permutation_num), function(x){
    # for each round of permutation, we permute the niche type labels within a sample
    permuted_cell_type_dt = cell_cell_type_dt[,.(
        cell_id,niche_type = sample(niche_type)
    ), .(sample_id, histology)]
    knnDT[,.(from_id,to_id,distance)][
        permuted_cell_type_dt[,.(cell_id, niche_type_to = niche_type)], 
        on = c('to_id' = 'cell_id'),
    ][
        # for the cells themselves, we do not permute the niche type labels
        permuted_cell_type_dt[,.(
            cell_id, niche_type_from = niche_type, sample_id, histology
        )], on = c('from_id' = 'cell_id')
    ][
        ,.(count = .N), .(niche_type_from, niche_type_to, sample_id, histology)
    ][, run := paste0('perm_',x)]
}))

adjacencies = rbind(observed_adjacency, permuted_adjacencies)[
    niche_type_from %in% niches_to_show & niche_type_to %in% niches_to_show
][
    # to each sample, we calculate the mean and sd of the count of the permuted adjacencies
    # and the z score of the observed adjacency
    ,.(
        mean_count = mean(count[run != 'observed']),
        sd_count = sd(count[run != 'observed']),
        observed_count = count[run == 'observed'],
        z_score = (
            count[run == 'observed'] - 
                mean(count[run != 'observed'])
        )/sd(count[run != 'observed'])
    ),
    .(niche_type_from, niche_type_to, sample_id, histology)
][
    ,.(
        # get the two tailed p values from the z scores
        p_value = 2*pnorm(-abs(z_score), lower.tail = TRUE),
        z_score
    ),.(niche_type_from, niche_type_to, sample_id, histology)
][
    !is.na(p_value),.(
        comb_p_value = ifelse(
            sum((p_value > 0) & (p_value <= 1)) >= 2,
            sumlog(p_value)$p, as.numeric(p_value)
        ), mean_z_score = mean(z_score)
    ), .(niche_type_from, niche_type_to, histology)
][, q_val := p.adjust(comb_p_value, method = 'BY')][
    observed_adjacency[,.(count = sum(count)), .(niche_type_from, niche_type_to, histology)], 
    on = c('niche_type_from','niche_type_to','histology')
]

ggplot(adjacencies[
    niche_type_from %in% niches_to_show & niche_type_to %in% niches_to_show
]) +
    geom_point(aes(
        x = niche_type_from, y = niche_type_to, #size = count, 
        fill = pmin(
            quantile(mean_z_score, 0.90, na.rm = TRUE), 
            mean_z_score, na.rm = TRUE
        ), color = ifelse(q_val < 0.05, 'significant','not significant'),
        shape = ifelse(mean_z_score > 0, 'Positive','Negative')
    ), stroke = 1, size = 5) +
    facet_wrap(~histology, nrow = 2) +
    theme(
        text = element_text(size = 15), legend.position = 'right'
    ) + scale_fill_gradient2(
        low = 'blue', mid = 'white', high = 'red', midpoint = 0
    ) + scale_shape_manual(values = c('Positive' = 21,'Negative' = 22)) +
    # scale_size_area(max_size = 10) +
    scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
    labs(
        title = 'Adjacency of Niches',
        x = 'Niche Type From',
        y = 'Niche Type To',
        size = 'Count',
        fill = 'Mean Z Score',
        color = 'Significance', 
        shape = 'Direction'
    ) 
```

## Visualizing the adjacency of niches with network graphs

```{r}
library(igraph)

x = 'central neurocytoma'
lapply(unique(adjacencies$histology), function(x){
    print(x)
    links = adjacencies[
        q_val <= 0.05 &  
            histology == x &
            niche_type_from %in% niches_to_show & 
            niche_type_to %in% niches_to_show &
            niche_type_from != niche_type_to & 
            mean_z_score >= 0, .(
            source = as.character(niche_type_from),
            target = as.character(niche_type_to),
            weight = ifelse(
                q_val == 0, -log10(min(q_val[q_val > 0])),
                -log10(q_val)
            )
        )
    ]
    g = graph_from_data_frame(links, directed = TRUE)
    # identify community in graph
    communities = cluster_walktrap(g, step = 2)
    data.table(
        vertex = V(g)$name, 
        community = communities$membership
    )[order(community)]
    set.seed(55555)
    plot(
        g, vertex.color = communities$membership,
        edge.width = E(g)$weight/10, edge.arrow.size = 0.5, 
        vertex.size = 10, vertex.label.cex = 1.5,
        vertex.label.font=3,
        layout = layout_with_fr(g),
        main = paste0('Niche Org. In  ',x)
    )
})

# recip_links = reciprocity[
#     , c('source','target') := 
#         tstrsplit(niche_combo, ' - ')
# ][
#     (reciprocal_sig),.(source, target)
# ]
# g = graph_from_data_frame(recip_links, directed = FALSE)
# plot(g, edge.width = 1, edge.arrow.size = 0.5, vertex.size = 10, vertex.label.cex = 1.5)
```
