---
title: "Finalizing the Gene Set"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 6
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---


```{r}
#| echo: FALSE
library(data.table)
library(ggplot2)
# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}
```

# Introduction

We have generated the final manually curated gene set here: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/markers/final_manual_markers.csv`

We have also generated the PERSIST gene set here: `/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/PERSIST_results_00062024.csv`

Finally, the Xenium Brain panel is here: `/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/Xenium_hBrain_v1_metadata.csv` and 5K panel is here: `/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/XeniumPrimeHuman5Kpan_tissue_pathways_metadata.csv`

# Load the gene sets

```{r}
manual_markers <- fread(paste0(workdir, "coja/Spatial_PLGG/data/markers/final_manual_markers.csv"))
persist_markers <- fread(paste0(workdir, "youyun/plgg/data/xenium_selection/PERSIST_results_00062024.csv"))
xenium_brain <- fread(paste0(workdir, "youyun/plgg/data/xenium_selection/Xenium_hBrain_v1_metadata.csv"))
xenium_5k <- fread(paste0(workdir, "youyun/plgg/data/xenium_selection/XeniumPrimeHuman5Kpan_tissue_pathways_metadata.csv"))
```

# Final Gene Set

```{r}
manual_marker_genes <- manual_markers$marker
persist_marker_genes <- persist_markers[(persist_set_46)]$name
xenium_brain_genes <- xenium_brain$Genes
final_gene_set <- unique(c(manual_marker_genes, persist_marker_genes, xenium_brain_genes))
add_on_geneset <- setdiff(final_gene_set, xenium_brain_genes)
writeLines(add_on_geneset, con = paste0(workdir, "youyun/plgg/data/xenium_selection/add_on_geneset.txt"))
```

# Overlap w the 5k set

```{r}
xenium_5k_genes <- xenium_5k$gene_name
# make a upset plot between xenium 5k, xenium brain, persist markers genes and manual marker genes
# use upsetr
library(UpSetR)
upsetr_set_dt = dcast(
    data.table(
        genes = c(xenium_5k_genes, xenium_brain_genes, persist_marker_genes, manual_marker_genes),
        set = c(
            rep("Xenium_5K", length(xenium_5k_genes)), 
            rep("Xenium_Brain", length(xenium_brain_genes)), 
            rep("PERSIST", length(persist_marker_genes)), 
            rep("Manual", length(manual_marker_genes))
        )
    ), genes ~ set, fun.aggregate = length
)
upsetr_set_dt[upsetr_set_dt == 2] <- 1
upset(
    upsetr_set_dt[,.(Xenium_5K, Xenium_Brain, PERSIST, Manual)],
    sets.bar.color = "#56B4E9", empty.intersections = TRUE,
    nsets = 4
)
```