---
title: "Manual Curation of Markers"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 6
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

We are trying to handpick some cell type and pathway markers for the PLGG Spatial Project. The goal is to have a list of markers that are specific to the cell types and pathways that we are interested in based on prior knowledge.

1. Xenium Panel
    
    `/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/Xenium_hBrain_v1_metadata.csv`
    
2. Cell type markers
    
    `/xchip/beroukhimlab/coja/Spatial_PLGG/data/markers/curated_markers.xlsx`
    
    1. Which of them already ovelap with the Xenium panel already? 

    2. Which of them are good to include? (2-3 per cell type, 10 cell type total)

3. pathway markers
    
    `/xchip/beroukhimlab/coja/Spatial_PLGG/data/markers/pathway_markers.csv`
    
    1. Which of them already overlap with the Xenium panel already?

    2. Which of them are good to include? (10 per pathway, 2 pathway total)

        Highly variable gene annotated here: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/markers/pathway_markers_vargene.csv`


```{r}
#| echo: FALSE
library(data.table)
library(ggplot2)
library(readxl)
library(knitr)
# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}
```

# Load Data

```{r}
# Load Xenium Panel
xenium_panel <- fread(paste0(workdir, "youyun/plgg/data/xenium_selection/Xenium_hBrain_v1_metadata.csv"))
# Load Cell Type Markers
cell_type_markers <- melt(data.table(
    read_xlsx(paste0(workdir, "coja/Spatial_PLGG/data/markers/curated_markers.xlsx"))
)[,id := c(1:.N)], id.vars = "id", variable.name = "cell_type", value.name = "marker")[
    !is.na(marker) & !cell_type %in% c('MAPK','Senmayo')
]
cell_type_markers[, id := NULL]
# Load Pathway Markers
pathway_markers <- fread(paste0(workdir, "coja/Spatial_PLGG/data/markers/pathway_markers.csv"))
# Load Highly Variable Genes
highly_variable_genes <- fread(paste0(workdir, "coja/Spatial_PLGG/data/markers/pathway_markers_vargene.csv"))
```

# Cell Type Markers

## Overlap with Xenium Panel

```{r}
#| results: hold
cell_type_markers[, xenium := marker %in% xenium_panel$Genes]
ggplot(cell_type_markers) +
    geom_bar(aes(x = cell_type, fill = xenium), position = "stack") +
    # add horizontal line at 3
    geom_hline(yintercept = 3, linetype = "dashed", color = "white") +
    labs(title = "Overlap of Cell Type Markers with Xenium Panel", x = "Overlap with Xenium Panel", y = "Count") +
    theme_minimal() + theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 15)
    ) + scale_fill_brewer(palette = "Set1")
kable(
    cell_type_markers[
        , .(
            `Markers included` = paste0(marker[xenium], collapse = ", "),
            `Markers not included` = paste0(marker[!xenium], collapse = ", "),
            `Need more` = 3 - sum(xenium)
        ), .(`Cell Type` = cell_type)
    ]
)
```

## Expert Curation of Cell Type Markers

Some were pick based on [Michelle's heatmap](https://www.notion.so/peteryzheng/Pan-PLGG-PERSIST-list-b0cc630a23f2406494fa2d0efdbc3a92?pvs=4#0dbf7f0e497248b79db0c09a45668a1d) showing some top cell type specific markers in the PLGG single cell data. 

1.  Stromal: PCOLCE (Michelle heatmap)
2.  OPC: no more needed
3.  Progenitor Glia: hard to pick
4.  Oligodendrocyte: PLP1 (Michelle heatmap)
5.  Endothelial: ESAM, VWF (Michelle heatmap)
6.  Neuron (mature neuronal body): RBFOX3 (Michelle heatmap), ENO2 (putative neuron markers), BDNF (from [Monje Paper](https://www.nature.com/articles/s41586-023-06678-1)), 
    -   glutamatergic (excitatory) and GABAergic (inhibitory) neurons already included in xenium panel mainly based off of [this paper](https://www.nature.com/articles/s41586-018-0654-5)
    -   [processes](https://www.biocompare.com/Editorial-Articles/589675-A-Guide-to-Neuronal-Cell-Markers/): TUBB3, NEFL, MAPT  
    -   [dendrites](https://www.biocompare.com/Editorial-Articles/589675-A-Guide-to-Neuronal-Cell-Markers/): MAP2
    -   [intermediate progenitor cell](https://www.biocompare.com/Editorial-Articles/592536-A-Guide-to-Neural-Stem-Cell-Markers/): EOMES (Tbr2) CUX2 (already included), EMX2, FEZF2, BCL11B (CTIP2) 
    -   [neurogenesis markers](https://www.biocompare.com/Editorial-Articles/589675-A-Guide-to-Neuronal-Cell-Markers/): GFAP, Nestin, SOX2 (all already included)
7.  Microglia: CD74 (Michelle heatmap), TMEM119 (putative microglia marker)
8.  Astrocyte: 	GFAP, CLU (Michelle heatmap)
9.  T cell: IL32 (Michelle heatmap), CD3D (putative T cell marker)
10. RGC: none
11. Macrophage: CD14 (putative macrophage marker already included in Xenium), other markers are in M1 and M2

```{r}
added_probes = data.table(
    marker = c(
        'PCOLCE', 'PLP1',
        'ESAM', 'VWF', 'RBFOX3','ENO2', 'BDNF',
        'TUBB3', 'NEFL', 'MAPT', 'MAP2',
        'EOMES', 'EMX2', 'FEZF2', 'BCL11B',
        'CD74', 'TMEM119', 'GFAP', 'CLU',
        'IL32', 'CD3D'
    ),
    annotation = c(
        'Stromal', 'Oligodendrocyte', 
        rep('Endothelial', 2), rep('Neuron', 11),
        rep('Microglia', 2), rep('Astrocyte', 2),
        rep('T_cell', 2)
    ),
    category = 'Cell Type',
    xenium = FALSE
)
if(any(added_probes$marker %in% xenium_panel$Genes)) {
    print(add_probes[marker %in% xenium_panel$Genes])
}else{
    print('All manually added markers are not in Xenium panel')
}
final_cell_type_markers = rbind(
    # markers already in xenium panel
    cell_type_markers[(xenium), .(marker, annotation = cell_type, category = 'Cell Type', xenium)], 
    # manually added markers
    added_probes
)
print(paste0('Adding ', nrow(added_probes), ' manually added markers'))
kable(
    final_cell_type_markers[
        ,.(
            `Markers Included` = paste0(marker[xenium], collapse = ", "),
            `Markers To Add` = paste0(marker[!xenium], collapse = ", ")
        ), 
        .(`Cell Type` = annotation)
    ],
    caption = "Final Cell Type Markers"
)
```

# Pathway Markers

## Overlap with Xenium Panel

```{r}
#| results: hold
highly_variable_genes$cohort_var_gene = ifelse(is.na(highly_variable_genes$cohort_var_gene), 0, highly_variable_genes$cohort_var_gene)
highly_variable_genes[, xenium := Gene %in% xenium_panel$Genes]
ggplot(highly_variable_genes) +
    geom_bar(aes(x = Pathway, fill = xenium), position = "stack") +
    # add horizontal line at 10
    geom_hline(yintercept = 10, linetype = "dashed", color = "white") +
    labs(title = "Overlap of Pathway Markers with Xenium Panel", x = "Overlap with Xenium Panel", y = "Count") +
    theme_minimal() + theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 15)
    ) + scale_fill_brewer(palette = "Set1")
kable(
    highly_variable_genes[
        , .(
            `Markers included` = paste0(Gene[xenium], collapse = ", "),
            `Need more` = 10 - sum(xenium),
            `Markers (var in 3)` = paste0(Gene[!xenium & cohort_var_gene == 3], collapse = ", "),
            `Markers (var in 2)` = paste0(Gene[!xenium & cohort_var_gene == 2], collapse = ", "),
            `Markers (var in 1)` = paste0(Gene[!xenium & cohort_var_gene == 1], collapse = ", "),
            `Markers (var in 0)` = paste0(Gene[!xenium & (
                cohort_var_gene == 0 | is.na(cohort_var_gene)
            )], collapse = ", ")
        ), .(Pathway)
    ]
)
```

## Expert Curation of Pathway Markers

1. senescent markers: CDKN1A, CDKN2A, CDKN2B, GLB1, B2M, CGAS, ARG2

2. MAPK [MSS Overlap gene set](https://www.nature.com/articles/s41467-023-40235-8#Fig1): GYPC, GPNMB, ETV5, TRIB2, EPB41L3, ST6GAL1, SPRY2, APOD, ETV4, SERPINA3 , ANKH, TRIB1, FCERIG, C3AR1 

3. M1: IL1B, IL6, TNFalpha, CD11b (ITGAM), CD16 (FCGR3A), CD32 (FCGR2A), CD86

    - ITGAM, FCGR3A, CD86 are already in xenium panel

4. M2: IL4, IL10, MRC1, Arg-1 (ARG1), CHI3L1, STAT6, CCL22, PPARG

```{r}
added_probes_pathway = data.table(
    marker = c(
        'CDKN1A', 'CDKN2A', 'CDKN2B', 'GLB1', 'B2M', 'CGAS', 'ARG2',
        'GYPC', 'GPNMB', 'ETV5', 'TRIB2', 'EPB41L3', 'ST6GAL1', 'SPRY2', 'APOD', 'ETV4', 'SERPINA3', 'ANKH', 'TRIB1', 'FCERIG', 'C3AR1',
        'IL1B', 'IL6', 'TNF', 'FCGR2A', 
        'IL4', 'IL10', 'MRC1', 'ARG1', 'CHI3L1', 'STAT6', 'CCL22', 'PPARG'
    ),
    annotation = c(
        rep('senescence', 7),
        rep('MAPK', 14),
        rep('M1', 4),
        rep('M2', 8)
    ),
    category = 'Pathway'
)[
    ,xenium := ifelse(marker %in% xenium_panel$Genes, TRUE, FALSE)
]

if(any(added_probes_pathway$marker %in% xenium_panel$Genes)) {
    print(added_probes_pathway[marker %in% xenium_panel$Genes])
}else{
    print('All manually added markers are not in Xenium panel')
}

final_pathway_markers = rbind(
    # markers already in xenium panel
    highly_variable_genes[(xenium),.(marker = Gene, annotation = Pathway, category = 'Pathway', xenium)], 
    # manually added markers
    added_probes_pathway
)
print(paste0('Adding ', nrow(added_probes_pathway), ' manually added markers'))

kable(
    final_pathway_markers[
        ,.(
            `Markers Included` = paste0(marker[xenium], collapse = ", "),
            `Markers To Add` = paste0(marker[!xenium], collapse = ", "),
            Total = .N
        ), 
        .(`Pathway` = annotation)
    ],
    caption = "Final Pathway Markers"
)
```

# Save Data

```{r}
final_markers = rbind(final_cell_type_markers, final_pathway_markers)
write.csv(final_markers, paste0(workdir, "coja/Spatial_PLGG/data/markers/final_manual_markers.csv"), row.names = FALSE)
print(paste0('Final markers saved to ', paste0(workdir, "coja/Spatial_PLGG/data/markers/final_manual_markers.csv")))
print(paste0('Adding in ', nrow(final_markers[!(xenium)]),' markers to the xenium panel'))
kable(
    final_markers[
        ,.(
            Total = .N,
            `Markers Included` = paste0(marker[xenium], collapse = ", "),
            `Markers To Add` = paste0(marker[!xenium], collapse = ", ")
        ), 
        .(category, annotation)
    ],
    caption = "Markers to Add to Xenium Panel"
)
```