---
title: "Manual Curation of Markers wrt 5K Xenium Panel"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 6
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

We are trying to handpick some cell type and pathway markers for the PLGG Spatial Project. The goal is to have a list of markers that are specific to the cell types and pathways that we are interested in based on prior knowledge.

1. Xenium Panel
    
    `/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/XeniumPrimeHuman5Kpan_tissue_pathways_metadata.csv`
    
2. Cell type markers
    
    `/xchip/beroukhimlab/coja/Spatial_PLGG/data/markers/curated_markers.xlsx`
    
    1. Which of them already ovelap with the Xenium panel already? 

    2. Which of them are good to include? (2-3 per cell type, 10 cell type total)

3. pathway markers
    
    `/xchip/beroukhimlab/coja/Spatial_PLGG/data/markers/pathway_markers.csv`
    
    1. Which of them already overlap with the Xenium panel already?

    2. Which of them are good to include? (10 per pathway, 2 pathway total)

        Highly variable gene annotated here: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/markers/pathway_markers_vargene.csv`


```{r}
#| echo: FALSE
library(data.table)
library(ggplot2)
library(readxl)
library(knitr)
# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}
```

# Load Data

```{r}
# Load Xenium Panel
xenium_panel <- fread(paste0(workdir, "youyun/plgg/data/xenium_selection/XeniumPrimeHuman5Kpan_tissue_pathways_metadata.csv"))
# Load Cell Type Markers
cell_type_markers <- melt(data.table(
    read_xlsx(paste0(workdir, "coja/Spatial_PLGG/data/markers/curated_markers.xlsx"))
)[,id := c(1:.N)], id.vars = "id", variable.name = "cell_type", value.name = "marker")[
    !is.na(marker) & !cell_type %in% c('MAPK','Senmayo')
]
cell_type_markers[, id := NULL]
# Load Pathway Markers
pathway_markers <- fread(paste0(workdir, "coja/Spatial_PLGG/data/markers/pathway_markers.csv"))
# Load Highly Variable Genes
highly_variable_genes <- fread(paste0(workdir, "coja/Spatial_PLGG/data/markers/pathway_markers_vargene.csv"))
```

# Cell Type Markers

## Overlap with Xenium Panel

```{r}
#| results: hold
cell_type_markers[, xenium := marker %in% xenium_panel$gene_name]
ggplot(cell_type_markers) +
    geom_bar(aes(x = cell_type, fill = xenium), position = "stack") +
    # add horizontal line at 3
    geom_hline(yintercept = 3, linetype = "dashed", color = "white") +
    labs(title = "Overlap of Cell Type Markers with Xenium Panel (5K)", x = "Overlap with Xenium Panel", y = "Count") +
    theme_minimal() + theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 15)
    ) + scale_fill_brewer(palette = "Set1")
kable(
    cell_type_markers[
        , .(
            `Markers included` = paste0(marker[xenium], collapse = ", "),
            `Markers not included` = paste0(marker[!xenium], collapse = ", "),
            `Need more` = 3 - sum(xenium)
        ), .(`Cell Type` = cell_type)
    ]
)
```

## Expert Curation of Cell Type Markers

There are 4 cell types we are missing markers for:

1.  Stromal -- VLMC cells?
2.  Microglia -- TMEM119 is included, and we have good coverage on M1 and M2
3.  Astrocyte -- according to the 5k panel annotation, they do have astrocyte markers?
4.  T cell -- lots of em

It seems reasonable that there are other markers in the 5k xenium panel that are not in the curated markers. We will find those here

```{r}
# get the VLMC markers from the xenium brain panel
VLMC_xenium_brain = fread(paste0(
    workdir, 'youyun/plgg/data/xenium_selection/Xenium_hBrain_v1_metadata.csv'
))[Annotation == 'VLMC']$Genes
print(paste0(
    'Additional VLMC markers from the xenium brain panel: ',
    sum(VLMC_xenium_brain %in% xenium_panel$gene_name)
))
print(paste0(
    'Additional Astrocyte markers in the xenium panel: ',
    paste0(
        xenium_panel[grepl('astrocyte', cell_type, ignore.case = TRUE)]$gene_name, collapse = ", "
    )
))
print(paste0(
    'Additional T cell markers in the xenium panel: ',
    paste0(
        xenium_panel[grepl('T cell', cell_type, ignore.case = TRUE)]$gene_name, collapse = ", "
    )
))
```

# Pathway Markers

## Overlap with Xenium Panel

```{r}
#| results: hold
highly_variable_genes$cohort_var_gene = ifelse(is.na(highly_variable_genes$cohort_var_gene), 0, highly_variable_genes$cohort_var_gene)
plgg_mapk = data.table(
    Gene = c(
        'GYPC', 'GPNMB', 'ETV5', 'TRIB2', 'EPB41L3', 'ST6GAL1', 'SPRY2', 'APOD', 'ETV4', 'SERPINA3', 'ANKH', 'TRIB1', 'FCERIG', 'C3AR1'
    ),
    Role = 'MAPK_plgg',
    Pathway = 'MAPK_plgg'
)
highly_variable_genes = rbind(
    highly_variable_genes[,.(Gene, Role, Pathway)],
    plgg_mapk
)
highly_variable_genes[, xenium := Gene %in% xenium_panel$gene_name]
ggplot(highly_variable_genes) +
    geom_bar(aes(x = Pathway, fill = xenium), position = "stack") +
    # add horizontal line at 10
    geom_hline(yintercept = 10, linetype = "dashed", color = "white") +
    labs(title = "Overlap of Pathway Markers with Xenium Panel (5K)", x = "Overlap with Xenium Panel", y = "Count") +
    theme_minimal() + theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 15)
    ) + scale_fill_brewer(palette = "Set1")
kable(
    highly_variable_genes[
        , .(
            `Markers included` = paste0(Gene[xenium], collapse = ", ")
            # `Need more` = 10 - sum(xenium),
            # `Markers (var in 3)` = paste0(Gene[!xenium & cohort_var_gene == 3], collapse = ", "),
            # `Markers (var in 2)` = paste0(Gene[!xenium & cohort_var_gene == 2], collapse = ", "),
            # `Markers (var in 1)` = paste0(Gene[!xenium & cohort_var_gene == 1], collapse = ", "),
            # `Markers (var in 0)` = paste0(Gene[!xenium & (
            #     cohort_var_gene == 0 | is.na(cohort_var_gene)
            # )], collapse = ", ")
        ), .(Pathway)
    ]
)
```

## Expert Curation of Pathway Markers

1. senescent markers: CDKN1A, CDKN2A, CDKN2B, GLB1, B2M, CGAS, ARG2

2. MAPK [MSS Overlap gene set](https://www.nature.com/articles/s41467-023-40235-8#Fig1): GYPC, GPNMB, ETV5, TRIB2, EPB41L3, ST6GAL1, SPRY2, APOD, ETV4, SERPINA3 , ANKH, TRIB1, FCERIG, C3AR1 

3. M1: IL1B, IL6, TNFalpha, CD11b (ITGAM), CD16 (FCGR3A), CD32 (FCGR2A), CD86

    - ITGAM, FCGR3A, CD86 are already in xenium panel

4. M2: IL4, IL10, MRC1, Arg-1 (ARG1), CHI3L1, STAT6, CCL22, PPARG

```{r}
added_probes_pathway = data.table(
    marker = c(
        'CDKN1A', 'CDKN2A', 'CDKN2B', 'GLB1', 'B2M', 'CGAS', 'ARG2',
        'GYPC', 'GPNMB', 'ETV5', 'TRIB2', 'EPB41L3', 'ST6GAL1', 'SPRY2', 'APOD', 'ETV4', 'SERPINA3', 'ANKH', 'TRIB1', 'FCERIG', 'C3AR1',
        'IL1B', 'IL6', 'TNF', 'FCGR2A', 
        'IL4', 'IL10', 'MRC1', 'ARG1', 'CHI3L1', 'STAT6', 'CCL22', 'PPARG'
    ),
    annotation = c(
        rep('senescence', 7),
        rep('MAPK', 14),
        rep('M1', 4),
        rep('M2', 8)
    ),
    category = 'Pathway'
)[
    ,xenium := ifelse(marker %in% xenium_panel$gene_name, TRUE, FALSE)
]

if(any(added_probes_pathway$marker %in% xenium_panel$gene_name)) {
    print(added_probes_pathway[marker %in% xenium_panel$gene_name])
}else{
    print('All manually added markers are not in Xenium panel')
}

final_pathway_markers = rbind(
    # markers already in xenium panel
    highly_variable_genes[(xenium),.(marker = Gene, annotation = Pathway, category = 'Pathway', xenium)], 
    # manually added markers
    added_probes_pathway
)
print(paste0('Adding ', nrow(added_probes_pathway), ' manually added markers'))

kable(
    final_pathway_markers[
        ,.(
            `Markers Included` = paste0(marker[xenium], collapse = ", "),
            `Markers To Add` = paste0(marker[!xenium], collapse = ", "),
            Total = .N
        ), 
        .(`Pathway` = annotation)
    ],
    caption = "Final Pathway Markers"
)
```

# Save Data

```{r}
# final_markers = rbind(final_cell_type_markers, final_pathway_markers)
# write.csv(final_markers, paste0(workdir, "coja/Spatial_PLGG/data/markers/final_manual_markers.csv"), row.names = FALSE)
# print(paste0('Final markers saved to ', paste0(workdir, "coja/Spatial_PLGG/data/markers/final_manual_markers.csv")))
# print(paste0('Adding in ', nrow(final_markers[!(xenium)]),' markers to the xenium panel'))
# kable(
#     final_markers[
#         ,.(
#             Total = .N,
#             `Markers Included` = paste0(marker[xenium], collapse = ", "),
#             `Markers To Add` = paste0(marker[!xenium], collapse = ", ")
#         ), 
#         .(category, annotation)
#     ],
#     caption = "Markers to Add to Xenium Panel"
# )
```