---
title: "Pan PLGG PERSIST Genes"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 6
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

We have a few gene sets that are related to the PERSIST project. Here we are going to look at the Pan PLGG PERSIST genes.

There is the pan PLGG PERSIST gene set that is trained on the entire dataset of PLGG samples. This is here `/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/PERSIST_results_00092024.csv`.

There is the curated gene list that is used to assign cell types. This is here `/xchip/beroukhimlab/coja/Spatial_PLGG/data/markers/curated_markers.xlsx`.

There is the xenium brain panel gene list. This is here `/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/Xenium_hBrain_v1_metadata.csv`.

There is also the old persist gene list trained on the PAs. This is here `/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/persist_100.csv` and here `/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/persist_480.csv`.

# Load Data

```{r}
library(data.table)
library(ggplot2)
library(readxl)
# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}
```

```{r}
# load the pan PLGG PERSIST gene set
pan_plgg_persist_genes = fread(gsub(
    '/xchip/beroukhimlab/',workdir,
    '/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/PERSIST_results_00092024.csv'
))
pan_plgg_P50_gene_list = pan_plgg_persist_genes[(persist_set_50)]$name
pan_plgg_P100_gene_list = pan_plgg_persist_genes[(persist_set_100)]$name

# load the curated gene list
curated_markers = melt(data.table(read_xlsx(gsub(
    '/xchip/beroukhimlab/',workdir,
    '/xchip/beroukhimlab/coja/Spatial_PLGG/data/markers/curated_markers.xlsx'
)))[, id := 1:.N], id.vars = "id", variable.name = "cell_type", value.name = "gene")[
    !is.na(gene), .(cell_type, gene)
]
curated_markers_gene_list = unique(curated_markers[! cell_type %in% c('MAPK','Senmayo')]$gene)
mapk_senmayo_markers_gene_list = unique(curated_markers[cell_type %in% c('MAPK','Senmayo')]$gene)

# load the xenium brain panel gene list
xenium_brain_panel = fread(gsub(
    '/xchip/beroukhimlab/',workdir,
    '/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/Xenium_hBrain_v1_metadata.csv'
))
xenium_brain_panel_gene_list = xenium_brain_panel$Genes

# load the old persist gene list
persist_100 = fread(gsub(
    '/xchip/beroukhimlab/',workdir,
    '/xchip/beroukhimlab/youyun/plgg/data/xenium_selection/persist_100.csv'
))
pa_P100_gene_list = persist_100$Genes
```

# Compare Gene Lists

```{r}
library(ggVennDiagram)

# compare the gene lists
gene_lists = list(
    `Markers` = curated_markers_gene_list,
    `Pathways` = mapk_senmayo_markers_gene_list,
    `PanPLGG P100` = pan_plgg_P100_gene_list,
    `PanPLGG P50` = pan_plgg_P50_gene_list,
    `PA P100` = pa_P100_gene_list,
    `Xenium` = xenium_brain_panel_gene_list
)

# plot the venn diagram
ggVennDiagram(gene_lists) + coord_cartesian(clip="off")
```