---
title: ""
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 6
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk:
        results: hold
        R.options:
            width: 120
---

```{r}
#| echo: FALSE
library(SingleCellExperiment)
library(data.table)
library(reticulate)
library(knitr)

# local vs remote
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usually at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (Sys.getenv("HOME") == "/PHShome/yz762"){
    # in eristwo, the home directory is at /PHShome/yz762
    workdir <- "/data/beroukhim1/"
} else if (Sys.getenv("HOME") == "/home/yz762"){
    # on the new server, the home directory is at /home/yz762
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    # in dipg or uger, the home directory is '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}
```

# Introduction

- Input single-nucleus AnnData: `youyun/plgg/data/single_cell/Extended_sndata_filtered.h5ad`
- Sample-to-histology mapping source:
  `youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds`
- Output directory: `youyun/plgg/data/sc_integration/`

This notebook reads one input h5ad, uses BANKSY metadata to map `sample_id -> histology`,
and writes one sample-specific h5ad per mapping row with filename:
`<sample_id>_<histology_with_underscores>_sn.h5ad`.

# Parameters

```{r}
input_h5ad <- paste0(workdir, "youyun/plgg/data/single_cell/Extended_sndata_filtered.h5ad")
output_dir <- paste0(workdir, "youyun/plgg/data/sc_integration/")
output_log_tsv <- paste0(output_dir, "h5ad_2_sample_h5ad_log.tsv")

banksy_rds <- ifelse(
    basename(workdir) != "dfci_mount",
    paste0(
        workdir,
        "youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/",
        "banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
    ),
    paste0(
        "/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/",
        "banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
    )
)
```

# Load sample-histology mapping

```{r}
stopifnot(file.exists(input_h5ad))
stopifnot(file.exists(banksy_rds))

banksy_embeddings <- readRDS(banksy_rds)

sample_histology <- as.data.table(unique(colData(banksy_embeddings)[, c("sample_id", "histology")]))
sample_histology[, sample_id := as.character(sample_id)]
sample_histology[, histology := as.character(histology)]
sample_histology <- unique(sample_histology[, .(sample_id, histology)])

# Ensure 1:1 sample_id -> histology mapping
conflicting <- sample_histology[, .(n_histology = uniqueN(histology)), by = sample_id][n_histology > 1]
if (nrow(conflicting) > 0) {
    stop(
        "Found sample_id values with multiple histology labels: ",
        paste(conflicting$sample_id, collapse = ", ")
    )
}

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

cat("Samples in mapping:", nrow(sample_histology), "\n")
kable(sample_histology[order(sample_id)])

use_condaenv("/mnt/storage/dept/medonc/beroukhim/youyun/util/miniforge3/envs/spatial/", required = TRUE)
```

# Split h5ad by sample and write outputs

```{python}
import anndata as ad
import pandas as pd
from pathlib import Path

input_h5ad = Path(r.input_h5ad)
output_dir = Path(r.output_dir)
output_log_tsv = Path(r.output_log_tsv)

sample_histology = r.sample_histology.copy()
sample_histology["sample_id"] = sample_histology["sample_id"].astype(str)
sample_histology["histology"] = sample_histology["histology"].astype(str)

adata = ad.read_h5ad(input_h5ad, backed="r")
if "orig.ident" not in adata.obs.columns:
    raise KeyError("Column 'orig.ident' not found in input h5ad obs.")

orig_ident = adata.obs["orig.ident"].astype(str).to_numpy()
logs = []

# try:
#     for row in sample_histology.itertuples(index=False):
#         sid = str(row.sample_id)
#         hist = str(row.histology)
#         hist_safe = hist.replace(" ", "_")
#         out_file = output_dir / f"{sid}_{hist_safe}_sn.h5ad"

#         mask = (orig_ident == sid)
#         n_cells = int(mask.sum())

#         if n_cells == 0:
#             logs.append({
#                 "sample_id": sid,
#                 "histology": hist,
#                 "n_cells": n_cells,
#                 "status": "missing_in_h5ad",
#                 "outfile": str(out_file),
#             })
#             continue

#         sub = adata[mask].to_memory()
#         sub_sample_ids = pd.Index(sub.obs["orig.ident"].astype(str).unique())
#         if len(sub_sample_ids) != 1 or sub_sample_ids[0] != sid:
#             logs.append({
#                 "sample_id": sid,
#                 "histology": hist,
#                 "n_cells": n_cells,
#                 "status": "sample_mismatch",
#                 "outfile": str(out_file),
#             })
#             continue

#         sub.write_h5ad(out_file, compression="gzip")
#         logs.append({
#             "sample_id": sid,
#             "histology": hist,
#             "n_cells": n_cells,
#             "status": "written",
#             "outfile": str(out_file),
#         })
# finally:
#     if hasattr(adata, "file") and adata.file is not None:
#         adata.file.close()

# log_df = pd.DataFrame(logs)
# log_df.to_csv(output_log_tsv, sep="\t", index=False)

# print(log_df)
# print("\nStatus counts:")
# print(log_df["status"].value_counts(dropna=False))
```

# Quick compatibility check (first 3 written files)

```{python}
# check_rows = []
# written = log_df.loc[log_df["status"] == "written", "outfile"].head(3).tolist()

# for f in written:
#     adx = ad.read_h5ad(f, backed="r")
#     try:
#         check_rows.append({
#             "outfile": f,
#             "n_obs": int(adx.n_obs),
#             "n_vars": int(adx.n_vars),
#             "n_obs_orig_ident": int(adx.obs["orig.ident"].astype(str).nunique()),
#             "n_layers": len(adx.layers.keys()),
#             "n_obsm": len(adx.obsm.keys()),
#         })
#     finally:
#         if hasattr(adx, "file") and adx.file is not None:
#             adx.file.close()

# pd.DataFrame(check_rows)
```

# Come up with a list of 2048 HVGs for PA Samples Together

we will subset the object to all PA samples together, and then use `scanpy.pp.highly_variable_genes` to get the top 2048 HVGs across all PA samples. This will be used for the PA spatial imputation step later on, where we will subset the object to only the PA samples and only these 2048 HVGs before running the imputation.

```{python}
import scanpy as sc

# Identify pilocytic astrocytoma (PA) sample IDs from BANKSY-derived mapping
sample_histology_pa = sample_histology.copy()
sample_histology_pa["histology"] = sample_histology_pa["histology"].astype(str)
sample_histology_pa["sample_id"] = sample_histology_pa["sample_id"].astype(str)

pa_sample_ids = (
    sample_histology_pa.loc[
        sample_histology_pa["histology"].str.contains("pilocytic astrocytoma", case=False, na=False),
        "sample_id"
    ]
    .drop_duplicates()
    .tolist()
)

if len(pa_sample_ids) == 0:
    raise ValueError("No PA samples found in sample_histology mapping.")

adata = ad.read_h5ad(input_h5ad, backed="r")
try:
    orig_ident = adata.obs["orig.ident"].astype(str).to_numpy()
    pa_mask = pd.Series(orig_ident).isin(pa_sample_ids).to_numpy()
    n_pa_cells = int(pa_mask.sum())
    if n_pa_cells == 0:
        raise ValueError("No PA cells found in input_h5ad after filtering with PA sample IDs.")

    adata_pa = adata[pa_mask].to_memory()
finally:
    if hasattr(adata, "file") and adata.file is not None:
        adata.file.close()

adata_pa.var_names_make_unique()
sc.pp.highly_variable_genes(adata_pa, n_top_genes=2048, inplace=True)

if "highly_variable_rank" in adata_pa.var.columns:
    pa_hvg_df = adata_pa.var.sort_values("highly_variable_rank")
else:
    pa_hvg_df = adata_pa.var.sort_index()

if pa_hvg_df.shape[0] != 2048:
    print(f"Warning: expected 2048 HVGs but got {pa_hvg_df.shape[0]}.")

pa_hvg_df.insert(0, "gene", pa_hvg_df.index.astype(str))

pa_hvg_tsv = output_dir / "PA_top2048_HVGs.tsv"

pa_hvg_df.to_csv(pa_hvg_tsv, sep="\t", index=False)

print(f"PA samples: {len(pa_sample_ids)}")
print(f"PA cells: {n_pa_cells}")
print(f"HVG list written: {pa_hvg_tsv}")
print(pa_hvg_df.head())
```
