---
title: ""
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 6
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

```{r}
#| echo: FALSE
library(SingleCellExperiment)
library(SpatialExperiment)
library(sceasy)

library(data.table)
library(ggplot2)
library(knitr)

# local vs remote
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (Sys.getenv("HOME") == "/PHShome/yz762"){
    # in eristwo, the home directory is at /PHShome/yz762
    workdir <- "/data/beroukhim1/"
} else if (Sys.getenv("HOME") == "/home/yz762"){
    # on the new server, the home directory is at /home/yz762
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    # in dipg or uger, the home directory is '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}
```

# Introduction

-   RDS object with Leiden Clusters: `/data/beroukhim1/youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/banksy_clusters_connected_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds`

-   subset version here: `/data/beroukhim1/youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds`

We will convert this into an AnnData object for use in Python

# Data Conversion

We will convert single cell experiment class in the rds object to an AnnData object in Python. 
More specifically, `SingleCellExperiment` class stores the single cell data in R and our spatial data is stored in a `SpatialExperiment` class.

ENVI takes in raw counts and log1p transform the expression

```{r}
.regularise_df <- function(df, drop_single_values = TRUE) {
    if (ncol(df) == 0) df[["name"]] <- rownames(df)
    if (drop_single_values) {
        k_singular <- sapply(df, function(x) length(unique(x)) == 1)
        if (sum(k_singular) > 0) {
        warning(
            paste("Dropping single category variables:"),
            paste(colnames(df)[k_singular], collapse = ", ")
        )
        }
        df <- df[, !k_singular, drop = F]
        if (ncol(df) == 0) df[["name"]] <- rownames(df)
    }
    return(df)
}
```

```{r}
spe2anndata <- function(obj, outFile = NULL, main_layer = "counts", transfer_layers = NULL, drop_single_values = TRUE) {
    if (!requireNamespace("SummarizedExperiment")) {
        stop("This function requires the 'SummarizedExperiment' package.")
    }
    if (!requireNamespace("SingleCellExperiment")) {
        stop("This function requires the 'SingleCellExperiment' package.")
    }
    if (!requireNamespace('SpatialExperiment')) {
        stop("This function requires the 'SpatialExperiment' package.")
    }

    assay_names <- SummarizedExperiment::assayNames(obj)
    main_layer <- match.arg(main_layer, assay_names)
    transfer_layers <- transfer_layers[transfer_layers %in% assay_names]
    transfer_layers <- transfer_layers[transfer_layers != main_layer]

    X <- SummarizedExperiment::assay(obj, main_layer)

    obs <- .regularise_df(as.data.frame(SummarizedExperiment::colData(obj)), drop_single_values = drop_single_values)

    var <- .regularise_df(as.data.frame(SummarizedExperiment::rowData(obj)), drop_single_values = drop_single_values)

    obsm <- NULL
    reductions <- SingleCellExperiment::reducedDimNames(obj)
    if (length(reductions) > 0) {
        obsm <- sapply(
            reductions,
            function(name) {
                as.matrix(
                    SingleCellExperiment::reducedDim(obj, type = name)
                )
            },
            simplify = FALSE
        )
        names(obsm) <- paste0(
            "X_", tolower(SingleCellExperiment::reducedDimNames(obj))
        )
    }
    # adding in spatial coordinates
    obsm[['spatial']] <- spatialCoords(obj)

    layers <- list()
    for (layer in transfer_layers) {
        mat <- SummarizedExperiment::assay(obj, layer)
        if (all(dim(mat) == dim(X))) layers[[layer]] <- Matrix::t(mat)
    }

    anndata <- reticulate::import("anndata", convert = FALSE)

    adata <- anndata$AnnData(
        X = Matrix::t(X),
        obs = obs,
        var = var,
        obsm = obsm,
        layers = layers
    )

    if (!is.null(outFile)) {
        adata$write(outFile, compression = "gzip")
    }

    adata
}
```

```{r}
banksy_embeddings <- readRDS(ifelse(
    basename(workdir) != 'dfci_mount', paste0(
        workdir, "youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/banksy_clusters_connected_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
    ), "/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
))
colData(banksy_embeddings)$sample_id <- factor(colData(banksy_embeddings)$sample_id)
colData(banksy_embeddings)$alteration = ifelse(is.na(colData(banksy_embeddings)$alteration), 'NA', colData(banksy_embeddings)$alteration)
cell_type_mapping = fread(paste0(
    workdir,'youyun/plgg/data/xenium_cell_types/banksy_proseg_cell_cluster_annotation.tsv'
))

sample_histology = as.data.table(unique(colData(banksy_embeddings)[,c('sample_id','histology')]))

mapply(
    function(x, y){
        # subset to one sample of interest
        banksy_embeddings_sample <- banksy_embeddings[,
            colData(banksy_embeddings)$sample_id == x
        ]

        # merge the metadata into the coldata
        original_order = colnames(banksy_embeddings_sample)
        merged_metadata = merge(
            as.data.table(colData(banksy_embeddings_sample))[
                ,.(clust_Harmony_BANKSY_lam0.2_k50_res1 = as.numeric(as.character(clust_Harmony_BANKSY_lam0.2_k50_res1)), cell_id_copy = cell_id)
            ], cell_type_mapping,
            by.x = "clust_Harmony_BANKSY_lam0.2_k50_res1",
            by.y = "cluster", all.x = TRUE
        )[match(gsub('__.*','', original_order), cell_id_copy)]
        colData(banksy_embeddings_sample) = cbind(
            colData(banksy_embeddings_sample),
            merged_metadata
        )

        if(!all(colData(banksy_embeddings_sample)$cell_id == colData(banksy_embeddings_sample)$cell_id_copy)){
            stop("cell_id does not match the original order")
        } else {
            message("cell_id matches the original order")
            rownames(colData(banksy_embeddings_sample)) <- original_order
        }


        use_condaenv('/mnt/storage/dept/medonc/beroukhim/youyun/util/miniforge3/envs/spatial')
        spe2anndata(
            banksy_embeddings_sample, outFile = paste0(
                workdir, "youyun/plgg/data/sc_integration/",
                x,'_',gsub(' ', '_', y), "_spatial.h5ad"
            ), main_layer = "normcounts",
            drop_single_values = FALSE
        )

        return(NULL)
    },
    sample_histology$sample_id,
    sample_histology$histology
)
```
