---
title: "Cell-type composition: healthy brain vs PLGG"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 10
        fig-height: 7
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk:
        results: hold
        R.options:
            width: 120
---

# Overview

Beta-binomial pseudobulk models comparing non-neoplastic cell-type proportions
between the healthy BCA atlas and PLGG tumors, controlling for age, sex, and
brain location.

Two analysis stages:

- **Stage 1**: healthy vs all tumor (maximum power).
- **Stage 2**: healthy vs each tumor entity (IntDx), one coefficient per entity.

The primary analysis uses a non-neoplastic (TME) denominator. A **sensitivity
analysis** at the end re-runs both stages with neoplastic cells kept in the
denominator, so TME fractions stay comparable across entities that differ in how
reliably neoplastic cells can be called.

Pseudobulk unit: `donor_proxy × location_broad`. Each row is a biological
donor (healthy) or patient sample (tumor) × brain region, with counts per cell
type and total non-neoplastic cells. `donor_proxy` collapses multi-batch and
per-cell-GSM healthy IDs to the biological-donor level; tumor `sample_ID`s are
used as-is.

```{r setup}
#| echo: FALSE
#| output: false
suppressPackageStartupMessages({
    library(arrow)
    library(dplyr)
    library(tidyr)
    library(purrr)
    library(stringr)
    library(ggplot2)
    library(ggrepel)
    library(ComplexHeatmap)
    library(circlize)
    library(glmmTMB)
    library(scales)
})

# Resolve data root from HOME (AGENTS.md convention)
home = Sys.getenv("HOME")
if (startsWith(home, "/Users/youyun") || startsWith(home, "/Users/youyunzheng")) {
    data_root = file.path(home, "Documents/HMS/PhD/beroukhimlab/dfci_mount")
} else if (home == "/PHShome/yz762") {
    data_root = "/data/beroukhim1"
} else if (home == "/home/yz762") {
    data_root = "/mnt/storage/dept/medonc/beroukhim"
} else {
    data_root = "/xchip/beroukhimlab"
}

sc_dir   = file.path(data_root, "youyun/plgg/data/single_cell")
code_dir = file.path(data_root, "youyun/plgg/code/healthy_comparisons")
in_path  = file.path(sc_dir, "combined_healthyTumor_obs_metadata_clean.parquet")

# Nature-style ggplot theme
theme_nature = function(base_size = 9) {
    theme_classic(base_size = base_size) +
    theme(
        text             = element_text(color = "black"),
        axis.text        = element_text(size = base_size, color = "black"),
        axis.title       = element_text(size = base_size + 1, color = "black"),
        axis.line        = element_line(color = "black", linewidth = 0.4),
        axis.ticks       = element_line(color = "black", linewidth = 0.4),
        panel.border     = element_rect(color = "black", fill = NA, linewidth = 0.5),
        strip.background = element_blank(),
        strip.text       = element_text(size = base_size, face = "bold"),
        legend.text      = element_text(size = base_size),
        legend.title     = element_text(size = base_size, face = "bold"),
        legend.key.size  = unit(0.4, "cm"),
        plot.title       = element_text(size = base_size + 2, face = "bold", hjust = 0.5),
        plot.subtitle    = element_text(size = base_size, hjust = 0.5, color = "gray40")
    )
}
```

# Data loading and pseudobulk

```{r load-pseudobulk}
df = read_parquet(in_path)
cat("Loaded:", nrow(df), "cells\n")

# --- Pseudobulk: counts per donor_proxy x location_broad x cell_type ----------
ct_counts = df |>
    group_by(donor_proxy, location_broad, cell_type_harmonised) |>
    summarise(n_ct = n(), .groups = "drop")

unit_totals = df |>
    group_by(donor_proxy, location_broad) |>
    summarise(n_total = n(), .groups = "drop")

# Sample-level covariates must be constant within donor_proxy x location_broad.
# Warn if any unit shows variation (e.g. mixed IntDx, which shouldn't occur).
sample_meta = df |>
    group_by(donor_proxy, location_broad) |>
    summarise(
        condition                    = first(condition),
        IntDx                        = first(IntDx),
        age_yr                       = first(age_yr),
        sex_combined                 = first(sex_combined),
        fraction_neoplastic_detected = first(fraction_neoplastic_detected),
        .groups = "drop"
    )

# Verify IntDx is constant within each tumor donor_proxy x location unit
intdx_check = df |>
    filter(condition == "tumor") |>
    group_by(donor_proxy, location_broad) |>
    summarise(n_intdx = n_distinct(IntDx), .groups = "drop") |>
    filter(n_intdx > 1)
if (nrow(intdx_check) > 0) warning("Some tumor units have mixed IntDx: ", paste(intdx_check$donor_proxy, collapse = ", "))

# Complete grid: all donor_proxy x location units x all cell types (fill zeros)
all_units = sample_meta |> select(donor_proxy, location_broad)
all_ct    = df |> distinct(cell_type_harmonised)
pseudobulk = cross_join(all_units, all_ct) |>
    left_join(ct_counts,    by = c("donor_proxy", "location_broad", "cell_type_harmonised")) |>
    left_join(unit_totals,  by = c("donor_proxy", "location_broad")) |>
    left_join(sample_meta,  by = c("donor_proxy", "location_broad")) |>
    mutate(n_ct = replace_na(n_ct, 0))

cat("Pseudobulk units (donor_proxy x location):", nrow(all_units), "\n")
cat("Pseudobulk rows (units x cell types):", nrow(pseudobulk), "\n")
cat("Healthy units:", sum(sample_meta$condition == "healthy"), "\n")
cat("Tumor units:", sum(sample_meta$condition == "tumor"), "\n")

# Cell types to model (Neoplastic already removed in clean parquet)
cell_types = sort(unique(df$cell_type_harmonised))
cat("Cell types:", paste(cell_types, collapse = ", "), "\n")
```

```{r model-helpers}
#| echo: FALSE

# Fit one beta-binomial model for a given cell type, formula, and dataset.
# Returns ALL non-intercept coefficients; caller filters to the terms of interest.
fit_betabin = function(ct, formula_str, dat) {
    dat_ct = dat |>
        filter(cell_type_harmonised == ct, !is.na(age_yr), !is.na(sex_combined)) |>
        mutate(n_other = n_total - n_ct)

    tryCatch({
        fit = glmmTMB(
            as.formula(formula_str),
            family = betabinomial(link = "logit"),
            data   = dat_ct
        )
        coef_tbl = as.data.frame(coef(summary(fit))$cond)
        coef_tbl$term = rownames(coef_tbl)
        rownames(coef_tbl) = NULL
        names(coef_tbl) = c("estimate", "std.error", "statistic", "p.value", "term")
        coef_tbl |>
            filter(term != "(Intercept)") |>
            mutate(
                cell_type  = ct,
                conf.low   = estimate - 1.96 * std.error,
                conf.high  = estimate + 1.96 * std.error,
                converged  = TRUE
            )
    }, error = function(e) {
        tibble(
            cell_type = ct, term = NA_character_,
            estimate = NA_real_, std.error = NA_real_,
            statistic = NA_real_, p.value = NA_real_,
            conf.low = NA_real_, conf.high = NA_real_,
            converged = FALSE, error_msg = conditionMessage(e)
        )
    })
}
```

---

# Stage 1: healthy vs all tumor

```{r stage1-model}
# Formula: proportion ~ condition (healthy/tumor) + age + location + sex
# Reference for condition: "healthy"
pseudobulk_s1 = pseudobulk |>
    mutate(condition = relevel(factor(condition), ref = "healthy"))

stage1_raw = map_dfr(cell_types, \(ct)
    fit_betabin(
        ct          = ct,
        formula_str = "cbind(n_ct, n_other) ~ condition + age_yr + location_broad + sex_combined",
        dat         = pseudobulk_s1
    )
)

# Group effect: healthy vs all tumor (FDR across cell types)
stage1 = stage1_raw |>
    filter(converged, !is.na(p.value), term == "conditiontumor") |>
    mutate(
        q.value   = p.adjust(p.value, method = "BH"),
        sig       = q.value < 0.1,
        direction = if_else(estimate > 0, "Enriched in tumor", "Depleted in tumor")
    )

# Covariate effects: age, sex, location — FDR within each term across cell types
covar_terms = c("age_yr", "sex_combinedMale",
                "location_broadPosterior_fossa", "location_broadSupratentorial")
stage1_covars = stage1_raw |>
    filter(converged, !is.na(p.value), term %in% covar_terms) |>
    group_by(term) |>
    mutate(q.value = p.adjust(p.value, method = "BH")) |>
    ungroup() |>
    mutate(sig = q.value < 0.1)

cat("Stage 1 convergence failures:",
    sum(!stage1_raw$converged | is.na(stage1_raw$p.value)), "/", length(cell_types), "\n")
print(stage1 |> select(cell_type, estimate, std.error, p.value, q.value, sig) |>
      arrange(q.value))

write.table(stage1, file.path(code_dir, "results_stage1_tumor_vs_healthy.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
write.table(stage1_covars, file.path(code_dir, "results_stage1_covariates.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r stage1-plot, fig.width=5, fig.height=5}
# Forest plot: log-OR per cell type, sorted by effect size.
# Error bars = 95% CI; color = significance at q < 0.1.
# x-axis clipped to 90th percentile of |estimate| so well-estimated cell types
# are legible — CI bars for rare/tumor-specific types may extend off-screen.
sig_colors = c("Enriched in tumor" = "#C0392B", "Depleted in tumor" = "#2980B9", "n.s." = "gray60")

stage1_plot = stage1 |>
    mutate(
        sig_label  = if_else(sig, direction, "n.s."),
        cell_type  = str_replace_all(cell_type, "_", " "),
        cell_type  = reorder(cell_type, estimate)
    )

ggplot(stage1_plot, aes(x = estimate, y = cell_type, color = sig_label)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.4) +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                   height = 0.25, linewidth = 0.5, alpha = 0.7) +
    geom_point(size = 2.2) +
    scale_color_manual(values = sig_colors, name = NULL) +
    scale_x_continuous(expand = expansion(mult = 0.05)) +
    coord_cartesian(xlim = c(-30, 30)) +
    labs(
        x        = "Log odds-ratio vs. healthy (beta-binomial)",
        y        = NULL,
        title    = "Cell-type proportions: tumor vs. healthy brain",
        subtitle = "Adjusted for age, sex, and brain location  |  BH q < 0.1\nCI bars clipped for cell types near-absent in healthy"
    ) +
    theme_nature(base_size = 9) +
    theme(legend.position = "bottom")
```

---

# Stage 2: healthy vs each tumor entity (IntDx)

```{r stage2-model}
# Minimum samples per IntDx entity to include in the model
MIN_SAMPLES_PER_ENTITY = 3

entity_counts = pseudobulk |>
    filter(condition == "tumor") |>
    distinct(donor_proxy, location_broad, IntDx) |>
    count(IntDx, name = "n_units")

entities_kept = entity_counts |> filter(n_units >= MIN_SAMPLES_PER_ENTITY) |> pull(IntDx)
cat("IntDx entities with >=", MIN_SAMPLES_PER_ENTITY, "units:", paste(sort(entities_kept), collapse = ", "), "\n")
cat("Excluded (too few units):", paste(setdiff(entity_counts$IntDx, entities_kept), collapse = ", "), "\n")

# Build Stage 2 pseudobulk: group = IntDx for tumor, "Healthy" for healthy
pseudobulk_s2 = pseudobulk |>
    filter(is.na(IntDx) | IntDx %in% entities_kept) |>
    mutate(
        group = if_else(condition == "healthy", "Healthy", as.character(IntDx)),
        group = relevel(factor(group), ref = "Healthy")
    )

stage2_raw = map_dfr(cell_types, \(ct)
    fit_betabin(
        ct          = ct,
        formula_str = "cbind(n_ct, n_other) ~ group + age_yr + location_broad + sex_combined",
        dat         = pseudobulk_s2
    )
)

# Group effects: each IntDx entity vs healthy (FDR across all cell_type x entity pairs)
stage2 = stage2_raw |>
    filter(converged, !is.na(p.value),
           str_detect(term, "^group"), term != "groupHealthy") |>
    mutate(
        entity  = str_remove(term, "^group"),
        q.value = p.adjust(p.value, method = "BH"),
        sig     = q.value < 0.1
    )

# Covariate effects from Stage 2 models — FDR within each term across cell types
stage2_covars = stage2_raw |>
    filter(converged, !is.na(p.value), term %in% covar_terms) |>
    group_by(term) |>
    mutate(q.value = p.adjust(p.value, method = "BH")) |>
    ungroup() |>
    mutate(sig = q.value < 0.1)

cat("\nStage 2 convergence failures:",
    sum(!stage2_raw$converged | is.na(stage2_raw$p.value)),
    "/", length(cell_types) * length(entities_kept), "\n")

write.table(stage2, file.path(code_dir, "results_stage2_by_entity.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
write.table(stage2_covars, file.path(code_dir, "results_stage2_covariates.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r stage2-heatmap, fig.width=8, fig.height=6}
# Heatmap: log-OR for each cell_type x entity. Rows = cell types, cols = entities.
# Significance overlay: * for q < 0.1, ** for q < 0.01.

# Pivot to matrix
hm_data = stage2 |>
    select(cell_type, entity, estimate, q.value) |>
    mutate(cell_type = str_replace_all(cell_type, "_", " "))

est_mat = hm_data |>
    select(cell_type, entity, estimate) |>
    pivot_wider(names_from = entity, values_from = estimate) |>
    tibble::column_to_rownames("cell_type") |>
    as.matrix()

sig_mat = hm_data |>
    select(cell_type, entity, q.value) |>
    pivot_wider(names_from = entity, values_from = q.value) |>
    tibble::column_to_rownames("cell_type") |>
    as.matrix()

# Align matrices (same rows/cols)
est_mat = est_mat[rownames(est_mat), colnames(est_mat)]
sig_mat = sig_mat[rownames(est_mat), colnames(est_mat)]

# Significance annotation: ** q<0.01, * q<0.1
sig_annot = matrix("", nrow = nrow(sig_mat), ncol = ncol(sig_mat),
                   dimnames = dimnames(sig_mat))
sig_annot[!is.na(sig_mat) & sig_mat < 0.01] = "**"
sig_annot[!is.na(sig_mat) & sig_mat >= 0.01 & sig_mat < 0.1] = "*"

# Hard cap at ±3 so cell types absent in healthy (which produce extreme ORs)
# don't wash out the color scale for well-estimated comparisons. Direction and
# rough magnitude are still communicated; asterisks mark significance.
cap_val = 3
col_fun = colorRamp2(
    c(-cap_val, 0, cap_val),
    c("#2166AC", "white", "#B2182B")   # blue-white-red, colorbrewer RdBu
)

# Sample size annotation for columns (n units per entity)
n_entity = entity_counts |>
    filter(IntDx %in% colnames(est_mat)) |>
    arrange(match(IntDx, colnames(est_mat)))

col_annot = HeatmapAnnotation(
    "n units" = anno_barplot(
        n_entity$n_units,
        gp       = gpar(fill = "gray40", col = NA),
        height   = unit(1.5, "cm"),
        axis_param = list(gp = gpar(fontsize = 7))
    ),
    annotation_name_gp = gpar(fontsize = 8)
)

Heatmap(
    est_mat,
    name             = "log-OR\nvs. healthy",
    col              = col_fun,
    na_col           = "gray90",
    cell_fun         = function(j, i, x, y, width, height, fill) {
        grid.text(sig_annot[i, j], x, y,
                  gp = gpar(fontsize = 9, col = "black", fontface = "bold"))
    },
    top_annotation   = col_annot,
    row_names_gp     = gpar(fontsize = 8),
    column_names_gp  = gpar(fontsize = 8, fontface = "bold"),
    column_names_rot = 45,
    cluster_rows     = TRUE,
    cluster_columns  = TRUE,
    show_row_dend    = TRUE,
    show_column_dend = FALSE,
    heatmap_legend_param = list(
        title_gp     = gpar(fontsize = 8, fontface = "bold"),
        labels_gp    = gpar(fontsize = 7),
        legend_height = unit(3, "cm")
    ),
    column_title     = "Cell-type enrichment per PLGG entity vs. healthy brain",
    column_title_gp  = gpar(fontsize = 10, fontface = "bold"),
    row_title_gp     = gpar(fontsize = 9)
)
```

---

# Covariate effects on cell-type proportions

Effects of age, sex, and brain location on cell-type proportions, extracted from
the same models. FDR is applied within each covariate term across cell types.
Location reference level is Intraventricular (alphabetically first).

```{r covariate-heatmap-helper}
#| echo: FALSE

# Build a log-OR heatmap for covariate terms.
# dat: long table with columns cell_type, term, estimate, q.value, sig
plot_covar_heatmap = function(dat, title_str) {
    hm = dat |>
        mutate(
            cell_type  = str_replace_all(cell_type, "_", " "),
            term_label = recode(term,
                "age_yr"                         = "Age (per yr)",
                "sex_combinedMale"               = "Sex: Male vs Female",
                "location_broadPosterior_fossa"  = "Location: Post. fossa vs Intraventricular",
                "location_broadSupratentorial"   = "Location: Supratentorial vs Intraventricular"
            )
        )

    sig_tbl = hm |> select(cell_type, term_label, q.value) |>
        pivot_wider(names_from = term_label, values_from = q.value) |>
        tibble::column_to_rownames("cell_type") |> as.matrix()

    est_tbl = hm |> select(cell_type, term_label, estimate) |>
        pivot_wider(names_from = term_label, values_from = estimate) |>
        tibble::column_to_rownames("cell_type") |> as.matrix()

    # Align to same rows
    est_tbl = est_tbl[rownames(est_tbl), colnames(est_tbl)]
    sig_tbl = sig_tbl[rownames(est_tbl), colnames(est_tbl)]

    sig_annot = matrix("", nrow = nrow(sig_tbl), ncol = ncol(sig_tbl),
                       dimnames = dimnames(sig_tbl))
    sig_annot[!is.na(sig_tbl) & sig_tbl < 0.01] = "**"
    sig_annot[!is.na(sig_tbl) & sig_tbl >= 0.01 & sig_tbl < 0.1] = "*"

    # Symmetric cap at 95th percentile of absolute estimates
    cap = quantile(abs(est_tbl), 0.95, na.rm = TRUE)
    col_fn = colorRamp2(c(-cap, 0, cap), c("#2166AC", "white", "#B2182B"))

    Heatmap(
        est_tbl,
        name             = "log-OR",
        col              = col_fn,
        na_col           = "gray90",
        cell_fun         = function(j, i, x, y, width, height, fill) {
            grid.text(sig_annot[i, j], x, y,
                      gp = gpar(fontsize = 9, col = "black", fontface = "bold"))
        },
        row_names_gp     = gpar(fontsize = 8),
        column_names_gp  = gpar(fontsize = 8),
        column_names_rot = 30,
        cluster_rows     = TRUE,
        cluster_columns  = FALSE,   # keep covariate order meaningful
        show_row_dend    = TRUE,
        show_column_dend = FALSE,
        heatmap_legend_param = list(
            title_gp      = gpar(fontsize = 8, fontface = "bold"),
            labels_gp     = gpar(fontsize = 7),
            legend_height = unit(2.5, "cm")
        ),
        column_title     = title_str,
        column_title_gp  = gpar(fontsize = 10, fontface = "bold")
    )
}
```

```{r stage1-covariate-heatmap, fig.width=7, fig.height=5.5}
plot_covar_heatmap(stage1_covars, "Stage 1: covariate effects on cell-type proportions")
```

```{r stage2-covariate-heatmap, fig.width=7, fig.height=5.5}
plot_covar_heatmap(stage2_covars, "Stage 2: covariate effects on cell-type proportions")
```

---

# Sensitivity: neoplastic cells kept in the denominator

The primary analysis above excludes neoplastic cells, so cell-type fractions are
expressed as a share of the **non-neoplastic (TME) compartment**. That choice is
confounded across entities because neoplastic cells can only be *called* where
tumor cells are aneuploid. Entities with strong aneuploidy (e.g. CN ~85%,
DLGNT ~70% neoplastic-called) have most tumor cells removed from the
denominator, inflating their TME fractions; entities with little detectable
aneuploidy (PA ~6%, RGNT ~2%, PGNT 0%) keep their tumor cells in the
denominator — mislabeled as OPC/Astrocyte — deflating the same TME fractions.
Immune shares are therefore not comparable across entities in the primary
analysis.

This section re-expresses every cell type as a **fraction of all cells**
(neoplastic included in the denominator) so the denominator is the whole tumor
for every entity, regardless of call rate.

The centerpiece is the **descriptive composition heatmap** below: mean per-donor
cell-type fraction for every cell type — including Neoplastic — across entities
and healthy, all on the common all-cell denominator. This is where the immune
signal the user asked about is actually readable and comparable across entities.
The beta-binomial "vs healthy" models that follow are supporting, and carry three
important caveats:

1. **"Vs healthy" odds ratios cannot carry the immune signal.** Healthy brain
   has zero B/T/NK cells, macrophages, monocytes, neutrophils, cDCs, and zero
   neoplastic cells. For those cell types the "vs healthy" comparison is complete
   separation (no denominator on the healthy side), so the model returns a huge
   positive coefficient that is a structural artifact, not enrichment. Read
   immune and neoplastic differences from the descriptive composition heatmap,
   not from the model odds ratios. The vs-healthy heatmap is therefore restricted
   to cell types actually present in healthy brain.
2. **This fixes only half the confound.** Putting neoplastic back in the
   denominator restores cross-entity comparability for immune/vascular/etc.
   fractions. It does **not** un-mislabel the uncalled tumor cells: in
   low-aneuploidy entities those cells remain counted as OPC/Astrocyte, so
   OPC and Astrocyte fractions there stay contaminated upward. Read OPC/
   Astrocyte in low-call entities (PA, RGNT, PGNT, MYXGNT) with caution.
3. **Effect sizes are not comparable to the primary analysis.** Cell types
   present in both arms (neurons, oligodendrocytes, astrocytes) shift toward
   "depleted in tumor" here because tumor units now spend denominator space on
   neoplastic cells that healthy units lack; types absent in healthy stay as
   complete-separation artifacts (see caveat 1). This is a denominator effect,
   not new biology: this section answers "fraction of *all* cells" while the
   primary analysis answers "fraction of the *TME*". `fraction_neoplastic_detected`
   is deliberately omitted as a covariate here — it is collinear with the
   denominator once neoplastic cells are counted.

```{r load-pseudobulk-neo}
# All-cell-denominator parquet (Neoplastic retained; see prepare_metadata.R)
in_path_neo = file.path(sc_dir, "combined_healthyTumor_obs_metadata_withneoplastic.parquet")
df_neo = read_parquet(in_path_neo)
cat("Loaded (with neoplastic):", nrow(df_neo), "cells\n")

# Denominator = ALL cells per unit (this is the correction — neoplastic counted)
unit_totals_neo = df_neo |>
    group_by(donor_proxy, location_broad) |>
    summarise(n_total = n(), .groups = "drop")

# Per-type counts, including Neoplastic (used for the descriptive barplot)
ct_counts_neo = df_neo |>
    group_by(donor_proxy, location_broad, cell_type_harmonised) |>
    summarise(n_ct = n(), .groups = "drop")

# Sample-level covariates, constant within donor_proxy x location_broad
sample_meta_neo = df_neo |>
    group_by(donor_proxy, location_broad) |>
    summarise(
        condition    = first(condition),
        IntDx        = first(IntDx),
        age_yr       = first(age_yr),
        sex_combined = first(sex_combined),
        .groups = "drop"
    )

# Modeled types = non-neoplastic TME (Neoplastic is degenerate vs healthy, shown
# descriptively). n_total still includes neoplastic, so these are all-cell shares.
cell_types_neo = setdiff(sort(unique(df_neo$cell_type_harmonised)), "Neoplastic")

# Cell types with ANY healthy cells — only these give a meaningful "vs healthy"
# odds ratio. Types absent in healthy (peripheral immune, neoplastic) are shown
# in the descriptive composition heatmap instead of the model heatmaps.
healthy_present = df_neo |>
    filter(condition == "healthy") |>
    distinct(cell_type_harmonised) |>
    pull(cell_type_harmonised)

all_units_neo = sample_meta_neo |> select(donor_proxy, location_broad)
all_ct_neo    = tibble(cell_type_harmonised = cell_types_neo)
pseudobulk_neo = cross_join(all_units_neo, all_ct_neo) |>
    left_join(ct_counts_neo,   by = c("donor_proxy", "location_broad", "cell_type_harmonised")) |>
    left_join(unit_totals_neo, by = c("donor_proxy", "location_broad")) |>
    left_join(sample_meta_neo, by = c("donor_proxy", "location_broad")) |>
    mutate(n_ct = replace_na(n_ct, 0))

# Full grid INCLUDING Neoplastic, with per-donor fractions, for the descriptive
# composition heatmap (the models below use the non-neoplastic subset).
all_ct_full = tibble(cell_type_harmonised = sort(unique(df_neo$cell_type_harmonised)))
pseudobulk_full = cross_join(all_units_neo, all_ct_full) |>
    left_join(ct_counts_neo,   by = c("donor_proxy", "location_broad", "cell_type_harmonised")) |>
    left_join(unit_totals_neo, by = c("donor_proxy", "location_broad")) |>
    left_join(sample_meta_neo, by = c("donor_proxy", "location_broad")) |>
    mutate(n_ct = replace_na(n_ct, 0), frac = n_ct / n_total)

# Entity filter (shared by the descriptive heatmap and Stage 2): same tumor units
# as the primary analysis, so the same entities pass the min-sample threshold.
entity_counts_neo = pseudobulk_neo |>
    filter(condition == "tumor") |>
    distinct(donor_proxy, location_broad, IntDx) |>
    count(IntDx, name = "n_units")
entities_kept_neo = entity_counts_neo |> filter(n_units >= MIN_SAMPLES_PER_ENTITY) |>
    pull(IntDx) |> as.character()

cat("Pseudobulk units (with-neoplastic):", nrow(all_units_neo), "\n")
cat("Modeled cell types:", paste(cell_types_neo, collapse = ", "), "\n")
cat("Present in healthy (modelable vs healthy):", paste(sort(healthy_present), collapse = ", "), "\n")
```

```{r pseudobulk-neo-verify}
#| echo: FALSE
# Sanity checks: only tumor denominators should change relative to the primary
# analysis; the unit set and healthy denominators must be identical.
denom_cmp = unit_totals |>
    rename(n_total_main = n_total) |>
    full_join(unit_totals_neo |> rename(n_total_neo = n_total),
              by = c("donor_proxy", "location_broad")) |>
    left_join(sample_meta_neo |> select(donor_proxy, location_broad, condition),
              by = c("donor_proxy", "location_broad"))

cat("Healthy units with changed denominator (expect 0):",
    sum(denom_cmp$condition == "healthy" &
        denom_cmp$n_total_main != denom_cmp$n_total_neo, na.rm = TRUE), "\n")
cat("Tumor units whose denominator grew (neoplastic added):",
    sum(denom_cmp$condition == "tumor" &
        denom_cmp$n_total_neo > denom_cmp$n_total_main, na.rm = TRUE), "\n")
cat("Units present in one parquet only (expect 0):",
    sum(is.na(denom_cmp$n_total_main) | is.na(denom_cmp$n_total_neo)), "\n")
cat("Tumor unit count — main:", sum(sample_meta$condition == "tumor"),
    " | with-neoplastic:", sum(sample_meta_neo$condition == "tumor"), "\n")
```

## Neoplastic proportion per entity (descriptive)

Fraction of all sequenced cells called Neoplastic, per entity. This is the
call-rate variable that drives the primary-analysis confound; it is shown
descriptively because a model coefficient vs healthy (structural zero) is
uninformative.

```{r neoplastic-fraction-barplot, fig.width=6, fig.height=4}
neo_frac_entity = df_neo |>
    filter(condition == "tumor") |>
    group_by(IntDx) |>
    summarise(
        n_cells = n(),
        n_neo   = sum(cell_type_harmonised == "Neoplastic"),
        frac    = n_neo / n_cells,
        .groups = "drop"
    ) |>
    arrange(desc(frac))

ggplot(neo_frac_entity, aes(x = reorder(IntDx, frac), y = frac)) +
    geom_col(fill = "#8E44AD", width = 0.7) +
    geom_text(aes(label = scales::percent(frac, accuracy = 1)),
              hjust = -0.15, size = 2.8) +
    coord_flip(ylim = c(0, 1)) +
    labs(
        x = NULL, y = "Fraction of cells called Neoplastic",
        title = "Neoplastic call rate varies widely across entities",
        subtitle = "Low-call entities keep tumor cells mislabeled in the denominator"
    ) +
    theme_nature(base_size = 9)
```

## Cell-type composition per entity (all-cell denominator, descriptive)

The core view. Each cell is counted against the whole-tumor denominator, so
fractions are directly comparable across entities and healthy — **Neoplastic
included as its own category**. Cells are the mean per-donor fraction within each
group (donor-weighted, matching the pseudobulk unit). Text is the actual mean
percentage; color is row-scaled (z-score across columns) so relative differences
are visible even for rare types like lymphocytes. This is the figure to read for
immune and neoplastic differences across entities.

```{r composition-heatmap-neo, fig.width=9, fig.height=7}
# Mean per-donor fraction per (group x cell type). group = Healthy or IntDx.
group_frac = pseudobulk_full |>
    mutate(group = if_else(condition == "healthy", "Healthy", as.character(IntDx))) |>
    filter(group == "Healthy" | group %in% entities_kept_neo) |>
    group_by(group, cell_type_harmonised) |>
    summarise(mean_frac = mean(frac), .groups = "drop")

# Fraction matrix: rows = cell types, cols = Healthy + entities.
frac_mat = group_frac |>
    mutate(cell_type_harmonised = str_replace_all(cell_type_harmonised, "_", " ")) |>
    pivot_wider(names_from = group, values_from = mean_frac) |>
    tibble::column_to_rownames("cell_type_harmonised") |>
    as.matrix()

# Column order: Healthy first, then entities by descending neoplastic fraction
# (so the call-rate gradient driving the confound reads left-to-right).
# IntDx is a factor (arrow dictionary) — coerce to character before c() with
# "Healthy", else the factor collapses to integer codes.
entity_order = neo_frac_entity |> filter(IntDx %in% entities_kept_neo) |>
    pull(IntDx) |> as.character()
col_order    = c("Healthy", entity_order)
frac_mat     = frac_mat[, col_order[col_order %in% colnames(frac_mat)], drop = FALSE]

# Row-scale (z-score across columns) for color; zero-variance rows -> 0.
frac_z = t(scale(t(frac_mat)))
frac_z[is.nan(frac_z)] = 0

z_col = colorRamp2(c(-1.5, 0, 1.5), c("#2166AC", "white", "#B2182B"))

Heatmap(
    frac_z,
    name             = "row z-score\n(fraction)",
    col              = z_col,
    cell_fun         = function(j, i, x, y, width, height, fill) {
        grid.text(scales::percent(frac_mat[i, j], accuracy = 0.1), x, y,
                  gp = gpar(fontsize = 6, col = "black"))
    },
    row_names_gp     = gpar(fontsize = 8),
    column_names_gp  = gpar(fontsize = 8, fontface = "bold"),
    column_names_rot = 45,
    cluster_rows     = TRUE,
    cluster_columns  = FALSE,   # keep Healthy-first, call-rate-ordered columns
    show_row_dend    = TRUE,
    show_column_dend = FALSE,
    heatmap_legend_param = list(
        title_gp      = gpar(fontsize = 8, fontface = "bold"),
        labels_gp     = gpar(fontsize = 7),
        legend_height = unit(3, "cm")
    ),
    column_title     = "Mean per-donor cell-type fraction (all cells) — Neoplastic included",
    column_title_gp  = gpar(fontsize = 10, fontface = "bold"),
    row_title_gp     = gpar(fontsize = 9)
)
```

## Stage 1 (all-cell denominator): tumor vs healthy

```{r stage1-neo-model}
# Same formula/reference as the primary Stage 1, but n_total now includes
# neoplastic cells. Reuses the fit_betabin helper.
pseudobulk_neo_s1 = pseudobulk_neo |>
    mutate(condition = relevel(factor(condition), ref = "healthy"))

stage1_neo_raw = map_dfr(cell_types_neo, \(ct)
    fit_betabin(
        ct          = ct,
        formula_str = "cbind(n_ct, n_other) ~ condition + age_yr + location_broad + sex_combined",
        dat         = pseudobulk_neo_s1
    )
)

stage1_neo = stage1_neo_raw |>
    filter(converged, !is.na(p.value), term == "conditiontumor") |>
    mutate(
        q.value   = p.adjust(p.value, method = "BH"),
        sig       = q.value < 0.1,
        direction = if_else(estimate > 0, "Enriched in tumor", "Depleted in tumor")
    )

cat("Stage 1 (neo) convergence failures:",
    sum(!stage1_neo_raw$converged | is.na(stage1_neo_raw$p.value)), "/", length(cell_types_neo), "\n")
print(stage1_neo |> select(cell_type, estimate, std.error, p.value, q.value, sig) |>
      arrange(q.value))

write.table(stage1_neo, file.path(code_dir, "results_stage1_tumor_vs_healthy_withneoplastic.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r stage1-neo-plot, fig.width=5, fig.height=5}
# Forest plot mirrors the primary Stage 1; note the shift toward "depleted in
# tumor" is expected (denominator now includes neoplastic cells).
sig_colors = c("Enriched in tumor" = "#C0392B", "Depleted in tumor" = "#2980B9", "n.s." = "gray60")

stage1_neo_plot = stage1_neo |>
    mutate(
        sig_label = if_else(sig, direction, "n.s."),
        cell_type = str_replace_all(cell_type, "_", " "),
        cell_type = reorder(cell_type, estimate)
    )

ggplot(stage1_neo_plot, aes(x = estimate, y = cell_type, color = sig_label)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.4) +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                   height = 0.25, linewidth = 0.5, alpha = 0.7) +
    geom_point(size = 2.2) +
    scale_color_manual(values = sig_colors, name = NULL) +
    scale_x_continuous(expand = expansion(mult = 0.05)) +
    coord_cartesian(xlim = c(-30, 30)) +
    labs(
        x        = "Log odds-ratio vs. healthy (beta-binomial)",
        y        = NULL,
        title    = "Cell-type proportions of ALL cells: tumor vs. healthy",
        subtitle = "Denominator includes neoplastic  |  adjusted for age, sex, location  |  BH q < 0.1"
    ) +
    theme_nature(base_size = 9) +
    theme(legend.position = "bottom")
```

## Stage 2 (all-cell denominator): each entity vs healthy

Formal per-entity odds ratios vs healthy, on the all-cell denominator. **Restricted
to cell types present in healthy brain** — peripheral-immune and neoplastic types
are structural zeros in healthy (complete separation) and belong to the
descriptive composition heatmap above, not here. OPC/Astrocyte rows in low-call
entities (see barplot) remain contaminated upward and should not be over-interpreted.

```{r stage2-neo-model}
# entities_kept_neo / entity_counts_neo computed in the load chunk above.
pseudobulk_neo_s2 = pseudobulk_neo |>
    filter(is.na(IntDx) | IntDx %in% entities_kept_neo) |>
    mutate(
        group = if_else(condition == "healthy", "Healthy", as.character(IntDx)),
        group = relevel(factor(group), ref = "Healthy")
    )

stage2_neo_raw = map_dfr(cell_types_neo, \(ct)
    fit_betabin(
        ct          = ct,
        formula_str = "cbind(n_ct, n_other) ~ group + age_yr + location_broad + sex_combined",
        dat         = pseudobulk_neo_s2
    )
)

stage2_neo = stage2_neo_raw |>
    filter(converged, !is.na(p.value),
           str_detect(term, "^group"), term != "groupHealthy") |>
    mutate(
        entity  = str_remove(term, "^group"),
        q.value = p.adjust(p.value, method = "BH"),
        sig     = q.value < 0.1
    )

cat("Stage 2 (neo) convergence failures:",
    sum(!stage2_neo_raw$converged | is.na(stage2_neo_raw$p.value)),
    "/", length(cell_types_neo) * length(entities_kept_neo), "\n")

write.table(stage2_neo, file.path(code_dir, "results_stage2_by_entity_withneoplastic.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r stage2-neo-heatmap, fig.width=8, fig.height=6}
# Heatmap mirrors the primary Stage 2 (log-OR per cell_type x entity, capped ±3),
# but restricted to cell types present in healthy so ORs are not complete-
# separation artifacts. Immune / neoplastic are in the composition heatmap above.
hm_data_neo = stage2_neo |>
    filter(cell_type %in% healthy_present) |>
    select(cell_type, entity, estimate, q.value) |>
    mutate(cell_type = str_replace_all(cell_type, "_", " "))

est_mat_neo = hm_data_neo |>
    select(cell_type, entity, estimate) |>
    pivot_wider(names_from = entity, values_from = estimate) |>
    tibble::column_to_rownames("cell_type") |>
    as.matrix()

# Drop any all-NA row/column (a cell type that failed to converge everywhere)
# so ComplexHeatmap's row clustering has a valid distance matrix.
est_mat_neo = est_mat_neo[rowSums(!is.na(est_mat_neo)) > 0,
                          colSums(!is.na(est_mat_neo)) > 0, drop = FALSE]

sig_mat_neo = hm_data_neo |>
    select(cell_type, entity, q.value) |>
    pivot_wider(names_from = entity, values_from = q.value) |>
    tibble::column_to_rownames("cell_type") |>
    as.matrix()
sig_mat_neo = sig_mat_neo[rownames(est_mat_neo), colnames(est_mat_neo)]

sig_annot_neo = matrix("", nrow = nrow(sig_mat_neo), ncol = ncol(sig_mat_neo),
                       dimnames = dimnames(sig_mat_neo))
sig_annot_neo[!is.na(sig_mat_neo) & sig_mat_neo < 0.01] = "**"
sig_annot_neo[!is.na(sig_mat_neo) & sig_mat_neo >= 0.01 & sig_mat_neo < 0.1] = "*"

cap_val = 3
col_fun = colorRamp2(c(-cap_val, 0, cap_val), c("#2166AC", "white", "#B2182B"))

n_entity_neo = entity_counts_neo |>
    filter(IntDx %in% colnames(est_mat_neo)) |>
    arrange(match(IntDx, colnames(est_mat_neo)))

col_annot_neo = HeatmapAnnotation(
    "n units" = anno_barplot(
        n_entity_neo$n_units,
        gp         = gpar(fill = "gray40", col = NA),
        height     = unit(1.5, "cm"),
        axis_param = list(gp = gpar(fontsize = 7))
    ),
    annotation_name_gp = gpar(fontsize = 8)
)

Heatmap(
    est_mat_neo,
    name             = "log-OR\nvs. healthy",
    col              = col_fun,
    na_col           = "gray90",
    cell_fun         = function(j, i, x, y, width, height, fill) {
        grid.text(sig_annot_neo[i, j], x, y,
                  gp = gpar(fontsize = 9, col = "black", fontface = "bold"))
    },
    top_annotation   = col_annot_neo,
    row_names_gp     = gpar(fontsize = 8),
    column_names_gp  = gpar(fontsize = 8, fontface = "bold"),
    column_names_rot = 45,
    cluster_rows     = TRUE,
    cluster_columns  = TRUE,
    show_row_dend    = TRUE,
    show_column_dend = FALSE,
    heatmap_legend_param = list(
        title_gp      = gpar(fontsize = 8, fontface = "bold"),
        labels_gp     = gpar(fontsize = 7),
        legend_height = unit(3, "cm")
    ),
    column_title     = "All-cell-denominator enrichment per entity vs. healthy",
    column_title_gp  = gpar(fontsize = 10, fontface = "bold"),
    row_title_gp     = gpar(fontsize = 9)
)
```

---

# Session info

```{r session-info}
#| code-fold: false
sessionInfo()
```
