---
title: ""
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 6
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk:
        results: hold
        R.options:
            width: 120
---

# Introduction

This notebook summarizes the exploratory DLGNT subtype pseudobulk analysis implemented in the colocated scripts:

```bash
bash sample_comparisons/dlgnt12_expression/run_dlgnt12_expression.sh
```

Equivalent manual steps:

```bash
conda run -n spatial python sample_comparisons/dlgnt12_expression/dlgnt12_prepare.py
conda run -n spatial Rscript sample_comparisons/dlgnt12_expression/dlgnt12_limma_fgsea.R
quarto render sample_comparisons/dlgnt12_expression/dlgnt12_expression.qmd
```

The prepare step resolves the single-cell H5AD from the same HOME-based `workdir` convention used in the repo’s other single-nucleus analyses:

- local mac: `~/Documents/HMS/PhD/beroukhimlab/dfci_mount/`
- `/PHShome/yz762`: `/data/beroukhim1/`
- `/home/yz762`: `/mnt/storage/dept/medonc/beroukhim/`
- fallback: `/xchip/beroukhimlab/`

The input file is always taken as:

```text
<workdir>/youyun/plgg/data/single_cell/Extended_sndata_filtered.csv_subset.h5ad
```

The primary unit of inference is the sample, not the cell. Raw counts are aggregated to one pseudobulk profile per `sample x cell_type`, differential expression is run with `edgeR`/`voom`/`limma`, and Hallmark `fgsea` is run on the post-`filterByExpr` tested-gene universe for each cell type.

```{r}
#| echo: false
library(data.table)
library(ggplot2)
library(knitr)

# The notebook only reads colocated analysis outputs, so keep all paths relative
# to this file instead of re-deriving workdir or raw-data locations.
analysis_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
handoff_dir <- file.path(analysis_dir, "handoff")
results_dir <- file.path(analysis_dir, "results")

required_files <- c(
    file.path(analysis_dir, "prepare_metadata.tsv"),
    file.path(analysis_dir, "annotation_level_support_summary.tsv"),
    file.path(analysis_dir, "annotation_level_support_detail.tsv"),
    file.path(analysis_dir, "dlgnt_sample_metadata.tsv"),
    file.path(analysis_dir, "eligible_cell_types.tsv"),
    file.path(results_dir, "cell_type_analysis_summary.tsv"),
    file.path(results_dir, "results_metadata.tsv")
)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0) {
    stop(
        "Missing analysis outputs. Run dlgnt12_prepare.py and dlgnt12_limma_fgsea.R first.\n",
        paste(missing_files, collapse = "\n")
    )
}

prepare_meta <- fread(file.path(analysis_dir, "prepare_metadata.tsv"))
results_meta <- fread(file.path(results_dir, "results_metadata.tsv"))
support_summary <- fread(file.path(analysis_dir, "annotation_level_support_summary.tsv"))
support_detail <- fread(file.path(analysis_dir, "annotation_level_support_detail.tsv"))
sample_meta <- fread(file.path(analysis_dir, "dlgnt_sample_metadata.tsv"))
eligible_dt <- fread(file.path(analysis_dir, "eligible_cell_types.tsv"))
analysis_summary <- fread(file.path(results_dir, "cell_type_analysis_summary.tsv"))

meta_value <- function(dt, meta_key) {
    # Metadata tables are stored in long format, so this helper keeps the report
    # code readable when pulling one value at a time into prose.
    hit <- dt[key == meta_key, value]
    if (length(hit) < 1) {
        return(NA_character_)
    }
    hit[[1]]
}

clean_hallmark_label <- function(x) {
    label <- gsub("^HALLMARK_", "", x)
    label <- gsub("_", " ", label)
    tools::toTitleCase(tolower(label))
}
```

# Run metadata

This run uses `r meta_value(prepare_meta, "cell_type_column")` as the default cell-type definition. The DLGNT cohort contains `r meta_value(prepare_meta, "n_dlgnt_samples")` samples and `r format(as.integer(meta_value(prepare_meta, "n_dlgnt_cells")), big.mark = ",")` cells across `DLGNT_1` and `DLGNT_2`.

The subtype comparison is exploratory because all `DLGNT_1` samples are spinal in this dataset, while `DLGNT_2` includes spinal, supratentorial, and pending-location samples. The subtype effect is therefore not separable from location.

```{r}
kable(prepare_meta)
kable(results_meta)
```

# Cohort support

```{r}
kable(sample_meta)
```

## Cell-type support across annotation levels

```{r}
kable(support_summary)
```

```{r}
ggplot(
    support_summary,
    aes(
        x = factor(sample_threshold_per_subtype),
        y = n_cell_types,
        fill = annotation_level
    )
) +
    geom_col(position = "dodge") +
    theme_minimal(base_size = 16) +
    labs(
        x = "Minimum samples per subtype",
        y = "Retained cell types",
        fill = "Annotation level",
        title = "Cell-type support across annotation levels"
    )
```

```{r}
kable(support_detail)
```

# Analyzed cell types

```{r}
# Order by available replication first, then tested-gene count, so the largest
# and best-supported cell types appear near the top of the summary table.
kable(analysis_summary[
    order(-n_samples_total, -n_tested_genes, cell_type),
    .(
        cell_type, n_samples_dlgnt1, n_samples_dlgnt2,
        total_cells_dlgnt1, total_cells_dlgnt2,
        n_genes_input, n_tested_genes, n_hallmark_pathways_retained
    )
])
```

```{r}
# Reshape to long format so the same plotting code can compare support, tested
# genes, and retained Hallmark pathways across cell types.
plot_dt <- melt(
    analysis_summary[
        ,
        .(cell_type, n_genes_input, n_tested_genes, n_hallmark_pathways_retained)
    ],
    id.vars = "cell_type",
    variable.name = "metric",
    value.name = "value"
)

ggplot(plot_dt, aes(x = reorder(cell_type, value), y = value, fill = metric)) +
    geom_col(position = "dodge") +
    coord_flip() +
    theme_minimal(base_size = 16) +
    labs(
        x = "Cell type",
        y = "Count",
        fill = "Metric",
        title = "Genes and Hallmark pathways retained per analyzed cell type"
    )
```

# Top differential expression

```{r}
# Load one DE table per analyzed cell type and combine them for shared ranking
# and plotting logic in the report.
de_list <- lapply(analysis_summary$cell_type_safe, function(cell_type_safe) {
    path <- file.path(results_dir, paste0(cell_type_safe, "_de.tsv"))
    if (!file.exists(path)) {
        return(data.table())
    }
    fread(path)
})
de_dt <- rbindlist(de_list, fill = TRUE)

# Within each cell type, rank by FDR first and use absolute moderated t as a
# tie-breaker so the table favors both significance and effect strength.
top_de <- de_dt[
    order(cell_type, fdr, -abs(t_stat)),
    head(.SD, 10),
    by = cell_type
]
kable(top_de)
```

```{r}
# Use a slightly smaller top-hit set for plotting so crowded facets stay legible.
top_de_plot <- de_dt[
    order(cell_type, fdr, -abs(t_stat)),
    head(.SD, 8),
    by = cell_type
]

ggplot(
    top_de_plot,
    aes(
        x = reorder(gene, t_stat),
        y = t_stat,
        fill = logFC_dlgnt2_vs_dlgnt1 > 0
    )
) +
    geom_col() +
    facet_wrap(~ cell_type, scales = "free_y") +
    coord_flip() +
    theme_minimal(base_size = 14) +
    scale_fill_manual(values = c("#1f77b4", "#d62728"), guide = "none") +
    labs(
        x = "Gene",
        y = "Moderated t statistic",
        title = "Top differential-expression hits by cell type"
    )
```

# Significant Hallmark pathways from `DLGNT_2 > DLGNT_1`

Positive NES values indicate pathways enriched in `DLGNT_2`. Negative NES values indicate pathways enriched on the `DLGNT_1` side within the same `DLGNT_2 > DLGNT_1` ranking.

```{r}
# As with DE, collect the per-cell-type GSEA outputs into one table for report
# summarization without rerunning the analysis.
gsea_list <- lapply(analysis_summary$cell_type_safe, function(cell_type_safe) {
    path <- file.path(results_dir, paste0(cell_type_safe, "_gsea.tsv"))
    if (!file.exists(path)) {
        return(data.table())
    }
    fread(path)
})
gsea_dt <- rbindlist(gsea_list, fill = TRUE)

# Use a single contrast to avoid mirrored duplicate results. The NES sign then
# carries direction: positive means DLGNT_2 up and negative means DLGNT_2 down.
sig_gsea_dlgnt2 <- gsea_dt[
    contrast == "DLGNT_2 > DLGNT_1" &
        !is.na(padj) &
        padj < 0.05
]
if (nrow(sig_gsea_dlgnt2) > 0) {
    sig_gsea_dlgnt2[, set_name_clean := clean_hallmark_label(set_name)]
    sig_gsea_dlgnt2[, direction := fifelse(NES > 0, "DLGNT_2 up", "DLGNT_2 down")]
    sig_gsea_table <- sig_gsea_dlgnt2[
        order(cell_type, padj, -abs(NES)),
        .(cell_type, set_name_clean, direction, NES, pval, padj, size)
    ]
    setnames(sig_gsea_table, "set_name_clean", "hallmark_pathway")
    kable(sig_gsea_table)
} else {
    cat("No significant Hallmark pathways reached padj < 0.05 in the DLGNT_2 > DLGNT_1 contrast.\n")
}
```

```{r}
#| results: asis
#| fig-height: 6
cell_types_for_gsea_plots <- analysis_summary[order(cell_type), unique(cell_type)]

if (length(cell_types_for_gsea_plots) > 0) {
    for (cell_type_name in cell_types_for_gsea_plots) {
        cell_dt <- copy(sig_gsea_dlgnt2[cell_type == cell_type_name][order(NES, padj)])

        if (nrow(cell_dt) < 1) {
            cat("### ", cell_type_name, "\n\n", sep = "")
            cat("No significant Hallmark pathways reached padj < 0.05 in the DLGNT_2 > DLGNT_1 contrast.\n\n")
            next
        }

        # Order pathways along the y-axis by NES so each panel reads cleanly from
        # DLGNT_2-down programs to DLGNT_2-up programs around the zero line.
        cell_dt[, pathway_plot := factor(set_name_clean, levels = set_name_clean)]
        cell_dt[, direction := fifelse(NES > 0, "DLGNT_2 up", "DLGNT_2 down")]

        cat("### ", cell_type_name, "\n\n", sep = "")
        print(
            ggplot(
                cell_dt,
                aes(
                    x = NES,
                    y = pathway_plot,
                    size = -log10(padj),
                    color = direction
                )
            ) +
                geom_vline(xintercept = 0, linetype = "dashed", color = "grey75") +
                geom_point(alpha = 0.85) +
                scale_color_manual(values = c("DLGNT_2 down" = "#1f77b4", "DLGNT_2 up" = "#d62728")) +
                scale_size_continuous(range = c(2.5, 7)) +
                theme_minimal(base_size = 13) +
                theme(
                    panel.grid.major.y = element_blank(),
                    axis.text.y = element_text(size = 10),
                    legend.position = "bottom"
                ) +
                labs(
                    x = "Normalized enrichment score",
                    y = "Hallmark pathway",
                    size = "-log10(padj)",
                    color = "Direction",
                    title = paste0(cell_type_name, ": significant pathways in DLGNT_2 > DLGNT_1")
                )
        )
        cat("\n\n")
    }
} else {
    cat("No significant Hallmark pathways reached padj < 0.05 in the DLGNT_2 > DLGNT_1 contrast.\n")
}
```
