---
title: "DLGNT_1 vs DLGNT_2: stress-adjustment comparison across 4 runs"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 14
        fig-height: 7
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk:
        results: hold
        R.options:
            width: 120
---

# Introduction

This notebook compares four configurations of the DLGNT_1 vs DLGNT_2 pseudobulk DE + Hallmark fgsea
pipeline to quantify the impact of (a) dropping sample 267134 and (b) adding a per-pseudobulk IEG
dissociation stress score as a covariate.

Render command:
```bash
conda run -n spatial quarto render sample_comparisons/dlgnt12_expression/dlgnt12_expression_stress_comparison.qmd
```

The four runs compared here are:

| Label | `--drop-samples` | `--no-stress-covariate` | `--keep-iegs-in-gsea` |
|---|---|---|---|
| Baseline | — | yes | yes |
| Drop 267134 | 267134 | yes | yes |
| Stress-adjusted | — | — | — |
| Stress-adj + Drop 267134 | 267134 | — | — |

The baseline is equivalent to the original `results/` pipeline (apart from floating-point non-determinism
at the 14th decimal place from MKL threading). All four runs read the same `handoff/` TSVs produced by
`dlgnt12_prepare.py`; no re-aggregation is needed.

**Interpretation caveat**: DLGNT_1 = 100% spinal and DLGNT_2 = mixed location, so subtype and location
remain confounded in every run. Stress adjustment removes one nuisance axis (per-sample handling
artifact), not the location axis.

```{r setup}
#| echo: false
library(data.table)
library(ggplot2)
library(knitr)

analysis_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)

# Run configurations in the order they should appear in all comparison plots.
run_configs <- list(
    list(
        label  = "Baseline",
        short  = "baseline",
        dir    = file.path(analysis_dir, "results_sanity_baseline"),
        color  = "#888888"
    ),
    list(
        label  = "Drop 267134",
        short  = "drop_267134",
        dir    = file.path(analysis_dir, "results_drop_267134"),
        color  = "#4daf4a"
    ),
    list(
        label  = "Stress-adjusted",
        short  = "stress_adj",
        dir    = file.path(analysis_dir, "results_stress_adjusted"),
        color  = "#e41a1c"
    ),
    list(
        label  = "Stress-adj + Drop 267134",
        short  = "stress_drop",
        dir    = file.path(analysis_dir, "results_stress_adjusted_drop_267134"),
        color  = "#984ea3"
    )
)

run_labels  <- sapply(run_configs, `[[`, "label")
run_colors  <- setNames(sapply(run_configs, `[[`, "color"), run_labels)

# Verify all result directories exist before loading.
missing_dirs <- sapply(run_configs, function(cfg) !dir.exists(cfg$dir))
if (any(missing_dirs)) {
    stop(
        "Missing result directories. Run dlgnt12_limma_fgsea_stress_adjusted.R for each configuration first.\n",
        paste(sapply(run_configs[missing_dirs], `[[`, "dir"), collapse = "\n")
    )
}

# Helper: read a TSV from a results directory; returns an empty data.table on miss.
load_tsv <- function(results_dir, filename) {
    path <- file.path(results_dir, filename)
    if (!file.exists(path)) return(data.table())
    fread(path)
}

clean_hallmark_label <- function(x) {
    label <- gsub("^HALLMARK_", "", x)
    label <- gsub("_", " ", label)
    tools::toTitleCase(tolower(label))
}

# Cell-type roster comes from the baseline run; all four runs use the same cell types.
ct_map <- load_tsv(run_configs[[1]]$dir, "cell_type_analysis_summary.tsv")[
    , .(cell_type, cell_type_safe)
]
cell_types <- ct_map$cell_type

# Load analysis summaries for all runs and stack with a run label.
summary_dt <- rbindlist(lapply(run_configs, function(cfg) {
    dt <- load_tsv(cfg$dir, "cell_type_analysis_summary.tsv")
    if (nrow(dt) > 0) dt[, run := cfg$label]
    dt
}), fill = TRUE)
summary_dt[, run := factor(run, levels = run_labels)]

# Load GSEA results for all runs across all cell types.
gsea_dt <- rbindlist(lapply(run_configs, function(cfg) {
    per_ct <- rbindlist(lapply(ct_map$cell_type_safe, function(cts) {
        load_tsv(cfg$dir, paste0(cts, "_gsea.tsv"))
    }), fill = TRUE)
    if (nrow(per_ct) > 0) per_ct[, run := cfg$label]
    per_ct
}), fill = TRUE)
gsea_dt[, run           := factor(run, levels = run_labels)]
gsea_dt[, set_name_clean := clean_hallmark_label(set_name)]

# Load DE results for all runs across all cell types.
de_dt <- rbindlist(lapply(run_configs, function(cfg) {
    per_ct <- rbindlist(lapply(ct_map$cell_type_safe, function(cts) {
        load_tsv(cfg$dir, paste0(cts, "_de.tsv"))
    }), fill = TRUE)
    if (nrow(per_ct) > 0) per_ct[, run := cfg$label]
    per_ct
}), fill = TRUE)
de_dt[, run := factor(run, levels = run_labels)]

# Per-pseudobulk stress diagnostics come from the stress-adjusted run only.
# The baseline and drop-only runs write the same columns but with stress_score = NA
# when --no-stress-covariate is set, so we read the canonical stress-adjusted version.
stress_diag <- load_tsv(
    run_configs[[which(sapply(run_configs, `[[`, "short") == "stress_adj")]]$dir,
    "stress_adjustment_diagnostics.tsv"
)
```

# Run metadata

```{r run-meta}
meta_dt <- rbindlist(lapply(run_configs, function(cfg) {
    dt <- load_tsv(cfg$dir, "results_metadata.tsv")
    if (nrow(dt) > 0) dt[, run := cfg$label]
    dt
}), fill = TRUE)

# Pivot to wide so each row is a metadata key and each column is a run.
meta_wide <- dcast(meta_dt, key ~ run, value.var = "value")
kable(meta_wide, caption = "Key metadata fields across all four runs")
```

# Stress score diagnostics

These plots use the per-pseudobulk data from the stress-adjusted run to establish whether the
IEG stress signal is real and how collinear it is with the DLGNT subtype label.

## Per-pseudobulk IEG stress score by subtype

DLGNT_1 pseudobulks should sit at a higher (less negative) stress score than DLGNT_2 pseudobulks
if sample-level handling/dissociation is the driver. Cell types where the two groups overlap
are less interpretable.

```{r stress-scatter}
#| fig-height: 8
ggplot(stress_diag, aes(x = IntDx, y = stress_score, color = IntDx)) +
    geom_jitter(width = 0.12, size = 2.8, alpha = 0.85) +
    facet_wrap(~ cell_type, ncol = 3) +
    scale_color_manual(
        values = c("DLGNT_1" = "#1f77b4", "DLGNT_2" = "#d62728"),
        guide  = "none"
    ) +
    theme_minimal(base_size = 13) +
    labs(
        x     = "Subtype",
        y     = "Stress score [log10(IEG fraction + 1e-9)]",
        title = "Per-pseudobulk IEG stress score by subtype"
    )
```

## Stress–subtype correlation per cell type

|Pearson r| > 0.9 (flagged as `adjustment_unstable = TRUE`) means stress and IntDx are too collinear
for the model to reliably separate them; adjusted GSEA results for those cell types should be treated
with extra scepticism.

```{r stress-corr-table}
# Correlation metrics repeat per-row within a cell type, so take one row per cell type.
diag_summary <- stress_diag[
    , .(
        pearson_r          = unique(stress_intdx_pearson),
        t_test_p           = unique(stress_intdx_t_pvalue),
        adjustment_unstable = unique(adjustment_unstable)
    ),
    by = cell_type
][order(-abs(pearson_r))]

kable(diag_summary, digits = 4, caption = "Stress–IntDx correlation per cell type")
```

## IEG fraction by subtype per cell type

The raw fraction (not log-scaled) shows the magnitude of the processing difference more intuitively.

```{r ieg-fraction-box}
#| fig-height: 7
ggplot(stress_diag, aes(x = IntDx, y = ieg_fraction, color = IntDx)) +
    geom_boxplot(outlier.shape = NA, width = 0.4) +
    geom_jitter(width = 0.1, size = 2.2, alpha = 0.7) +
    facet_wrap(~ cell_type, ncol = 3, scales = "free_y") +
    scale_color_manual(
        values = c("DLGNT_1" = "#1f77b4", "DLGNT_2" = "#d62728"),
        guide  = "none"
    ) +
    theme_minimal(base_size = 13) +
    labs(
        x     = "Subtype",
        y     = "IEG fraction (IEG counts / total counts)",
        title = "Per-pseudobulk IEG fraction by subtype"
    )
```

# Significant pathway counts across runs

The heatmap below shows how many Hallmark pathways reach padj < 0.05 in the DLGNT_2 > DLGNT_1 contrast
for each cell type × run combination. Large drops from left to right indicate stress-driven inflation
in the baseline.

```{r sig-count-heatmap}
sig_counts <- gsea_dt[
    contrast == "DLGNT_2 > DLGNT_1" & !is.na(padj) & padj < 0.05,
    .N,
    by = .(cell_type, run)
]

# Fill zeros for cell-type × run combinations with no significant pathways.
all_combos <- CJ(
    cell_type = cell_types,
    run       = factor(run_labels, levels = run_labels)
)
sig_counts <- merge(all_combos, sig_counts, by = c("cell_type", "run"), all.x = TRUE)
sig_counts[is.na(N), N := 0L]

ggplot(sig_counts, aes(x = run, y = cell_type, fill = N)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = N), size = 5.5, fontface = "bold") +
    scale_fill_gradient(low = "#f7fbff", high = "#08306b") +
    theme_minimal(base_size = 14) +
    theme(
        axis.text.x     = element_text(angle = 20, hjust = 1),
        panel.grid      = element_blank(),
        legend.position = "right"
    ) +
    labs(
        x    = NULL,
        y    = "Cell type",
        fill = "# sig\npathways",
        title = "Significant Hallmark pathways per cell type × run (padj < 0.05, DLGNT_2 > DLGNT_1)"
    )
```

# Per-cell-type pathway dot grid

Each panel shows all Hallmark pathways that are significant (padj < 0.05) in **at least one run**.
Dot size = −log10(padj); grey dots = pathway not significant in that run; color = enrichment direction.
This makes it easy to see which pathways are robust vs artifacts.

```{r dot-grid}
#| results: asis
#| fig-height: 9
for (ct in cell_types) {
    # Union of significant pathways for this cell type across all runs.
    sig_union <- gsea_dt[
        contrast == "DLGNT_2 > DLGNT_1" & cell_type == ct &
            !is.na(padj) & padj < 0.05,
        unique(set_name_clean)
    ]

    cat("## ", ct, "\n\n", sep = "")

    if (length(sig_union) == 0) {
        cat("No significant Hallmark pathways in any run.\n\n")
        next
    }

    # Pull GSEA rows for those pathways in all runs (even if not significant).
    plot_dt <- gsea_dt[
        contrast == "DLGNT_2 > DLGNT_1" & cell_type == ct &
            set_name_clean %in% sig_union
    ]

    # For pathways present in the union but absent in a run, there will simply be
    # no row -- that is fine; ggplot omits them, leaving a visual gap.
    plot_dt[, is_sig := !is.na(padj) & padj < 0.05]
    # Replace NA neg_log10_padj with 0 so scale_size_continuous works without na.value
    # (not supported in older ggplot2); size=0 points are still drawn via color to show presence.
    plot_dt[, neg_log10_padj := fifelse(is_sig, -log10(padj), 0)]
    plot_dt[, direction := fcase(
        is_sig & NES > 0,  "DLGNT_2 up",
        is_sig & NES <= 0, "DLGNT_2 down",
        default            = "not sig (padj ≥ 0.05)"
    )]

    # Order pathways by mean NES across all runs so the plot reads DLGNT_2-down
    # at the bottom and DLGNT_2-up at the top.
    path_order <- plot_dt[
        , .(mean_nes = mean(NES, na.rm = TRUE)),
        by = set_name_clean
    ][order(mean_nes), set_name_clean]
    plot_dt[, pathway_ord := factor(set_name_clean, levels = path_order)]

    n_pathways <- length(sig_union)

    print(
        ggplot(
            plot_dt,
            aes(x = run, y = pathway_ord, size = neg_log10_padj, color = direction)
        ) +
            geom_point(alpha = 0.85) +
            scale_color_manual(
                values = c(
                    "DLGNT_2 up"            = "#d62728",
                    "DLGNT_2 down"          = "#1f77b4",
                    "not sig (padj ≥ 0.05)" = "#cccccc"
                )
            ) +
            scale_size_continuous(range = c(1.5, 8)) +
            theme_minimal(base_size = 12) +
            theme(
                panel.grid.major.y = element_blank(),
                axis.text.x        = element_text(angle = 20, hjust = 1),
                axis.text.y        = element_text(size = 9),
                legend.position    = "bottom"
            ) +
            labs(
                x     = NULL,
                y     = "Hallmark pathway",
                size  = "−log10(padj)\n(0 = not sig)",
                color = "Direction",
                title = paste0(ct, ": pathway significance across runs")
            )
    )
    cat("\n\n")
}
```

# Top DE genes: baseline vs stress-adjusted

For each cell type, the top 8 genes by FDR (union across both runs) are shown. Genes whose ranking
or direction changes substantially between the two runs are likely stress-collinear.

```{r de-comparison}
#| results: asis
#| fig-height: 6
for (ct in cell_types) {
    ct_de <- de_dt[
        cell_type == ct & run %in% c("Baseline", "Stress-adjusted")
    ]

    # Top 8 within each run; take the union across both runs for a shared gene set.
    top8 <- ct_de[
        order(run, fdr, -abs(t_stat)),
        head(.SD, 8),
        by = run
    ]
    union_genes <- unique(top8$gene)

    plot_de <- ct_de[gene %in% union_genes]
    plot_de[, direction := fifelse(t_stat > 0, "DLGNT_2 up", "DLGNT_2 down")]

    # Order genes by t_stat in the stress-adjusted run so both panels share the same y-axis order.
    gene_ord <- plot_de[
        run == "Stress-adjusted", .(mean_t = mean(t_stat)), by = gene
    ][order(mean_t), gene]
    # Append any genes only in baseline (not in stress-adjusted) at the bottom.
    gene_ord <- c(setdiff(union_genes, gene_ord), gene_ord)
    plot_de[, gene_ord := factor(gene, levels = gene_ord)]

    cat("## ", ct, "\n\n", sep = "")
    print(
        ggplot(plot_de, aes(x = t_stat, y = gene_ord, fill = direction)) +
            geom_col() +
            facet_wrap(~ run, ncol = 2) +
            scale_fill_manual(
                values = c("DLGNT_2 down" = "#1f77b4", "DLGNT_2 up" = "#d62728"),
                guide  = "none"
            ) +
            theme_minimal(base_size = 12) +
            theme(axis.text.y = element_text(size = 9)) +
            labs(
                x     = "Moderated t statistic",
                y     = "Gene",
                title = paste0(ct, ": top DE genes — baseline vs stress-adjusted")
            )
    )
    cat("\n\n")
}
```

# Pathway persistence table

Pathways significant in all 4 runs are the most robust (surviving every intervention). Pathways
only significant in the baseline are the most likely to be stress-driven artifacts.

```{r persistence-table}
persistence <- gsea_dt[
    contrast == "DLGNT_2 > DLGNT_1" & !is.na(padj) & padj < 0.05,
    .(
        n_runs_sig = uniqueN(run),
        runs_sig   = paste(sort(unique(as.character(run))), collapse = " | "),
        # Mean NES and mean -log10(padj) summarise direction and strength.
        mean_NES      = round(mean(NES), 3),
        mean_neg_logp = round(mean(-log10(padj)), 2)
    ),
    by = .(cell_type, set_name_clean)
][order(cell_type, -n_runs_sig, -mean_neg_logp)]

setnames(persistence, "set_name_clean", "hallmark_pathway")
kable(
    persistence,
    caption = "Hallmark pathway persistence across runs (DLGNT_2 > DLGNT_1, padj < 0.05)"
)
```

```{r persistence-summary}
# Quick summary: how many pathways survive 4 / 3 / 2 / 1 runs per cell type.
pers_summary <- persistence[
    , .N, by = .(cell_type, n_runs_sig)
][order(cell_type, -n_runs_sig)]
setnames(pers_summary, "N", "n_pathways")

ggplot(pers_summary, aes(x = factor(n_runs_sig), y = n_pathways, fill = factor(n_runs_sig))) +
    geom_col() +
    facet_wrap(~ cell_type, ncol = 3) +
    scale_fill_brewer(palette = "Blues", direction = 1) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "none") +
    labs(
        x     = "Number of runs pathway is significant in",
        y     = "Number of Hallmark pathways",
        title = "Pathway persistence by cell type"
    )
```
