---
title: "Cell Cell Adjacencies in Banksy Niches"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

Here we will use a neighborhood adjacency analysis to look at the adjacency of cell types within niches. 


```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(dbscan)
library(circlize)
library(Seurat)
library(scales)
library(htmlwidgets)

# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
```


# Preperations

## Loading the data

The objects are here:

RDS object with Leiden Clusters: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds`

    -   `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_embeddings/total_spatial_banksy_clusters_subset_20240719_092550.rds`


Cell type clusters annotation: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv`  


```{r}
# load the banksy embeddings
banksy_embeddings = readRDS(ifelse(
    workdir == '/xchip/beroukhimlab/',
    paste0(workdir,'coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds'),
    '/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/total_spatial_banksy_clusters_subset_20240719_092550.rds'
))
colData(banksy_embeddings)$sample_id = factor(colData(banksy_embeddings)$sample_id)

# Load the cell type clusters annotation
prelim_annotation_res1 = fread(paste0(
    workdir,'coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv'
))

# specify the cluster names 
# this is mainly to keep track of which resolution we are using
cell_res = 1
cell_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k50_res%.1f',cell_res))
niche_res = 1
niche_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.8_k50_res%.1f',niche_res))

# # filter the banksy object to only include the cells with annotated cell types
# annotated_cell_clusters = sort(unique(prelim_annotation_res1$cluster))
# banksy_embeddings = banksy_embeddings[,colData(banksy_embeddings)[,cell_clust_name] %in% annotated_cell_clusters]
```


## Build the color palette for all niches and cell types downstream

```{r}
set.seed(5555)
cluster_classes = levels(colData(banksy_embeddings)[[niche_clust_name]])
niche_color_manual = structure(
    pals::polychrome()[c(1:length(cluster_classes))],
    names = sample(cluster_classes)
)
# for niche resolution 0.5
# niches_to_show = c(1:11, 13, 14, 16, 17)
# for niche resolution 1
niches_to_show = c(1:20, 22:25)

cell_type_to_test = 'Coarse Cell Type'
cell_type_to_test = 'Fine Cell Type'
cell_type_to_test = 'Fine Cell Type UMAP'

cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))
cell_type_color_manual = structure(
    pals::alphabet()[c(1:length(cell_type_class))],
    names = cell_type_class
)
```


## Setting up Neighborhood Analysis

Here we will look at the niches in the data and annotate them using the cell types of self and neighbors.

```{r}
banksy_embeddings = banksy_embeddings[,colData(banksy_embeddings)$sample_id %in% c(258482,240468)]
table(colData(banksy_embeddings)$sample_id)
k_geom = 30
locs = spatialCoords(banksy_embeddings)
knn = dbscan::kNN(x = locs, k = k_geom)
knnDT <- data.table(
    from = rep(seq_len(nrow(knn$id)), k_geom),
    from_id = rep(rownames(knn$id), k_geom),
    to = as.vector(knn$id),
    to_id = rownames(knn$id)[as.vector(knn$id)],
    distance = as.vector(knn$dist)
)
cell_cell_type_dt = data.table(
    cell_id = rownames(colData(banksy_embeddings)),
    cell_type = as.numeric(colData(banksy_embeddings)[,cell_clust_name]),
    niche_type = as.factor(colData(banksy_embeddings)[,niche_clust_name]),
    sample_id = colData(banksy_embeddings)$sample_id,
    histology = colData(banksy_embeddings)$histology
)
```

cell annotation 

```{r}
gg_annotation = fread('~/Downloads/pathology_GG240468_meta.csv')[
    ,.(cell_id = paste0(cell_id,'__240468'), annotation)
]
pa_annotation = fread('~/Downloads/pathology_PA258482_meta.csv')[
    ,.(cell_id = paste0(cell_id,'__258482'), annotation)
]
message(paste0(
    'For GG: there are ', nrow(gg_annotation), ' cells with annotations, ',
    'and ', nrow(cell_cell_type_dt[sample_id == '240468']), ' cells in the banksy object after QC.',
    '\nTotal intersection: ', length(intersect(gg_annotation$cell_id, cell_cell_type_dt[sample_id == '240468']$cell_id)),
    ' and setdiff: ', length(setdiff(gg_annotation$cell_id, cell_cell_type_dt[sample_id == '240468']$cell_id))
))
message(paste0(
    'For PA: there are ', nrow(pa_annotation), ' cells with annotations, ',
    'and ', nrow(cell_cell_type_dt[sample_id == '258482']), ' cells in the banksy object after QC.',
    '\nTotal intersection: ', length(intersect(pa_annotation$cell_id, cell_cell_type_dt[sample_id == '258482']$cell_id)),
    ' and setdiff: ', length(setdiff(pa_annotation$cell_id, cell_cell_type_dt[sample_id == '258482']$cell_id))
))
```

# Cell Cell Adjacency Analysis in GG

```{r}
gg_knnDT = knnDT[from_id %in% cell_cell_type_dt[sample_id == '240468']$cell_id]
gg_cell_cell_type_dt = cell_cell_type_dt[sample_id == '240468']
gg_cel_type_prop = gg_cell_cell_type_dt[, .(prop = round(.N/nrow(gg_cell_cell_type_dt),4)), .(cell_type)][order(-prop)]
kable(gg_cel_type_prop)
gg_cell_cell_type_dt[
    ,cell_type := ifelse(cell_type %in% gg_cel_type_prop[prop < 0.01]$cell_type, 'Other', cell_type)
]

gg_dt_to_wilcox_test = dcast(
    gg_knnDT[,.(from_id,to_id,distance)][
        gg_cell_cell_type_dt[,.(cell_id,cell_type_to = cell_type)], 
        on = c('to_id' = 'cell_id'), nomatch = NULL
    ][
        gg_cell_cell_type_dt[,.(
            cell_id, niche_type_from = niche_type, 
            cell_type_from = cell_type, sample_id, histology
        )], on = c('from_id' = 'cell_id'), nomatch = NULL
    ][
        gg_annotation, on = c('from_id' = 'cell_id'), nomatch = NULL
    ][
        ,.(neighbor_proportions = .N/k_geom), .(from_id, cell_type_from, cell_type_to, annotation)
    ], from_id + cell_type_from + annotation ~ cell_type_to, 
    value.var = 'neighbor_proportions', fill = 0
)

gg_cca_enrichment = rbindlist(lapply(
    sort(unique(gg_dt_to_wilcox_test$cell_type_from)), function(x){
        print(x)
        pairwise_categorical_continuous_enrichment_wilcox_test(
            gg_dt_to_wilcox_test[cell_type_from == x, -2], num_cores = 1
        )[, cell_type_from := x]
    }
))

cell_type_replacement = structure(
    c(unlist(prelim_annotation_res1[, .(get(cell_type_to_test))]),'Other'),
    names = c(prelim_annotation_res1$cluster,'Other')
)
gg_cca_enrichment$cell_type_name_from = cell_type_replacement[gg_cca_enrichment$cell_type_from]
gg_cca_enrichment$cell_type_name_to = cell_type_replacement[gg_cca_enrichment$continuous_cols]
gg_cca_enrichment$q_value = p.adjust(gg_cca_enrichment$p_value, method = 'fdr')

ggplot(gg_cca_enrichment) +
    geom_point(aes(
        x = cell_type_name_from, y = cell_type_name_to, size = N_var2,
        fill = delta, 
        color = ifelse(q_value <= 0.05 & as.numeric(delta) > 0, 'significant','not significant')
    ), shape = 21, stroke = 1) + 
    facet_wrap(~annotation, nrow = 1) +
    theme_linedraw() +
    theme(
        text = element_text(size = 15), 
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'right'
    ) + scale_fill_viridis_c() +
    scale_size_area(max_size = 10) +
    scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
    labs(
        title = 'Enrichment of CCA By Annotation in GG 240468',
        x = 'Cell Type (Self)',
        y = 'Cell Type (Neighbor)',
        size = '% of Cells',
        fill = 'Delta',
        color = 'Significance'
    ) + guides(
        fill = guide_legend(title = 'Delta'),
        color = guide_legend(title = 'Significance')
    )

```

# Cell Cell Adjacency Analysis in PA

```{r}
pa_knnDT = knnDT[from_id %in% cell_cell_type_dt[sample_id == '258482']$cell_id]
pa_cell_cell_type_dt = cell_cell_type_dt[sample_id == '258482']
pa_cel_type_prop = pa_cell_cell_type_dt[, .(prop = round(.N/nrow(pa_cell_cell_type_dt),4)), .(cell_type)][order(-prop)]
kable(pa_cel_type_prop)
pa_cell_cell_type_dt[
    ,cell_type := ifelse(cell_type %in% pa_cel_type_prop[prop < 0.001]$cell_type, 'Other', cell_type)
]

pa_dt_to_wilcox_test = dcast(
    pa_knnDT[,.(from_id,to_id,distance)][
        pa_cell_cell_type_dt[,.(cell_id,cell_type_to = cell_type)], 
        on = c('to_id' = 'cell_id'), nomatch = NULL
    ][
        pa_cell_cell_type_dt[,.(
            cell_id, niche_type_from = niche_type, 
            cell_type_from = cell_type, sample_id, histology
        )], on = c('from_id' = 'cell_id'), nomatch = NULL
    ][
        pa_annotation, on = c('from_id' = 'cell_id'), nomatch = NULL
    ][
        ,.(neighbor_proportions = .N/k_geom), .(from_id, cell_type_from, cell_type_to, annotation)
    ], from_id + cell_type_from + annotation ~ cell_type_to, 
    value.var = 'neighbor_proportions', fill = 0
)

pa_cca_enrichment = rbindlist(lapply(
    sort(unique(pa_dt_to_wilcox_test$cell_type_from)), function(x){
        print(x)
        pairwise_categorical_continuous_enrichment_wilcox_test(
            pa_dt_to_wilcox_test[cell_type_from == x, -2], num_cores = 1
        )[, cell_type_from := x]
    }
))

cell_type_replacement = structure(
    c(unlist(prelim_annotation_res1[, .(get(cell_type_to_test))]),'Other'),
    names = c(prelim_annotation_res1$cluster,'Other')
)
pa_cca_enrichment$cell_type_name_from = cell_type_replacement[pa_cca_enrichment$cell_type_from]
pa_cca_enrichment$cell_type_name_to = cell_type_replacement[pa_cca_enrichment$continuous_cols]
pa_cca_enrichment$q_value = p.adjust(pa_cca_enrichment$p_value, method = 'fdr')

ggplot(pa_cca_enrichment) +
    geom_point(aes(
        x = cell_type_name_from, y = cell_type_name_to, size = N_var2,
        fill = delta, 
        color = ifelse(q_value <= 0.05 & as.numeric(delta) > 0, 'significant','not significant')
    ), shape = 21, stroke = 1) + 
    facet_wrap(~annotation, nrow = 1) +
    theme_linedraw() +
    theme(
        text = element_text(size = 15), 
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'right'
    ) + scale_fill_viridis_c() +
    scale_size_area(max_size = 10) +
    scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
    labs(
        title = 'Enrichment of CCA By Annotation in PA 258482',
        x = 'Cell Type (Self)',
        y = 'Cell Type (Neighbor)',
        size = '% of Cells',
        fill = 'Delta',
        color = 'Significance'
    ) + guides(
        fill = guide_legend(title = 'Delta'),
        color = guide_legend(title = 'Significance')
    )



```