---
title: "Sample-Level Modeling of Cell Type Proportions in Pathology Regions"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk:
        results: hold
        R.options:
            width: 120
---

# Introduction

This notebook is a temporary sample-level comparison against the region-level workflow in `annotations/linear_models.qmd`.

Here we will be using linear models to test the association between cell types and pathology annotation regions. For each histology type, we have 2-4 samples. In this sample-level version, all cells from the same sample and pathology annotation are aggregated together, so there is one observation per `sample_id x pathology_annotation`.

Previously, what we have done is using fisher exact tests to test the association between every single cell type and every single pathology region within each sample, then combine p values across samples. This works in principle, but is not ideal since p values do not convey the direction of the association. So combining an enriched p value and depleted p value will still yield a significant combined p value. Additionally, we are not accounting for the sample level variability due to covariates such as age, location of the tumor, etc.

So to improve on this short coming, we use count-based beta-binomial mixed-effect models to test association between cell types and pathology annotation regions. We fit these models with `glmmTMB`, using a sample-level random intercept to account for baseline between-sample composition differences. The model is:

$$
\begin{align}
k_{ij} &\sim \text{BetaBinomial}(n_{ij}, p_{ij}, \phi) \\\\
\text{logit}(p_{ij}) &= \beta_0 + \beta_1 x_{ij} + b_{s(i)}, \quad b_{s(i)} \sim \mathcal{N}(0,\sigma^2_s)
\end{align}
$$

Where $k_{ij}$ is the count of cells from cell type $i$ in observation $j$, $n_{ij}$ is the total number of cells in that sample-pathology observation, and $x_{ij}$ is the pathology annotation region. This setup models morphology effects while adjusting for sample-to-sample baseline shifts.

However, this approach does have some limitations. Since we only have 2-4 independent samples per histology type, power remains limited. Additionally, we are doing this individually for each cell type independently, so we are not accounting for the correlation between cell types.

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(dbscan)
library(circlize)
library(Seurat)
library(scales)
library(htmlwidgets)
library(glmmTMB)


# local vs remote
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usually at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (Sys.getenv("HOME") == "/PHShome/yz762"){
    # in eristwo, the home directory is at /PHShome/yz762
    workdir <- "/data/beroukhim1/"
} else if (Sys.getenv("HOME") == "/home/yz762"){
    # on the new server, the home directory is at /home/yz762
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    # in dipg or uger, the home directory is '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
source(paste0(workdir,'/youyun/plgg/code/helpers/enrichment_models.R'))
```


# Preperations

## Loading the data

```{r}
banksy_subset_rds_path = file.path(
    workdir, "youyun/plgg/data/Xenium_annotations",
    "banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds"
)
banksy_full_rds_path = file.path(
    workdir, "youyun/plgg/data/Xenium_annotations",
    "banksy_clusters_connected_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds"
)
banksy_rds_path = if (file.exists(banksy_subset_rds_path)) banksy_subset_rds_path else banksy_full_rds_path

banksy_embeddings <- readRDS(banksy_rds_path)
message("Using BANKSY annotated object: ", banksy_rds_path)

if (!"pathology_annotation" %in% colnames(colData(banksy_embeddings))) {
    stop("Missing required annotation metadata column: pathology_annotation")
}

colData(banksy_embeddings)$sample_id = factor(colData(banksy_embeddings)$sample_id)
colData(banksy_embeddings)$pathology_annotation = ifelse(
    !grepl('Compact', colData(banksy_embeddings)$pathology_annotation), colData(banksy_embeddings)$pathology_annotation,
    gsub('\\.', ',', colData(banksy_embeddings)$pathology_annotation)
)

prelim_annotation_res1 = fread(paste0(
    workdir,'youyun/plgg/data/xenium_cell_types/banksy_proseg_cell_cluster_annotation.tsv'
))

# specify the cluster names
cell_res = 1
cell_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k50_res%.1f',cell_res))
niche_res = 1
niche_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.8_k50_res%.1f',niche_res))
```


## Build the color palette for all niches and cell types downstream

```{r}
set.seed(5555)
cluster_classes = levels(colData(banksy_embeddings)[[niche_clust_name]])
niche_color_manual = structure(
    pals::polychrome()[c(1:length(cluster_classes))],
    names = sample(cluster_classes)
)

cell_type_to_test = 'Coarse Cell Type'
# cell_type_to_test = 'Fine Cell Type'
# cell_type_to_test = 'Fine Cell Type UMAP'

cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))
cell_type_color_manual = structure(
    pals::alphabet()[c(1:length(cell_type_class))],
    names = cell_type_class
)
```

## Visualize Pathology Annotations in Space

```{r}
#| fig-height: 12
#| fig-width: 24
graph_dt = as.data.table(cbind(
    spatialCoords(banksy_embeddings),
    colData(banksy_embeddings)[c(
        'sample_id', 'cell_id', 'histology', 'pathology_annotation'
    )]
))

graph_dt[, pathology_annotation := relevel(factor(ifelse(
    !grepl('Compact', pathology_annotation), pathology_annotation,
    gsub('\\.', ',', pathology_annotation)
)), ref = 'Compact, fibrillary component')]

pathology_color_scale = if (requireNamespace("ggthemes", quietly = TRUE)) {
    ggthemes::scale_color_carto_d(
        name = '',
        palette = 'Safe'
    )
} else {
    scale_color_brewer(
        name = '',
        palette = 'Set2'
    )
}

ggplot(
    graph_dt[grepl('pilo', histology) & !grepl('dominant', pathology_annotation)],
    aes(x = sdimy, y = sdimx, color = pathology_annotation)
) +
    facet_wrap(~ sample_id, scales = 'free_x', nrow = 1) +
    geom_point(size = 0.00001, alpha = 0.7) +
    pathology_color_scale +
    theme_void(base_size = 40) +
    theme(
        legend.position = "bottom",
        text = element_text(size = 40),
        strip.text = element_text(size = 30),
        strip.background = element_blank()
    ) +
    guides(
        color = guide_legend(override.aes = list(size = 5, alpha = 1), nrow = 3, byrow = TRUE)
    ) +
    labs(
        title = 'Pathology Annotations on H&E Overlaid on Cell Locations',
        x = '',
        y = ''
    ) +
    coord_flip()
```

# Modeling Data Manipulation

We will then build the data for modeling from the spatial data.

First we will clean up the data. Then we will calculate the proportion of each cell type in each pathology region for each sample.

```{r}
sparsity_threshold = 0.0025
modeling_data_ct = as.data.table(merge(
    colData(banksy_embeddings)[
        c('cell_id','sample_id', 'histology','alteration', cell_clust_name, 'pathology_annotation')
    ], prelim_annotation_res1[, .(cluster, cell_type = get(cell_type_to_test))],
    by.x = cell_clust_name, by.y = 'cluster', all.x = TRUE
))[, `:=`(
    cell_id = paste0(cell_id, '__', sample_id),
    cell_type = fifelse(is.na(cell_type), 'Other', cell_type),
    pathology_annotation = relevel(factor(pathology_annotation), ref = "unannotated")
)]

ct_prop_dt = dcast(
    unique(modeling_data_ct[
        , num_cells_per_sample := .N, .(sample_id)
    ][
        , num_cells_per_cell_type := .N, .(cell_type, sample_id)
    ][
        , .(prop_cells_per_cell_type = num_cells_per_cell_type / num_cells_per_sample),
        .(cell_type, sample_id, histology)
    ]), sample_id + histology ~ cell_type, value.var = 'prop_cells_per_cell_type', fill = 0
)
melted_ct_prop_dt = melt(
    ct_prop_dt, id.vars = c('sample_id', 'histology'),
    variable.name = 'cell_type', value.name = 'prop_cells_per_cell_type'
)

modeling_data_ct = merge(
    modeling_data_ct,
    melted_ct_prop_dt[
        , .(cell_type_cleaned_up = fifelse(
            sum(prop_cells_per_cell_type < sparsity_threshold) / .N >= 0.5,
            'Other', as.character(cell_type)
        )), .(cell_type, histology)
    ],
    by = c('cell_type', 'histology'), all.x = TRUE
)[
    , cell_type := cell_type_cleaned_up
][
    , cell_type_cleaned_up := NULL
]

modeling_dt_ct = dcast(
    unique(modeling_data_ct[
        , num_cells_per_sample_region := .N, .(sample_id, pathology_annotation)
    ][
        , .(prop_cell_type_per_sample_region = .N / num_cells_per_sample_region), .(
            sample_id, pathology_annotation, cell_type,
            histology, alteration, num_cells_per_sample_region
        )
    ]),
    sample_id + pathology_annotation + histology + alteration + num_cells_per_sample_region ~ cell_type,
    value.var = 'prop_cell_type_per_sample_region', fill = 0
)
```

Then we will do the same for niches

```{r}
sparsity_threshold = 0.0025
modeling_data_niche = as.data.table(colData(banksy_embeddings)[
    c('cell_id','sample_id', 'histology','alteration', niche_clust_name, 'pathology_annotation')
])[, cell_id := paste0(cell_id, '__', sample_id)]
setnames(modeling_data_niche, niche_clust_name, 'niche_type')

niche_prop_dt = dcast(
    unique(modeling_data_niche[
        , num_cells_per_sample := .N, .(sample_id)
    ][
        , num_cells_per_niche_type := .N, .(niche_type, sample_id)
    ][
        , .(prop_cells_per_niche_type = num_cells_per_niche_type / num_cells_per_sample),
        .(niche_type, sample_id, histology)
    ]), sample_id + histology ~ niche_type, value.var = 'prop_cells_per_niche_type', fill = 0
)
melted_niche_prop_dt = melt(
    niche_prop_dt, id.vars = c('sample_id', 'histology'),
    variable.name = 'niche_type', value.name = 'prop_cells_per_niche_type'
)

modeling_data_niche = merge(
    modeling_data_niche,
    melted_niche_prop_dt[
        , .(niche_type_cleaned_up = fifelse(
            sum(prop_cells_per_niche_type < sparsity_threshold) / .N >= 0.5,
            'Other', as.character(niche_type)
        )), .(niche_type, histology)
    ],
    by = c('niche_type', 'histology'), all.x = TRUE
)[
    , niche_type := niche_type_cleaned_up
][
    , niche_type_cleaned_up := NULL
]

modeling_dt_niche = dcast(
    unique(modeling_data_niche[
        , num_cells_per_sample_region := .N, .(sample_id, pathology_annotation)
    ][
        , .(prop_niche_type_per_sample_region = .N / num_cells_per_sample_region), .(
            sample_id, pathology_annotation, niche_type,
            histology, alteration, num_cells_per_sample_region
        )
    ]),
    sample_id + pathology_annotation + histology + alteration + num_cells_per_sample_region ~ niche_type,
    value.var = 'prop_niche_type_per_sample_region', fill = 0
)
```

## Overview of the Cell and Niche Types we are modeling

```{r}
kable(modeling_data_ct[
    , .(
        number_of_cell_types = length(unique(cell_type)),
        cell_types = paste(sort(unique(cell_type)), collapse = ', ')
    ), .(sample_id, histology, alteration)
][order(-histology)])
kable(modeling_data_niche[
    , .(
        number_of_niche_types = length(unique(niche_type)),
        niche_types = paste(sort(unique(niche_type)), collapse = ', ')
    ), .(sample_id, histology, alteration)
][order(-histology)])
```


# Mixed-Effects Modeling Framework

We now use a unified count-based beta-binomial mixed-effects pipeline with explicit QC gating. Pathology filtering rules are centralized below and reused for both modalities.

```{r}
pathology_rules = data.table(
    analysis_id = c("PA", "CLN", "CN", "GG"),
    ref_region = c("unannotated", "Cerebellum, molecular layer", "Neuropil zone", "unannotated"),
    histology_regex = c("pilo", NA_character_, "central", NA_character_),
    pathology_exclude_regex = c("dominant", NA_character_, "Lipid", NA_character_),
    pathology_include_regex = c(NA_character_, NA_character_, NA_character_, NA_character_)
)
pathology_rules[, sample_ids_include := vector("list", .N)]
pathology_rules[analysis_id == "CLN", sample_ids_include := list(as.character(c(67068, 42112, 87352)))]
pathology_rules[analysis_id == "GG", sample_ids_include := list(as.character(240468))]

cell_type_count_dt = build_feature_count_dt(
    coldata_dt = modeling_data_ct,
    feature_col = "cell_type"
)

niche_type_count_dt = build_feature_count_dt(
    coldata_dt = modeling_data_niche,
    feature_col = "niche_type"
)

stack_or_empty = function(x) {
    if (length(x) == 0) return(data.table())
    rbindlist(x, use.names = TRUE, fill = TRUE)
}

run_rule_set = function(count_dt, modality_label) {
    all_results_list = list()
    primary_results_list = list()
    qc_summary_list = list()
    qc_details_list = list()
    filter_log_list = list()
    feature_qc_list = list()
    data_checks_list = list()

    for (i in seq_len(nrow(pathology_rules))) {
        rule = pathology_rules[i]

        candidate_histologies = unique(count_dt$histology)

        if (!is.na(rule$histology_regex)) {
            candidate_histologies = candidate_histologies[
                grepl(rule$histology_regex, candidate_histologies, ignore.case = TRUE)
            ]
        }

        if (!is.null(rule$sample_ids_include[[1]])) {
            candidate_histologies = intersect(
                candidate_histologies,
                unique(count_dt[sample_id %in% rule$sample_ids_include[[1]], histology])
            )
        }

        if (length(candidate_histologies) == 0) next

        for (histology_value in candidate_histologies) {
            fit_obj = fit_feature_set(
                count_dt = count_dt,
                feature_col = "feature",
                histology_value = histology_value,
                ref_region = rule$ref_region,
                filters = list(
                    sample_ids_include = rule$sample_ids_include[[1]],
                    pathology_exclude_regex = rule$pathology_exclude_regex,
                    pathology_include_regex = rule$pathology_include_regex,
                    modality_label = modality_label
                ),
                fdr_scope = "histology_modality"
            )

            if (nrow(fit_obj$all_results) > 0) {
                fit_obj$all_results[
                    , `:=`(analysis_id = rule$analysis_id, ref_region = rule$ref_region)
                ]
                all_results_list[[length(all_results_list) + 1]] = fit_obj$all_results
            }

            if (nrow(fit_obj$primary_results) > 0) {
                fit_obj$primary_results[
                    , `:=`(analysis_id = rule$analysis_id, ref_region = rule$ref_region)
                ]
                primary_results_list[[length(primary_results_list) + 1]] = fit_obj$primary_results
            }

            if (nrow(fit_obj$model_qc_summary) > 0) {
                fit_obj$model_qc_summary[
                    , analysis_id := rule$analysis_id
                ]
                qc_summary_list[[length(qc_summary_list) + 1]] = fit_obj$model_qc_summary
            }

            if (nrow(fit_obj$model_qc_details) > 0) {
                fit_obj$model_qc_details[
                    , analysis_id := rule$analysis_id
                ]
                qc_details_list[[length(qc_details_list) + 1]] = fit_obj$model_qc_details
            }

            if (nrow(fit_obj$filter_log) > 0) {
                fit_obj$filter_log[
                    , analysis_id := rule$analysis_id
                ]
                filter_log_list[[length(filter_log_list) + 1]] = fit_obj$filter_log
            }

            if (nrow(fit_obj$feature_qc) > 0) {
                fit_obj$feature_qc[
                    , `:=`(analysis_id = rule$analysis_id, histology = histology_value, modality = modality_label)
                ]
                feature_qc_list[[length(feature_qc_list) + 1]] = fit_obj$feature_qc
            }

            if (nrow(fit_obj$data_checks) > 0) {
                fit_obj$data_checks[
                    , analysis_id := rule$analysis_id
                ]
                data_checks_list[[length(data_checks_list) + 1]] = fit_obj$data_checks
            }
        }
    }

    list(
        all_results = stack_or_empty(all_results_list),
        primary_results = stack_or_empty(primary_results_list),
        model_qc_summary = stack_or_empty(qc_summary_list),
        model_qc_details = stack_or_empty(qc_details_list),
        filter_log = stack_or_empty(filter_log_list),
        feature_qc = stack_or_empty(feature_qc_list),
        data_checks = stack_or_empty(data_checks_list)
    )
}

cell_type_results = run_rule_set(
    count_dt = cell_type_count_dt,
    modality_label = "cell_type"
)

niche_type_results = run_rule_set(
    count_dt = niche_type_count_dt,
    modality_label = "niche_type"
)
```

## QC Summary

```{r}
kable(cell_type_results$model_qc_summary)
kable(niche_type_results$model_qc_summary)
```

```{r}
kable(cell_type_results$data_checks)
kable(niche_type_results$data_checks)
```

## Export Standardized Outputs

```{r}
out_dir = file.path("annotations", "model_outputs", "mixed_models_sample_level")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

fwrite(cell_type_results$primary_results, file.path(out_dir, "linear_models_sample_level_cell_type_primary_results.tsv"), sep = "\t")
fwrite(cell_type_results$all_results, file.path(out_dir, "linear_models_sample_level_cell_type_all_results_with_qc.tsv"), sep = "\t")
fwrite(cell_type_results$model_qc_summary, file.path(out_dir, "linear_models_sample_level_cell_type_model_qc_summary.tsv"), sep = "\t")

fwrite(niche_type_results$primary_results, file.path(out_dir, "linear_models_sample_level_niche_type_primary_results.tsv"), sep = "\t")
fwrite(niche_type_results$all_results, file.path(out_dir, "linear_models_sample_level_niche_type_all_results_with_qc.tsv"), sep = "\t")
fwrite(niche_type_results$model_qc_summary, file.path(out_dir, "linear_models_sample_level_niche_type_model_qc_summary.tsv"), sep = "\t")
```

# Visualize Primary-Pass Results

## Cell Types

```{r}
plot_cell_primary = summarize_results_for_plot(
    results_dt = cell_type_results$all_results,
    primary_only = TRUE
)

if (nrow(plot_cell_primary) > 0) {
    ggplot(plot_cell_primary, aes(
        x = estimate_log_odds,
        y = feature,
        color = factor(sign_call, levels = c("enriched", "depleted", "not_significant"))
    )) +
        geom_point() +
        geom_errorbarh(aes(
            xmin = estimate_log_odds - 1.96 * se,
            xmax = estimate_log_odds + 1.96 * se
        ), height = 0.25) +
        geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
        facet_wrap(~ paste0(analysis_id, " | ", annotation_region), scales = "free_x") +
        theme_bw(base_size = 12) +
        theme(
            legend.position = "bottom",
            legend.title = element_blank(),
            panel.grid.minor = element_blank()
        ) +
        labs(
            title = "Cell Type Enrichment (Primary-Pass Mixed Models, Sample-Level)",
            x = "Log-odds estimate",
            y = "Cell Type"
        ) +
        scale_color_brewer(palette = "Set1")
} else {
    message("No primary-pass cell-type models to plot.")
}
```

## Niche Types

```{r}
plot_niche_primary = summarize_results_for_plot(
    results_dt = niche_type_results$all_results,
    primary_only = TRUE
)

if (nrow(plot_niche_primary) > 0) {
    ggplot(plot_niche_primary, aes(
        x = estimate_log_odds,
        y = feature,
        color = factor(sign_call, levels = c("enriched", "depleted", "not_significant"))
    )) +
        geom_point() +
        geom_errorbarh(aes(
            xmin = estimate_log_odds - 1.96 * se,
            xmax = estimate_log_odds + 1.96 * se
        ), height = 0.25) +
        geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
        facet_wrap(~ paste0(analysis_id, " | ", annotation_region), scales = "free_x") +
        theme_bw(base_size = 12) +
        theme(
            legend.position = "bottom",
            legend.title = element_blank(),
            panel.grid.minor = element_blank()
        ) +
        labs(
            title = "Niche Type Enrichment (Primary-Pass Mixed Models, Sample-Level)",
            x = "Log-odds estimate",
            y = "Niche Type"
        ) +
        scale_color_brewer(palette = "Set1")
} else {
    message("No primary-pass niche-type models to plot.")
}
```

# Flagged or Exploratory Results (Warn/Fail)

```{r}
plot_flagged = summarize_results_for_plot(
    results_dt = rbindlist(list(cell_type_results$all_results, niche_type_results$all_results), use.names = TRUE, fill = TRUE),
    primary_only = FALSE
)[model_status != "pass"]

if (nrow(plot_flagged) > 0) {
    ggplot(plot_flagged, aes(
        x = estimate_log_odds,
        y = feature,
        color = analysis_id,
        shape = model_status
    )) +
        geom_point(size = 2) +
        geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
        facet_wrap(~ modality, scales = "free_y") +
        theme_bw(base_size = 12) +
        labs(
            title = "Flagged / Exploratory Results (Sample-Level)",
            x = "Log-odds estimate",
            y = "Feature"
        )
} else {
    message("No warn/fail rows available for flagged view.")
}
```
