---
title: "Interpreting BANKSY Niches"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

In this doc, we will be looking at the banksy clusters and try to interpret them based on their identity, and other characteristics.

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(dbscan)
library(circlize)
library(Seurat)
library(plotly)
library(scales)
library(htmlwidgets)
library(metap)

# local vs remote
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (Sys.getenv("HOME") == "/PHShome/yz762"){
    # in eristwo, the home directory is at /PHShome/yz762
    workdir <- "/data/beroukhim1/"
} else if (Sys.getenv("HOME") == "/home/yz762"){
    # on the new server, the home directory is at /home/yz762
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    # in dipg or uger, the home directory is '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
# glm_family = 'quasibinomial'
glm_family = 'binomial'
```


# Load the data

The objects are here:

RDS object with Leiden Clusters: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds`

    -   `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_embeddings/total_spatial_banksy_clusters_subset_20240719_092550.rds`


Cell type clusters annotation: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv`  


```{r}
banksy_embeddings <- readRDS(ifelse(
    basename(workdir) != 'dfci_mount', paste0(
        workdir, "youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/banksy_clusters_connected_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
    ), "/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
))
colData(banksy_embeddings)$sample_id <- factor(colData(banksy_embeddings)$sample_id)

prelim_annotation_res1 <- fread(paste0(
    workdir,'youyun/plgg/data/xenium_cell_types/banksy_proseg_cell_cluster_annotation.tsv'
))

# specify the cluster names 
# this is mainly to keep track of which resolution we are using
cell_res = 1
cell_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k50_res%.1f',cell_res))
niche_res = 1
niche_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.8_k50_res%.1f',niche_res))

# # filter the banksy object to only include the cells with annotated cell types
# annotated_cell_clusters = sort(unique(prelim_annotation_res1$cluster))
# banksy_embeddings = banksy_embeddings[,colData(banksy_embeddings)[,cell_clust_name] %in% annotated_cell_clusters]
```


Build the color palette for all niches and cell types downstream

```{r}
set.seed(5555)
cluster_classes = unique(as.character(colData(banksy_embeddings)[[niche_clust_name]]))
niche_color_manual = structure(
    pals::polychrome()[c(1:length(cluster_classes))],
    names = cluster_classes
)
# for niche resolution 0.5
# niches_to_show = c(1:11, 13, 14, 16, 17)
# for niche resolution 1
niches_to_show = cluster_classes

cell_type_to_test = 'Coarse Cell Type'
cell_type_to_test = 'Fine Cell Type'
cell_type_to_test = 'Fine Cell Type UMAP'

cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))
cell_type_color_manual = structure(
    pals::alphabet()[c(1:length(cell_type_class))],
    names = cell_type_class
)

```

# What are the niches?

## Setting up Neighborhood Analysis

Here we will look at the niches in the data and annotate them using the cell types of self and neighbors.

```{r}
k_geom = 15
locs = spatialCoords(banksy_embeddings)
knn = dbscan::kNN(x = locs, k = k_geom)
knnDT <- data.table(
    from = rep(seq_len(nrow(knn$id)), k_geom),
    from_id = rep(rownames(knn$id), k_geom),
    to = as.vector(knn$id),
    to_id = rownames(knn$id)[as.vector(knn$id)],
    distance = as.vector(knn$dist)
)
cell_cell_type_dt = data.table(
    cell_id = rownames(colData(banksy_embeddings)),
    cell_type = as.numeric(colData(banksy_embeddings)[,cell_clust_name]),
    niche_type = as.factor(colData(banksy_embeddings)[,niche_clust_name]),
    sample_id = colData(banksy_embeddings)$sample_id,
    histology = colData(banksy_embeddings)$histology
)

cell_type_cols = setdiff(colnames(prelim_annotation_res1),'cluster')
niche_self_neighbor_dt = knnDT[,c('from','to') := NULL][
    cell_cell_type_dt[,.(cell_id,cell_type)], on = c('to_id' = 'cell_id'),
    cell_type_to := cell_type
][cell_cell_type_dt, on = c('from_id' = 'cell_id')][
    ,.(
    cell_id = from_id, cell_type, niche_type, sample_id, 
    histology, neighbor_cell_type = cell_type_to
)][
    prelim_annotation_res1, on = c('cell_type' = 'cluster')
][
    prelim_annotation_res1, on = c('neighbor_cell_type' = 'cluster'),
    paste0(cell_type_cols,' Neighbors') := mget(paste0('i.',cell_type_cols))
][order(cell_id)]
niche_self_dt = unique(niche_self_neighbor_dt[
    ,mget(c('cell_id','cell_type','niche_type','sample_id','histology',cell_type_cols))
])
```

# GLM Analysis of Niche Enrichment

We will set up a glm to test for enrichment of cell types in each niche.

niche = b1 * ct1 + b2 * ct2 + ... + bn * ctn + histology

```{r}
# modeling_dt = dcast(
#     niche_self_neighbor_dt[
#         !is.na(get(paste0(cell_type_to_test,' Neighbors'))),
#         .(proportion = .N/15), .(
#             cell_id, niche_type, histology,
#             neighbor_cell_type = get(paste0(cell_type_to_test,' Neighbors'))
#         )
#     ],
#     cell_id + niche_type + histology ~ neighbor_cell_type, 
#     value.var = 'proportion', fill = 0
# )
# # run 1 vs rest models for each niche
# cell_type_enrichment = rbindlist(lapply(
#     niches_to_show,
#     function(niche_i){
#         print(niche_i)
#         formula_str = paste0(
#             'I(niche_type == ', niche_i, ') ~ ',
#             paste0(
#                 paste0(
#                     '`', setdiff(
#                         colnames(modeling_dt),
#                         c('cell_id','niche_type','histology')
#                     ), '`'
#                 ),
#                 collapse = ' + '
#             ),
#             ' + histology'
#         )
#         glm_fit = glm(
#             formula = as.formula(formula_str),
#             data = modeling_dt,
#             family = glm_family
#         )
#         glm_summary = summary(glm_fit)
#         coef_dt = as.data.table(summary(glm_fit)$coefficients, keep.rownames = 'cell_type')[
#             , niche_type := niche_i
#         ]
#         if(glm_family == 'binomial'){
#             setnames(coef_dt, 'Pr(>|z|)', 'p_val')
#         }else if(glm_family == 'quasibinomial'){
#             setnames(coef_dt, 'Pr(>|t|)', 'p_val')
#         }
#         coef_dt[!grepl('(Intercept)', cell_type), q_val := p.adjust(p_val, method = 'BH')]
#         return(coef_dt)
#     }
# ))

# write.table(
#     cell_type_enrichment,
#     file = paste0(
#         workdir,'youyun/plgg/code/niche/proseg_output_analysis/niche_',
#         gsub(' ','_',tolower(cell_type_to_test)),'_', glm_family,'_enrichment.tsv'
#     ),
#     sep = '\t', quote = FALSE, row.names = FALSE
# )
```

```{r}
#| fig-width: 15
#| fig-height: 6
cell_type_enrichment = fread(paste0(
    workdir,'youyun/plgg/code/niche/proseg_output_analysis/niche_',
    gsub(' ','_',tolower(cell_type_to_test)),'_', glm_family,'_enrichment.tsv'
))

ggplot(
    cell_type_enrichment[
        # !grepl('Intercept', cell_type) & 
        # niche_type %in% as.character(c(7,9,10,11,16,18,19,22,27))
        grepl('Astro|OPC|Mye', cell_type) &
        niche_type %in% as.character(c(7,18,22))
    ], aes(
        x = Estimate, y = gsub('`','', cell_type), color = factor(ifelse(
            q_val >= 0.05, 'not significant', ifelse(
                Estimate > 0, 'enriched', 'depleted'
            )
        ), levels = c('enriched', 'depleted', 'not significant'))
    )
) +
    geom_point() +
    geom_errorbarh(aes(
        xmin = Estimate - 1.96 * `Std. Error`,
        xmax = Estimate + 1.96 * `Std. Error`
    ), height = 0.4) + 
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
    facet_wrap(
        ifelse(niche_type == 7,'Compact, fibrillary\ncomponent' ,'Loose, myxoid\ncomponent') ~ 
            paste0('niche ', niche_type), 
        nrow = 1, scales = "free"
    ) + 
    theme_bw(base_size = 20) +
    theme(
        legend.position = 'bottom',
        legend.title = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 30),
        strip.background = element_rect(fill = "grey90", color = NA),
        strip.text = element_text(color = "black", face = "bold"),
    ) +
    # xlim(-10, 10) +
    labs(
        title = 'Niche Type Enrichment by Cell Types',
        x = 'Estimate',
        y = 'Cell Type'
    ) + scale_color_brewer(palette = 'Set1')

```

# Dirichlet Multinomial to fit cell type proportions as a function of covariates

```{r}
# modeling_dt_mglm = dcast(
#     niche_self_neighbor_dt[
#         !is.na(get(paste0(cell_type_to_test,' Neighbors'))),
#         .(count = .N), .(
#             cell_id, niche_type, histology,
#             neighbor_cell_type = get(paste0(cell_type_to_test,' Neighbors'))
#         )
#     ],
#     cell_id + niche_type + histology ~ neighbor_cell_type, 
#     value.var = 'count', fill = 0
# )
# modeling_dt_mglm = rbind(
#     modeling_dt_mglm, copy(modeling_dt_mglm)[
#         , c('niche_type','histology') := list(
#             'baseline', 'baseline'
#         )
#     ]
# )[
#     , c('niche_type', 'histology') := list(
#         relevel(factor(niche_type), ref = 'baseline'),
#         relevel(factor(histology), ref = 'baseline')
#     )
# ]
# library(MGLM)
# mglm_fit = MGLMreg(
#     as.matrix(modeling_dt_mglm[,.SD, .SDcols = setdiff(
#         colnames(modeling_dt_mglm),
#         c('cell_id','niche_type','histology')
#     )]) ~ niche_type + histology,
#     data = modeling_dt_mglm,
#     dist = 'DM'
# )
```