---
title: "Analyzing BANKSY Embedding for Niche Identities"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

In this doc, we will be looking at the banksy clusters and try to interpret them based on their identity, and other characteristics.

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(dbscan)
library(circlize)
library(Seurat)
library(plotly)
library(scales)
library(htmlwidgets)
library(metap)

# local vs remote
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (Sys.getenv("HOME") == "/PHShome/yz762"){
    # in eristwo, the home directory is at /PHShome/yz762
    workdir <- "/data/beroukhim1/"
} else if (Sys.getenv("HOME") == "/home/yz762"){
    # on the new server, the home directory is at /home/yz762
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    # in dipg or uger, the home directory is '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
```


# Load the data

The objects are here:

RDS object with Leiden Clusters: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds`

    -   `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_embeddings/total_spatial_banksy_clusters_subset_20240719_092550.rds`


Cell type clusters annotation: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv`  


```{r}
banksy_embeddings <- readRDS(ifelse(
    basename(workdir) != 'dfci_mount', paste0(
        workdir, "youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/banksy_clusters_connected_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
    ), "/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
))
colData(banksy_embeddings)$sample_id <- factor(colData(banksy_embeddings)$sample_id)

prelim_annotation_res1 <- fread(paste0(
    workdir,'youyun/plgg/data/xenium_cell_types/banksy_proseg_cell_cluster_annotation.tsv'
))[
    , `Fine Cell Type UMAP` := ifelse(
        `Fine Cell Type UMAP` == 'Stromal/Endothelial/VLMC 1',
        'Stromal 1', `Fine Cell Type UMAP`
    )
]

# specify the cluster names 
# this is mainly to keep track of which resolution we are using
cell_res = 1
cell_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k50_res%.1f',cell_res))
niche_res = 1
niche_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.8_k50_res%.1f',niche_res))

# # filter the banksy object to only include the cells with annotated cell types
# annotated_cell_clusters = sort(unique(prelim_annotation_res1$cluster))
# banksy_embeddings = banksy_embeddings[,colData(banksy_embeddings)[,cell_clust_name] %in% annotated_cell_clusters]

# sample-histology reference table
sample_histology = as.data.table(unique(colData(banksy_embeddings)[,c('sample_id','histology')]))
```


Build the color palette for all niches and cell types downstream

```{r}
set.seed(5555)
cluster_classes = unique(as.character(colData(banksy_embeddings)[[niche_clust_name]]))
niche_color_manual = structure(
    pals::polychrome()[c(1:length(cluster_classes))],
    names = cluster_classes
)
# for niche resolution 0.5
# niches_to_show = c(1:11, 13, 14, 16, 17)
# for niche resolution 1
niches_to_show = cluster_classes

cell_type_to_test = 'Coarse Cell Type'
cell_type_to_test = 'Fine Cell Type'
cell_type_to_test = 'Fine Cell Type UMAP'

cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))
cell_type_color_manual = structure(
    pals::alphabet()[c(1:length(cell_type_class))],
    names = cell_type_class
)

```

# Looking at the niche embeddings

```{r}
#| fig-height: 12
#| fig-width: 30
#| warning: false
plot_grid(
    ggdraw() + draw_label("BANKSY Niche Embeddings", fontface='bold') +
        theme(
            plot.title = element_text(hjust = 0.5),
            text = element_text(size = 30)
        ),
    plot_grid(
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.8", 
            point_size = 0.00001, 
            color_by = niche_clust_name
        ) + theme(
            text = element_text(size = 20),
            legend.title = element_blank(),
            legend.position = 'bottom'
        ) + scale_color_manual(values = niche_color_manual) +
            guides(colour = guide_legend(
            override.aes = list(size = 5, alpha = 1),
            nrow = 4
        )) + labs(
            title = 'Niche Type',
            x = 'UMAP1', y = 'UMAP2'
        ),
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.8", 
            point_size = 0.00001, 
            color_by = "sample_id"
        ) + theme(
            text = element_text(size = 20),
            legend.title = element_blank(),
            legend.position = 'bottom',
            legend.justification = "right"
        ) + guides(colour = guide_legend(
            override.aes = list(size = 5, alpha = 1),
            nrow = 4
        )) + labs(
            title = 'Sample IDs',
            x = 'UMAP1', y = 'UMAP2'
        ),
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.8", 
            point_size = 0.00001, 
            color_by = "histology"
        ) + theme(
            text = element_text(size = 20),
            legend.title = element_blank(),
            legend.position = 'bottom'
        ) + guides(colour = guide_legend(
            override.aes = list(size = 5, alpha = 1),
            nrow = 4
        )) + labs(
            title = 'Histology',
            x = 'UMAP1', y = 'UMAP2'
        ),
        
        nrow = 1,
        rel_widths = c(1, 1, 1)
    ), ncol = 1,rel_heights = c(0.5, 5)
)
```

```{r}
#| fig-height: 12
#| fig-width: 30
#| warning: false
niche_umap = lapply(niches_to_show, function(x){
    new_niche_color_manual = structure(c(
        rep('gray', length(niche_color_manual) - 1), 
        niche_color_manual[names(niche_color_manual) == x]
    ), names = c(setdiff(names(niche_color_manual), x), x))
    assign(
        paste0('niche_',x,'_umap'),
        plotReducedDim(
            banksy_embeddings, 
            "UMAP_Harmony_BANKSY_lam0.8", 
            point_size = 0.00001, 
            color_by = niche_clust_name
        ) + theme(
            text = element_text(size = 20),
            legend.title = element_blank(),
            legend.position = 'none'
        ) + theme_void() +
            scale_color_manual(values = new_niche_color_manual) +
            guides(colour = 'none') + labs(
            title = paste0('Niche ',x)
        ),envir = .GlobalEnv
    )
})
plot_grid(
    ggdraw() + draw_label("BANKSY Niche Embeddings", fontface='bold') +
        theme(
            plot.title = element_text(hjust = 0.5),
            text = element_text(size = 30)
        ),
    plot_grid(
        plotlist = mget(paste0('niche_',niches_to_show,'_umap')),
        nrow = round(sqrt(length(niches_to_show)/2))
    ), ncol = 1,rel_heights = c(0.5, 5)
)
```


# Looking at the niches in space

```{r}
#| fig-height: 12
#| fig-width: 24
graph_dt = melt(
    as.data.table(cbind(
        spatialCoords(banksy_embeddings),
        colData(banksy_embeddings)[c(
            'sample_id','cell_id', 'histology',
            grep('Harmony.*.res',clusterNames(banksy_embeddings), value = TRUE)
        )]
    )),id.vars = c('sample_id','cell_id','histology','sdimx','sdimy'),
    variable.name = 'Cluster Type',value.name = 'Cluster ID'
)[,`Cluster ID` := as.factor(as.numeric(`Cluster ID`))]

g = ggplot(
    graph_dt[`Cluster Type` == niche_clust_name], 
    aes(x=sdimy, y=sdimx, color=`Cluster ID`)
) +
    facet_wrap(histology ~ sample_id, nrow = 2, scale = 'free_x') +
    # subset cells
    # geom_point(size = 0.3, alpha = 0.5) + 
    # all cells
    geom_point(size = 0.00001) + 
    scale_color_manual(values = niche_color_manual) +
    theme_classic() + 
    theme(
        legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        text = element_text(size = 30),
        strip.text = element_text(size = 20),
        strip.background = element_blank()
    ) + guides(
        color = guide_legend(override.aes = list(size=5, alpha = 1), nrow = 3, byrow = TRUE)
    ) + labs(
        title = 'BANKSY Niches in Space',
        x = '',y = ''
    ) +
    coord_flip()
g
# p = ggplotly(g)
# saveWidget(p, paste0(workdir,'/youyun/plgg/code/niche/niche_identity_spatial.html'), selfcontained = TRUE)
```


# What are the niches?

## Setting up Neighborhood Analysis

Here we will look at the niches in the data and annotate them using the cell types of self and neighbors.

```{r}
k_geom = 15
locs = spatialCoords(banksy_embeddings)
knn = dbscan::kNN(x = locs, k = k_geom)
knnDT <- data.table(
    from = rep(seq_len(nrow(knn$id)), k_geom),
    from_id = rep(rownames(knn$id), k_geom),
    to = as.vector(knn$id),
    to_id = rownames(knn$id)[as.vector(knn$id)],
    distance = as.vector(knn$dist)
)
cell_cell_type_dt = data.table(
    cell_id = rownames(colData(banksy_embeddings)),
    cell_type = as.numeric(colData(banksy_embeddings)[,cell_clust_name]),
    niche_type = as.factor(colData(banksy_embeddings)[,niche_clust_name]),
    sample_id = colData(banksy_embeddings)$sample_id,
    histology = colData(banksy_embeddings)$histology
)

cell_type_cols = setdiff(colnames(prelim_annotation_res1),'cluster')
niche_self_neighbor_dt = knnDT[,c('from','to') := NULL][
    cell_cell_type_dt[,.(cell_id,cell_type)], on = c('to_id' = 'cell_id'),
    cell_type_to := cell_type
][cell_cell_type_dt, on = c('from_id' = 'cell_id')][
    ,.(
    cell_id = from_id, cell_type, niche_type, sample_id, 
    histology, neighbor_cell_type = cell_type_to
)][
    prelim_annotation_res1, on = c('cell_type' = 'cluster')
][
    prelim_annotation_res1, on = c('neighbor_cell_type' = 'cluster'),
    paste0(cell_type_cols,' Neighbors') := mget(paste0('i.',cell_type_cols))
][order(cell_id)]
niche_self_dt = unique(niche_self_neighbor_dt[
    ,mget(c('cell_id','cell_type','niche_type','sample_id','histology',cell_type_cols))
])
```


## What is the distribution of niches?

Across samples and histologies

```{r}
#| layout-ncol: 1
ggplot(niche_self_dt) + 
    geom_bar(aes(x = sample_id, fill = as.factor(niche_type)), position = 'fill') + 
    facet_wrap(~histology, scale = 'free_x') +
    theme_minimal() + theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 15), legend.position = 'bottom'
    ) + scale_fill_manual(values = niche_color_manual) +
    labs(
        title = 'Distribution of Niches Across Samples',
        x = 'Sample ID',
        y = 'Proportion of Cells'
    ) + guides(
        fill = guide_legend(title = 'Niche Type', nrow = 2)
    )

ggplot(niche_self_dt) + 
    geom_bar(aes(x = sample_id, fill = as.factor(niche_type)), position = 'stack') + 
    facet_wrap(~histology, scale = 'free') +
    theme_minimal() + theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 15), legend.position = 'bottom'
    ) + scale_fill_manual(values = niche_color_manual) +
    labs(
        title = 'Number of Niches Across Samples',
        x = 'Sample ID',
        y = 'Proportion of Cells'
    ) + guides(
        fill = guide_legend(title = 'Niche Type', nrow = 2)
    )
```

## Are there niches exclusive or prevalent across samples and histologies?

Some statistical testing

```{r}
histology_niche_enrichment = pairwise_categorical_continuous_enrichment_wilcox_test(dcast(
    niche_self_dt[
        !is.na(get(cell_type_to_test)),.(proportion = as.numeric(.N)), .(
            sample_id, histology, niche_type 
        )
    ][,proportion := proportion/sum(proportion), .(sample_id, histology)],
    sample_id + histology ~ niche_type, 
    value.var = 'proportion', fill = 0
), padj_method = 'BH')
setnames(
    histology_niche_enrichment, 1:2,
    c('histology','niche_type')
)
ggplot(histology_niche_enrichment[niche_type %in% niches_to_show]) +
    geom_point(aes(
        x = as.factor(as.numeric(niche_type)), y = histology, size = N_var2, 
        # color by odds_ratios column capped at percentile 95
        fill = delta, color = ifelse(
            q_value < 0.25 & as.numeric(delta) > 0, 
            'significant','not significant'
        )
    ), shape = 21, stroke = 1) +
    theme(
        text = element_text(size = 15), legend.position = 'right'
    ) + scale_fill_viridis_c() +
    scale_size_area(max_size = 20) +
    scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
    labs(
        title = 'Enrichment of Niches Across Histologies',
        x = 'Niche Type',
        y = 'Histology',
        size = 'Average Proportion',
        fill = 'Delta Proportions',
        color = 'Significance'
    ) + guides(
        fill = guide_legend(title = 'Avg Delta'),
        color = guide_legend(title = 'Significance')
    )
```

## Is each niche exclusive to one self-cell-type?

```{r}
ggplot(niche_self_dt) + 
    geom_bar(aes(x = niche_type, fill = as.factor(get(cell_type_to_test))), position = 'fill') + 
    theme_minimal() + theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 15), legend.position = 'bottom'
    ) + scale_fill_manual(values = cell_type_color_manual) +
    labs(
        title = 'Distribution of Self Cell Type By Niche',
        x = 'Niche Type',
        y = 'Proportion of Cell Type'
    ) + guides(
        fill = guide_legend(title = cell_type_to_test, nrow = 4)
    )
ggplot(
    unique(niche_self_neighbor_dt[
        ,mget(c('cell_id','cell_type','niche_type','sample_id','histology',cell_type_cols))
    ])
) + geom_bar(aes(x = niche_type, fill = as.factor(get(cell_type_to_test))), position = 'stack') + 
    theme_minimal() + theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 15), legend.position = 'bottom'
    ) + scale_fill_manual(values = cell_type_color_manual) +
    labs(
        title = 'Distribution of Self Cell Type By Niche',
        x = 'Niche Type',
        y = 'Proportion of Cell Type'
    ) + guides(
        fill = guide_legend(title = cell_type_to_test, nrow = 4)
    )
```

```{r}
ggplot(niche_self_dt[niche_type %in% niches_to_show]) + 
    geom_bar(aes(x = niche_type, fill = as.factor(get(cell_type_to_test))), position = 'fill') + 
    theme_minimal() + theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 15), legend.position = 'bottom'
    ) + scale_fill_manual(values = cell_type_color_manual) +
    labs(
        title = 'Distribution of Self Cell Type By Niche',
        x = 'Niche Type',
        y = 'Proportion of Cell Type'
    ) + guides(
        fill = guide_legend(title = cell_type_to_test, nrow = 4)
    )
ggplot(
    unique(niche_self_neighbor_dt[
        ,mget(c('cell_id','cell_type','niche_type','sample_id','histology',cell_type_cols))
    ])[niche_type %in% niches_to_show]
) + geom_bar(aes(x = niche_type, fill = as.factor(get(cell_type_to_test))), position = 'stack') + 
    theme_minimal() + theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 15), legend.position = 'bottom'
    ) + scale_fill_manual(values = cell_type_color_manual) +
    labs(
        title = 'Distribution of Self Cell Type By Niche',
        x = 'Niche Type',
        y = 'Proportion of Cell Type'
    ) + guides(
        fill = guide_legend(title = cell_type_to_test, nrow = 4)
    )
```

Some statistical testing

We are using metap here for fisher's method for combining p-values. It is important to note that here, `metap` is throwing out p-values == 0 because this would give -Inf after log transformation required for the fisher's method. There is no real p-value of 0 but this does happen because of the numeric lower bound of R, and so even though they are thrown out, they still contain meaningful information and also very significant asssociation. Therefore, we are likely underestimating the significance of the combined p-values. 

```{r}
#| fig-width: 20
#| fig-height: 12
self_cell_type_niche_enrichment_per_sample = do.call('rbind',mapply(, 
    function(x, y ){pairwise_categorical_enrichment_test(
        niche_self_dt[sample_id == x]$niche_type, niche_self_dt[sample_id == x][[cell_type_to_test]]
    )[,c('sample_id','histology') := list(x, y)]},
    sample_histology$sample_id, sample_histology$histology, 
    SIMPLIFY = FALSE
))
self_cell_type_niche_enrichment_by_histology = self_cell_type_niche_enrichment_per_sample[
    ,.(
        N_var1 = mean(N_var1), N_var2 = mean(N_var2),
        N_var1var2 = mean(N_var1var2/N_var1), odds_ratios = mean(odds_ratios),
        p_value = ifelse(
            sum((p_value > 0) & (p_value <= 1)) >= 2,
            sumlog(p_value)$p, p_value
        )
    ),.(var1, var2, histology)
][
    ,q_value := p.adjust(p_value, method = 'BY')
]
setnames(
    self_cell_type_niche_enrichment_by_histology, 1:2,
    c('niche_type', 'cell_type')
)
ggplot(self_cell_type_niche_enrichment_by_histology[niche_type %in% niches_to_show]) +
    geom_point(aes(
        x = as.factor(as.numeric(niche_type)), y = cell_type, size = N_var1var2, 
        # color by odds_ratios column capped at percentile 95
        fill = pmin(
            quantile(log(odds_ratios), 0.90, na.rm = TRUE), 
            log(odds_ratios), na.rm = TRUE
        ), color = ifelse(
            q_value < 0.05 & odds_ratios > 1, 
            'significant','not significant'
        ), alpha = ifelse(
            q_value < 0.05 & odds_ratios > 1, 
            'significant','not significant'
        )
    ), shape = 21, stroke = 1) +
    facet_wrap(~histology, axes = 'all') +
    theme(
        text = element_text(size = 15), legend.position = 'right'
    ) + scale_fill_viridis_c() + scale_alpha_manual(
        values = c('significant' = 1,'not significant' = 0.3)
    ) + scale_size_area(
        max_size = 10,
        breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5),
        labels = c('0%','10%','20%','30%','40%','50%')
    ) + scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
    labs(
        title = 'Enrichment of Self Cell Type By Niche',
        x = 'Niche Type',
        y = 'Cell Type',
        size = '%. Niche\n Identity',
        fill = 'log OR',
        color = 'Significance'
    ) + guides(
        fill = guide_legend(title = 'log OR'),
        color = guide_legend(title = 'Significance'),
        alpha = 'none'
    )

ggplot(self_cell_type_niche_enrichment_by_histology[niche_type %in% niches_to_show]) +
    geom_point(aes(
        x = as.factor(as.numeric(niche_type)), y = cell_type, size = N_var1var2, 
        # color by odds_ratios column capped at percentile 95
        fill = pmin(
            quantile(log(odds_ratios), 0.90, na.rm = TRUE), 
            log(odds_ratios), na.rm = TRUE
        ), color = ifelse(
            q_value < 0.05 & odds_ratios > 1, 
            'significant','not significant'
        )
    ), shape = 21, stroke = 1) +
    facet_wrap(~histology, axes = 'all') +
    theme(
        text = element_text(size = 15), legend.position = 'right'
    ) + scale_fill_viridis_c() +
    scale_size_area(
        max_size = 10,
        breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5),
        labels = c('0%','10%','20%','30%','40%','50%')
    ) + scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
    labs(
        title = 'Enrichment of Self Cell Type By Niche',
        x = 'Niche Type',
        y = 'Cell Type',
        size = '%. Niche\n Identity',
        fill = 'log OR',
        color = 'Significance'
    ) + guides(
        fill = guide_legend(title = 'log OR'),
        color = guide_legend(title = 'Significance'),
    )
```

## Does each niche represent a group of cells that are isotropic or anisotropic?

```{r}
ggplot(
    niche_self_neighbor_dt[
        niche_type %in% niches_to_show,.(proportion = .N/k_geom), .(
            cell_id, niche_type, 
            neighbor_cell_type = get(paste0(cell_type_to_test,' Neighbors'))
        )
    ]
) + geom_boxplot(aes(
    x = neighbor_cell_type, y = proportion, 
    fill = neighbor_cell_type
), outlier.size = 0.1) +
    facet_wrap(~niche_type) +
    theme_minimal() + theme(
        axis.text.x = element_blank(),
        text = element_text(size = 15), legend.position = 'bottom'
    ) + scale_fill_manual(values = cell_type_color_manual) +
    labs(
        title = 'Distribution of Neighbor Cell Type By Niche',
        x = 'Neighbor Cell Type',
        y = 'Proportion of Neighbor Cell Type'
    ) + guides(
        fill = guide_legend(title = cell_type_to_test, nrow = 4)
    )
```

Some statistical testing

```{r}
#| fig-width: 20
#| fig-height: 12
neighbor_cell_type_niche_enrichment_per_sample = do.call('rbind',mapply( 
    function(x, y){pairwise_categorical_continuous_enrichment_t_test(dcast(
        niche_self_neighbor_dt[
            !is.na(get(paste0(cell_type_to_test,' Neighbors'))) & sample_id == x,
            .(proportion = .N/k_geom), .(
                cell_id, niche_type, 
                neighbor_cell_type = get(paste0(cell_type_to_test,' Neighbors'))
            )
        ],
        cell_id + niche_type ~ neighbor_cell_type, 
        value.var = 'proportion', fill = 0
    ))[, c('sample_id','histology') := list(x, y)]},
    sample_histology$sample_id, sample_histology$histology,
    SIMPLIFY = FALSE
))
neighbor_cell_type_niche_enrichment_by_histology = neighbor_cell_type_niche_enrichment_per_sample[
    ,.(
        N_var1 = mean(N_var1), N_var2 = mean(N_var2),
        N_var1var2 = mean(N_var1var2), delta = mean(delta),
        p_value = ifelse(
            sum((p_value > 0) & (p_value <= 1)) >= 2,
            sumlog(p_value)$p, p_value
        )
    ),.(niche_type, continuous_cols, histology)
][
    ,q_value := p.adjust(p_value, method = 'BY')
]
neighbor_cell_type_niche_enrichment_by_histology[
    ,.(sum_prop = sum(N_var2)),.(niche_type, histology)
]
ggplot(neighbor_cell_type_niche_enrichment_by_histology[niche_type %in% niches_to_show]) +
    geom_point(aes(
        x = as.factor(as.numeric(niche_type)), y = continuous_cols, 
        size = N_var2,fill = delta, color = ifelse(
            q_value < 0.05 & as.numeric(delta) > 0, 
            'significant','not significant'
        ), alpha = ifelse(
            q_value < 0.05 & as.numeric(delta) > 0, 
            'significant','not significant'
        )
    ), shape = 21, stroke = 1) + 
    facet_wrap(~histology, axes = 'all') +
    theme(
        text = element_text(size = 15), legend.position = 'right'
    ) + scale_fill_viridis_c() + scale_alpha_manual(
        values = c('significant' = 1,'not significant' = 0.3)
    ) + scale_size_area(
        max_size = 10,
        breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5),
        labels = c('0%','10%','20%','30%','40%','50%')
    ) + scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
    labs(
        title = 'Enrichment of Neighbor Cell Type By Niche',
        x = 'Niche Type',
        y = 'Neighbor Cell Type',
        size = '% Neighboring \nIdentity',
        fill = 'Delta',
        color = 'Significance',
    ) + guides(
        fill = guide_legend(title = 'Delta'),
        color = guide_legend(title = 'Significance'),
        alpha = 'none'
    )

ggplot(neighbor_cell_type_niche_enrichment_by_histology[niche_type %in% niches_to_show]) +
    geom_point(aes(
        x = as.factor(as.numeric(niche_type)), y = continuous_cols, 
        size = N_var2,fill = delta, color = ifelse(
            q_value < 0.05 & as.numeric(delta) > 0, 
            'significant','not significant'
        )
    ), shape = 21, stroke = 1) + 
    facet_wrap(~histology, axes = 'all') +
    theme(
        text = element_text(size = 15), legend.position = 'right'
    ) + scale_fill_viridis_c() + scale_size_area(
        max_size = 10,
        breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5),
        labels = c('0%','10%','20%','30%','40%','50%')
    ) + scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
    labs(
        title = 'Enrichment of Neighbor Cell Type By Niche',
        x = 'Niche Type',
        y = 'Neighbor Cell Type',
        size = '% Neighboring \nIdentity',
        fill = 'Delta',
        color = 'Significance',
    ) + guides(
        fill = guide_legend(title = 'Delta'),
        color = guide_legend(title = 'Significance'),
    )
```

## Pulling everything together to understand the niches components

```{r}
#| fig-width: 20
#| fig-height: 12
library(scales)
including_niches = intersect(
    intersect(
        unique(histology_niche_enrichment$niche_type),
        unique(self_cell_type_niche_enrichment_by_histology$niche_type)
    ),
    unique(neighbor_cell_type_niche_enrichment_by_histology$niche_type) 
)
enrichment_overview_dt = merge(
    rbind(
        self_cell_type_niche_enrichment_by_histology[
            niche_type %in% including_niches,
            .(cell_type, size = N_var1var2/sum(N_var1var2), q_value, test_statistic = odds_ratios),
            .(niche_type,histology)
        ][, data_type := 'Self Cell Type'],
        neighbor_cell_type_niche_enrichment_by_histology[
            niche_type %in% including_niches,
            .(cell_type = continuous_cols, size = N_var2, q_value, test_statistic = delta),
            .(niche_type,histology)
        ][, data_type := 'Neighbor Cell Type']
    ),histology_niche_enrichment[
        niche_type %in% including_niches
        ,.(q_value_histo = q_value, test_statistic_histo = delta), 
        .(niche_type, histology)
    ], by = c('niche_type','histology')
)

histo_vector = structure(
    c('CN','CLN','GG','PA'), 
    names = c('central neurocytoma','cerebellar liponeurocytoma','ganglioglioma','pilocytic astrocytoma')
)
enrichment_overview_dt[, histology := histo_vector[histology]]

ggplot(enrichment_overview_dt[q_value_histo <= 0.25 & as.numeric(test_statistic_histo) > 0]) + 
    geom_point(aes(
        x = data_type, y = cell_type, size = size, 
        # fill = q_value, 
        color = ifelse(
            # if it's the fishers test, we use a odds ratio > 1 as positive enrichment
            (q_value < 0.05 & data_type == 'Self Cell Type' & test_statistic > 1) |
            # if it's the continuous test, we use a delta > 0 as positive enrichment
                (q_value < 0.05 & data_type == 'Neighbor Cell Type' & test_statistic > 0), 
            'sig. enriched',ifelse(
                # if it's the fishers test, we use a odds ratio > 1 as positive enrichment
                (q_value < 0.05 & data_type == 'Self Cell Type' & test_statistic < 1) |
                # if it's the continuous test, we use a delta > 0 as positive enrichment
                    (q_value < 0.05 & data_type == 'Neighbor Cell Type' & test_statistic < 0), 
                'sig. depleted','not sig.'
            )
        )
    )) + 
    facet_wrap(as.numeric(niche_type) ~ histology, nrow = 2) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 20), 
        legend.position = 'bottom',
        strip.text = element_text(size = 15),
        strip.background = element_blank(),
        panel.spacing = unit(1, "lines")
    ) + #scale_fill_viridis_c()  +
    scale_size_area(max_size = 10) +
    scale_x_discrete(
        labels = c('Self','Neighbor'),
        limits = c('Self Cell Type','Neighbor Cell Type')
    ) +
    scale_color_manual(values = c(
        'sig. enriched' = 'firebrick3', 'sig. depleted' = 'dodgerblue2',
        'not sig.' = 'slategrey'
    )) +
    labs(
        title = 'Enrichment of Cell Types By Niche',
        x = 'Niche Type',
        y = 'Cell Type',
        size = 'Proportion Cell Type',
        fill = 'q-value',
        color = 'Significance'
    ) + guides(
        size = guide_legend(title = 'Proportion', nrow = 2),
        color = guide_legend(title = 'Significance', nrow = 2)
    )
```

### Only Significantly Enriched Niches

```{r}
lapply(
    histo_vector, function(x){
        ggplot(enrichment_overview_dt[q_value_histo <= 0.25 & as.numeric(test_statistic_histo) > 0 & histology == x]) + 
            geom_point(aes(
                x = data_type, y = cell_type, size = size, 
                # fill = q_value, 
                color = ifelse(
                    # if it's the fishers test, we use a odds ratio > 1 as positive enrichment
                    (q_value < 0.05 & data_type == 'Self Cell Type' & test_statistic > 1) |
                    # if it's the continuous test, we use a delta > 0 as positive enrichment
                        (q_value < 0.05 & data_type == 'Neighbor Cell Type' & test_statistic > 0), 
                    'sig. enriched',ifelse(
                        # if it's the fishers test, we use a odds ratio > 1 as positive enrichment
                        (q_value < 0.05 & data_type == 'Self Cell Type' & test_statistic < 1) |
                        # if it's the continuous test, we use a delta > 0 as positive enrichment
                            (q_value < 0.05 & data_type == 'Neighbor Cell Type' & test_statistic < 0), 
                        'sig. depleted','not sig.'
                    )
                )
            )) + 
            facet_wrap(histology ~ as.numeric(niche_type), nrow = 1) +
            theme(
                axis.text.x = element_text(angle = 45, hjust = 1),
                text = element_text(size = 20), 
                legend.position = 'bottom',
                strip.text = element_text(size = 15),
                strip.background = element_blank(),
                panel.spacing = unit(1, "lines")
            ) + #scale_fill_viridis_c()  +
            scale_size_area(max_size = 10) +
            scale_x_discrete(
                labels = c('Self','Neighbor'),
                limits = c('Self Cell Type','Neighbor Cell Type')
            ) +
            scale_color_manual(values = c(
                'sig. enriched' = 'firebrick3', 'sig. depleted' = 'dodgerblue2',
                'not sig.' = 'slategrey'
            )) +
            labs(
                title = 'Enrichment of Cell Types By Niche',
                x = 'Niche Type',
                y = 'Cell Type',
                size = 'Proportion Cell Type',
                fill = 'q-value',
                color = 'Significance'
            ) + guides(
                size = guide_legend(title = 'Proportion', nrow = 2 ),
                color = guide_legend(title = 'Significance', nrow = 2)
            )
    }
)

```

### All Niches by Histology

```{r}
#| fig-width: 20
#| fig-height: 12
lapply(
    histo_vector, function(x){
        ggplot(enrichment_overview_dt[histology == x]) + 
            geom_point(aes(
                x = data_type, y = cell_type, size = size, 
                # fill = q_value, 
                color = ifelse(
                    # if it's the fishers test, we use a odds ratio > 1 as positive enrichment
                    (q_value < 0.05 & data_type == 'Self Cell Type' & test_statistic > 1) |
                    # if it's the continuous test, we use a delta > 0 as positive enrichment
                        (q_value < 0.05 & data_type == 'Neighbor Cell Type' & test_statistic > 0), 
                    'sig. enriched',ifelse(
                        # if it's the fishers test, we use a odds ratio > 1 as positive enrichment
                        (q_value < 0.05 & data_type == 'Self Cell Type' & test_statistic < 1) |
                        # if it's the continuous test, we use a delta > 0 as positive enrichment
                            (q_value < 0.05 & data_type == 'Neighbor Cell Type' & test_statistic < 0), 
                        'sig. depleted','not sig.'
                    )
                )
            )) + 
            facet_wrap(histology ~ as.numeric(niche_type), nrow = 2) +
            theme(
                axis.text.x = element_text(angle = 45, hjust = 1),
                text = element_text(size = 20), 
                legend.position = 'bottom',
                strip.text = element_text(size = 15),
                strip.background = element_blank(),
                panel.spacing = unit(1, "lines")
            ) + #scale_fill_viridis_c()  +
            scale_size_area(max_size = 10) +
            scale_x_discrete(
                labels = c('Self','Neighbor'),
                limits = c('Self Cell Type','Neighbor Cell Type')
            ) +
            scale_color_manual(values = c(
                'sig. enriched' = 'firebrick3', 'sig. depleted' = 'dodgerblue2',
                'not sig.' = 'slategrey'
            )) +
            labs(
                title = 'Enrichment of Cell Types By Niche',
                x = 'Niche Type',
                y = 'Cell Type',
                size = 'Proportion Cell Type',
                fill = 'q-value',
                color = 'Significance'
            ) + guides(
                size = guide_legend(title = 'Proportion', nrow = 2),
                color = guide_legend(title = 'Significance', nrow = 2)
            )
    }
)
```