---
title: "BANKSY Cell Type in Space"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

This doc is solely dedicated to visualization of cell types in space

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(dbscan)
library(circlize)
library(Seurat)
library(plotly)
library(scales)
library(htmlwidgets)
library(metap)

# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
```


# Load the data

The objects are here:

RDS object with Leiden Clusters: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds`

    -   `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_embeddings/total_spatial_banksy_clusters_subset_20240719_092550.rds`


Cell type clusters annotation: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv`  


```{r}
# load the banksy embeddings
banksy_embeddings <- readRDS(ifelse(
    workdir == "/xchip/beroukhimlab/", paste0(
        workdir, "youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/banksy_clusters_connected_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
    ), "/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_20_30_res_0.75_1_20241223_222156.rds"
))
colData(banksy_embeddings)$sample_id = factor(colData(banksy_embeddings)$sample_id)

# specify the cluster names 
# this is mainly to keep track of which resolution we are using
cell_res = 1
kc = 50
cell_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k%.1i_res%.1f', kc ,cell_res))

# # for when there is new manual annotations
# # Load the cell type clusters annotation
# prelim_annotation_res1 = fread(paste0(
#     workdir,'coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv'
# ))
# for now
cell_cluster_count = as.data.table(colData(banksy_embeddings))[
    ,.(N = .N),.(cell_clusters = get(cell_clust_name))
][order(N, decreasing = TRUE)]
kable(t(cell_cluster_count))
cell_clusters = cell_cluster_count[N/ncol(banksy_embeddings) > 0.001]$cell_clusters
prelim_annotation_res1 = data.table(
    cluster = cell_clusters,
    `Coarse Cell Type` = cell_clusters,
    `Fine Cell Type` = cell_clusters,
    `Fine Cell Type UMAP` = cell_clusters
)
```


Build the color palette for all niches and cell types downstream

```{r}
set.seed(5555)
cell_type_to_test = 'Coarse Cell Type'
cell_type_to_test = 'Fine Cell Type'
cell_type_to_test = 'Fine Cell Type UMAP'

cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))

cell_type_color_manual = structure(
    pals::polychrome()[c(1:length(cell_type_class))],
    names = as.character(cell_type_class)
)

```

# Looking at the cell types in space all together

```{r}
#| fig-height: 12
#| fig-width: 24
graph_dt = melt(
    as.data.table(cbind(
        spatialCoords(banksy_embeddings),
        colData(banksy_embeddings)[c(
            'sample_id','cell_id', 'histology',
            grep('Harmony.*.res',clusterNames(banksy_embeddings), value = TRUE)
        )]
    )),id.vars = c('sample_id','cell_id','histology','sdimx','sdimy'),
    variable.name = 'Cluster Type',value.name = 'Cluster ID'
)[`Cluster Type` == cell_clust_name][
    prelim_annotation_res1[, cluster := as.character(cluster)], 
    on = c('Cluster ID' = 'cluster')
]


g = ggplot(
    graph_dt[`Cluster Type` == cell_clust_name], 
    aes(x=sdimy, y=sdimx, color=get(cell_type_to_test))
) +
    facet_wrap(histology ~ sample_id, nrow = 2, scale = 'free_x') +
    # subset cells
    # geom_point(size = 0.3, alpha = 0.5) + 
    # all cells
    geom_point(size = 0.00001) + 
    scale_color_manual(values = cell_type_color_manual) +
    theme_classic() + 
    theme(
        legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        text = element_text(size = 30),
        strip.text = element_text(size = 20),
        strip.background = element_blank()
    ) + guides(
        color = guide_legend(
            override.aes = list(size=5, alpha = 1), nrow = 4, byrow = TRUE,
            title = 'Cell Type'
        )
    ) + labs(
        title = 'BANKSY Cell Types in Space',
        x = '',y = ''
    ) +
    coord_flip()
g
```


# Looking at the niches in space separately

```{r}
#| fig-height: 32
#| fig-width: 24
x = unique(graph_dt[order(-histology)]$sample_id)[1]
lapply(unique(graph_dt[order(-histology)]$sample_id), function(x){
    ggplot(
        graph_dt[`Cluster Type` == cell_clust_name & sample_id == x], 
        aes(x=sdimy, y=sdimx, color=get(cell_type_to_test))
    ) +
        facet_wrap(
            histology ~ sample_id + get(cell_type_to_test), labeller = label_wrap_gen(), 
            ncol = 6, scale = 'free_x'
        ) +
        # subset cells
        # geom_point(size = 0.3, alpha = 0.5) + 
        # all cells
        geom_point(size = 0.00001) + 
        scale_color_manual(values = cell_type_color_manual) +
        theme_classic() + 
        theme(
            legend.position = "bottom",
            axis.text.x=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks=element_blank(),
            text = element_text(size = 30),
            strip.text = element_text(size = 20),
            strip.background = element_blank()
        ) + guides(
            color = guide_legend(
                override.aes = list(size=5, alpha = 1), nrow = 4, byrow = TRUE,
                title = 'Cell Type'
            )
        ) + labs(
            title = paste0('BANKSY Cell Types in Space: ',x),
            x = '',y = ''
        ) +
        coord_flip()
})
```

```{r}
#| fig-height: 12
#| fig-width: 24
x = unique(graph_dt[`Cluster Type` == cell_clust_name][[cell_type_to_test]])[1]
lapply(sort(unique(graph_dt[`Cluster Type` == cell_clust_name][[cell_type_to_test]])), function(x){
    ggplot(
        graph_dt[`Cluster Type` == cell_clust_name & get(cell_type_to_test) == x], 
        aes(x=sdimy, y=sdimx, color=get(cell_type_to_test))
    ) +
        facet_wrap(
            histology ~ sample_id, 
            nrow = 2, scale = 'free_x'
        ) +
        # subset cells
        # geom_point(size = 0.3, alpha = 0.5) + 
        # all cells
        geom_point(size = 0.00001) + 
        scale_color_manual(values = cell_type_color_manual) +
        theme_classic() + 
        theme(
            legend.position = "bottom",
            axis.text.x=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks=element_blank(),
            text = element_text(size = 30),
            strip.text = element_text(size = 20),
            strip.background = element_blank()
        ) + guides(
            color = guide_legend(
                override.aes = list(size=5, alpha = 1), nrow = 1, byrow = TRUE,
                title = 'Cell Type'
            )
        ) + labs(
            title = paste0('BANKSY Cell Types in Space: ',x),
            x = '',y = ''
        ) +
        coord_flip()
})
```

