---
title: "Analyzing BANKSY Embedding for Cell Identities"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

In this doc, we will be looking at the banksy clusters and try to interpret them based on their identity, and other characteristics.

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(ComplexHeatmap)
library(circlize)


# local vs remote
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (Sys.getenv("HOME") == "/PHShome/yz762"){
    # in eristwo, the home directory is at /PHShome/yz762
    workdir <- "/data/beroukhim1/"
} else if (Sys.getenv("HOME") == "/home/yz762"){
    # on the new server, the home directory is at /home/yz762
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    # in dipg or uger, the home directory is '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

```


# Load the data

The objects are here:

1.  RDS object with Leiden Clusters: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240711_135041.rds`

    -   `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_embeddings/total_spatial_banksy_clusters_subset_20240711_135041.rds`

Metadata is here:

1.  `/Users/youyun/Documents/HMS/PhD/beroukhimlab/broad_mount/coja/Spatial_PLGG/data/metadata/Xenium_PS.xlsx`

```{r}
banksy_embeddings <- readRDS(ifelse(
    basename(workdir) != 'dfci_mount', paste0(
        workdir, "youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
    ), "/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
))
colData(banksy_embeddings)$sample_id <- factor(colData(banksy_embeddings)$sample_id)

cell_type_markers <- fread(paste0(
    workdir, "youyun/plgg/data/markers/final_manual_markers.csv"
))

xenium_markers <- fread(paste0(
    workdir, "youyun/plgg/data/xenium_selection/Xenium_hBrain_v1_metadata.csv"
))
```

Some additional neuronal markers 

```{r}
# General neuronal markers
general_neuronal_markers <- c(
    "RELN", "SYNPR", "CUX2", "RORB", "CHODL", 
    "BCAN", "NTNG1", "NTNG2", "NXPH2", "APP"
)
xenium_markers[Genes %in% general_neuronal_markers]

# Excitatory neuronal markers
excitatory_neuronal_markers <- c(
    "SLC17A6", "SLC17A7", "RORB", "CUX2", 
    "RELN", "TAC1", "TACR1"
)
xenium_markers[Genes %in% excitatory_neuronal_markers]

# Inhibitory neuronal markers
inhibitory_neuronal_markers <- c(
    "GAD1", "GAD2", "PVALB", "VIP", "SST", 
    "LHX6", "CHODL", "RELN", "HTR2A", "HTR2C"
)
xenium_markers[Genes %in% inhibitory_neuronal_markers]

cell_type_markers = rbind(
    cell_type_markers,
    data.table(
        marker = general_neuronal_markers, annotation = 'Neuronal',
        category = 'Cell Type', xenium = TRUE
    ),
    data.table(
        marker = excitatory_neuronal_markers, annotation = 'Excitatory Neuronal',
        category = 'Cell Type', xenium = TRUE
    ),
    data.table(
        marker = inhibitory_neuronal_markers, annotation = 'Inhibitory Neuronal',
        category = 'Cell Type', xenium = TRUE
    )
)
```

# Looking at the harmony corrected embeddings

We will first look at the extent of batch effect in the cohort by looking at the embeddings before and after harmony correction. 

```{r}
plot_grid(
    ggdraw() + draw_label("BANKSY Embedding UMAP Cell Identity and Niches", fontface='bold'),
    plot_grid(
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.2", 
            point_size = 0.1,point_alpha = 0.5, color_by = "sample_id"
        ) +
            theme(legend.title = element_blank()) +
            guides(colour = 'none'),
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.8", 
            point_size = 0.1,point_alpha = 0.5, color_by = "sample_id"
        ) +
            theme(legend.title = element_blank()) +
            guides(colour = guide_legend(override.aes = list(size = 5, alpha = 1))),
        nrow = 1,
        rel_widths = c(1, 1.2)
    ), ncol = 1,rel_heights = c(0.1, 5)
)
```

# What do the cell clusters look like?

```{r}
res = 1
kc = 50
clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k%.1i_res%.1f',kc, res))
```

We will be looking at resolution = `r res` for the clusters using the `r clust_name` column.

```{r}
plot_grid(
    ggdraw() + draw_label("BANKSY Embedding UMAP for Cell Identity(L) and Niche(R)", fontface='bold'),
    plot_grid(
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.2", 
            point_size = 0.1, point_alpha = 0.5, 
            color_by = clust_name
        ) +
            theme(legend.title = element_blank()) +
            guides(colour = guide_legend(override.aes = list(size = 5, alpha = 1))),
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.8", 
            point_size = 0.1, point_alpha = 0.5, 
            color_by = gsub('lam0.2','lam0.8',clust_name)
        ) +
            theme(legend.title = element_blank()) +
            guides(colour = guide_legend(override.aes = list(size = 5, alpha = 1))),
        nrow = 1,
        rel_widths = c(1, 1)
    ), ncol = 1,rel_heights = c(0.1, 5)
)
```

Just to get an idea of where each cluster is, we can look at the UMAPs colored individually by the clusters.

```{r}
cell_cluster_count = as.data.table(colData(banksy_embeddings))[
    ,.(N = .N),.(cell_clusters = get(clust_name))
][order(N, decreasing = TRUE)]
kable(t(cell_cluster_count))
cell_clusters = cell_cluster_count[N/ncol(banksy_embeddings) > 0.001]$cell_clusters
# one hot encoded columns
one_hot_encoded_clusters = data.frame(dcast(data.table(
    cell_sample_id = rownames(colData(banksy_embeddings)),
    clusters = paste0('cell_in_',colData(banksy_embeddings)[,clust_name])
), cell_sample_id ~ clusters, length)[order(match(cell_sample_id,rownames(colData(banksy_embeddings))))])
# cbind into the colData
if(all(rownames(colData(banksy_embeddings)) == colData(banksy_embeddings)$cell_sample_id)){
    colData(banksy_embeddings) = cbind(
        colData(banksy_embeddings),
        one_hot_encoded_clusters
    )
}else{
    stop('Cell IDs do not match')
}
```

```{r}
# make the plots
a = lapply(cell_clusters, function(x){
    assign(
        paste0('cell_type_plot_',x),
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.2", 
            point_size = 0.1, point_alpha = 0.5, 
            color_by = paste0('cell_in_',x)
        )  + theme(
            legend.position = "none",
            axis.text=element_blank(),
            axis.ticks=element_blank(),
            axis.title=element_blank()
        ) + labs(title = paste0('Cluster ',x)),
        envir = .GlobalEnv
    )
})

plot_grid(
    ggdraw() + draw_label("BANKSY Embedding UMAP by Cluster", fontface='bold'),
    plot_grid(
        plotlist = mget(paste0('cell_type_plot_',cell_clusters)),
        ncol = 6,
        rel_widths = rep(1,length(cell_clusters))
    ), ncol = 1,rel_heights = c(0.1, 5)
)

```


# What are the cell types in the clusters?

## Manually- and Xenium- annotated cell type markers

```{r}
markers_to_score = rbind(cell_type_markers[
    category == 'Cell Type' & xenium == TRUE,
    .(markers = marker),.(annotation = paste0(annotation, '_manual'))
],xenium_markers[
    ,.(markers = Genes),.(annotation = paste0(Annotation, '_xenium'))
])[
    ,annotation := make.names(annotation)
]
markers_to_score = markers_to_score[annotation %in% markers_to_score[,.N, .(annotation)][N>=2]$annotation]

# Scoring the cells by cell type markers
cell_types = unique(markers_to_score$annotation)
marker_score = do.call('cbind',lapply(cell_types, function(x){
    genes_of_interest = markers_to_score[annotation == x]$markers
    cell_scores = log10(colSums(
        assay(banksy_embeddings[genes_of_interest,], 'normcounts')
    )/length(genes_of_interest) + 1)
    cell_scores_df = data.frame(cell_scores)
    rownames(cell_scores_df) = names(cell_scores)
    colnames(cell_scores_df) = x
    cell_scores_df
}))
colData(banksy_embeddings)[,cell_types] = NULL
if(all(rownames(colData(banksy_embeddings)) == rownames(colData(banksy_embeddings)))){
    colData(banksy_embeddings) = cbind(colData(banksy_embeddings),marker_score)
}else{
    stop('Cell IDs do not match')
}

# make the plots
a = lapply(cell_types, function(x){
    assign(
        paste0('plot_',x),
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.2", 
            point_size = 0.1, point_alpha = 0.5, 
            color_by = x
        ) +
            theme(legend.title = element_blank(), legend.position = 'bottom') +
            guides(colour = guide_legend(override.aes = list(size = 5, alpha = 1), nrow = 1)) +
            labs(title = x),
        envir = .GlobalEnv
    )
})
```

We are going to annotate `{r} length(cell_types)` cell types. and we will now visualize the marker scores. 

```{r}
#| fig-width: 15
#| fig-height: 10
plot_grid(
    ggdraw() + draw_label("BANKSY Embedding UMAP by Manually Curated Markers", fontface='bold'),
    plot_grid(
        plotlist = mget(paste0('plot_',grep('_manual',cell_types, value = TRUE))),
        ncol = 4,
        rel_widths = rep(1,sum(grepl('_manual',cell_types)))
    ), ncol = 1,rel_heights = c(0.1, 5)
)
```

```{r}
#| fig-width: 20
#| fig-height: 20
plot_grid(
    ggdraw() + draw_label("BANKSY Embedding UMAP by Xenium Markers", fontface='bold'),
    plot_grid(
        plotlist = mget(paste0('plot_',grep('_xenium',cell_types, value = TRUE))),
        ncol = 5,
        rel_widths = rep(1,sum(grepl('_xenium',cell_types)))
    ), ncol = 1,rel_heights = c(0.1, 5)
)
```

We should complement the manual approach with more statistics

## Looking at these 'metagenes' statistically

A better way to do this is to test statistically, which clusters are enriched for which signature. We will do that here

```{r}
cell_type_cluster_sig = do.call('rbind',lapply(paste0('cell_in_',cell_clusters), function(x){
    scran_cell_markers = findMarkers(
        as.matrix(t(colData(banksy_embeddings)[,cell_types])),
        groups = colData(banksy_embeddings)[,x],
        test.type="wilcox", direction="up"
    )    
    delta_dt = melt(setDT(
        as.data.frame(colData(banksy_embeddings)[,c(cell_types,x)]),
        keep.rownames = 'cell_sample_id'
    ), id.vars = c('cell_sample_id',x), variable.name = 'cell_type', value.name = 'score')[
        ,.(delta_score = mean(score[get(x) == 1]) - mean(score[get(x) == 0])),.(cell_type)
    ]
    cluster_cell_type = setDT(
        data.frame(scran_cell_markers[[2]]),
        keep.rownames = 'cell_type'
    )[,.(cell_type = make.names(cell_type), Top, p.value, FDR, cluster = x)]
    merge(
        cluster_cell_type, delta_dt, by = 'cell_type'
    )
}))

# make a hierarchically clustered heatmap
cell_type_cluster_sig_matrix = dcast(
    cell_type_cluster_sig[,.(cell_type, cluster, FDR = ifelse(FDR<0.05,'*',''))],
    cluster ~ cell_type, value.var = 'FDR'
)
cell_type_cluster_delta_matrix = dcast(
    cell_type_cluster_sig[,.(cell_type, cluster, delta_score)],
    cluster ~ cell_type, value.var = 'delta_score'
)
plot_sig_mt = as.matrix(cell_type_cluster_sig_matrix[,c(2:ncol(cell_type_cluster_sig_matrix)), with = FALSE])
plot_delta_mt = as.matrix(cell_type_cluster_delta_matrix[,c(2:ncol(cell_type_cluster_delta_matrix)), with = FALSE])
rownames(plot_sig_mt) = cell_type_cluster_sig_matrix$cluster
rownames(plot_delta_mt) = cell_type_cluster_delta_matrix$cluster
col_fun = colorRamp2(c(-0.3,0,0.3), c("red", "white", "green"))
Heatmap(
    plot_delta_mt, name = "Delta Mean Norm Counts", col = col_fun,
    cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%s", plot_sig_mt[i, j]), x, y, gp = gpar(fontsize = 10))
})
```

It is clear that:

1.  Cluster 3 is OPC
2.  Cluster 7 and 10 are astrocytes
3.  Cluster 9 is stromal cells
4.  Cluster 8 and 22 are myeloid cells
5.  Cluster 2 and 6 are RGCs (and are sample specific to the 2 CNs respectively)
6.  Cluster 19 is oligodendrocytes
7.  Cluster 18 are proliferative cells
8.  Cluster 14 is lymphoid cells

This leaves:

1.  Clusters 1, 4, 5, 11, 13, 15, 16, 20, 21, 34 which should all be neurons of some sort. 
2.  Cluster 23 has oligodendrocytic and GBM-TME (xenium) markers
3.  Cluster 24 has GBM-Cancer Cell (xenium) markers


## Looking at the DE Genes in each cluster

```{r}
ggplot(colData(banksy_embeddings)) + 
    geom_bar(aes(x = get(clust_name), fill = as.numeric(get(clust_name)))) + 
    facet_wrap(~sample_id, nrow = 3, scale = 'free_y') +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 15)
    ) + labs(
        x = 'Cluster ID', y = 'Number of Cells (log10)'
    ) + scale_fill_viridis_c(option = 'D') +
    guides(fill = guide_legend(title = 'Cluster ID')) +
    scale_y_log10()
```

```{r}
#| warning: false
banksy_embeddings_seurat = as.Seurat(banksy_embeddings, counts = 'normcounts', data = NULL)
# banksy_embeddings_seurat <- SCTransform(
#     banksy_embeddings_seurat, assay = "originalexp", 
#     verbose = FALSE, variable.features.n = 10000
# )
Idents(banksy_embeddings_seurat) = clust_name
```

### Wilcox Test

```{r}
#| fig-width: 12
#| fig-height: 12
DE_genes_wilcox = data.table(FindAllMarkers(
    banksy_embeddings_seurat[
        ,which(colData(banksy_embeddings)[, clust_name] %in% cell_clusters)
    ], slot = 'data',
    test.use = 'wilcox',
    only.pos = TRUE,
    verbose = FALSE
))[,cluster := as.character(cluster)]

DE_genes_wilcox = merge(
    DE_genes_wilcox,
    xenium_markers[,.(gene = Genes, Annotation)],
    by = 'gene'
)

xenium_markers[,.N,.(Annotation)][order(-N)]
paste0(xenium_markers$Genes, collapse = ', ')

kable(DE_genes_wilcox[order(p_val_adj)][
    p_val_adj <= 0.05 & avg_log2FC >= 1, .(
        N = .N, 
        `Markers` = paste0(
            gene, ' (', Annotation, ')', 
            collapse = ', '
        )
        # `Markers` = paste0(
        #     gene[avg_log2FC >= 1], ' (', 
        #     round(avg_log2FC[avg_log2FC >= 1], 2),  ', ',
        #     Annotation[avg_log2FC >= 1] ,
        #     ')', collapse = ', '
        # )
    ), 
    .(cluster = as.character(cluster))
][order(as.numeric(cluster))])
```

```{r}
DoHeatmap(
    subset(banksy_embeddings_seurat, downsample = 100),
    features = unique(DE_genes_wilcox[
        p_val_adj < 0.05, 
        .(gene = head(gene[order(-avg_log2FC)], 10)), 
        .(cluster = as.numeric(cluster))
    ][order(cluster)]$gene),
    slot = 'data'
) + NoLegend()
```


```{r}
neuronal_markers = FindMarkers(
    banksy_embeddings_seurat[
        ,which(colData(banksy_embeddings)[, clust_name] %in% cell_clusters)
    ], slot = 'data', ident.1 = 11,
    ident.2 = c(1, 4, 5, 13, 15, 16, 20, 21, 34),
    test.use = 'wilcox',
    only.pos = TRUE,
    verbose = FALSE
)
setDT(neuronal_markers, keep.rownames = TRUE)
neuronal_markers$gene_annotation = xenium_markers[match(neuronal_markers$rn, xenium_markers$Genes),]$Annotation
neuronal_markers[p_val_adj <= 0.05][,.N, .(gene_annotation)][order(-N)]

```

```{r}
plotReducedDim(
    banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.2", 
    point_size = 0.1,point_alpha = 0.5, color_by = "SOX4",
    by_exprs_values = 'normcounts'
) +
    theme(legend.title = element_blank()) +
    guides(colour = 'none')
```


### MAST Test

```{r}
# DE_genes_mast = data.table(FindAllMarkers(
#     banksy_embeddings_seurat[
#         ,which(colData(banksy_embeddings)[, clust_name] %in% cell_clusters)
#     ], slot = 'counts',
#     test.use = 'MAST',
#     only.pos = TRUE,
#     verbose = FALSE
# ))

# DE_genes_mast = merge(
#     DE_genes_mast,
#     xenium_markers[,.(gene = Genes, Annotation)],
#     by = 'gene'
# )

# kable(DE_genes_mast[order(p_val_adj)][
#     p_val_adj <= 0.05 & avg_log2FC >= 1, .(
#         N = .N, 
#         `Markers` = paste0(
#             gene[avg_log2FC >= 1], ' (', 
#             round(avg_log2FC[avg_log2FC >= 1], 2),  ', ',
#             Annotation[avg_log2FC >= 1] ,
#             ')', collapse = ', '
#         )
#     ), 
#     .(cluster)
# ][order(cluster)])
```

### Comparing the two tests

```{r}
# de_genes_count = merge(
#     DE_genes_wilcox[
#         p_val_adj <= 0.05 & avg_log2FC >= 1,
#         .N, .(cluster)
#     ][order(cluster)],
#     DE_genes_mast[
#         p_val_adj <= 0.05 & avg_log2FC >= 1,
#         .N, .(cluster)
#     ][order(cluster)],
#     by = 'cluster', all = TRUE,
#     suffixes = c('_wilcox','_mast')
# )
```

### GSEA

code to subset the pathways to only include the genes in the panel and then run fgsea. The results are not very good except for identifying coarse immune cells types

```{r}
# library(fgsea)
# library(msigdbr)
# # https://bioconductor.org/packages/release/bioc/vignettes/fgsea/inst/doc/geseca-tutorial.html#analysis-of-single-cell-rna-seq

# # get human gene set
# pathwaysDF <- msigdbr("human")
# pathways <- split(pathwaysDF$gene_symbol, pathwaysDF$gs_name)
# # Filter pathways to include only genes in the panel
# pathways <- lapply(pathways, function(genes) intersect(genes, xenium_markers$Genes))
# logfc_genes = data.table(FindAllMarkers(
#     banksy_embeddings_seurat[
#         ,which(colData(banksy_embeddings)[, clust_name] %in% cell_clusters)
#     ], slot = 'data',
#     test.use = 'wilcox',
#     logfc.threshold = 0,  # Include all genes regardless of log fold change
#     min.pct = 0,          # Include all genes regardless of expression percentage
#     verbose = FALSE
# ))[,cluster := as.character(cluster)]

# # perform GSEA
# fgsea_results = rbindlist(lapply(
#     c(1,4,5,11,15,16,20,21,34), 
#     function(x){
#         sig_genes = logfc_genes[
#             cluster == x,
#             .(gene, score = avg_log2FC)
#         ][order(-score)]
#         ranked_list <- setNames(sig_genes$score, sig_genes$gene)
#         # Perform GSEA
#         fgsea_results <- fgsea(
#             pathways = pathways,
#             stats = ranked_list,
#             minSize = 2,  # Minimum size of gene sets
#             maxSize = 266,  # Maximum size of gene sets
#             scoreType = "pos"
#         )[,cluster := x]

#         # View significant results
#         fgsea_results[padj <= 0.05]
#     }
# ))

# kable(fgsea_results[,.(
#     pathway, padj, size, cluster = as.character(cluster)
# )][order(as.numeric(cluster))])

```

# Preliminary annotation

With the information above, we can begin to piece together some preliminary definition for each cluster. 

This [Cell Paper](https://www.sciencedirect.com/science/article/pii/S0092867421005018?via%3Dihub#fig1) and this [Nature Paper](https://www.nature.com/articles/s41586-019-1506-7/figures/1)

```{r}
prelim_annotation_res1 = data.table(t(setDT(list(
    `3` = c('OPC','OPC'),
    `10` = c('Astrocyte','Astrocyte'),
    `7` = c('Astrocyte','Astrocyte (VLMC, myeloid)'),
    `9` = c('Stromal','Stromal/Endothelial/VLMC'),
    `8` = c('Myeloid','Myeloid (Endothelial)'),
    `22` = c('Myeloid','Myeloid'),
    `6` = c('RGC','RGC (Pax6)'),
    `2` = c('RGC','RGC'),
    `19` = c('Oligodendrocytes','Oligodendrocytes'),
    `18` = c('Proliferation','Proliferation'),
    `14` = c('Lymphoid','Lymphoid'),
    `1` = c('Neuron','Neuron (RELN)'),
    `4` = c('Neuron','Neuron (NNAT, SOX11)'),
    `5` = c('Neuron','Neuron (52 DEG)'),
    `11` = c('Neuron','Neuron (APP, RELN, SOX4)'),
    `13` = c('Neuron','Neuron (SLC17A7, C1QL3, CRYM, HS3ST2, SLIT3)'),
    `15` = c('Neuron','Neuron (NTNG2)'),
    `16` = c('Neuron','Neuron (PROX1)'),
    `20` = c('Neuron','Neuron (FBLN, EGFR, MYO16)'),
    `21` = c('Neuron','Neuron (SST, GAD1, LHX6)'),
    `34` = c('Neuron','Neuron (SOX9)'),
    `23` = c('Neuron','Neuron (NR4A2)'),
    `24` = c('Neuron','Neuron (IGFBP5)')
))), keep.rownames = 'cluster')[
    , c('Coarse Cell Type', 'Fine Cell Type Annotation') := list(V1, V2)
][,c('V1','V2') := NULL][
    ,`Fine Cell Type` := gsub(' \\(.*','', `Fine Cell Type Annotation`)
][
    order(as.numeric(cluster)),
    `Fine Cell Type UMAP` := paste0(
        `Fine Cell Type`, ' ', c(1:.N)
    ),.(`Fine Cell Type`)
]
setcolorder(
    prelim_annotation_res1,
    c(
        'cluster','Coarse Cell Type',
        'Fine Cell Type','Fine Cell Type UMAP',
        'Fine Cell Type Annotation'
    )
)
kable(prelim_annotation_res1)
write.table(
    prelim_annotation_res1,
    file = paste0(
        workdir,'youyun/plgg/data/xenium_cell_types/banksy_proseg_cell_cluster_annotation.tsv'
    ),
    sep = '\t', quote = FALSE, row.names = FALSE
)
```

## Plotting out the preliminary annotation on UMAP

Plotting out some cohort information:

1.  UMAP before and after harmony correction
2.  UMAP colored by the preliminary annotation
3.  UMAP colored by the sample ID
4.  UMAP colored by the histology

we will first merge in the preliminary annotation into the banksy embeddings

```{r}
original_order = rownames(colData(banksy_embeddings))
merged_coldata = merge(
    colData(banksy_embeddings),
    prelim_annotation_res1[,cluster := as.character(cluster)],
    by.x = clust_name, by.y = 'cluster', all.x = TRUE
)
rownames(merged_coldata) = paste0(merged_coldata$cell_id, '__',merged_coldata$sample_id)
colData(banksy_embeddings) = merged_coldata[original_order,]
```

```{r}
cell_type_to_test = 'Coarse Cell Type'
cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))
cell_type_color_manual = structure(
    pals::alphabet()[c(1:length(cell_type_class))],
    names = cell_type_class
)

plot_grid(
    ggdraw() + draw_label("BANKSY Cell Identity Embedding", fontface='bold'),
    plot_grid(
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.2", 
            point_size = 0.1,point_alpha = 0.5, color_by = cell_type_to_test
        ) + theme(
            legend.title = element_blank(),
            legend.position = 'bottom'
        ) + scale_color_manual(values = cell_type_color_manual) +
            guides(colour = guide_legend(
            override.aes = list(size = 5, alpha = 1),
            nrow = 4
        )) + labs(
            title = 'Cell Type',
            x = 'UMAP1', y = 'UMAP2'
        ),
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.2", 
            point_size = 0.1,point_alpha = 0.5, color_by = "sample_id"
        ) + theme(
            legend.title = element_blank(),
            legend.position = 'bottom',
            legend.justification = "right"
        ) + guides(colour = guide_legend(
            override.aes = list(size = 5, alpha = 1),
            nrow = 4
        )) + labs(
            title = 'Sample IDs',
            x = 'UMAP1', y = 'UMAP2'
        ),
        plotReducedDim(
            banksy_embeddings, "UMAP_Harmony_BANKSY_lam0.2", 
            point_size = 0.1,point_alpha = 0.5, color_by = "histology"
        ) + theme(
            legend.title = element_blank(),
            legend.position = 'bottom'
        ) + guides(colour = guide_legend(
            override.aes = list(size = 5, alpha = 1),
            nrow = 4
        )) + labs(
            title = 'Histology',
            x = 'UMAP1', y = 'UMAP2'
        ),
        
        nrow = 1,
        rel_widths = c(1, 1, 1)
    ), ncol = 1,rel_heights = c(0.1, 5)
)
```

# Cell Type Enrichment Analysis

```{r}
source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
cell_type_to_test = 'Fine Cell Type UMAP'
histology_cell_type_enrichment = pairwise_categorical_continuous_enrichment_wilcox_test(dcast(
    data.table(
        sample_id = colData(banksy_embeddings)$sample_id,
        histology = colData(banksy_embeddings)$histology,
        cluster = as.character(colData(banksy_embeddings)[,clust_name]),
        cell_type = colData(banksy_embeddings)[,cell_type_to_test]
    )[
        cluster %in% cell_clusters, 
        .(proportion = as.numeric(.N)), 
        .(sample_id, histology, cell_type)
    ][, proportion := proportion/sum(proportion), .(sample_id, histology)],
    sample_id + histology ~ cell_type,
    value.var = 'proportion', fill = 0
),padj_method = 'BH')
setnames(
    histology_cell_type_enrichment, 1:2,
    c('histology','cell_type')
)
head(histology_cell_type_enrichment)
```

```{r}
ggplot(histology_cell_type_enrichment) +
    geom_point(aes(
        x = cell_type, 
        y = histology, size = N_var2, 
        # color by odds_ratios column capped at percentile 95
        fill = delta, color = ifelse(
            q_value < 0.25 & as.numeric(delta) > 0, 
            'significant','not significant'
        )
    ), shape = 21, stroke = 1) +
    theme(
        text = element_text(size = 15), legend.position = 'right',
        axis.text.x = element_text(angle = 45, hjust = 1)
    ) + scale_fill_viridis_c() +
    scale_size_area(max_size = 20) +
    scale_color_manual(values = c('significant' = 'red','not significant' = 'black')) +
    labs(
        title = 'Enrichment of Cell Types Across Histologies',
        x = 'Cell Type',
        y = 'Histology',
        size = 'Average Proportion',
        fill = 'Delta Proportions',
        color = 'Significance'
    ) + guides(
        fill = guide_legend(title = 'Avg Delta'),
        color = guide_legend(title = 'Significance')
    )
```