---
title: "Cell-type and Niche-type Enrichment by Tumor Histology"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 13
        fig-height: 5.5
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk:
        results: hold
        R.options:
            width: 120
---

# Introduction

Which BANKSY cell types and niche types are enriched or depleted across
tumor histology, on the final res_ct=0.2/res_ni=0.5 annotated object (see
`XENIUM_RUNBOOK.md` §2a/§2b-2 for how that object was produced). Three
tests, in order:

1. Cell-type proportion vs. histology.
2. Niche-type proportion vs. histology.
3. Niche *composition* (which cell types make up each niche) vs. histology.

Tests 1 and 2 use a **unified beta-binomial one-vs-rest model** (the same engine
as the pathology analyses; `helpers/enrichment_models.R`), replacing the
earlier Wilcoxon-on-proportions. No sample random intercept: histology is one
value per sample, so the beta-binomial overdispersion carries the sample-level
variance (adding `(1|sample_id)` would be unidentifiable). Test 3 is a separate,
within-sample niche↔cell-type association (Fisher + Fisher's method) and is left
as-is.

**This cohort has only 11 samples across 4 histology groups** (central
neurocytoma n=2, ganglioglioma n=2, cerebellar liponeurocytoma n=3,
pilocytic astrocytoma n=4). Read the **effect size (`estimate_log_odds`) and its
95% CI** as the headline; `q_value` is for **ranking only** and is
*anti-conservative* at these group sizes -- with 2-4 samples the CI leans on a
single shared dispersion pooled across all samples, so n=2 groups (CN, GG) can
show q~1e-27 with intervals tighter than 2 tumors truly justify. The **direction**
of a call is trustworthy and describes this cohort faithfully; the extreme q
magnitude does not generalize to the disease. Each call is worth a glance at
`n_samples` (how many samples drove it) and the per-sample composition bars above.

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(SpatialExperiment)
library(SummarizedExperiment)
library(metap)
library(pals)   # polychrome()/alphabet() categorical palettes for the composition bars

# Path resolution per AGENTS.md so this notebook works across mac/cluster mounts.
home <- Sys.getenv("HOME")
if (home %in% c("/Users/youyun", "/Users/youyunzheng")) {
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (home == "/PHShome/yz762") {
    workdir <- "/data/beroukhim1/"
} else if (home == "/home/yz762") {
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    workdir <- "/xchip/beroukhimlab/"
}
source(paste0(workdir, "youyun/plgg/code/helpers/test_enrichment.R"))
# lineage_display: snake_case cell-type key -> human-readable label, applied at
# plot time only (keeps the canonical marker_panel keys stable everywhere else).
source(paste0(workdir, "youyun/plgg/code/helpers/cell_type_display.R"))
# unified beta-binomial enrichment engine (build_feature_count_dt,
# fit_feature_histology_set) -- replaces the old Wilcoxon histology tests.
source(paste0(workdir, "youyun/plgg/code/helpers/enrichment_models.R"))
```

# Load data

```{r}
outputdir = paste0(workdir, "youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kct_50_resct_0.5,1_kni_50_resni_0.5,1")
rds = paste0(outputdir, "/banksy_clusters_connected_annotated_k_geom_15_30_pc_20_lam_0.2_0.8_kct_50_resct_0.2_kni_50_resni_0.5_20260826_115134.rds")
banksy_spe = readRDS(rds)

# cell_type: this object's own automated marker-AUC call (score_lineages_top_gene(),
# min_gene_auc=0.55, includes "undetermined") -- not the older manually curated
# banksy_proseg_cell_cluster_annotation.tsv table, which was never regenerated
# at this resolution and would duplicate work already validated this session.
# niche_type: the raw niche cluster ID -- niches aren't given marker-based labels
# anywhere in this pipeline, so the numbered cluster IS the "niche type".
cd = as.data.table(colData(banksy_spe))
cd[, cell_type := cell_type_clust_Harmony_BANKSY_lam0.2_k50_res0.2]
cd[, niche_type := as.character(clust_Harmony_BANKSY_lam0.8_k50_res0.5)]

sample_histology = unique(cd[, .(sample_id, histology)])
print(sample_histology[order(histology)])
```

# Raw composition per sample

Before the statistical tests: what does each sample actually look like? These
stacked bars show the raw proportion of every cell type (and every niche type)
per sample, each bar summing to 1, faceted by histology so the four groups are
visually separated. There's no modeling here -- it's the underlying composition
the enrichment tests below are run on, so read these first to sanity-check what
drives any "enriched"/"depleted" call.

```{r}
#| fig-height: 5.4
# Work on a copy so setting factor orders for plotting doesn't affect the
# enrichment sections below, which recompute proportions from `cd`.
comp_dt = copy(cd)
# sample_id is stored numeric in this object (e.g. 53398, 241670) -- coerce to a
# factor so the bar x-axis is categorical (one discrete slot per sample) rather
# than a continuous scale, which would place all 11 bars on a ~40k-330k axis and
# collapse each to an invisible sliver under width = 0.5.
comp_dt[, sample_id := factor(sample_id)]

# Stacking / legend order: most abundant cell type at the bottom, "undetermined"
# last so the "no confident call" bucket doesn't sit among the real lineages.
ct_levels = comp_dt[, .N, by = cell_type][order(-N)]$cell_type
ct_levels = c(setdiff(ct_levels, "undetermined"), "undetermined")
comp_dt[, cell_type := factor(cell_type, levels = ct_levels)]
# Niches ordered numerically so the legend reads 1..N, not string-sorted (1,10,11,...).
niche_levels = as.character(sort(as.integer(unique(comp_dt$niche_type))))
comp_dt[, niche_type := factor(niche_type, levels = niche_levels)]
# Histology facet order: small-n groups first, matching the enrichment panels.
hist_levels = c("central neurocytoma", "ganglioglioma",
                "cerebellar liponeurocytoma", "pilocytic astrocytoma")
comp_dt[, histology := factor(histology, levels = hist_levels)]

# Palettes keyed to those levels (pals::alphabet for cell types, polychrome for
# niches -- same families as banksy_niches_by_histology_proseg.qmd). "undetermined"
# is forced grey so it reads as "unlabeled" rather than a distinct lineage.
cell_type_color_manual = setNames(pals::alphabet()[seq_along(ct_levels)], ct_levels)
cell_type_color_manual["undetermined"] = "grey75"
niche_color_manual = setNames(pals::polychrome()[seq_along(niche_levels)], niche_levels)

# Shared bar builder. facet_grid(space="free_x") keeps bar width constant across
# panels (facet_wrap would stretch a 2-sample panel to a 4-sample panel's width);
# label_wrap_gen wraps long histology names so narrow panels don't truncate their
# titles; the c(0, 0.02) lower expansion lets bars sit flush on the x axis.
composition_bar = function(fill_col, palette, fill_labels, legend_title, title_text) {
    ggplot(comp_dt, aes(x = sample_id, fill = .data[[fill_col]])) +
        geom_bar(position = "fill", width = 0.5) +
        facet_grid(cols = vars(histology), scales = "free_x", space = "free_x",
                   labeller = label_wrap_gen(width = 14)) +
        scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
        scale_fill_manual(values = palette, labels = fill_labels, name = legend_title) +
        labs(title = title_text, x = NULL, y = "Proportion of cells") +
        theme_classic(base_size = 13) +
        theme(
            axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
            strip.background = element_rect(fill = "grey95", color = "grey80", linewidth = 0.3),
            strip.text = element_text(face = "bold", size = 9.5, margin = margin(4, 2, 4, 2)),
            strip.clip = "off",
            panel.spacing.x = unit(0.7, "lines"),
            plot.title = element_text(face = "bold", size = 15)
        )
}

composition_bar("cell_type", cell_type_color_manual, lineage_display, "Cell type",
                "Cell-type composition per sample")
```

```{r}
#| fig-height: 5.6
composition_bar("niche_type", niche_color_manual, ggplot2::waiver(), "Niche type",
                "Niche-type composition per sample")
```

# Cell-type enrichment by histology

Unified **beta-binomial** one-vs-rest test (`fit_feature_histology_set` in
`helpers/enrichment_models.R`): for each (cell type, histology) we fit
`cbind(k, n-k) ~ is_histology` on the per-sample counts (k = that cell type's
cells in a sample, n = total cells in the sample). **No random intercept** --
histology is one value per sample, so a sample RI is unidentifiable; the
beta-binomial overdispersion carries the sample-level variance. This replaces
the earlier Wilcoxon-on-proportions test, unifying with the pathology models.

**Read the effect size, not the q-value.** logOR + 95% CI is the headline; q is
for ranking only and is *anti-conservative* at small n -- with 2-4 samples per
histology, the CI leans on a single shared dispersion pooled across all samples,
so n=2 groups (CN, GG) can show q~1e-27 with CIs that are tighter than 2 tumors
can truly justify. The direction generalizes; the tight magnitude/certainty does
not. See the per-sample composition bars above for the raw evidence.

```{r}
# per-sample counts: one row per (sample, cell type), k / n from build_feature_count_dt
celltype_counts = build_feature_count_dt(
    cd, feature_col = "cell_type",
    observation_col = "sample_id", region_col = "histology", extra_cols = "histology"
)
celltype_histology_enrichment = fit_feature_histology_set(celltype_counts)
setnames(celltype_histology_enrichment, "feature", "cell_type")
fwrite(celltype_histology_enrichment, paste0(workdir, "youyun/plgg/code/niche/proseg_output_analysis/celltype_histology_enrichment.tsv"), sep = "\t")
print(celltype_histology_enrichment[order(q_value)][, .(cell_type, histology, n_samples,
    estimate_log_odds = round(estimate_log_odds, 2), q_value = signif(q_value, 2), sign_call)])
```

# Niche-type enrichment by histology

Same unified beta-binomial one-vs-rest test, on niche-cluster counts instead of
cell-type counts. Rare / near-empty niches (e.g. 15, 17, 19) hit separation at
these small group sizes and come back as degenerate models (huge |logOR|, flagged
`model_status != "pass"`) -- the forest below caps the axis and draws them as
arrows so they're not mistaken for null.

```{r}
niche_counts = build_feature_count_dt(
    cd, feature_col = "niche_type",
    observation_col = "sample_id", region_col = "histology", extra_cols = "histology"
)
niche_histology_enrichment = fit_feature_histology_set(niche_counts)
setnames(niche_histology_enrichment, "feature", "niche_type")
fwrite(niche_histology_enrichment, paste0(workdir, "youyun/plgg/code/niche/proseg_output_analysis/niche_histology_enrichment.tsv"), sep = "\t")
print(niche_histology_enrichment[order(q_value)][, .(niche_type, histology, n_samples,
    estimate_log_odds = round(estimate_log_odds, 2), q_value = signif(q_value, 2), sign_call, model_status)])
```

# Niche composition (cell-type makeup within each niche) by histology

Does which cell types make up a given niche change by histology? Per-sample
Fisher's exact test between `niche_type` and `cell_type`
(`pairwise_categorical_enrichment_test`), then p-values from samples of the
*same* histology combined via `metap::sumlog` (Fisher's method) when >=2
valid p-values exist, BY-adjusted (matches
`banksy_niches_by_histology_proseg.qmd`'s existing block exactly, re-pointed
at the new object). Note: `metap::sumlog` drops p==0 values, which can bias
the combined p slightly upward for very strong per-sample associations --
same caveat the source notebook already carries.

```{r}
self_cell_type_niche_enrichment_per_sample = do.call('rbind', mapply(
    function(x, y) {
        pairwise_categorical_enrichment_test(
            cd[sample_id == x]$niche_type, cd[sample_id == x]$cell_type
        )[, c('sample_id', 'histology') := list(x, y)]
    },
    sample_histology$sample_id, sample_histology$histology, SIMPLIFY = FALSE
))

niche_composition_histology_enrichment = self_cell_type_niche_enrichment_per_sample[
    , .(
        N_var1 = mean(N_var1), N_var2 = mean(N_var2),
        N_var1var2 = mean(N_var1var2 / N_var1), odds_ratios = mean(odds_ratios),
        p_value = ifelse(sum((p_value > 0) & (p_value <= 1)) >= 2, metap::sumlog(p_value)$p, p_value)
    ), .(var1, var2, histology)
][, q_value := p.adjust(p_value, method = 'BY')]
setnames(niche_composition_histology_enrichment, c('var1', 'var2'), c('niche_type', 'cell_type'))
fwrite(niche_composition_histology_enrichment, paste0(workdir, "youyun/plgg/code/niche/proseg_output_analysis/niche_composition_histology_enrichment.tsv"), sep = "\t")
print(niche_composition_histology_enrichment[order(q_value)][1:15])
```

# Visualization

Forest plot: point = beta-binomial one-vs-rest **log-odds** (histology vs rest),
whisker = 95% CI (`estimate ± 1.96·SE`), color = direction only when
`q_value < 0.25` (red = enriched, blue = depleted, grey = not significant / not
`pass`). The x-axis is capped at ±6 with sideways arrows for models that overflow
(separation / near-empty niches), so a few degenerate fits don't blow out the
scale. Facet labels carry the per-histology sample count `n`.

**Effect size is the headline, q is for ranking only.** As discussed above, the
q-values are anti-conservative at small n -- the direction of each colored call
is trustworthy and the logOR/CI describe the cohort faithfully, but the extreme
q magnitudes (n=2 groups can reach ~1e-27) lean on a shared-dispersion assumption
and should not be read as population-level certainty. Ranking calls by q within
the cohort is fine; treating q=1e-27 as "27× more certain" than q=1e-5 is not.

```{r}
#| fig-height: 5.2
# Forest plot for the unified beta-binomial histology results. Effect size
# (estimate_log_odds) is the headline; q_cutoff colors direction only. cap
# bounds the x-axis so separation/near-empty models (huge |logOR|) don't blow
# out the scale -- overflow points are drawn as sideways arrows. q is NOT
# printed on the plot (anti-conservative at small n; see section intro).
plot_enrichment_forest = function(dt, feature_col, title_text, q_cutoff = 0.25,
                                  feature_labels = ggplot2::waiver(), cap = 6) {
    dt = copy(dt)
    dt = dt[!is.na(estimate_log_odds)]
    # direction requires a converged (pass) model AND q under the cutoff
    dt[, direction := fifelse(model_status == "pass" & q_value < q_cutoff & estimate_log_odds > 0, "Enriched",
                        fifelse(model_status == "pass" & q_value < q_cutoff & estimate_log_odds < 0, "Depleted", "Not significant"))]
    dt[, direction := factor(direction, levels = c("Enriched", "Not significant", "Depleted"))]
    dt[, hist_label := paste0(histology, " (n=", n_samples, ")")]
    # cap the estimate + CI for display; overflow points become arrows
    dt[, est_cap := pmax(pmin(estimate_log_odds, cap), -cap)]
    dt[, lo_cap := pmax(pmin(ci_low, cap), -cap)]
    dt[, hi_cap := pmax(pmin(ci_high, cap), -cap)]
    dt[, shape_code := fifelse(abs(estimate_log_odds) <= cap, "point",
                         fifelse(estimate_log_odds > cap, "hi", "lo"))]
    # order features by largest capped |effect| so the strongest sit at the top
    feature_order = dt[, .(m = max(pmin(abs(estimate_log_odds), cap), na.rm = TRUE)), by = feature_col][order(m)][[feature_col]]
    dt[[feature_col]] = factor(dt[[feature_col]], levels = feature_order)

    dir_colors = c(Enriched = "#B2182B", `Not significant` = "grey65", Depleted = "#2166AC")
    ggplot(dt, aes(x = est_cap, y = .data[[feature_col]], color = direction)) +
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.4) +
        geom_errorbar(aes(xmin = lo_cap, xmax = hi_cap, linewidth = direction != "Not significant"),
                      orientation = "y", width = 0) +
        geom_point(aes(shape = shape_code, size = direction != "Not significant")) +
        facet_wrap(~ hist_label, nrow = 1) +
        coord_cartesian(xlim = c(-cap, cap)) +
        scale_shape_manual(values = c(point = 16, hi = 62, lo = 60), guide = "none") +  # 62='>' 60='<'
        scale_color_manual(values = dir_colors, name = NULL,
                            guide = guide_legend(override.aes = list(linewidth = 1.2, size = 3))) +
        scale_linewidth_manual(values = c(`TRUE` = 0.9, `FALSE` = 0.5), guide = "none") +
        scale_size_manual(values = c(`TRUE` = 2.6, `FALSE` = 1.8), guide = "none") +
        # cell-type panel: relabel keys -> display names via feature_labels;
        # niche panel passes waiver() so numeric niche IDs stay as-is
        scale_y_discrete(labels = feature_labels) +
        labs(
            title = title_text,
            subtitle = paste0("Beta-binomial one-vs-rest log-odds (95% CI; x capped at ±", cap,
                              ", arrows = overflow). Colored at q < ", q_cutoff, " (ranking only)."),
            x = "Log-odds (histology − rest)", y = NULL
        ) +
        theme_classic(base_size = 13) +
        theme(
            panel.grid.major.x = element_line(color = "grey92", linewidth = 0.3),
            strip.background = element_rect(fill = "grey95", color = "grey80", linewidth = 0.3),
            strip.text = element_text(face = "bold", size = 10.5),
            axis.line = element_line(color = "black", linewidth = 0.35),
            axis.ticks = element_line(color = "black", linewidth = 0.35),
            axis.text.y = element_text(size = 10),
            axis.text.x = element_text(angle = 30, hjust = 1),
            panel.spacing.x = unit(0.9, "lines"),
            axis.title.x = element_text(size = 11, margin = margin(t = 8)),
            plot.title = element_text(face = "bold", size = 15, margin = margin(b = 2)),
            plot.subtitle = element_text(color = "grey30", size = 10.5, margin = margin(b = 10)),
            legend.position = "bottom",
            legend.text = element_text(size = 10),
            legend.margin = margin(t = -4),
            plot.margin = margin(10, 14, 10, 10)
        )
}

plot_enrichment_forest(celltype_histology_enrichment, "cell_type", "Cell-type enrichment by tumor histology", feature_labels = lineage_display)
```

```{r}
#| fig-height: 6.5
plot_enrichment_forest(niche_histology_enrichment, "niche_type", "Niche-type enrichment by tumor histology")
```
