---
title: "Niche interpretation: neighborhood composition + pathology + histology"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 13
        fig-height: 7
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk:
        results: hold
        R.options:
            width: 120
---

# Introduction

The BANKSY niches (lam0.8, res_ni=0.5) are numbered 1-19 with no biological
identity. This notebook builds a per-niche "dictionary" on the **final object**
(res_ct=0.2/res_ni=0.5) by characterizing each niche along orthogonal lenses and
synthesizing a candidate identity, keeping the niches numbered.

Lenses:

1. **Neighborhood cell-type composition** (the backbone). A niche is a *context*
   cluster, so we describe it by the cell types in each cell's local
   neighborhood (self + 15 nearest spatial neighbors) pooled into one proportion
   vector per niche -- essentially what BANKSY clustered on. This keeps the
   spatial-context signal while giving a single interpretable vector per niche.
2. **Pathology correspondence** -- beta-binomial niche x pathology-region model
   (`annotations/linear_models_niches.qmd`, re-run on this object), the external
   ground-truth anchor.
3. **Histology association** -- from `niche_histology_enrichment.tsv`
   (sample-level, n=11, underpowered; secondary).

The synthesis is a `niche_dictionary.tsv` with a free-text `candidate_identity`.

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(SpatialExperiment)
library(SummarizedExperiment)
library(dbscan)
library(pals)

# Path resolution per AGENTS.md (works across mac/cluster mounts).
home <- Sys.getenv("HOME")
if (home %in% c("/Users/youyun", "/Users/youyunzheng")) {
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (home == "/PHShome/yz762") {
    workdir <- "/data/beroukhim1/"
} else if (home == "/home/yz762") {
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    workdir <- "/xchip/beroukhimlab/"
}
source(paste0(workdir, "youyun/plgg/code/helpers/test_enrichment.R"))
# lineage_display: snake_case cell-type key -> human-readable label (display only)
source(paste0(workdir, "youyun/plgg/code/helpers/cell_type_display.R"))
```

# Load data

We load the **pathology-annotated** copy of the final object (produced by the
re-pointed `add_annotations.R`), so cell type, niche, pathology, histology and
spatial coordinates all travel together.

```{r}
obj_rds = paste0(
    workdir,
    "youyun/plgg/data/Xenium_annotations/",
    "banksy_clusters_connected_annotated_k_geom_15_30_pc_20_lam_0.2_0.8_kct_50_resct_0.2_kni_50_resni_0.5_20260826_115134_pathology.rds"
)
banksy_spe = readRDS(obj_rds)

cd = as.data.table(colData(banksy_spe))
# cell type: this object's marker-AUC lineage call (same column used elsewhere)
cd[, cell_type := as.character(cell_type_clust_Harmony_BANKSY_lam0.2_k50_res0.2)]
# niche type: the raw niche cluster id (numeric, kept as the "niche type")
cd[, niche_type := as.character(clust_Harmony_BANKSY_lam0.8_k50_res0.5)]

# niche display order = numeric
niche_levels = as.character(sort(as.integer(unique(cd$niche_type))))
cat("n cells:", nrow(cd), " | niches:", paste(niche_levels, collapse = ","), "\n")
```

# Neighborhood cell-type composition (backbone)

For every cell we take its own type plus the cell types of its 15 nearest
spatial neighbors (k_geom=15, the BANKSY geometry), and pool these over all
cells in a niche to get one cell-type proportion vector per niche. Samples are
spatially staggered in the object, so a global kNN does not cross samples.

```{r}
k_geom = 15
locs = spatialCoords(banksy_spe)
knn = dbscan::kNN(x = locs, k = k_geom)

# cell types indexed the same as the object columns
ct_vec = cd$cell_type
niche_vec = cd$niche_type

# long table: one row per (cell, neighbor) edge carrying the NEIGHBOR's cell type,
# plus one self row per cell -- so each cell contributes k_geom+1 rows to its niche
nb_dt = data.table(
    from = rep(seq_len(nrow(knn$id)), times = k_geom),
    member_ct = ct_vec[as.vector(knn$id)]
)
self_dt = data.table(from = seq_len(length(ct_vec)), member_ct = ct_vec)
neighborhood = rbind(nb_dt, self_dt)
neighborhood[, niche_type := niche_vec[from]]

# per-niche neighborhood composition (proportions sum to 1 within niche)
niche_comp = neighborhood[, .(N = .N), by = .(niche_type, member_ct)]
niche_comp[, prop := N / sum(N), by = niche_type]
niche_comp[, niche_type := factor(niche_type, levels = niche_levels)]

# order cell types by overall neighborhood abundance, undetermined last
ct_order = niche_comp[, .(m = sum(N)), by = member_ct][order(-m)]$member_ct
ct_order = c(setdiff(ct_order, "undetermined"), "undetermined")
niche_comp[, member_ct := factor(member_ct, levels = ct_order)]

# persist the full niche x cell-type composition so downstream figures (e.g. the
# focused presentation panel below, or a log2-OR variant) can be built from this
# TSV without reloading the 5GB object.
fwrite(niche_comp[, .(niche_type, member_ct, N, prop)], paste0(workdir,
    "youyun/plgg/code/niche/proseg_output_analysis/niche_celltype_composition.tsv"), sep = "\t")
```

## Niche x cell-type neighborhood-composition heatmap

```{r}
#| fig-height: 6.5
ggplot(niche_comp, aes(x = member_ct, y = niche_type, fill = prop)) +
    geom_tile(color = "grey92", linewidth = 0.3) +
    geom_text(aes(label = ifelse(prop >= 0.08, sprintf("%.0f", prop * 100), "")),
              size = 3, color = "grey15") +
    scale_x_discrete(labels = lineage_display) +
    scale_fill_gradient(low = "white", high = "#2166AC", name = "% of\nneighborhood",
                        labels = scales::percent) +
    labs(title = "Niche neighborhood composition (self + 15 neighbors, pooled)",
         subtitle = "Each row sums to 100%; numbers are % of the niche's pooled neighborhood cells",
         x = "Cell type (neighborhood member)", y = "Niche") +
    theme_classic(base_size = 13) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          plot.title = element_text(face = "bold", size = 15))
```

## Histology-signature niches (presentation figure)

Focused 4x4 view for the talk: each histology's characteristic niche and the
cell type that anchors it, on the diagonal. Fill = % of that niche's pooled
neighborhood. Niche 1 is enriched in both CN and CLN; shown here as the
neuronal-rich CL niche.

```{r}
#| fig-height: 4.5
#| fig-width: 9
sig_niches = c("2", "5", "6", "1")                                  # GG, PA, CN, CL
niche_hist = c("2" = "Niche 2 · Ganglioglioma", "5" = "Niche 5 · Pilocytic astrocytoma",
               "6" = "Niche 6 · Central neurocytoma", "1" = "Niche 1 · Cerebellar liponeurocytoma")
sig_ct = c("astro_glial", "opc", "radial_glia", "neuronal")          # cols in diagonal order

sig_dt = niche_comp[as.character(niche_type) %in% sig_niches &
                    as.character(member_ct) %in% sig_ct]
sig_dt[, niche_lab := factor(niche_hist[as.character(niche_type)],
       levels = rev(niche_hist[sig_niches]))]                        # top row = niche 2
sig_dt[, ct_lab := factor(as.character(member_ct), levels = sig_ct)]

sig_fig = ggplot(sig_dt, aes(x = ct_lab, y = niche_lab, fill = prop)) +
    geom_tile(color = "white", linewidth = 1) +
    geom_text(aes(label = sprintf("%.0f%%", prop * 100),
                  color = prop > 0.35), size = 5, fontface = "bold") +
    scale_x_discrete(labels = lineage_display, position = "top") +
    scale_fill_gradient(low = "white", high = "#2166AC", name = "% of niche\nneighborhood",
                        labels = scales::percent) +
    scale_color_manual(values = c(`TRUE` = "white", `FALSE` = "grey20"), guide = "none") +
    labs(title = "Histology-specific niches are ecosystems defined by characteristic cell types",
         subtitle = "Tile = % of the niche's pooled (self + 15-neighbor) neighborhood",
         x = NULL, y = NULL) +
    theme_minimal(base_size = 15) +
    theme(panel.grid = element_blank(),
          axis.text.x = element_text(face = "bold"),
          plot.title = element_text(face = "bold", size = 15),
          plot.subtitle = element_text(color = "grey30", size = 11))
sig_fig
# slide-ready standalone PNG
ggsave(paste0(workdir, "youyun/plgg/code/niche/proseg_output_analysis/niche_histology_signature.png"),
       sig_fig, width = 9, height = 4.5, dpi = 200)
```

## Per-niche composition as stacked bars

Same neighborhood proportions as the heatmap, as one stacked bar per niche --
easier to read each niche's dominant type(s) at a glance.

```{r}
#| fig-height: 5.5
ct_levels_bar = levels(niche_comp$member_ct)   # abundance-ordered, undetermined last
bar_cols = setNames(pals::alphabet()[seq_along(ct_levels_bar)], ct_levels_bar)
bar_cols["undetermined"] = "grey75"
ggplot(niche_comp, aes(x = niche_type, y = prop, fill = member_ct)) +
    geom_col(width = 0.7) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    scale_fill_manual(values = bar_cols, labels = lineage_display, name = "Cell type") +
    labs(title = "Per-niche neighborhood composition (stacked)",
         x = "Niche", y = "Proportion of neighborhood") +
    theme_classic(base_size = 13) +
    theme(plot.title = element_text(face = "bold", size = 15))
```

# Neighborhood cell-type enrichment vs cohort

Which cell types are over-/under-represented in each niche's neighborhood
relative to the whole-cohort baseline -- the statistical complement to the raw
proportions above. We form a one-vs-rest 2x2 from the pooled neighborhood counts
per (niche, cell type) and take the log2 odds ratio: > 0 = enriched (red),
< 0 = depleted (blue).

**Caveat:** neighborhoods overlap (each cell is counted in ~16 of them), so a
formal p-value here is pseudo-replicated and would be spuriously tiny. We
therefore read the **effect size** (log2 OR), not significance.

```{r}
#| fig-height: 6.5
grand = sum(niche_comp$N)
ct_tot = niche_comp[, .(N_ct = sum(N)), by = member_ct]
ni_tot = niche_comp[, .(N_ni = sum(N)), by = niche_type]
enr = merge(merge(niche_comp[, .(niche_type, member_ct, N)], ct_tot, by = "member_ct"),
            ni_tot, by = "niche_type")
# one-vs-rest 2x2 per (niche, cell type), Haldane 0.5 correction for empty cells:
# a=in-niche&type, b=in-niche&other, c=out-niche&type, d=out-niche&other
enr[, a := N][, b := N_ni - a][, c := N_ct - a][, d := grand - N_ni - c]
enr[, log2_or := log2(((a + 0.5) * (d + 0.5)) / ((b + 0.5) * (c + 0.5)))]
enr[, member_ct := factor(member_ct, levels = levels(niche_comp$member_ct))]
enr[, niche_type := factor(niche_type, levels = niche_levels)]
lim = max(abs(enr$log2_or))
ggplot(enr, aes(x = member_ct, y = niche_type, fill = log2_or)) +
    geom_tile(color = "grey92", linewidth = 0.3) +
    geom_text(aes(label = sprintf("%+.1f", log2_or)), size = 2.6, color = "grey15") +
    scale_x_discrete(labels = lineage_display) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
                         limits = c(-lim, lim), name = "log2 OR\nvs cohort") +
    labs(title = "Neighborhood cell-type enrichment/depletion by niche",
         subtitle = "One-vs-rest log2 odds ratio (effect size; red = enriched, blue = depleted)",
         x = "Cell type", y = "Niche") +
    theme_classic(base_size = 13) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          plot.title = element_text(face = "bold", size = 15))
```

# Sample occupancy: is a niche private to one tumor?

A niche that is **private to a single sample** is a candidate **tumor-specific
(neoplastic) niche** -- shared/normal niches recur across samples and
histologies, so privacy is itself a signal, not a nuisance. It is also exactly
why such niches come back NA/untestable in the histology enrichment (no other
sample has them, so there is no cross-sample contrast). We surface this
directly here instead of dropping it. (Confirming a private niche is truly
neoplastic vs. a sample-specific normal structure would need orthogonal
evidence, e.g. inferCNV -- so we flag it as a *candidate*.)

```{r}
#| fig-height: 6
# per-niche sample occupancy: fraction of each niche's cells coming from each sample
occ = cd[, .N, by = .(niche_type, sample_id)]
occ[, frac := N / sum(N), by = niche_type]
occ[, niche_type := factor(niche_type, levels = niche_levels)]

# per-niche summary: total cells, # samples holding >=5% of the niche, and the
# single largest sample's share. A niche is "sample-private" if >=80% of its
# cells come from one sample.
niche_sample_summary = occ[, .(
    n_cells = sum(N),
    n_samples_5pct = sum(frac >= 0.05),
    top_sample_frac = max(frac)
), by = niche_type][order(niche_type)]
niche_sample_summary[, sample_private := top_sample_frac >= 0.8]
print(niche_sample_summary)

ggplot(occ, aes(x = factor(sample_id), y = niche_type, fill = frac)) +
    geom_tile(color = "grey92", linewidth = 0.3) +
    geom_text(aes(label = ifelse(frac >= 0.1, sprintf("%.0f", frac * 100), "")),
              size = 2.6, color = "grey15") +
    scale_fill_gradient(low = "white", high = "#762A83", name = "% of niche's\ncells",
                        labels = scales::percent) +
    labs(title = "Niche sample occupancy (fraction of each niche's cells per sample)",
         subtitle = "A niche concentrated in one column is sample-private -> candidate tumor niche",
         x = "Sample", y = "Niche") +
    theme_classic(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(face = "bold", size = 15))
```

# Assemble the niche dictionary

Per niche we take the **most enriched** neighborhood cell type(s) (highest log2
OR vs the cohort, from the enrichment above -- not merely the most *abundant*,
which is usually Neuronal everywhere), the top enriched pathology region (from
the beta-binomial model), the histology it is enriched in, and whether it is
sample-private (candidate tumor niche).

```{r}
# --- most ENRICHED neighborhood cell type(s): top 2 by log2 OR vs cohort (from
# `enr` above), NOT the most abundant. Surfaces what makes a niche distinctive
# (e.g. Endothelial in a vascular niche) instead of whatever is abundant
# everywhere. Require prop (a / N_ni) >= 2% so a rare cell type with a large
# Haldane-corrected OR can't hijack the label. Show both log2 OR and %.
top_ct = enr[log2_or > 0 & (a / N_ni) >= 0.02][order(niche_type, -log2_or),
    .(top_enriched_celltypes = paste0(
        lineage_display[as.character(member_ct[1:min(2, .N)])], " (",
        sprintf("%+.1f log2OR, %.0f%%", log2_or[1:min(2, .N)],
                (a[1:min(2, .N)] / N_ni[1:min(2, .N)]) * 100), ")",
        collapse = "; ")
    ), by = niche_type]

# --- pathology: top enriched region per niche (sign_call == enriched, min q) ---
path_tsv = paste0(workdir,
    "youyun/plgg/code/annotations/annotations/model_outputs/mixed_models/linear_models_niches_primary_results.tsv")
path_res = fread(path_tsv)
# real anatomic regions only: "unannotated" = cells outside any drawn polygon
# (absence of annotation, not a region), and "Other" = the model's collapsed
# minority-niche bucket -- neither belongs as a niche's "top pathology".
# sign_call == "enriched" already encodes q < 0.25 (the model's discovery cutoff),
# so no extra q filter -- this surfaces e.g. Lipidized-cells enrichments too.
path_top = path_res[
    model_status == "pass" & sign_call == "enriched" &
    annotation_region != "unannotated" & as.character(feature) %in% niche_levels
][order(feature, q_value),
    .(top_pathology = paste0(annotation_region[1], " (logOR=", round(estimate_log_odds[1], 1), ")")),
    by = .(niche_type = as.character(feature))]

# --- histology: top enriched histology per niche (beta-binomial logOR > 0, min q) ---
# uses estimate_log_odds now that the histology test is beta-binomial (not Wilcoxon
# delta); require model_status == "pass" so degenerate niches (15/17/19) don't get
# a spurious top-histology from a separation artifact.
histo_tsv = paste0(workdir,
    "youyun/plgg/code/niche/proseg_output_analysis/niche_histology_enrichment.tsv")
histo_res = fread(histo_tsv)
# abs(logOR) < 6 guard: model_status=="pass" does NOT catch quasi-separation
# (numerically converged but degenerate), which produces huge logORs (13, 22)
# for sample-private niches -- those are artifacts of the niche living in one
# sample, not real histology enrichment, so exclude them. Real effects top out
# ~5 (radial_glia in CN), so 6 cleanly separates signal from separation.
histo_top = histo_res[
    model_status == "pass" & !is.na(estimate_log_odds) &
    estimate_log_odds > 0 & estimate_log_odds < 6 & !is.na(q_value)
][order(niche_type, q_value),
    .(top_histology = paste0(histology[1], " (logOR=", round(estimate_log_odds[1], 1), ")")),
    by = .(niche_type = as.character(niche_type))]

# --- merge into the dictionary ---
niche_dict = data.table(niche_type = niche_levels)
niche_dict = merge(niche_dict, top_ct[, .(niche_type = as.character(niche_type), top_enriched_celltypes)],
                   by = "niche_type", all.x = TRUE)
niche_dict = merge(niche_dict, path_top, by = "niche_type", all.x = TRUE)
niche_dict = merge(niche_dict, histo_top, by = "niche_type", all.x = TRUE)
# sample occupancy: n_cells, breadth, and the sample-private (candidate tumor) flag
niche_dict = merge(niche_dict, niche_sample_summary[, .(
    niche_type = as.character(niche_type), n_cells, n_samples_5pct,
    top_sample_frac = round(top_sample_frac, 2), sample_private)],
    by = "niche_type", all.x = TRUE)
niche_dict[, niche_type := factor(niche_type, levels = niche_levels)]
setorder(niche_dict, niche_type)

# --- candidate identity: a compact auto-descriptor from the lenses ---
# (heuristic starting point -- refine by eye against the heatmap/spatial maps)
niche_dict[, candidate_identity := paste0(
    top_enriched_celltypes,
    ifelse(!is.na(top_pathology), paste0(" | region: ", top_pathology), ""),
    ifelse(!is.na(top_histology), paste0(" | hist: ", top_histology), ""),
    ifelse(sample_private, " | SAMPLE-PRIVATE (candidate tumor niche)", "")
)]

fwrite(niche_dict, paste0(workdir,
    "youyun/plgg/code/niche/proseg_output_analysis/niche_dictionary.tsv"), sep = "\t")
print(niche_dict[, .(niche_type, top_enriched_celltypes, top_pathology, top_histology,
                     n_cells, n_samples_5pct, sample_private)])
```

# Niche x pathology region

The niche's top enriched pathology region is summarized in the `top_pathology`
column of the dictionary above (from the beta-binomial niche x region models).
The full forest plot with 95% CIs lives in its natural home,
`annotations/linear_models_niches.qmd` -- not duplicated here.

# Spatial reference (downsampled)

Where the niches sit in tissue (5% downsample for legibility). Samples are
staggered, so this is the whole cohort in one panel set.

```{r}
#| fig-height: 9
#| fig-width: 14
set.seed(5555)
sub_idx = sample(seq_len(nrow(cd)), min(80000, nrow(cd)))
spat = data.table(
    sdimx = spatialCoords(banksy_spe)[sub_idx, 1],
    sdimy = spatialCoords(banksy_spe)[sub_idx, 2],
    niche_type = factor(cd$niche_type[sub_idx], levels = niche_levels),
    sample_id = cd$sample_id[sub_idx],
    histology = cd$histology[sub_idx]
)
niche_cols = setNames(pals::polychrome()[seq_along(niche_levels)], niche_levels)
ggplot(spat, aes(x = sdimx, y = sdimy, color = niche_type)) +
    geom_point(size = 0.15, alpha = 0.6) +
    facet_wrap(~ histology + sample_id, scales = "free", ncol = 4) +
    scale_color_manual(values = niche_cols, name = "Niche") +
    guides(color = guide_legend(override.aes = list(size = 3, alpha = 1), ncol = 1)) +
    labs(title = "BANKSY niches in space (5% downsample)") +
    theme_void(base_size = 12) +
    theme(strip.text = element_text(size = 8), legend.position = "right",
          plot.title = element_text(face = "bold", size = 15))
```
