---
title: "Evaluate Leiden resolution sweep"
format:
  html:
    embed-resources: true   # inline figures/libs into one standalone .html (easy to copy off the cluster)
    fig-dpi: 96
editor_options:
  chunk_output_type: console
---

# Purpose

Pick data-backed defaults for cell-type (lam0.2) and niche (lam0.8) Leiden
clustering from the (k_leiden x resolution) sweep produced by
`leiden_resolution_sweep.R`.

Evaluation runs in three layers, cheapest first:

1. **Layer 1** — n_clusters heatmaps from the sweep TSV. Filter combos by
   plausible cluster counts and look for plateaus (stable structure across
   adjacent resolutions).
2. **Layer 2** — biology for the finalists: re-run Leiden on the finalist
   (k, res) pairs only (same seed as the sweep, so labels are reproducible),
   then check cluster size distribution, sample mixing, marker expression
   (lam0.2), and spatial coherence (lam0.8).
3. **Decision** — pick the smallest cluster count per lambda that still
   resolves the biology you care about (parsimony wins ties).

Inputs : `leiden_resolution_sweep_counts.tsv` (from the sweep) and the same
         90k-cell BANKSY subset RDS the sweep used.
Outputs: a single decision table at the bottom; no objects saved to disk.

# Setup

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
    library(SummarizedExperiment)
    library(SpatialExperiment)
    library(data.table)
    library(ggplot2)
    library(Banksy)
    library(cowplot)   # composite plot_grid layouts (repo convention over patchwork)
    library(pals)      # polychrome() / alphabet() categorical palettes
    library(presto)    # wilcoxauc(): fast single-cell Wilcoxon-AUC, used in 2c/2c.5
})

# Path resolution mirrors the sweep script and AGENTS.md.
home = Sys.getenv("HOME")
if (home %in% c("/Users/youyun", "/Users/youyunzheng")) {
    workdir = "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
    subset_path = "/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds"
} else if (home == "/PHShome/yz762") {
    workdir = "/data/beroukhim1/"
    subset_path = paste0(workdir, "youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds")
} else if (home == "/home/yz762") {
    workdir = "/mnt/storage/dept/medonc/beroukhim/"
    subset_path = paste0(workdir, "youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds")
} else {
    workdir = "/xchip/beroukhimlab/"
    subset_path = paste0(workdir, "youyun/plgg/data/banksy_param_search/k1_15_k2_30_lambda1_0.2_lambda2_0.8_npcs_20_kc1_30_kc2_50_res1_0.75_res2_1/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215.rds")
}

# Sweep TSV lives next to this qmd (Quarto sets cwd to the qmd's dir).
sweep_tsv = "leiden_resolution_sweep_counts.tsv"

# Same seed as the sweep so Leiden labels are reproducible when we re-run
# the finalist combos in Layer 2.
seed_val = 55555

# PLGG marker panel for Layer 2 cell-type identification, restricted to genes
# that are actually on this cohort's custom 266-gene Xenium panel (verified
# against rownames(o), 2026-08-03). The prior version of this list leaned on
# canonical markers that are NOT on the panel and were silently dropped by the
# intersect() in the 2c chunk -- GFAP, S100B (astro), the entire pan-neuronal
# set (RBFOX3, SYT1, STMN2, SNAP25, MAP2, SYP), MBP, PLP1 (mature oligo), CD3D,
# CD8A (T cell), CLDN5 (endothelial), FOXJ1 (ependymal) -- so the old heatmap
# was blank for neurons, mature oligo, T cells, and ependymal.
#
# PLGG is not only glial tumors: ganglioglioma / DNET carry neuronal components
# and PA cells resemble radial glia, so we keep a radial-glia (progenitor) group
# plus a single merged `neuronal` group to read the glioneuronal axis. We do NOT
# split excitatory / inhibitory / immature neuron subtypes: the panel has no
# pan-neuronal anchor (RBFOX3/SYT1/SNAP25/MAP2 absent), the neuronal signal is
# concentrated in patient-private glioneuronal-tumor samples, and the subtype
# argmax was near-zero-margin / resolution-unstable. radial_glia markers are
# pan-neural-progenitor (not RGC-specific) and light up across glial tumors.
# Myeloid cells are one `myeloid` group (microglia + macrophage markers); we do
# not resolve microglia vs bone-marrow-derived macrophages. PTPRC (CD45) is
# dropped: it is pan-immune and would leak a myeloid signal onto T cells.
# Groups with zero panel coverage are omitted entirely: pan-neuronal structural
# genes, B/plasma, mural/pericyte, and ependymal (FOXJ1 absent).
marker_panel = list(
    astro_glial   = c("AQP4", "APOE", "GJA1"),
    radial_glia   = c("NES", "PAX6", "SOX2", "HES1", "SOX9"),
    opc           = c("OLIG1", "OLIG2", "PDGFRA", "CSPG4", "SOX10", "BCAN", "PTPRZ1"),
    oligo_mature  = c("MOG", "MOBP", "CLDN11", "MAG", "ST18"),
    neuronal      = c("SLC17A7", "SLC17A6", "RORB", "GAD1", "GAD2", "SOX11"),
    myeloid       = c("P2RY12", "CX3CR1", "AIF1", "CD68", "ITGAM", "CD163"),
    t_cell        = c("CD4", "CD2", "TRAC", "IL7R"),
    endothelial   = c("PECAM1", "FLT1"),
    proliferation = c("MKI67", "TOP2A", "PCNA", "CENPF")
)
```

# Layer 1 — n_clusters filter

Read the sweep output and look at how cluster count varies across the grid.
The bounds below are loose first-pass filters: cell types need at least the
major lineages separated (>=8) without fragmenting (<=25); niches need to
keep the major spatial environments distinct (>=5) without becoming a
patchwork (<=12).

```{r layer1-load, message=FALSE}
sweep = fread(sweep_tsv)
# Defensive: enforce ordering for heatmap axes.
sweep[, k   := factor(k,   levels = sort(unique(k)))]
sweep[, res := factor(res, levels = sort(unique(as.numeric(as.character(res)))))]
sweep[]
```

```{r layer1-heatmap, fig.width=9, fig.height=4}
# Heatmap of n_clusters across the (k, res) grid, faceted by lambda.
# Annotate cells that fall inside the plausibility band so the eye lands
# on the candidate region immediately.
plausible_band = list(`0.2` = c(8, 25), `0.8` = c(5, 12))
sweep[, plausible := {
    band = plausible_band[[as.character(lambda)]]
    n_clusters >= band[1] & n_clusters <= band[2]
}, by = lambda]

ggplot(sweep, aes(x = res, y = k, fill = n_clusters)) +
    geom_tile() +
    geom_text(aes(label = n_clusters,
                  colour = ifelse(plausible, "in band", "out")),
              size = 3) +
    scale_colour_manual(values = c("in band" = "black", "out" = "grey60"),
                        name = NULL) +
    scale_fill_viridis_c() +
    facet_wrap(~ paste0("lambda = ", lambda)) +
    labs(x = "Leiden resolution", y = "Leiden k_neighbors",
         title = "Cluster counts across sweep grid",
         caption = "Black labels = inside plausibility band; grey = outside") +
    theme_bw()
```

```{r layer1-plateaus}
# A plateau = two or more adjacent (k or res) cells with the same n_clusters.
# Plateaus indicate the structure is stable to small parameter perturbations
# at that point, which is the signal we want for a defensible default.
plateau_rows = sweep[plausible == TRUE][
    , .(combos = .N, ks = paste(unique(k), collapse = ",")),
    by = .(lambda, n_clusters)
][combos >= 2]
plateau_rows[order(lambda, n_clusters)]
```

# Layer 1.5 — pick finalists

After eyeballing the heatmap and plateau table above, fill in 2–4 candidates
per lambda. Each row is `(k, res)` — the sweep's seed makes the labels
reproducible when we re-cluster.

```{r layer1-pick-finalists}
# Picks span the in-band cluster counts at the strongest plateaus found in
# the sweep (counts.tsv from leiden_resolution_sweep.R, 2026-05-11).
# lam0.2 cell-type axis is very stable across (k, res); lam0.8 niche axis
# is more sensitive, so we anchor each finalist on a plateau cell.
#
# lam0.2:
#   12 clusters @ (k=80, res=0.2) — early-stable, coarsest defensible
#   14 clusters @ (k=80, res=0.5) — widest plateau across (k, res), the
#                                    strongest data-backed default
#   18 clusters @ (k=50, res=0.8) — fine end of plausibility band
# lam0.8:
#    5 niches  @ (k=50, res=0.2) — coarsest defensible niche set
#    8 niches  @ (k=120, res=0.5) — mid, sits on the 8-plateau
#   11 niches  @ (k=80, res=0.6) — widest plateau, finest defensible
finalists = data.table(
    lambda = c(0.2, 0.2, 0.2,
               0.8, 0.8, 0.8),
    k      = c( 80,  80,  50,
                50, 120,  80),
    res    = c(0.2, 0.5, 0.8,
               0.2, 0.5, 0.6)
)
finalists
```

# Layer 2 — biology for finalists

## Load subset and re-cluster finalists

We re-run Leiden only on the chosen combos — `clusterBanksy` expands a
(k_neighbors, resolution) vector internally, so two calls (one per lambda)
cover everything. Identical seed to the sweep keeps labels reproducible.

```{r layer2-recluster, message=FALSE}
# File-based cache: the recluster step is the only expensive chunk in this
# qmd (~10 min for 6 combos). Cache to an RDS so re-renders are instant;
# delete this file (or change its name) to force a fresh recluster after
# editing the finalists table.
recluster_cache = "evaluate_leiden_sweep_recluster.rds"

if (file.exists(recluster_cache)) {
    cat(sprintf("[%s] Loading cached recluster: %s\n",
                format(Sys.time()), recluster_cache))
    o = readRDS(recluster_cache)
} else {
    cat(sprintf("[%s] Loading subset object...\n", format(Sys.time())))
    o = readRDS(subset_path)
    colData(o)$sample_id = factor(colData(o)$sample_id)

    # Drop any clust_* columns that came along with the saved object so the
    # new clustering grid lands in clean colData slots.
    existing_clust = grep("^clust_", colnames(colData(o)), value = TRUE)
    if (length(existing_clust) > 0) colData(o)[, existing_clust] = NULL

    set.seed(seed_val)
    for (lam in unique(finalists$lambda)) {
        f = finalists[lambda == lam]
        cat(sprintf("[%s] Re-clustering lam%s for %d finalist combos...\n",
                    format(Sys.time()), lam, nrow(f)))
        o = Banksy::clusterBanksy(
            o, dimred = paste0("Harmony_BANKSY_lam", lam),
            k_neighbors = unique(f$k),
            resolution  = unique(f$res),
            algo = "leiden",
            seed = seed_val
        )
    }
    saveRDS(o, recluster_cache)
    cat(sprintf("[%s] Cached to: %s\n", format(Sys.time()), recluster_cache))
}

# Build the list of columns we'll iterate over below.
finalist_cols = finalists[, .(
    cluster_col = sprintf("clust_Harmony_BANKSY_lam%s_k%d_res%s",
                          lambda, k, res),
    lambda = lambda, k = k, res = res
)]
finalist_cols
```

## 2a — cluster size distribution

A healthy combo has no single cluster eating >40% of cells (under-segmented)
and no clusters with <50 cells (noise). Combos that fail either bound are
poor candidates regardless of marker / spatial story.

```{r layer2-cluster-sizes}
cluster_sizes = rbindlist(lapply(finalist_cols$cluster_col, function(cc) {
    tab = table(colData(o)[[cc]])
    data.table(
        cluster_col = cc,
        n_clusters  = length(tab),
        min_size    = min(tab),
        median_size = median(tab),
        max_size    = max(tab),
        max_frac    = max(tab) / sum(tab),
        n_small     = sum(tab < 50)   # under-50-cell clusters
    )
}))
cluster_sizes[, flag := fcase(
    max_frac > 0.40,           "under-segmented (max_frac>0.40)",
    n_small  >= 1,             "noise cluster(s) (<50 cells)",
    default                  = "ok"
)]
cluster_sizes
```

## 2b — sample mixing (Shannon entropy)

Per cluster, entropy of cell counts across samples normalized by
`log(n_samples)`. 1 = perfectly mixed across all samples; near 0 = the
cluster is essentially one sample. A single-sample cluster is **not**
automatically an artifact: in this histologically heterogeneous n=11 cohort,
patient-private clusters are expected — whether neoplastic or histology-specific
normal populations — so they legitimately fail to mix while normal lineages
(immune / vascular / glia / oligo) do integrate across patients. We therefore
(i) record each cluster's median
`transcript_counts` as a QC readout — a single-sample cluster with counts on
par with the clean lineages is a real patient-private population, not
low-quality noise — and (ii) defer the actual mixing judgement to the
lineage-aware step (2c.5) that feeds the scorecard.

```{r layer2-sample-entropy}
n_samples = nlevels(colData(o)$sample_id)
entropy_norm = function(x) {
    p = prop.table(table(x))
    p = p[p > 0]
    -sum(p * log(p)) / log(n_samples)
}

# Per-cluster table (not just a combo summary): we need per-cluster entropy to
# join against per-cluster lineage in 2c.5. med_counts is the artifact check.
lib = colData(o)$transcript_counts
entropy_bycluster = rbindlist(lapply(finalist_cols$cluster_col, function(cc) {
    by_clust  = split(colData(o)$sample_id, colData(o)[[cc]])
    by_counts = split(lib,                  colData(o)[[cc]])
    ent = vapply(by_clust, entropy_norm, numeric(1))
    data.table(
        cluster_col = cc,
        cluster     = names(by_clust),
        entropy     = ent,
        low_entropy = ent < 0.30,                        # near-single-sample
        med_counts  = vapply(by_counts, median, numeric(1))
    )
}))

# Combo-level summary carried into the scorecard (same columns as before).
sample_entropy = entropy_bycluster[, .(
    min_entropy    = min(entropy),
    median_entropy = median(entropy),
    n_low_entropy  = sum(low_entropy)
), by = cluster_col]
sample_entropy
```

Visual companion to the table: for each finalist combo, a UMAP colored by cluster
and a UMAP colored by sample (top row) over a per-cluster entropy barchart. This
lets you read off directly whether the low-entropy (firebrick) bars are the same
clusters that sit apart / are dominated by one sample on the UMAPs.

```{r layer2-entropy-plots, fig.width=13, fig.height=9, message=FALSE, warning=FALSE}
# Optional downsample for plotting ONLY (1.0 = all cells). Lower if the HPC
# render of ~12 large scatters is slow; does not touch any computed table.
plot_frac = 0.15
set.seed(seed_val)
plot_idx = if (plot_frac < 1) sort(sample(ncol(o), round(ncol(o) * plot_frac))) else seq_len(ncol(o))

# Precomputed lambda-matched UMAP coords, subset to the plotted cells.
umap_coords = list(
    `0.2` = as.data.table(reducedDim(o, "UMAP_Harmony_BANKSY_lam0.2"))[plot_idx, 1:2],
    `0.8` = as.data.table(reducedDim(o, "UMAP_Harmony_BANKSY_lam0.8"))[plot_idx, 1:2]
)
for (nm in names(umap_coords)) setnames(umap_coords[[nm]], c("umap_1", "umap_2"))

# Shared scatter builder (reused by the 2c.5 lineage UMAP in the same render
# session): color points by `vals`, optionally label centroids. `label_vals`
# lets us color by one variable (e.g. lineage) but label centroids by another
# (e.g. cluster id). All value vectors are full-length; we index by plot_idx.
umap_scatter = function(coords, vals, title, palette,
                        show_labels = TRUE, label_vals = NULL) {
    d = copy(coords)[, grp := factor(vals[plot_idx])]
    if (!is.null(label_vals)) d[, lab := factor(label_vals[plot_idx])]
    p = ggplot(d, aes(umap_1, umap_2, colour = grp)) +
        geom_point(size = 0.1, alpha = 0.5) +
        scale_colour_manual(values = palette, name = NULL, na.value = "grey80") +
        guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
        labs(title = title, x = "UMAP 1", y = "UMAP 2") +
        theme_bw()
    if (show_labels) {
        lab_col = if (!is.null(label_vals)) "lab" else "grp"
        cent = d[, .(x = median(umap_1), y = median(umap_2)), by = c(lab_col)]
        setnames(cent, lab_col, "lab")
        p = p + geom_text(data = cent, aes(x = x, y = y, label = lab),
                          inherit.aes = FALSE, colour = "black", size = 3.2)
    }
    p
}

# Sample palette shared across all combos so sample colors stay stable.
samp_levels = levels(colData(o)$sample_id)
sample_pal  = structure(pals::polychrome(length(samp_levels)), names = samp_levels)

# One composite per finalist combo (both lambdas), lambda-matched UMAP.
for (i in seq_len(nrow(finalist_cols))) {
    cc     = finalist_cols$cluster_col[i]
    lam    = as.character(finalist_cols$lambda[i])
    coords = umap_coords[[lam]]

    cl        = as.character(colData(o)[[cc]])
    cl_levels = sort(unique(cl))
    clpal     = structure(pals::polychrome(length(cl_levels)), names = cl_levels)

    p_cluster = umap_scatter(coords, cl, paste0(cc, "\nclusters"),
                             clpal, show_labels = TRUE)                     # label = cluster id
    p_sample  = umap_scatter(coords, as.character(colData(o)$sample_id),
                             "sample_id", sample_pal, show_labels = FALSE)

    # Barchart: clusters ordered by ascending entropy; 0.30 gate dashed.
    eb = entropy_bycluster[cluster_col == cc]
    eb[, cluster := factor(cluster, levels = cluster[order(entropy)])]
    p_entropy = ggplot(eb, aes(cluster, entropy, fill = low_entropy)) +
        geom_col() +
        geom_hline(yintercept = 0.30, linetype = "dashed") +
        scale_fill_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick"),
                          name = "entropy < 0.30") +
        labs(x = "cluster", y = "normalized entropy",
             title = "per-cluster sample-mixing entropy") +
        theme_bw()

    print(cowplot::plot_grid(
        cowplot::plot_grid(p_cluster, p_sample, nrow = 1),
        p_entropy, ncol = 1, rel_heights = c(2, 1)))
}
```

## 2c — marker expression per cluster (lam0.2 only)

Two views of the same marker panel: raw mean log-expression per cluster
(is the gene expressed at all?), and single-cell Wilcoxon AUC per cluster
(one-vs-rest specificity -- this is the exact metric 2c.5's lineage argmax
uses below, so it's the view to actually check the pick against). Good
lam0.2 combos light up each lineage in a distinct cluster on both.

```{r layer2-markers, fig.width=10, fig.height=5}
# Use logcounts if present, else fall back to counts on a log1p.
aname = if ("logcounts" %in% assayNames(o)) "logcounts" else "counts"
genes_all = unlist(marker_panel, use.names = FALSE)
genes_present = intersect(genes_all, rownames(o))
missing_genes = setdiff(genes_all, rownames(o))
if (length(missing_genes) > 0) {
    message("Genes not in panel rownames (skipped): ",
            paste(missing_genes, collapse = ", "))
}
lineage_of = setNames(rep(names(marker_panel), lengths(marker_panel)),
                       genes_all)

# Long-format mean per (cluster, gene), faceted by combo. Means computed
# directly to avoid coercing the SCE to Seurat for a quick diagnostic.
marker_mean = function(cc) {
    g = colData(o)[[cc]]
    expr = assay(o, aname)[genes_present, , drop = FALSE]
    if (aname == "counts") expr = log1p(expr)
    m = vapply(split(seq_along(g), g), function(idx) {
        Matrix::rowMeans(expr[, idx, drop = FALSE])
    }, numeric(length(genes_present)))
    rownames(m) = genes_present
    melt(as.data.table(m, keep.rownames = "gene"),
         id.vars = "gene", variable.name = "cluster", value.name = "mean_expr")
}
lam02_cols = finalist_cols[startsWith(cluster_col,
                                      "clust_Harmony_BANKSY_lam0.2"),
                           cluster_col]
marker_df = rbindlist(lapply(lam02_cols, function(cc) {
    out = marker_mean(cc); out[, cluster_col := cc]; out
}))
marker_df[, lineage := lineage_of[gene]]
marker_df[, gene := factor(gene, levels = genes_all)]

ggplot(marker_df, aes(x = cluster, y = gene, fill = mean_expr)) +
    geom_tile() +
    facet_grid(lineage ~ cluster_col, scales = "free", space = "free") +
    scale_fill_viridis_c() +
    labs(title = "Marker mean expression per cluster (lam0.2 finalists)",
         x = "cluster", y = NULL) +
    theme_bw() +
    theme(axis.text.x = element_text(size = 6),
          strip.text.y = element_text(angle = 0))
```

```{r layer2-markers-auc, fig.width=10, fig.height=5}
# Single-cell effect size (Wilcoxon AUC, one-vs-rest) per (cluster_col,
# cluster, gene), computed once here from the raw single cells and reused
# by both this heatmap and the 2c.5 lineage argmax below -- so what you
# eyeball here is exactly what drives the lineage call. AUC=0.5 means no
# separation between the cluster and every other cell; AUC=1 means the
# gene perfectly separates the cluster from the rest.
#
# Chosen over z-scoring per-cluster pseudobulk means (the prior approach):
# pseudobulk z-scoring only has as many points as there are clusters
# (12-18) to estimate a gene's mean/SD from, which penalizes markers with
# a naturally wide dynamic range across clusters (e.g. PAX6, which varies
# a lot cluster-to-cluster for real biological reasons) relative to
# markers with a narrow baseline -- even when the wide-range marker is far
# more specific at the single-cell level. AUC is rank-based and
# scale-invariant, so it doesn't have that failure mode, and it uses every
# cell in a cluster instead of collapsing to one pseudobulk point.
expr_all = assay(o, aname)[genes_present, , drop = FALSE]
if (aname == "counts") expr_all = log1p(expr_all)
auc_long = rbindlist(lapply(lam02_cols, function(cc) {
    res = as.data.table(presto::wilcoxauc(expr_all, factor(colData(o)[[cc]])))
    res[, cluster_col := cc]
    res
}))
setnames(auc_long, c("feature", "group"), c("gene", "cluster"))
auc_long[, lineage := lineage_of[gene]]
auc_long[, gene := factor(gene, levels = genes_all)]

ggplot(auc_long, aes(x = cluster, y = gene, fill = auc)) +
    geom_tile() +
    facet_grid(lineage ~ cluster_col, scales = "free", space = "free") +
    scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                         midpoint = 0.5, limits = c(0, 1),
                         name = "AUC\n(one-vs-rest)") +
    labs(title = "Marker specificity per cluster (single-cell Wilcoxon AUC, lam0.2 finalists)",
         x = "cluster", y = NULL) +
    theme_bw() +
    theme(axis.text.x = element_text(size = 6),
          strip.text.y = element_text(angle = 0))
```

## 2c.5 — per-cluster lineage assignment (lam0.2)

Assign each lam0.2 cluster a top lineage so the scorecard can treat sample
mixing correctly. A single-sample cluster is a red flag **only** when it maps to
a *normal* lineage (immune / vascular / glia / oligo — these should integrate
across patients). The glioneuronal / progenitor compartment is **expected** to be
patient-specific in this histologically heterogeneous cohort — whether those
cells are neoplastic or histology-specific normal populations — so it must not
gate the cell-type pick.

The signature is the mean single-cell Wilcoxon AUC minus 0.5 (one-vs-rest,
computed once in the 2c chunk above and reused here) over each lineage's
panel-present markers; the argmax labels the cluster. AUC replaced z-scoring
per-cluster pseudobulk means: pseudobulk z-scoring only has ~12-18 points
(one per cluster) to estimate a gene's mean/SD from, which penalized markers
with a wide dynamic range across clusters (e.g. PAX6) relative to markers
with a narrow baseline, even when the wide-range marker was the more specific
one at the single-cell level. These labels are still **directional, not
definitive cell-type calls**: the margins between the top and runner-up
lineage are often near zero for the glioneuronal clusters -- radial_glia and
neuronal markers genuinely co-occur there (e.g. PAX6 alongside GAD2/SOX11) --
so which subtype "wins" is sensitive to noise and can shift with resolution.
Read them as a coarse normal-vs-glioneuronal split, not a fine cell-type
annotation.

```{r layer2-lineage-assign}
# Lineages we expect to integrate across patients. radial_glia and the merged
# `neuronal` group are deliberately LEFT OUT: in this histologically heterogeneous
# cohort a patient-private glioneuronal / progenitor cluster is expected --
# whether neoplastic or a histology-specific normal population -- so a
# single-sample cluster mapping to one of those is not a mixing failure.
normal_lineages = c("astro_glial", "opc", "oligo_mature",
                    "myeloid", "t_cell", "endothelial")

# Per-cluster top lineage for one finalist column: from the shared single-cell
# AUC table (auc_long, computed once in the 2c chunk) -> mean(auc - 0.5)
# within each lineage's panel-present markers -> argmax. Ties broken by
# marker_panel list order (astro_glial first).
lineage_score = function(cc) {
    d = auc_long[cluster_col == cc]
    sig_long = d[, .(score = mean(auc - 0.5)), by = .(cluster, lineage)]
    sig_wide = dcast(sig_long, cluster ~ lineage, value.var = "score")
    lineage_names = names(marker_panel)
    m = as.matrix(sig_wide[, ..lineage_names])
    top = lineage_names[max.col(replace(m, is.na(m), -Inf),
                                ties.method = "first")]
    data.table(cluster_col = cc, cluster = sig_wide$cluster, top_lineage = top)
}

lam02_lineage = rbindlist(lapply(lam02_cols, lineage_score))
lam02_lineage[, is_normal_lineage := top_lineage %in% normal_lineages]
# Merge in entropy + QC so the private-vs-artifact call is visible in one table.
lam02_lineage = merge(
    lam02_lineage,
    entropy_bycluster[, .(cluster_col, cluster, entropy, low_entropy, med_counts)],
    by = c("cluster_col", "cluster"), sort = FALSE
)
setorder(lam02_lineage, cluster_col, entropy)
lam02_lineage
```

Preliminary cell-type UMAP: each lam0.2 finalist, cells colored by their cluster's
assigned `top_lineage`, with the cluster id labeled at each centroid. The low-entropy
clusters from 2b should land in the neuronal / radial-glia (glioneuronal / progenitor)
lineages, while the normal lineages (myeloid, OPC, oligo, astro, T cell, endothelial)
should spread across the UMAP and across samples.

```{r layer2-lineage-umap, fig.width=15, fig.height=5.5, message=FALSE, warning=FALSE}
# Fixed palette over all lineage names so colors are identical across the three
# resolutions. Reuses umap_scatter / umap_coords / plot_idx from the 2b plot chunk.
lineage_pal = structure(pals::alphabet(length(marker_panel)), names = names(marker_panel))

lineage_umaps = lapply(lam02_cols, function(cc) {
    # cluster -> top_lineage lookup for this combo, then map to each cell.
    lin_map  = lam02_lineage[cluster_col == cc, setNames(top_lineage, cluster)]
    cell_lin = lin_map[as.character(colData(o)[[cc]])]
    # color by lineage, but label centroids by cluster id for cross-reference to 2b.
    umap_scatter(umap_coords[["0.2"]], cell_lin,
                 sub("clust_Harmony_BANKSY_", "", cc), lineage_pal,
                 show_labels = TRUE, label_vals = as.character(colData(o)[[cc]]))
})
print(cowplot::plot_grid(plotlist = lineage_umaps, nrow = 1))
```

## 2d — spatial coherence (lam0.8 only)

Niches should form contiguous patches in xy, not salt-and-pepper. Eyeball
2–3 samples per finalist. If multiple samples disagree on which combo looks
contiguous, weight the lam0.8 decision toward the combo that holds up
across samples.

```{r layer2-spatial, fig.width=12, fig.height=8}
lam08_cols = finalist_cols[startsWith(cluster_col,
                                      "clust_Harmony_BANKSY_lam0.8"),
                           cluster_col]
# Pick 3 samples with the largest cell counts so the spatial pattern is
# visible at small point sizes; swap for specific samples of interest later.
sample_pick = names(sort(table(colData(o)$sample_id), decreasing = TRUE))[1:3]

coords = as.data.table(spatialCoords(o))
setnames(coords, c("x", "y"))
coords[, sample_id := colData(o)$sample_id]
for (cc in lam08_cols) coords[[cc]] = factor(colData(o)[[cc]])

for (cc in lam08_cols) {
    # facet_wrap(scales="free") is incompatible with coord_equal (fixed
    # ratio). For spatial coherence we care more about contiguity than
    # true isotropy, so use theme(aspect.ratio=1) which works with free
    # scales and keeps each facet roughly square.
    p = ggplot(coords[sample_id %in% sample_pick],
               aes(x = x, y = y, colour = .data[[cc]])) +
        geom_point(size = 0.1, alpha = 0.6) +
        facet_wrap(~ sample_id, scales = "free", nrow = 1) +
        guides(colour = guide_legend(override.aes = list(size = 2))) +
        labs(title = cc) +
        theme_bw() +
        theme(legend.position = "bottom",
              aspect.ratio = 1)
    print(p)
}
```

# Decision scorecard

Roll the diagnostics into one table so the final pick is mechanical. Sample
mixing is applied **per axis**, reflecting what 2b / 2c.5 showed:

- **lam0.8 (niches):** keep the original hard gate — a single-sample niche is
  most likely an artifact, and there is no marker-based lineage to reason about.
- **lam0.2 (cell types):** mixing is a red flag **only** for clusters that map
  to a *normal* lineage. Patient-private glioneuronal / progenitor clusters are
  expected (whether neoplastic or histology-specific normal), so they do not gate
  the pick. We count
  only low-entropy clusters whose top lineage is normal (`n_low_entropy_normal`);
  the cell-type resolution is then decided by size sanity + marker separation
  (2c) + parsimony, not by raw sample mixing.

If two combos tie, prefer the lower-res / lower-k one (parsimony).

```{r decision-scorecard}
# lam0.2 mixing failures = low-entropy clusters that map to a NORMAL lineage.
# (lam0.8 combos aren't in lam02_lineage, so they get 0 here and fall back to
# the original n_low_entropy hard gate below.)
low_entropy_normal = merge(
    entropy_bycluster[low_entropy == TRUE, .(cluster_col, cluster)],
    lam02_lineage[is_normal_lineage == TRUE, .(cluster_col, cluster)],
    by = c("cluster_col", "cluster")
)[, .(n_low_entropy_normal = .N), by = cluster_col]

scorecard = merge(merge(
    cluster_sizes[, .(cluster_col, n_clusters, max_frac, n_small, size_flag = flag)],
    sample_entropy[, .(cluster_col, min_entropy, n_low_entropy)],
    by = "cluster_col"
), finalist_cols, by = "cluster_col")
scorecard = merge(scorecard, low_entropy_normal, by = "cluster_col", all.x = TRUE)
scorecard[is.na(n_low_entropy_normal), n_low_entropy_normal := 0]

scorecard[, passes_size := size_flag == "ok"]
# lam0.2: lineage-aware mixing; lam0.8: original hard gate on all low-entropy.
scorecard[, passes_mixing := fifelse(lambda == 0.2,
                                     n_low_entropy_normal == 0,
                                     n_low_entropy == 0)]
scorecard[, recommend := passes_size & passes_mixing]
scorecard[order(lambda, n_clusters)]
```

Once a winner per lambda is picked from this table, update the downstream
notebooks (`banksy_clusters_proseg.qmd`, `banksy_niches_proseg.qmd`) to use
those (k, res) defaults and rerun. For lam0.2, confirm the pick on the 2c AUC
heatmap (each lineage specific to a distinct cluster) before committing, since
the scorecard no longer gates cell types on sample mixing.
