---
title: "Modeling of Cell Type Proportions with MAPK Activation"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

Here we will be using linear models to test the association between cell types and MAPK activation. We previously generated imputed MAPK overlap geneset score on the xenium data.

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(dbscan)
library(circlize)
library(Seurat)
library(scales)
library(htmlwidgets)
library(ggrepel)
library(rcartocolor)


# local vs remote
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (Sys.getenv("HOME") == "/PHShome/yz762"){
    # in eristwo, the home directory is at /PHShome/yz762
    workdir <- "/data/beroukhim1/"
} else if (Sys.getenv("HOME") == "/home/yz762"){
    # on the new server, the home directory is at /home/yz762
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    # in dipg or uger, the home directory is '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
```


# Preperations

## Loading the data

The objects are here:

RDS object with Leiden Clusters: `/data/beroukhim1/youyun/plgg/data/Xenium_annotations/banksy_clusters_connected_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds`

    -   `/data/beroukhim1/youyun/plgg/data/Xenium_annotations/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds`

Cell type clusters annotation: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv`  


```{r}
# load the banksy embeddings
banksy_embeddings <- readRDS(ifelse(
    !Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng"), paste0(
        workdir, "youyun/plgg/data/Xenium_annotations/banksy_clusters_connected_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds"
    ), "/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds"
))
colData(banksy_embeddings)$sample_id = factor(colData(banksy_embeddings)$sample_id)

# Load the cell type clusters annotation
prelim_annotation_res1 = fread(paste0(
    workdir,'youyun/plgg/data/xenium_cell_types/banksy_proseg_cell_cluster_annotation.tsv'
))

# specify the cluster names 
# this is mainly to keep track of which resolution we are using
cell_res = 1
cell_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k50_res%.1f',cell_res))
niche_res = 1
niche_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.8_k50_res%.1f',niche_res))

# # filter the banksy object to only include the cells with annotated cell types
# annotated_cell_clusters = sort(unique(prelim_annotation_res1$cluster))
# banksy_embeddings = banksy_embeddings[,colData(banksy_embeddings)[,cell_clust_name] %in% annotated_cell_clusters]
```


## Build the color palette for all niches and cell types downstream

```{r}
set.seed(5555)
cluster_classes = levels(colData(banksy_embeddings)[[niche_clust_name]])
niche_color_manual = structure(
    pals::polychrome()[c(1:length(cluster_classes))],
    names = sample(cluster_classes)
)
# for niche resolution 0.5
# niches_to_show = c(1:11, 13, 14, 16, 17)
# for niche resolution 1
niches_to_show = c(1:20, 22:25)

cell_type_to_test = 'Fine Cell Type UMAP'
cell_type_to_test = 'Fine Cell Type'
cell_type_to_test = 'Coarse Cell Type'

cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))
cell_type_color_manual = structure(
    pals::alphabet()[c(1:length(cell_type_class))],
    names = cell_type_class
)
```

## MAPK score

```{r}
mapk_scores = fread(paste0(workdir,'youyun/plgg/code/sc_integration/ENVI/2_spatial_imputation/PA/MAPK_overlap_scores.csv'))
```

# Which cell types are significantly MAPK-high (or -low), within each patient

`sc.tl.score_genes` fits its control-gene baseline independently per sample (see
`1_PA_MAPK_scores.qmd`), so raw score magnitude is not comparable *across*
patients -- a patient with a systematically shifted baseline (e.g. from
per-sample SCTransform scaling) will look "MAPK-high" everywhere even with no
real pathway difference. To sidestep that, we test each cell type against the
*other* cells in the *same* patient/type (two-sided Mann-Whitney -- higher OR
lower than the rest), which only relies on that patient's own internal
ranking, not its absolute scale.

This section only needs `mapk_scores`, so it runs without loading the (large)
BANKSY object below.

```{r}
# sample id + driver alteration lookup. `sample` is formatted as e.g.
# "188570 pilocytic astrocytoma"; the id is always the leading digits.
mapk_scores[, sample_id := gsub('^([0-9]+).*', '\\1', sample)]
mapk_scores[, alteration := fcase(
    sample_id == '188570', 'KIAA1549::BRAF',
    sample_id == '258482', 'FGFR1 ITD',
    sample_id == '320784', 'FAM131B::BRAF',
    sample_id == '328368', 'RAF1::TOP2B'
)]
# fixed sample/alteration display order (numeric sample id order), reused by the heatmaps below
alteration_levels = c('KIAA1549::BRAF', 'FGFR1 ITD', 'FAM131B::BRAF', 'RAF1::TOP2B')

# NOTE: spatial cell types come from `Fine.Cell.Type.UMAP` (Xenium/BANKSY
# clustering, with subtypes e.g. "Astrocyte 1"/"Astrocyte 2", "OPC 1"); sn cell
# types come from `annot_v1` (a coarser Seurat reference annotation: just
# "Astrocytes", "OPC"). These are two different annotation pipelines at two
# different resolutions -- they are not the same vocabulary, so don't expect
# spatial/sn categories to line up 1:1 below; compare shape/ranking within a
# panel, not exact labels across panels.

MIN_N_GROUP = 20  # drop (patient, cell type) groups too small to test reliably

# fread reads a blank CSV field as "" rather than NA for character columns, so
# is.na(cell_type) alone would silently keep these rows in the "rest of the
# patient" background group (unlike pandas' dropna, which treats blanks as NaN)
mapk_scores[cell_type == '', cell_type := NA]

patient_sig_combos = unique(mapk_scores[!is.na(cell_type), .(type, sample_id, cell_type)])

patient_sig_results = rbindlist(mapply(function(ty, sid, ct) {
    in_group = mapk_scores[type == ty & sample_id == sid & cell_type == ct, MAPK_overlap_score]
    out_group = mapk_scores[type == ty & sample_id == sid & cell_type != ct, MAPK_overlap_score]
    if (length(in_group) < MIN_N_GROUP || length(out_group) < MIN_N_GROUP) return(NULL)

    # exact = FALSE: use the normal approximation, since the exact Mann-Whitney
    # null distribution is infeasible at these group sizes (up to ~1e5 cells).
    # two-sided: we want to flag cell types that are significantly MAPK-LOW
    # too, not just MAPK-high -- see effect below.
    test = wilcox.test(in_group, out_group, alternative = 'two.sided', exact = FALSE)
    n1 = length(in_group); n2 = length(out_group)
    # rank-biserial effect size from the U statistic: -1 = in_group always
    # lower than the rest, 0 = no difference, +1 = in_group always higher.
    # Unlike a one-sided test, a two-sided-significant row CAN have a negative
    # effect (genuinely MAPK-low) -- the heatmap below uses a diverging
    # red/grey/blue scale specifically so those aren't collapsed into "not
    # significant" grey.
    # as.numeric() before multiplying: n1*n2 as plain R integers can silently
    # overflow 32-bit int (NA, no warning surfaced downstream) once group
    # sizes climb into the tens of thousands -- doesn't happen at current
    # per-patient group sizes, but cheap to guard against for good.
    effect = 2 * unname(test$statistic) / (as.numeric(n1) * as.numeric(n2)) - 1

    data.table(
        type = ty, sample_id = sid, cell_type = ct,
        n = n1, median = median(in_group), median_rest = median(out_group),
        p = test$p.value, effect = effect
    )
}, patient_sig_combos$type, patient_sig_combos$sample_id, patient_sig_combos$cell_type, SIMPLIFY = FALSE))

# BH-correct within each patient x type family -- that's the natural "family of
# tests" here (a per-patient discovery list), not a correction across patients
patient_sig_results[, q := p.adjust(p, method = 'BH'), .(type, sample_id)]
patient_sig_results = merge(patient_sig_results, unique(mapk_scores[, .(sample_id, alteration)]), by = 'sample_id')
patient_sig_results[, alteration := factor(alteration, levels = alteration_levels)]

fwrite(
    patient_sig_results,
    paste0(workdir, 'youyun/plgg/code/sc_integration/ENVI/2_spatial_imputation/PA/mapk_patient_cell_type_significance.tsv'),
    sep = '\t'
)
```

## Heatmap: patient x cell type

```{r}
# Patient (row) x cell type (column) heatmap, one ggplot per grouping, faceted
# by data modality (spatial vs sn) instead of two separate plots. Cell type is
# now on the same (coarse, harmonized) vocabulary in both modalities -- see
# 1_PA_MAPK_scores.qmd -- so a single shared column order/x-axis is meaningful
# across facets, not just within one.
#
# Fill = rank-biserial effect size for cell types significant at q < 0.05
# (diverging blue/white/red: blue = significantly MAPK-low, red = MAPK-high,
# saturation = how separated); grey = tested but not significant. A cell type
# absent from a modality's vocabulary (e.g. "RGC" in sn) simply has no tile in
# that facet. Columns are ranked by mean effect (descending), so the most
# consistently MAPK-high cell types read leftmost and the most consistently
# MAPK-low ones read rightmost -- ties (e.g. never significant either way)
# broken by total number of patient x modality significant hits.
plot_patient_sig_heatmap = function(
    dt, subtitle = '', title_size = 15, title_wrap = NULL, legend_position = 'right',
    legend_breaks = waiver(), title_override = NULL
) {
    dt_sub = copy(dt)
    col_rank = dt_sub[, .(
        n_sig = sum(q < 0.05), mean_effect = mean(effect)
    ), cell_type][order(-mean_effect, -n_sig)]
    dt_sub[, cell_type := factor(cell_type, levels = col_rank$cell_type)]
    # only cells significant at q < 0.05 get a graded fill (can be negative --
    # two-sided test, see above); everything else (tested-but-not-significant)
    # falls back to `na.value` below
    dt_sub[, fill_effect := ifelse(q < 0.05, effect, NA_real_)]
    dt_sub[, modality := factor(
        fifelse(type == 'spatial', 'Spatial (ENVI-imputed)', 'Single-nucleus (observed)'),
        levels = c('Spatial (ENVI-imputed)', 'Single-nucleus (observed)')
    )]

    # title_override bypasses the constructed "MAPK-high/-low..." title
    # entirely -- for narrow figures a short purpose-built title is more
    # reliable than wrapping the (long, subtitle-dependent) default one
    title_text = if (!is.null(title_override)) {
        title_override
    } else {
        paste0('MAPK-high/-low cell types by patient', subtitle)
    }
    # for narrow figures, wrap the title so it can't run past the plot edge
    # (base strwrap rather than stringr, no new dependency needed for this)
    if (!is.null(title_wrap)) {
        title_text = paste(strwrap(title_text, width = title_wrap), collapse = '\n')
    }

    ggplot(dt_sub, aes(x = cell_type, y = alteration, fill = fill_effect)) +
        geom_tile(color = 'white', linewidth = 1.1) +
        facet_wrap(~ modality, ncol = 1) +
        scale_fill_gradient2(
            low = '#2166ac', mid = '#f7f7f7', high = '#b2182b', midpoint = 0,
            na.value = 'grey92', limits = c(-1, 1), breaks = legend_breaks,
            name = 'Effect size\n(q < 0.05)'
        ) +
        scale_x_discrete(expand = expansion(add = 0.6)) +
        scale_y_discrete(expand = expansion(add = 0.6)) +
        labs(title = title_text) +
        theme_minimal(base_size = 14) +
        theme(
            axis.title = element_blank(),
            axis.text.x = element_text(angle = 45, hjust = 1, size = 11, color = 'grey20'),
            axis.text.y = element_text(size = 11.5, face = 'bold', color = 'grey20'),
            panel.grid = element_blank(),
            strip.text = element_text(face = 'bold', size = 12.5, hjust = 0, color = 'grey15'),
            strip.background = element_blank(),
            panel.spacing.y = unit(1.6, 'lines'),
            plot.title = element_text(face = 'bold', size = title_size, margin = margin(b = 10)),
            legend.title = element_text(size = 10.5),
            legend.key.height = unit(1.4, 'lines'),
            legend.position = legend_position
        )
}
```

```{r}
#| fig-width: 8
#| fig-height: 7
plot_patient_sig_heatmap(patient_sig_results)
```

```{r}
#| fig-width: 7.5
#| fig-height: 6
# Zoomed: OPC/astrocyte lineage only -- the tumor-relevant populations in PA
# (an OPC/astrocytic-lineage tumor), as opposed to microenvironment/immune cell
# types. "undetermined" is included here too: it's an sn-only cell-typing
# artifact category (never appears in spatial data), not a real cell type, but
# it's still part of the same "poorly-typed tumor-adjacent" grouping question.
tumor_lineage_regex = 'Astro|OPC|undetermined'
plot_patient_sig_heatmap(
    patient_sig_results[grepl(tumor_lineage_regex, cell_type)],
    ' -- OPC/astrocyte lineage', title_wrap = 46
)
```

```{r}
#| fig-width: 4.2
#| fig-height: 6
# Same OPC/astrocyte zoom, spatial only, at half the width of the combined
# figure above -- title is wrapped and the legend moved below the tiles so
# neither runs past the narrower plot edge.
plot_patient_sig_heatmap(
    patient_sig_results[type == 'spatial' & grepl('Astro|OPC', cell_type)],
    title_size = 13, legend_position = 'bottom', legend_breaks = c(-1, 0, 1),
    title_override = 'OPC/astrocyte -- spatial'
)
```

# Visualize MAPK by niche

```{r}
cell_niches = as.data.table(colData(banksy_embeddings[
    , colData(banksy_embeddings)$histology == 'pilocytic astrocytoma'
])[, c('alteration',niche_clust_name)], keep.rownames = 'cell_id')[
    , alteration := fcase(
        grepl('188570', cell_id), 'KIAA1549::BRAF',
        grepl('258482', cell_id), 'FGFR1 ITD',
        grepl('320784', cell_id), 'FAM131B::BRAF',
        grepl('328368', cell_id), 'RAF1::TOP2B'
    )
]
setnames(cell_niches, c(niche_clust_name), c('niche_type'))
# select only the original columns from mapk_scores: it now also carries
# sample_id/alteration (added above), which would otherwise collide with
# cell_niches' own (independently-derived) versions of those same columns
mapk_scores_w_niches = merge(
    mapk_scores[, .(cell_id, sample, type, cell_type, MAPK_overlap_score)], cell_niches,
    by.x = 'cell_id', by.y = 'cell_id', all.y = TRUE
)[
    , sample_id := gsub('.*__','',cell_id)
][
    , niche_type := ifelse(
        # niche_type %in% c(1,2,3,5,7,9,19,11,14,16,18,19,22,27),
        niche_type %in% c(7, 18, 22),
        niche_type, 'Other'
    )
]

# Save the per cell type x niche summary (mean/SE) so downstream reporting
# can reuse it without re-loading the full BANKSY object.
mapk_by_niche_summary = mapk_scores_w_niches[
    grepl('Astro|OPC', cell_type) & niche_type != 'Other',
    .(
        MAPK_overlap_score_mean = mean(MAPK_overlap_score),
        MAPK_overlap_score_se = sd(MAPK_overlap_score)/sqrt(.N),
        n_cells = .N
    ), .(cell_type, niche_type = factor(as.numeric(niche_type)))
]
fwrite(
    mapk_by_niche_summary,
    paste0(workdir, 'youyun/plgg/code/sc_integration/ENVI/2_spatial_imputation/PA/mapk_by_niche_summary.tsv'),
    sep = '\t'
)

ggplot(
    mapk_by_niche_summary, aes(
        x = factor(cell_type), y = MAPK_overlap_score_mean
    )
) +
    geom_point(
        aes(group = niche_type, color = niche_type),
        position = position_dodge(width = 0.8)
    ) +
    geom_errorbar(
        aes(
            y = MAPK_overlap_score_mean,
            ymin = MAPK_overlap_score_mean - MAPK_overlap_score_se,
            ymax = MAPK_overlap_score_mean + MAPK_overlap_score_se,
            color = niche_type,
            group = interaction(cell_type, niche_type)
        ),
        position = position_dodge(width = 0.8),
        width = 0.2, size = 1
    ) +
    theme_bw(base_size = 20) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'bottom'
    ) +
    labs(
        title = "MAPK Activation by Niche Type",
        x = "Cell Type",
        y = "MAPK Activation Score"
    ) +
    scale_color_carto_d(name = "Niche Type", palette = "Vivid") +
    guides(
        fill = guide_legend(title = "Niche Type", nrow = 2, byrow = TRUE)
    )
```

# Neighborhood analysis

## Setting up Neighborhood Analysis

Here we will look at the niches in the data and annotate them using the cell types of self and neighbors.

```{r}
k_geom = 15
locs = spatialCoords(banksy_embeddings[, colData(banksy_embeddings)$histology == 'pilocytic astrocytoma'])
knn = dbscan::kNN(x = locs, k = k_geom)
knnDT <- data.table(
    from = rep(seq_len(nrow(knn$id)), k_geom),
    from_id = rep(rownames(knn$id), k_geom),
    to = as.vector(knn$id),
    to_id = rownames(knn$id)[as.vector(knn$id)],
    distance = as.vector(knn$dist)
)
cell_cell_type_dt = data.table(
    cell_id = rownames(colData(banksy_embeddings[])),
    cell_type = as.character(colData(banksy_embeddings)[,cell_clust_name]),
    sample_id = colData(banksy_embeddings)$sample_id,
    histology = colData(banksy_embeddings)$histology
)[histology == 'pilocytic astrocytoma']

sparsity_threshold = 0.01
cell_cell_type_dt[
    , num_cells_per_sample := .N, .(sample_id)
][
    , cell_type := ifelse(
        .N/num_cells_per_sample < sparsity_threshold |
            is.na(cell_type), 'Other', cell_type
    ),.(sample_id, cell_type)
]
```

# Cell Cell Adjacency Analysis

To do this, we compare each niche's cell cell adjacency to all other niches' cell cell adjacency. 

```{r}
cell_type_assignment_vector = structure(
    c(prelim_annotation_res1[[cell_type_to_test]], 'Other'),
    names = c(prelim_annotation_res1$cluster,'Other')
)
adjacencies = knnDT[,.(from_id,to_id,distance)][
    cell_cell_type_dt[,.(cell_id,cell_type_to = cell_type)], 
    on = c('to_id' = 'cell_id')
][
    cell_cell_type_dt[,.(
        cell_id, cell_type_from = cell_type, sample_id, histology
    )], on = c('from_id' = 'cell_id')
][
    , c('cell_type_test_from','cell_type_test_to') := list(
        cell_type_assignment_vector[cell_type_from],
        cell_type_assignment_vector[cell_type_to]
    )
]
```

# Modeling cell type proportions with MAPK activation

```{r}
adjacencies_and_mapk = merge(
    dcast(
        adjacencies[,.(neighbor_prop = .N/k_geom),.(from_id, cell_type_test_from, cell_type_test_to, sample_id)],
        from_id + cell_type_test_from + sample_id ~ cell_type_test_to,
        value.var = 'neighbor_prop', fill = 0
    ),
    mapk_scores[,.(cell_id, MAPK_overlap_score)], by.x = 'from_id', by.y = 'cell_id'
)[, weights := k_geom]
cell_types_in_dt = setdiff(colnames(adjacencies_and_mapk), c('from_id', 'cell_type_test_from', 'sample_id', 'MAPK_overlap_score', 'weights'))
cell_type_pairs_to_test = expand.grid(cell_type_from = cell_types_in_dt, cell_type_to = cell_types_in_dt)

mapk_adj_results = rbindlist(mapply(
    function(ct1, ct2) {
        print(paste0("Testing ", ct1, " -> ", ct2))
        if (nrow(adjacencies_and_mapk[cell_type_test_from == ct1]) == 0) {
            message(sprintf("Cell type %s not found in the data!", ct1))
            return(NULL)
        }
        if (all(adjacencies_and_mapk[[ct2]] == 0)) {
            message(sprintf("Cell type %s has all zeros!", ct))
            return(NULL)
        }
        
        tmp_dt = adjacencies_and_mapk[cell_type_test_from == ct1]

        if (length(unique(tmp_dt$sample_id)) == 1){
            model = glm(
                as.formula(paste0('`',ct2, "` ~ MAPK_overlap_score")),
                data = tmp_dt, family = binomial(link = 'logit'), weights = weights
            )   
        } else {
            model = glm(
                as.formula(paste0('`',ct2, "` ~ MAPK_overlap_score + sample_id")),
                data = tmp_dt, family = binomial(link = 'logit'), weights = weights
            )
        }
        model_coef = as.data.frame(summary(model)$coefficients)
        model_coef$variables = rownames(model_coef)
        model_coef = as.data.table(model_coef)[variables == 'MAPK_overlap_score'][
            ,c('neg_log10_p_value', 'effect','ct_from','ct_to') := list(
                -log10(`Pr(>|z|)`), exp(Estimate),
                ct1, ct2
            )
        ]
        return(model_coef)
    }, 
    cell_type_pairs_to_test$cell_type_from, 
    cell_type_pairs_to_test$cell_type_to, 
    SIMPLIFY = FALSE
))[
    , q_value := p.adjust(`Pr(>|z|)`, method = 'BH')
]

# save so downstream reporting can reuse this without re-loading the BANKSY
# object / re-fitting ~n_cell_types^2 GLMs
fwrite(
    mapk_adj_results,
    paste0(workdir, 'youyun/plgg/code/sc_integration/ENVI/2_spatial_imputation/PA/mapk_adj_results.tsv'),
    sep = '\t'
)
```

# Results

```{r}
ggplot(mapk_adj_results[grepl('Astro|OPC', ct_from)]) +
    geom_point(aes(
        x = Estimate, y = neg_log10_p_value,
        color = ifelse(
            q_value >= 0.05,'not significant',
            ifelse(
                as.numeric(effect) > 1, 'enriched','depleted'
            )
        ), alpha = 0.5
    )) +
    geom_text_repel(aes(
        x = Estimate, y = neg_log10_p_value,
        label = ifelse(q_value < 0.05, gsub('`','',ct_to), '')
    ), max.overlaps = 100, force = 3) +
    theme_linedraw() +
    theme(
        text = element_text(size = 15), 
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'right', 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()
    ) + scale_color_manual(values = c(
        'enriched' = 'red', 'depleted' = 'blue',
        'not significant' = 'grey'
    )) +
    geom_vline(xintercept = 0, linetype = 'dashed', color = 'black') +
    xlim(-1, 0.5) +
    labs(
        title = "Effect of Imputed MAPK on Neighbor Cell Type Proportions",
        x = "Effect size (log coefficient)",
        y = "-log10(p-value)",
        color = "Significance"
    ) + facet_wrap(~ ct_from, nrow = 1)
```

## Per-patient neighborhood composition vs. MAPK activation

The volcano above pools all 4 patients into one model per cell-type pair
(with `sample_id` as a covariate), which asks "on average across patients,
does a cell's neighbor composition track its MAPK score." Here we instead fit
one model per (cell type, patient) -- no pooling, no covariate, since each
fit is already a single patient -- to ask how that relationship varies by
driver alteration. Restricted to OPC/Astrocyte (the tumor-lineage cell types)
as the focal cell whose neighbors we're examining.

```{r}
tumor_lineage_cell_types = c('Astrocyte', 'OPC')
# adjacencies_and_mapk$sample_id carries factor levels from the full (11
# sample, 5 histology) banksy_embeddings object; as.character + unique here
# restricts to the samples actually present after the PA-only filter above
sample_ids_present = sort(unique(as.character(adjacencies_and_mapk$sample_id)))
sample_alteration_map = unique(mapk_scores[, .(sample_id, alteration)])

by_patient_combos = expand.grid(
    cell_type_from = tumor_lineage_cell_types,
    cell_type_to = cell_types_in_dt,
    sample_id = sample_ids_present,
    stringsAsFactors = FALSE
)

mapk_adj_results_by_patient = rbindlist(mapply(
    function(ct1, ct2, sid) {
        tmp_dt = adjacencies_and_mapk[cell_type_test_from == ct1 & as.character(sample_id) == sid]
        if (nrow(tmp_dt) == 0 || all(tmp_dt[[ct2]] == 0)) return(NULL)

        # single-predictor model: no sample_id covariate needed since tmp_dt
        # is already restricted to one patient
        model = glm(
            as.formula(paste0('`', ct2, '` ~ MAPK_overlap_score')),
            data = tmp_dt, family = binomial(link = 'logit'), weights = weights
        )
        model_coef = as.data.frame(summary(model)$coefficients)
        model_coef$variables = rownames(model_coef)
        model_coef = as.data.table(model_coef)[variables == 'MAPK_overlap_score'][
            # -log10(p) from the z-score directly in log-space (pnorm(...,
            # log.p=TRUE)), not from Pr(>|z|) -- with these group sizes Pr(>|z|)
            # can underflow to exactly 0 (double precision floor ~1e-308) for
            # large |z|, which would otherwise make -log10(p) = Inf and break
            # any plot axis built on it. The true -log10(p) is still finite
            # (just very large) -- only the intermediate p-value underflows.
            , c('neg_log10_p_value', 'effect', 'ct_from', 'ct_to', 'sample_id') := list(
                -(log(2) + pnorm(-abs(`z value`), log.p = TRUE)) / log(10),
                exp(Estimate), ct1, ct2, sid
            )
        ]
        return(model_coef)
    },
    by_patient_combos$cell_type_from,
    by_patient_combos$cell_type_to,
    by_patient_combos$sample_id,
    SIMPLIFY = FALSE
))

# BH-correct within each (cell type, patient) facet cell -- same convention as
# the per-patient significance heatmap above: each facet is its own discovery
# list, not corrected against the other 7 panels
mapk_adj_results_by_patient[, q_value := p.adjust(`Pr(>|z|)`, method = 'BH'), .(ct_from, sample_id)]
mapk_adj_results_by_patient = merge(mapk_adj_results_by_patient, sample_alteration_map, by = 'sample_id')
mapk_adj_results_by_patient[, alteration := factor(alteration, levels = alteration_levels)]

fwrite(
    mapk_adj_results_by_patient,
    paste0(workdir, 'youyun/plgg/code/sc_integration/ENVI/2_spatial_imputation/PA/mapk_adj_results_by_patient.tsv'),
    sep = '\t'
)
```

```{r}
#| fig-width: 13
#| fig-height: 7.5
# colors match the diverging red/blue used by the significance heatmaps above
# (red = enriched, blue = depleted).
#
# facet_wrap, not facet_grid, for the free scales: facet_grid's scales='free'
# only frees x PER COLUMN and y PER ROW (panels in the same row still share
# one y-axis, same for columns and x) -- so one outlier panel (e.g. RAF1::TOP2B
# / Astrocyte reaching -log10(p) ~800) was still stretching every other panel
# in its row/column flat. facet_wrap's scales='free' gives each of the 8
# panels its own fully independent x AND y range, which is what's actually
# needed here (checked directly: Estimate range from [-0.03,0.01] to
# [-1.46,0.14], -log10(p) up to >800 in some panels vs <10 in others).
plot_dt = copy(mapk_adj_results_by_patient)

# per-panel context: how many focal-cell-type (OPC/Astrocyte) cells are in
# this patient's spatial sample, and what fraction of that sample they make
# up -- shown in the strip subtitle so a null result can be read against how
# much statistical power/tissue prevalence backed it
sp_composition = mapk_scores[type == 'spatial' & !is.na(cell_type)]
sample_totals = sp_composition[, .(total_n = .N), sample_id]
focal_composition = sp_composition[
    cell_type %in% tumor_lineage_cell_types, .(n = .N), .(sample_id, cell_type)
]
focal_composition = merge(focal_composition, sample_totals, by = 'sample_id')
focal_composition[, pct := 100 * n / total_n]
setnames(focal_composition, 'cell_type', 'ct_from')

plot_dt[, sample_id := as.character(sample_id)]
plot_dt = merge(plot_dt, focal_composition[, .(sample_id, ct_from, n, pct)], by = c('sample_id', 'ct_from'))
plot_dt[, panel_subtitle := sprintf('n=%s (%.1f%% of sample)', format(n, big.mark = ','), pct)]
plot_dt[, panel_label := paste0(ct_from, '  ·  ', alteration, '\n', panel_subtitle)]

panel_levels = unique(plot_dt[order(
    factor(ct_from, levels = tumor_lineage_cell_types), alteration
), panel_label])
plot_dt[, panel_label := factor(panel_label, levels = panel_levels)]

ggplot(plot_dt, aes(x = Estimate, y = neg_log10_p_value)) +
    geom_hline(
        yintercept = -log10(0.05), linetype = 'dotted', color = 'grey55', linewidth = 0.4
    ) +
    geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey40', linewidth = 0.5) +
    geom_point(aes(color = ifelse(
        q_value >= 0.05, 'not significant',
        ifelse(as.numeric(effect) > 1, 'enriched', 'depleted')
    )), size = 2, alpha = 0.7) +
    geom_text_repel(aes(
        label = ifelse(q_value < 0.05, gsub('`', '', ct_to), ''),
        color = ifelse(as.numeric(effect) > 1, 'enriched', 'depleted')
    ), max.overlaps = 100, force = 4, size = 2.8, fontface = 'bold',
        segment.size = 0.3, segment.color = 'grey60', show.legend = FALSE) +
    facet_wrap(~ panel_label, scales = 'free', ncol = 4) +
    scale_color_manual(values = c(
        'enriched' = '#b2182b', 'depleted' = '#2166ac', 'not significant' = 'grey75'
    )) +
    labs(
        title = 'Neighbor cell-type composition vs. MAPK activation, per patient',
        subtitle = 'Panel: focal cell type (OPC/Astrocyte) · driver alteration  ·  dotted line: q = 0.05',
        x = 'Effect size (log coefficient)', y = expression(-log[10](p)), color = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
        axis.text = element_text(size = 8.5, color = 'grey40'),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = 'grey92', linewidth = 0.4),
        panel.background = element_rect(fill = 'grey98', color = NA),
        strip.text = element_text(face = 'bold', size = 10.5, color = 'grey15', lineheight = 1.15),
        strip.background = element_blank(),
        panel.spacing = unit(0.8, 'lines'),
        plot.title = element_text(face = 'bold', size = 16, margin = margin(b = 3)),
        plot.subtitle = element_text(color = 'grey40', size = 11, margin = margin(b = 10)),
        legend.position = 'bottom',
        legend.text = element_text(size = 10)
    )
```

# Myeloid glial-exposure analysis

How does being near MAPK-high glial cells (OPC/Astrocyte) alter myeloid cell gene programs? We answer this in two parts using the k-NN graph already built above:

- **Part A** — binary adjacency: does having *any* glial neighbor associate with a myeloid signature score? (all myeloid cells)
- **Part B** — continuous MAPK dose: among myeloid cells with ≥2 glial neighbors, does the *mean MAPK level* of those neighbors track the myeloid signature score? (weighted least squares, `weights = n_glial_neighbors`, to down-weight cells with noisy exposure estimates)

Signature scores come from `myeloid_signature_scores.csv` (56 signatures: 50 MSigDB Hallmark + 6 myeloid/immune literature sets; generated by `build_myeloid_signatures.py` + `1_PA_MAPK_scores.qmd`). Scores are z-scored within each patient before fitting so effect sizes are comparable across signatures.

## Build glial exposure table

```{r}
glial_cell_types_target = c('Astrocyte', 'OPC')

# MAPK scores of spatial cells, for looking up glial neighbors' activation levels
glial_mapk_lookup = mapk_scores[type == 'spatial', .(cell_id, glial_mapk = MAPK_overlap_score)]

# All unique myeloid cells present in the PA spatial kNN graph
all_myeloid_cells = unique(
    adjacencies[cell_type_test_from == 'Myeloid', .(cell_id = from_id, sample_id)]
)

# Myeloid→glial directed edges enriched with the glial neighbor's MAPK score
myeloid_glial_edges = adjacencies[
    cell_type_test_from == 'Myeloid' & cell_type_test_to %in% glial_cell_types_target
][
    glial_mapk_lookup, on = c('to_id' = 'cell_id'), nomatch = NULL
]

# Aggregate per myeloid cell: count glial neighbors, average their MAPK
glial_exposure_agg = myeloid_glial_edges[
    , .(n_glial_neighbors = .N, mean_glial_mapk = mean(glial_mapk))
    , .(cell_id = from_id, sample_id)
]

# Full join to keep myeloid cells with zero glial neighbors (important for Part A)
myeloid_glial_exposure = merge(
    all_myeloid_cells, glial_exposure_agg,
    by = c('cell_id', 'sample_id'), all.x = TRUE
)
myeloid_glial_exposure[
    is.na(n_glial_neighbors),
    `:=`(n_glial_neighbors = 0L, mean_glial_mapk = NA_real_)
][
    , has_glial_neighbor := n_glial_neighbors >= 1
]

cat(sprintf(
    'Myeloid cells total: %d | with ≥1 glial neighbor: %d (%.1f%%) | with ≥2: %d (%.1f%%)\n',
    nrow(myeloid_glial_exposure),
    sum(myeloid_glial_exposure$has_glial_neighbor),
    100 * mean(myeloid_glial_exposure$has_glial_neighbor),
    sum(myeloid_glial_exposure$n_glial_neighbors >= 2),
    100 * mean(myeloid_glial_exposure$n_glial_neighbors >= 2)
))

fwrite(
    myeloid_glial_exposure,
    paste0(workdir, 'youyun/plgg/code/sc_integration/ENVI/2_spatial_imputation/PA/myeloid_glial_exposure.tsv'),
    sep = '\t'
)
```

## Load myeloid signature scores and join

```{r}
# myeloid_signature_scores.csv: 45k myeloid cells × 56 score columns + cell_id + sample_id.
# Generated by 1_PA_MAPK_scores.qmd (sc.tl.score_genes on spatial ENVI imputation).
sig_scores = fread(paste0(
    workdir, 'youyun/plgg/code/sc_integration/ENVI/2_spatial_imputation/PA/myeloid_signature_scores.csv'
))
score_cols = grep('_score$', names(sig_scores), value = TRUE)

# Join with glial exposure; warn if match rate is low
myeloid_data = merge(
    myeloid_glial_exposure,
    sig_scores[, c('cell_id', 'sample_id', score_cols), with = FALSE],
    by = c('cell_id', 'sample_id')
)
cat(sprintf(
    'Joined %d / %d myeloid cells to signature scores (%.1f%% match).\n',
    nrow(myeloid_data), nrow(myeloid_glial_exposure),
    100 * nrow(myeloid_data) / nrow(myeloid_glial_exposure)
))

# Z-score each signature within each patient before fitting.
# sc.tl.score_genes baselines are per-sample, so raw magnitudes are not
# cross-patient comparable (same rationale as the MAPK significance section above).
myeloid_data[
    , (score_cols) := lapply(.SD, function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))
    , .SDcols = score_cols, by = sample_id
]
```

## Part A: binary glial adjacency → myeloid signature scores

```{r}
#| results: hold
fit_lm_row = function(dt, sig_col, predictor_col, weight_col = NULL) {
    fml = as.formula(paste0('`', sig_col, '` ~ `', predictor_col, '`'))
    wts = if (!is.null(weight_col)) dt[[weight_col]] else NULL
    m = tryCatch(lm(fml, data = dt, weights = wts), error = function(e) NULL)
    if (is.null(m)) return(NULL)
    coefs = as.data.frame(summary(m)$coefficients)
    coefs$term = rownames(coefs)
    row = as.data.table(coefs)[term == paste0('`', predictor_col, '`')]
    if (nrow(row) == 0) return(NULL)
    data.table(
        signature   = sig_col,
        Estimate    = row$Estimate,
        SE          = row[['Std. Error']],
        t_value     = row[['t value']],
        p_value     = row[['Pr(>|t|)']],
        r_squared   = summary(m)$r.squared,
        n_cells     = nrow(dt)
    )
}

combos_a = CJ(sample_id = unique(myeloid_data$sample_id), signature = score_cols)

parta_results = rbindlist(mapply(
    function(sid, sig) {
        dt = myeloid_data[sample_id == sid, .SD, .SDcols = c(sig, 'has_glial_neighbor')]
        dt = dt[complete.cases(dt)]
        # require ≥30 cells and both glial/non-glial groups represented
        if (nrow(dt) < 30 || length(unique(dt$has_glial_neighbor)) < 2) return(NULL)
        res = fit_lm_row(dt, sig, 'has_glial_neighbor')
        if (!is.null(res)) res[, sample_id := sid]
        res
    },
    combos_a$sample_id, combos_a$signature, SIMPLIFY = FALSE
))

parta_results[, q_value := p.adjust(p_value, method = 'BH'), by = sample_id]
parta_results = merge(parta_results, unique(mapk_scores[, .(sample_id, alteration)]), by = 'sample_id')
parta_results[, alteration := factor(alteration, levels = alteration_levels)]

fwrite(
    parta_results,
    paste0(workdir, 'youyun/plgg/code/sc_integration/ENVI/2_spatial_imputation/PA/myeloid_glial_exposure_parta_results.tsv'),
    sep = '\t'
)
cat('Part A: significant (q<0.05) signature-patient hits:', sum(parta_results$q_value < 0.05, na.rm = TRUE), '\n')
```

## Part B: continuous glial MAPK dose → myeloid signature scores

```{r}
#| results: hold
# Restrict to myeloid cells with ≥2 glial neighbors; weight by n_glial_neighbors
# so cells with more neighbors (lower-noise mapk exposure estimates) pull harder.
partb_base = myeloid_data[n_glial_neighbors >= 2]

combos_b = CJ(sample_id = unique(partb_base$sample_id), signature = score_cols)

partb_results = rbindlist(mapply(
    function(sid, sig) {
        dt = partb_base[sample_id == sid, .SD,
                        .SDcols = c(sig, 'mean_glial_mapk', 'n_glial_neighbors')]
        dt = dt[complete.cases(dt)]
        if (nrow(dt) < 30) return(NULL)
        res = fit_lm_row(dt, sig, 'mean_glial_mapk', weight_col = 'n_glial_neighbors')
        if (!is.null(res)) res[, sample_id := sid]
        res
    },
    combos_b$sample_id, combos_b$signature, SIMPLIFY = FALSE
))

partb_results[, q_value := p.adjust(p_value, method = 'BH'), by = sample_id]
partb_results = merge(partb_results, unique(mapk_scores[, .(sample_id, alteration)]), by = 'sample_id')
partb_results[, alteration := factor(alteration, levels = alteration_levels)]

fwrite(
    partb_results,
    paste0(workdir, 'youyun/plgg/code/sc_integration/ENVI/2_spatial_imputation/PA/myeloid_glial_exposure_partb_results.tsv'),
    sep = '\t'
)
cat('Part B: significant (q<0.05) signature-patient hits:', sum(partb_results$q_value < 0.05, na.rm = TRUE), '\n')
```

## Heatmap: glial MAPK dose effect on myeloid signatures (Part B)

Signatures are sorted by mean |estimate| across patients (most-affected at top). Bold labels = myeloid/immune-specific signatures; plain = generic Hallmark (cell cycle, metabolism, etc.) included as specificity controls.

```{r}
#| fig-width: 9
#| fig-height: 14
# Signatures considered myeloid/immune-specific (bold in plot)
immune_sigs = c(
    'HALLMARK_INFLAMMATORY_RESPONSE_score',
    'HALLMARK_INTERFERON_GAMMA_RESPONSE_score',
    'HALLMARK_INTERFERON_ALPHA_RESPONSE_score',
    'HALLMARK_TNFA_SIGNALING_VIA_NFKB_score',
    'HALLMARK_COMPLEMENT_score',
    'HALLMARK_IL6_JAK_STAT3_SIGNALING_score',
    'HALLMARK_IL2_STAT5_SIGNALING_score',
    'HALLMARK_ALLOGRAFT_REJECTION_score',
    'MG_Act_Tripathi_score',
    'PA_TIM3pos_Microglia_score',
    'PA_TIM3neg_Microglia_score',
    'Bernstein_Complement_Immunosuppressive_score',
    'Bernstein_Scavenger_Immunosuppressive_score',
    'FOSL2_Regulon_Small_score'
)

plot_exposure_heatmap = function(results_dt, title) {
    dt = copy(results_dt)
    dt[, sig_label := sub('_score$', '', signature)]
    row_order = dt[
        , .(mean_abs = mean(abs(Estimate), na.rm = TRUE)), sig_label
    ][order(-mean_abs)]$sig_label
    dt[, sig_label := factor(sig_label, levels = row_order)]
    # only fill cells significant at q < 0.05; grey otherwise (matching
    # the patient-significance heatmap convention above)
    dt[, fill_val := ifelse(q_value < 0.05, Estimate, NA_real_)]

    is_immune = sub('_score$', '', immune_sigs)
    face_vec = ifelse(levels(dt$sig_label) %in% is_immune, 'bold', 'plain')

    ggplot(dt, aes(x = alteration, y = sig_label, fill = fill_val)) +
        geom_tile(color = 'white', linewidth = 0.6) +
        scale_fill_gradient2(
            low = '#2166ac', mid = '#f7f7f7', high = '#b2182b', midpoint = 0,
            na.value = 'grey92', limits = c(-1, 1),
            name = 'Standardized\nbeta (q<0.05)'
        ) +
        scale_x_discrete(expand = expansion(add = 0.6)) +
        scale_y_discrete(expand = expansion(add = 0.6)) +
        labs(
            title = title,
            subtitle = 'Fill: β from z-scored signature ~ mean glial MAPK (q<0.05 BH per patient); grey = n.s.\nBold rows = myeloid/immune-specific signatures; plain = generic Hallmark controls'
        ) +
        theme_minimal(base_size = 12) +
        theme(
            axis.title = element_blank(),
            axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
            axis.text.y = element_text(size = 8, face = face_vec),
            panel.grid = element_blank(),
            plot.title = element_text(face = 'bold', size = 14, margin = margin(b = 4)),
            plot.subtitle = element_text(size = 9, color = 'grey40', margin = margin(b = 8)),
            legend.title = element_text(size = 9),
            legend.key.height = unit(1.4, 'lines')
        )
}

plot_exposure_heatmap(
    partb_results,
    'Effect of glial MAPK exposure on myeloid gene program scores (Part B)'
)
```

```{r}
#| fig-width: 9
#| fig-height: 14
# Part A for comparison: binary adjacency effect (same layout, different predictor)
plot_exposure_heatmap(
    parta_results,
    'Effect of glial adjacency (binary) on myeloid gene program scores (Part A)'
)
```