---
title: "Myeloid & MAPK programs across PA glial niches (Q1 + Q2)"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk:
        results: hold
        R.options:
            width: 120
---

# Introduction

This notebook works on the finalized (res_ct = 0.2 / res_ni = 0.5) pathology-annotated
BANKSY object, restricted to the 4 pilocytic astrocytoma (PA) samples. It covers:

- **Q1 — how does MAPK differ between the two PA glial niches?** Within astrocytes and
  OPCs (the PA tumor lineage) we compare MAPK across three groups — **niche 3 vs niche 5
  vs everything else** — because the two PA glial niches behave oppositely, so lumping
  them (or a single PA-niche-vs-rest contrast) masks the difference. (Testing niche-level
  *mean* MAPK would be circular: glia are MAPK-high regardless of niche.)
- **MAPK vs. neighbor composition (restored)** — as a focal astrocyte/OPC becomes more
  MAPK-active, does the cell-type make-up of its local neighborhood change? The original
  pooled weighted binomial GLM `neighbor_prop ~ MAPK_overlap_score + sample_id` (volcano),
  plus a per-patient panel showing how consistent each shift is across the four drivers.
- **Q2 — do myeloid gene programs shift with the glial neighborhood?** For each
  myeloid cell we classify its glial neighborhood into
  {near PA-niche glia / near other glia / no glial neighbor} and compare 56
  myeloid programs across the three, then ask whether PA-niche proximity adds
  anything beyond the glial cells' own MAPK dose. *(Model design under revision.)*

All cell-type labels are shown through the finalized display vocabulary
(`helpers/cell_type_display.R`): spatial snake_case via `lineage_display` and
single-nucleus labels via `sn_lineage_display`, so the spatial and single-nucleus facets
of the patient x cell-type heatmap share columns.

It is **object-free**: every cell annotation (cell type, niche, sample, histology,
alteration, x/y centroid) is read from the Phase-0 metadata TSV
(`export_new_object_cell_metadata.R`), and the imputed scores from
`MAPK_overlap_scores.csv` / `myeloid_signature_scores.csv` (Phase 1,
`1_PA_MAPK_scores.qmd`). Nothing here re-loads the 5 GB BANKSY object.

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)
library(dbscan)
library(ggrepel)
library(scales)
library(cowplot)
library(rcartocolor)
library(pals)       # polychrome() niche palette, keyed like celltype_niche_histology_enrichment.qmd

# workdir resolves per-machine from HOME (argos = /mnt/storage/dept/medonc/beroukhim/)
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (Sys.getenv("HOME") == "/PHShome/yz762") {
    workdir <- "/data/beroukhim1/"
} else if (Sys.getenv("HOME") == "/home/yz762") {
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    workdir <- "/xchip/beroukhimlab/"
}

# absolute path to this analysis dir (all output TSVs land here)
out_dir = paste0(workdir, 'youyun/plgg/code/sc_integration/ENVI/2_spatial_imputation/PA/')

# Finalized display-name lookups (spatial snake_case + single-nucleus labels -> the same
# plotting vocabulary) so every figure shows finalized names AND the spatial vs
# single-nucleus facets of the patient x cell-type heatmap share columns. Lightweight
# source: cell_type_display.R has no library() calls. Provides lineage_display (spatial)
# and sn_lineage_display (single-nucleus, force-aligned Lymphoid->T cells, Stromal->Endothelial).
source(paste0(workdir, 'youyun/plgg/code/helpers/cell_type_display.R'))
# disp(): spatial snake_case key -> display name; unknown keys (e.g. the sparsity-collapse
# 'Other') pass through unchanged. disp_sn(): single-nucleus label -> same display vocabulary.
disp    = function(x) ifelse(x %in% names(lineage_display),    unname(lineage_display[x]),    x)
disp_sn = function(x) ifelse(x %in% names(sn_lineage_display), unname(sn_lineage_display[x]), x)

# Shared figure theme, matching the polished per-patient neighbor-composition volcano:
# light grey panels, faint major grid, bold blank-background facet strips. Applied to both
# the volcano and the Q1 niche plots so the whole notebook reads consistently.
ref_theme = theme_minimal(base_size = 13) +
    theme(axis.text = element_text(size = 9, color = 'grey40'),
          panel.grid.minor = element_blank(),
          panel.grid.major = element_line(color = 'grey92', linewidth = 0.4),
          panel.background = element_rect(fill = 'grey98', color = NA),
          strip.text = element_text(face = 'bold', size = 11, color = 'grey15', lineheight = 1.15),
          strip.background = element_blank(),
          panel.spacing = unit(0.8, 'lines'),
          plot.title = element_text(face = 'bold', size = 15, margin = margin(b = 3)),
          plot.subtitle = element_text(color = 'grey40', size = 11, margin = margin(b = 10)),
          legend.position = 'bottom', legend.text = element_text(size = 10))
```

# Load metadata + imputed scores

```{r}
pa_ids = c('188570', '258482', '320784', '328368')   # the 4 PA samples; all analysis restricted here
glial_types = c('astro_glial', 'opc', 'radial_glia')  # glial lineages, finalized lowercase labels

# Phase-0 metadata TSV: single source of truth for cell_type / niche / sample /
# histology / alteration / spatial centroids, all read straight from the finalized
# object's own colData (no external cluster-annotation join). See
# export_new_object_cell_metadata.R and memory reference_new_banksy_object_columns.
cell_meta = fread(
    paste0(workdir, 'youyun/plgg/data/xenium_cell_types/cell_metadata_resct0.2_resni0.5_20260826.tsv'),
    colClasses = list(character = c('sample_id', 'niche', 'ct_cluster'))
)
cell_meta_pa = cell_meta[sample_id %in% pa_ids]        # every downstream table derives from this
cat('Cohort cells:', nrow(cell_meta), '| PA cells:', nrow(cell_meta_pa), '\n')

# Niche palette: identical construction to celltype_niche_histology_enrichment.qmd
# (pals::polychrome() keyed to cohort niches, numerically sorted) so niche colors match
# that figure. Used for the Q1 niche plots below.
niche_levels_all = as.character(sort(as.integer(unique(cell_meta$niche))))
niche_color_manual = setNames(pals::polychrome()[seq_along(niche_levels_all)], niche_levels_all)
cat('PA cell-type counts:\n'); print(cell_meta_pa[, .N, cell_type][order(-N)])
```

## MAPK score

```{r}
# MAPK_overlap_scores.csv: spatial (ENVI-imputed) + single-nucleus rows, each with
# cell_id, MAPK_overlap_score, cell_type (spatial = new lowercase labels; sn =
# annot_v0). Regenerated in Phase 1 against the new object's own labels.
mapk_scores = fread(paste0(out_dir, 'MAPK_overlap_scores.csv'))

# sample id + driver alteration lookup. `sample` is formatted "188570 pilocytic
# astrocytoma"; the id is always the leading digits.
mapk_scores[, sample_id := gsub('^([0-9]+).*', '\\1', sample)]
mapk_scores[, alteration := fcase(
    sample_id == '188570', 'KIAA1549::BRAF',
    sample_id == '258482', 'FGFR1 ITD',
    sample_id == '320784', 'FAM131B::BRAF',
    sample_id == '328368', 'RAF1::TOP2B'
)]
# fixed sample/alteration display order (numeric sample id order), reused by heatmaps below
alteration_levels = c('KIAA1549::BRAF', 'FGFR1 ITD', 'FAM131B::BRAF', 'RAF1::TOP2B')

# fread reads a blank CSV field as "" not NA, so normalize before is.na() filtering
mapk_scores[cell_type == '', cell_type := NA]
```

# Which cell types are significantly MAPK-high (or -low), within each patient

`sc.tl.score_genes` fits its control-gene baseline independently per sample, so
raw score magnitude is not comparable *across* patients. We test each cell type
against the *other* cells in the *same* patient/modality (two-sided Mann-Whitney),
which relies only on that patient's internal ranking, not its absolute scale.

```{r}
MIN_N_GROUP = 20   # drop (patient, cell type) groups too small to test reliably

patient_sig_combos = unique(mapk_scores[!is.na(cell_type), .(type, sample_id, cell_type)])

patient_sig_results = rbindlist(mapply(function(ty, sid, ct) {
    in_group  = mapk_scores[type == ty & sample_id == sid & cell_type == ct, MAPK_overlap_score]
    out_group = mapk_scores[type == ty & sample_id == sid & cell_type != ct, MAPK_overlap_score]
    if (length(in_group) < MIN_N_GROUP || length(out_group) < MIN_N_GROUP) return(NULL)

    # normal approximation (exact null infeasible at ~1e5 cells); two-sided so we
    # also flag significantly MAPK-LOW cell types
    test = wilcox.test(in_group, out_group, alternative = 'two.sided', exact = FALSE)
    n1 = length(in_group); n2 = length(out_group)
    # rank-biserial effect size: -1 = in_group always lower, 0 = no diff, +1 = always higher.
    # as.numeric() guards against 32-bit int overflow of n1*n2 at large group sizes.
    effect = 2 * unname(test$statistic) / (as.numeric(n1) * as.numeric(n2)) - 1

    data.table(
        type = ty, sample_id = sid, cell_type = ct,
        n = n1, median = median(in_group), median_rest = median(out_group),
        p = test$p.value, effect = effect
    )
}, patient_sig_combos$type, patient_sig_combos$sample_id, patient_sig_combos$cell_type, SIMPLIFY = FALSE))

# BH within each patient x modality family (the natural per-patient discovery list)
patient_sig_results[, q := p.adjust(p, method = 'BH'), .(type, sample_id)]
patient_sig_results = merge(patient_sig_results, unique(mapk_scores[, .(sample_id, alteration)]), by = 'sample_id')
patient_sig_results[, alteration := factor(alteration, levels = alteration_levels)]

# Reconcile the two modality vocabularies to a single display vocabulary so the same
# biological cell type lands in the same heatmap column across the spatial and
# single-nucleus facets: spatial (lowercase snake_case) via disp(), single-nucleus
# (annot_v0) via disp_sn() (force-aligned Lymphoid->T cells, Stromal->Endothelial).
patient_sig_results[, cell_type_disp := fifelse(type == 'spatial', disp(cell_type), disp_sn(cell_type))]

fwrite(patient_sig_results, paste0(out_dir, 'mapk_patient_cell_type_significance.tsv'), sep = '\t')
```

## Heatmap: patient x cell type

```{r}
# Patient (row) x cell type (column) heatmap, faceted by modality (spatial vs sn).
# Fill = rank-biserial effect for cell types significant at q<0.05 (blue = MAPK-low,
# red = MAPK-high); grey = tested but n.s. Columns ranked by mean effect.
plot_patient_sig_heatmap = function(
    dt, subtitle = '', title_size = 15, title_wrap = NULL, legend_position = 'right',
    legend_breaks = waiver(), title_override = NULL
) {
    dt_sub = copy(dt)
    # rank/plot on the reconciled display label so spatial & sn facets share columns
    col_rank = dt_sub[, .(n_sig = sum(q < 0.05), mean_effect = mean(effect)), cell_type_disp][order(-mean_effect, -n_sig)]
    dt_sub[, cell_type_disp := factor(cell_type_disp, levels = col_rank$cell_type_disp)]
    dt_sub[, fill_effect := ifelse(q < 0.05, effect, NA_real_)]
    dt_sub[, modality := factor(
        fifelse(type == 'spatial', 'Spatial (ENVI-imputed)', 'Single-nucleus (observed)'),
        levels = c('Spatial (ENVI-imputed)', 'Single-nucleus (observed)')
    )]

    title_text = if (!is.null(title_override)) title_override else paste0('MAPK-high/-low cell types by patient', subtitle)
    if (!is.null(title_wrap)) title_text = paste(strwrap(title_text, width = title_wrap), collapse = '\n')

    ggplot(dt_sub, aes(x = cell_type_disp, y = alteration, fill = fill_effect)) +
        geom_tile(color = 'white', linewidth = 1.1) +
        facet_wrap(~ modality, ncol = 1) +
        scale_fill_gradient2(
            low = '#2166ac', mid = '#f7f7f7', high = '#b2182b', midpoint = 0,
            na.value = 'grey92', limits = c(-1, 1), breaks = legend_breaks, name = 'Effect size\n(q < 0.05)'
        ) +
        scale_x_discrete(expand = expansion(add = 0.6)) +
        scale_y_discrete(expand = expansion(add = 0.6)) +
        labs(title = title_text) +
        theme_minimal(base_size = 14) +
        theme(
            axis.title = element_blank(),
            axis.text.x = element_text(angle = 45, hjust = 1, size = 11, color = 'grey20'),
            axis.text.y = element_text(size = 11.5, face = 'bold', color = 'grey20'),
            panel.grid = element_blank(),
            strip.text = element_text(face = 'bold', size = 12.5, hjust = 0, color = 'grey15'),
            strip.background = element_blank(),
            panel.spacing.y = unit(1.6, 'lines'),
            plot.title = element_text(face = 'bold', size = title_size, margin = margin(b = 10)),
            legend.title = element_text(size = 10.5),
            legend.key.height = unit(1.4, 'lines'),
            legend.position = legend_position
        )
}
```

```{r}
#| fig-width: 8
#| fig-height: 7
plot_patient_sig_heatmap(patient_sig_results)
```

```{r}
#| fig-width: 7.5
#| fig-height: 6
# Zoomed to the OPC/astrocyte lineage -- the tumor-relevant populations in PA
# (an OPC/astrocytic-lineage tumor). ignore.case + the underscore labels so this
# matches BOTH spatial (astro_glial/opc) and sn (Astrocyte/OPC). radial_glia is
# excluded here (user request). "undetermined" is a genuine spatial cell type here
# (not an sn-only artifact), kept in this "poorly-typed tumor-adjacent" grouping.
tumor_lineage_regex = 'astro|opc|undetermined'
plot_patient_sig_heatmap(
    patient_sig_results[grepl(tumor_lineage_regex, cell_type, ignore.case = TRUE)],
    ' -- OPC/astrocyte lineage', title_wrap = 46
)
```

```{r}
#| fig-width: 4.2
#| fig-height: 6
# Same lineage zoom, spatial only, narrower. radial_glia excluded (user request).
plot_patient_sig_heatmap(
    patient_sig_results[type == 'spatial' & grepl('astro|opc', cell_type, ignore.case = TRUE)],
    title_size = 13, legend_position = 'bottom', legend_breaks = c(-1, 0, 1),
    title_override = 'OPC/astrocyte -- spatial'
)
```

# Derive PA glial niches (programmatic, not hardcoded)

A niche qualifies as a **PA glial niche** if it is BOTH:

- **(a) PA-enriched at q < 0.05** in `niche_histology_enrichment.tsv` (glmmTMB
  log-odds vs cohort baseline). We use the odds-ratio enrichment, not raw PA
  fraction, so a niche that is glial simply because a PA sample contributes many
  of its cells does not spuriously qualify.
- **(b) glial-rich**: its PA glial fraction (astro_glial + opc + radial_glia)
  exceeds the whole-cohort glial baseline.

```{r}
cohort_glial_frac = cell_meta[, mean(cell_type %in% glial_types)]   # baseline across all 11 samples

# (b) per-niche PA glial composition
niche_comp = cell_meta_pa[, .(
    n_cells    = .N,
    glial_frac = mean(cell_type %in% glial_types),
    opc_frac   = mean(cell_type == 'opc'),
    astro_frac = mean(cell_type == 'astro_glial'),
    myeloid_frac = mean(cell_type == 'myeloid')
), niche][order(-glial_frac)]

# (a) PA enrichment from the existing niche-histology model
enr = fread(paste0(workdir, 'youyun/plgg/code/niche/proseg_output_analysis/niche_histology_enrichment.tsv'))
pa_enr = enr[histology == 'pilocytic astrocytoma',
             .(niche = as.character(niche_type), logOR = estimate_log_odds,
               q = q_value, sign_call, model_status)]

enriched_niches   = pa_enr[sign_call == 'enriched' & q < 0.05 & model_status == 'pass', niche]
glial_rich_niches = niche_comp[glial_frac > cohort_glial_frac, niche]
pa_glial_niches   = intersect(enriched_niches, glial_rich_niches)   # the derived set, used everywhere below

cat('Cohort glial baseline fraction:', round(cohort_glial_frac, 3), '\n')
cat('PA-enriched niches (q<0.05):', paste(sort(as.integer(enriched_niches)), collapse = ', '), '\n')
cat('==> PA GLIAL NICHES (enriched AND glial-rich):', paste(sort(as.integer(pa_glial_niches)), collapse = ', '), '\n')

# report the qualifying niches with their enrichment + composition
kable(
    merge(niche_comp, pa_enr, by = 'niche')[niche %in% pa_glial_niches][order(-logOR)],
    digits = 3, caption = 'Derived PA glial niches: enrichment (logOR, q) + PA composition'
)
```

# Q1 -- MAPK across PA glial niches: niche 3 vs niche 5 vs everything else

The two PA glial niches (3 and 5) behave differently, so rather than lumping them we
compare MAPK *within* astrocytes and OPCs across three mutually exclusive groups:
**niche 3**, **niche 5**, and **everything else** (all niches other than 3 and 5 = the
non-PA-glial background). MAPK is z-scored per patient before fitting (scores are
per-sample-baselined, so raw magnitudes are not cross-patient comparable). Niche
membership comes from BANKSY spatial clustering while MAPK comes from ENVI imputation --
different matrices, so this contrast is not mechanically circular.

**Interpretation frame:** MAPK-high glia in a PA-enriched niche is *consistent with*
tumor identity, not proof (inferCNV unconfirmed; the panel is shared upstream input to
both BANKSY and ENVI). We restrict to astrocytes and OPCs (the PA tumor lineage); with
n = 4 PA samples the readout is effect size + directional consistency across patients,
not cohort-level significance.

```{r}
focal_glia = c('astro_glial', 'opc')   # user: only astrocytes and OPCs (drop sparse radial_glia)

# focal glia (PA) joined to spatial MAPK; z-score MAPK per patient
q1_dt = merge(
    cell_meta_pa[cell_type %in% focal_glia, .(cell_id, sample_id, cell_type, niche)],
    mapk_scores[type == 'spatial', .(cell_id, MAPK_overlap_score)],
    by = 'cell_id'
)
q1_dt[, mapk_z := (MAPK_overlap_score - mean(MAPK_overlap_score)) / sd(MAPK_overlap_score), by = sample_id]
# driver alteration on each focal cell (x-axis + per-driver brackets in the niche plots below)
q1_dt = merge(q1_dt, unique(mapk_scores[, .(sample_id, alteration)]), by = 'sample_id')
q1_dt[, alteration := factor(alteration, levels = alteration_levels)]

# three mutually exclusive groups. 'background' (all niches other than 3 and 5) is the
# reference level, so each niche's lm coefficient is that niche's MAPK shift vs the
# non-PA-glial background, and a relevel gives the direct niche-5-vs-niche-3 contrast.
q1_dt[, niche_group := fcase(niche == '3', 'niche3', niche == '5', 'niche5', default = 'background')]
q1_dt[, niche_group := factor(niche_group, levels = c('background', 'niche3', 'niche5'))]

MIN_N_Q1 = 20   # need >=20 focal cells in EACH of the three groups to fit the contrast
q1_combos = unique(q1_dt[, .(sample_id, cell_type)])

# per (focal cell type, sample): lm(mapk_z ~ niche_group). Pull both niche-vs-background
# coefficients, plus a refit with niche3 as reference for the niche-5-vs-niche-3 contrast.
q1_results = rbindlist(mapply(function(sid, ct) {
    dt = q1_dt[sample_id == sid & cell_type == ct]
    n_bg = dt[niche_group == 'background', .N]
    n_n3 = dt[niche_group == 'niche3', .N]
    n_n5 = dt[niche_group == 'niche5', .N]
    if (n_bg < MIN_N_Q1 || n_n3 < MIN_N_Q1 || n_n5 < MIN_N_Q1) return(NULL)   # all 3 groups required
    m   = lm(mapk_z ~ niche_group, data = dt)                             # ref = background
    m35 = lm(mapk_z ~ relevel(niche_group, ref = 'niche3'), data = dt)    # ref = niche3 (n5 vs n3)
    co   = as.data.table(summary(m)$coefficients,   keep.rownames = 'term')
    co35 = as.data.table(summary(m35)$coefficients, keep.rownames = 'term')
    pick = function(tab, key) tab[grepl(key, term, fixed = TRUE)][1]       # single matching coefficient row
    r3 = pick(co, 'niche3'); r5 = pick(co, 'niche5'); r53 = pick(co35, 'niche5')
    data.table(
        sample_id = sid, cell_type = ct,
        n_background = n_bg, n_niche3 = n_n3, n_niche5 = n_n5,
        beta_niche3 = r3$Estimate, se_niche3 = r3[['Std. Error']], p_niche3 = r3[['Pr(>|t|)']],
        beta_niche5 = r5$Estimate, se_niche5 = r5[['Std. Error']], p_niche5 = r5[['Pr(>|t|)']],
        beta_n5_vs_n3 = r53$Estimate, p_n5_vs_n3 = r53[['Pr(>|t|)']]
    )
}, q1_combos$sample_id, q1_combos$cell_type, SIMPLIFY = FALSE))

q1_results = merge(q1_results, unique(mapk_scores[, .(sample_id, alteration)]), by = 'sample_id')
q1_results[, alteration := factor(alteration, levels = alteration_levels)]
q1_results[, cell_type_disp := disp(cell_type)]
fwrite(q1_results, paste0(out_dir, 'mapk_q1_niche_three_group.tsv'), sep = '\t')

# long form of the two niche-vs-background effects, for BH, the summary, and plotting
q1_long = melt(
    q1_results,
    id.vars = c('sample_id', 'cell_type', 'cell_type_disp', 'alteration'),
    measure.vars = list(beta = c('beta_niche3', 'beta_niche5'),
                        se   = c('se_niche3',   'se_niche5'),
                        p    = c('p_niche3',    'p_niche5')),
    variable.name = 'niche_idx'
)
q1_long[, niche := fifelse(as.integer(niche_idx) == 1L, 'niche 3', 'niche 5')]   # melt index 1->niche3, 2->niche5
q1_long[, q := p.adjust(p, method = 'BH'), by = sample_id]   # per-sample BH over both contrasts

# across-patient readout (n=4): directional consistency of each niche-vs-background effect
q1_consistency = q1_long[!is.na(beta), .(
    n_samples = .N,
    n_pos     = sum(beta > 0),
    n_neg     = sum(beta < 0),
    mean_beta = mean(beta)
), by = .(cell_type_disp, niche)]
q1_consistency[, consistent := n_samples > 0 & (n_pos == n_samples | n_neg == n_samples)]
cat('Q1 across-patient consistency (beta = niche MAPK-z vs non-PA-glial background):\n')
print(q1_consistency[order(cell_type_disp, niche)])
kable(q1_results[order(cell_type_disp, alteration),
      .(cell_type_disp, alteration, n_niche3, n_niche5, n_background,
        beta_niche3, beta_niche5, beta_n5_vs_n3)],
      digits = 3, caption = 'Q1: MAPK-z per group vs background, and niche-5-vs-niche-3, per patient')
```

```{r}
# Reusable niche plot (shared by the 3- and 4-group versions).
#   groups_map : named char vector niche_id -> group label; niches not listed fall into
#                'everything else' (so the reference set differs between the 3- and 4-group calls).
# x = driver; within each driver a pointrange per niche (dot = mean, line = +/- SD of the
# within-patient MAPK z-score); brackets over each driver compare each niche vs 'everything
# else' by a per-driver per-cell Wilcoxon on mapk_z (user's chosen test). Niche colors from
# niche_color_manual (polychrome), matching celltype_niche_histology_enrichment.qmd.
q1_niche_plot = function(groups_map, group_levels, title, out_stub) {
    G = length(group_levels); W = 0.75
    d = copy(q1_dt)
    d[, grp := fifelse(as.character(niche) %in% names(groups_map),
                       groups_map[as.character(niche)], 'everything else')]
    d[, grp := factor(grp, levels = group_levels)]
    d[, cell_type_disp := disp(cell_type)]

    # per (cell type, driver, niche) mean +/- SE of within-patient MAPK z
    gm = d[, .(mean_z = mean(mapk_z), se = sd(mapk_z) / sqrt(.N), n = .N), by = .(cell_type_disp, alteration, grp)]

    # per-driver per-cell Wilcoxon: each niche vs 'everything else', within (cell type, driver)
    niches = setdiff(group_levels, 'everything else')
    br = rbindlist(lapply(unique(gm$cell_type_disp), function(ctd)
        rbindlist(lapply(alteration_levels, function(drv) {
            y = d[cell_type_disp == ctd & alteration == drv & grp == 'everything else', mapk_z]
            rbindlist(lapply(niches, function(g) {
                x = d[cell_type_disp == ctd & alteration == drv & grp == g, mapk_z]
                if (length(x) < 10 || length(y) < 10) return(NULL)   # too few cells to test
                data.table(cell_type_disp = ctd, alteration = drv, grp = g, p = wilcox.test(x, y)$p.value)
            }))
        }))))
    br[, p.signif := fcase(p < 1e-3, '***', p < 1e-2, '**', p < 5e-2, '*', default = 'ns')]
    br[, alteration := factor(alteration, levels = alteration_levels)]
    # dodge geometry: group g over total width W sits at category center + this offset;
    # used to draw each bracket from its niche's dodged point across to the everything-else point.
    dodge_off = function(g, gg, ww) ww * ((2 * g - 1 - gg) / (2 * gg))
    br[, xc := as.integer(alteration)]; br[, g_idx := match(grp, group_levels)]
    br[, `:=`(x1 = xc + dodge_off(g_idx, G, W), x2 = xc + dodge_off(G, G, W))]
    # bracket heights above the data, staggered by niche; na.rm since a sparse (driver x niche)
    # group can have n=1 -> se=NA, which would otherwise NA out the whole facet's y-range
    yr = gm[, .(hi = max(mean_z + se, na.rm = TRUE), lo = min(mean_z - se, na.rm = TRUE)), by = cell_type_disp]
    yr[, span := hi - lo]; br = merge(br, yr, by = 'cell_type_disp')
    br[, ytick := 0.025 * span]; br[, y := hi + 0.08 * span + (g_idx - 1) * 0.10 * span]

    fwrite(gm, paste0(out_dir, out_stub, '_means.tsv'), sep = '\t')
    fwrite(br, paste0(out_dir, out_stub, '_brackets.tsv'), sep = '\t')

    # niche group colors from the cohort polychrome palette (+ grey background)
    q1_pal = c(niche_color_manual[c('3', '5', '7')], 'grey70')
    names(q1_pal) = c('niche 3', 'niche 5', 'niche 7', 'everything else')
    ggplot(gm, aes(x = alteration, y = mean_z, color = grp)) +
        geom_hline(yintercept = 0, linetype = 'dashed', color = 'grey60', linewidth = 0.5) +
        geom_pointrange(aes(ymin = mean_z - se, ymax = mean_z + se), position = position_dodge(width = W), size = 0.45) +
        # geom_blank forces the free y-scale to include the top brackets
        geom_blank(data = br, aes(x = xc, y = y + 0.04 * span), inherit.aes = FALSE) +
        geom_segment(data = br, aes(x = x1, xend = x2, y = y, yend = y), inherit.aes = FALSE, linewidth = 0.35, color = 'grey35') +
        geom_segment(data = br, aes(x = x1, xend = x1, y = y, yend = y - ytick), inherit.aes = FALSE, linewidth = 0.35, color = 'grey35') +
        geom_segment(data = br, aes(x = x2, xend = x2, y = y, yend = y - ytick), inherit.aes = FALSE, linewidth = 0.35, color = 'grey35') +
        geom_text(data = br, aes(x = (x1 + x2) / 2, y = y, label = p.signif), inherit.aes = FALSE, vjust = -0.1, size = 3, color = 'grey25') +
        facet_wrap(~ cell_type_disp, nrow = 1, scales = 'free_y') +
        scale_color_manual(values = q1_pal, name = NULL, breaks = group_levels) +
        labs(title = title,
             subtitle = 'dot = mean, line = +/- SE of within-patient MAPK z; brackets = per-driver per-cell Wilcoxon vs everything else (*** <.001 ** <.01 * <.05)',
             x = NULL, y = 'MAPK (z-scored within patient)') +
        ref_theme + theme(axis.text.x = element_text(angle = 25, hjust = 1, size = 10, color = 'grey20'))
}
```

```{r}
#| fig-width: 10
#| fig-height: 6
# niche 3 vs niche 5 vs everything else (everything else = niches other than 3 and 5)
q1_niche_plot(
    groups_map  = c('3' = 'niche 3', '5' = 'niche 5'),
    group_levels = c('niche 3', 'niche 5', 'everything else'),
    title = 'MAPK Activity by Niche',
    out_stub = 'mapk_q1_niche_three_group'
)
```

```{r}
#| fig-width: 11
#| fig-height: 6
# niche 3 vs niche 5 vs niche 7 vs everything else (everything else = niches other than 3,5,7).
# Niche 7 is NOT a PA-enriched glial niche -- included as a comparison group (user request).
q1_niche_plot(
    groups_map  = c('3' = 'niche 3', '5' = 'niche 5', '7' = 'niche 7'),
    group_levels = c('niche 3', 'niche 5', 'niche 7', 'everything else'),
    title = 'MAPK Activity by Niche',
    out_stub = 'mapk_q1_niche_four_group'
)
```

# Neighborhood graph (per-sample kNN)

```{r}
k_geom = 15
# Build the kNN graph PER SAMPLE. Xenium coordinates are per-slide-local, so a
# single graph over all PA cells would create cross-sample false neighbors and
# corrupt every myeloid->glial edge. dbscan::kNN$id is n x k (column-major), so
# rep(cell_id, k) aligns row-i's k neighbors with as.vector(knn$id).
build_sample_knn = function(sdt) {
    locs = as.matrix(sdt[, .(x, y)])
    knn = dbscan::kNN(x = locs, k = k_geom)
    data.table(
        from_id   = rep(sdt$cell_id, k_geom),
        to_id     = sdt$cell_id[as.vector(knn$id)],
        distance  = as.vector(knn$dist),
        sample_id = sdt$sample_id[1]
    )
}
knnDT = rbindlist(lapply(split(cell_meta_pa, by = 'sample_id'), build_sample_knn))

# annotate each edge with from/to cell type, straight from the TSV (no cluster->label map)
adjacencies = merge(knnDT, cell_meta_pa[, .(to_id = cell_id, cell_type_to = cell_type)], by = 'to_id')
adjacencies = merge(adjacencies, cell_meta_pa[, .(from_id = cell_id, cell_type_from = cell_type)], by = 'from_id')
# sanity: per-sample construction must produce zero cross-sample edges
stopifnot(adjacencies[gsub('.*__', '', from_id) != gsub('.*__', '', to_id), .N] == 0)
cat('kNN edges:', nrow(adjacencies), '| myeloid source cells:',
    adjacencies[cell_type_from == 'myeloid', uniqueN(from_id)], '\n')
```

# MAPK activation vs. neighbor composition

The original neighbor-composition analysis: as a focal astrocyte/OPC becomes more
MAPK-active, does the cell-type make-up of its local (k = 15) neighborhood change? For
each focal cell we take the proportion of its neighbors of each (collapsed) type and fit
the exact model + volcano we had before -- a weighted binomial GLM
`neighbor_prop ~ MAPK_overlap_score + sample_id` (RAW imputed MAPK; `sample_id` pools the
4 patients), one model per neighbor cell type -- ported object-free (rebuilt from the
`adjacencies` kNN table + `MAPK_overlap_scores.csv`; no BANKSY object).

**Note on inference.** These GLMs are fit *per cell* (tens of thousands), so the p-values
are anticonservative (pseudoreplication, per `feedback_percell_pseudoreplication`); read
the volcano for the *direction and relative size* of neighbor-composition shifts, and the
per-patient panel below for how consistent each shift is across the four drivers.

```{r}
# Object-free rebuild of the original `adjacencies_and_mapk`. Sparsity-collapse cell-type
# labels (<1% of a sample's cells -> 'Other' for that sample) on BOTH the focal (from) and
# neighbor (to) side, exactly as the original did before building the graph.
sparsity_threshold = 0.01
prev = cell_meta_pa[, .(n = .N), .(sample_id, cell_type)]
prev[, frac := n / sum(n), sample_id]
prev[, ct_c := ifelse(frac < sparsity_threshold | is.na(cell_type), 'Other', cell_type)]
collapse_map = prev[, .(sample_id, cell_type, ct_c)]

adj = merge(adjacencies, collapse_map[, .(sample_id, cell_type_from = cell_type, cff = ct_c)],
            by = c('sample_id', 'cell_type_from'))
adj = merge(adj, collapse_map[, .(sample_id, cell_type_to = cell_type, ctt = ct_c)],
            by = c('sample_id', 'cell_type_to'))

# per focal cell: proportion of its k neighbors of each collapsed type (wide), + RAW MAPK
adjacencies_and_mapk = merge(
    dcast(adj[, .(neighbor_prop = .N / k_geom), .(from_id, cff, sample_id, ctt)],
          from_id + cff + sample_id ~ ctt, value.var = 'neighbor_prop', fill = 0),
    mapk_scores[type == 'spatial', .(cell_id, MAPK_overlap_score)],
    by.x = 'from_id', by.y = 'cell_id'
)[, weights := k_geom]
cell_types_in_dt = setdiff(colnames(adjacencies_and_mapk),
                           c('from_id', 'cff', 'sample_id', 'MAPK_overlap_score', 'weights'))

# --- POOLED model (the original): neighbor_prop[ct2] ~ MAPK + sample_id, over ALL pairs ---
pooled_pairs = CJ(ct_from = cell_types_in_dt, ct_to = cell_types_in_dt, unique = TRUE)
mapk_adj_results = rbindlist(mapply(function(ct1, ct2) {
    tmp = adjacencies_and_mapk[cff == ct1]
    if (nrow(tmp) == 0 || all(tmp[[ct2]] == 0)) return(NULL)
    # original guard: a focal type present in only one sample can't take the sample_id
    # covariate (single factor level), so drop it there
    rhs = if (length(unique(tmp$sample_id)) == 1) '`%s` ~ MAPK_overlap_score'
          else '`%s` ~ MAPK_overlap_score + sample_id'
    m = glm(as.formula(sprintf(rhs, ct2)),
            data = tmp, family = binomial(link = 'logit'), weights = weights)
    co = as.data.table(summary(m)$coefficients, keep.rownames = 'variables')[variables == 'MAPK_overlap_score']
    co[, .(ct_from = ct1, ct_to = ct2, Estimate, effect = exp(Estimate),
           p_value = `Pr(>|z|)`, neg_log10_p_value = -log10(`Pr(>|z|)`))]
}, pooled_pairs$ct_from, pooled_pairs$ct_to, SIMPLIFY = FALSE))
mapk_adj_results[, q_value := p.adjust(p_value, method = 'BH')]
mapk_adj_results[, `:=`(ct_from_disp = disp(ct_from), ct_to_disp = disp(ct_to))]
fwrite(mapk_adj_results, paste0(out_dir, 'mapk_adj_results.tsv'), sep = '\t')
```

```{r}
#| fig-width: 12
#| fig-height: 6
# Pooled volcano, reference theme (ref_theme; enriched=red #b2182b, depleted=blue #2166ac,
# n.s.=grey75; bold colored labels for q<0.05; xlim(-1,0.5)); focal astro/OPC, display names.
vdt = mapk_adj_results[ct_from %in% focal_glia]
vdt[, sig := ifelse(q_value >= 0.05, 'not significant', ifelse(as.numeric(effect) > 1, 'enriched', 'depleted'))]
ggplot(vdt, aes(x = Estimate, y = neg_log10_p_value)) +
    geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey40', linewidth = 0.5) +
    geom_point(aes(color = sig), size = 2, alpha = 0.8) +
    geom_text_repel(aes(label = ifelse(q_value < 0.05, ct_to_disp, ''), color = sig),
                    max.overlaps = 100, force = 4, size = 2.9, fontface = 'bold',
                    segment.size = 0.3, segment.color = 'grey60', show.legend = FALSE) +
    scale_color_manual(values = c('enriched' = '#b2182b', 'depleted' = '#2166ac', 'not significant' = 'grey75'), name = NULL) +
    xlim(-1, 0.5) +
    facet_wrap(~ ct_from_disp, nrow = 1) +
    labs(title = 'Neighbor cell-type composition vs. MAPK activation (pooled)',
         subtitle = 'Pooled binomial GLM (neighbor_prop ~ MAPK + sample_id); x>0 = neighbor enriches as focal MAPK rises',
         x = 'Effect size (log coefficient)', y = expression(-log[10](p))) +
    ref_theme
```

## Per-patient neighborhood composition vs. MAPK activation

The pooled volcano fits one model per neighbor type with `sample_id` as a covariate
("on average across patients, does neighbor composition track MAPK"). Here we instead fit
one model per (focal cell type, patient) -- no covariate, a single patient each -- to see
how consistent the relationship is across the four driver alterations. The per-panel
subtitle gives how many focal (OPC/astrocyte) cells backed that fit.

```{r}
# per-patient GLM: neighbor_prop[ct2] ~ MAPK_overlap_score (RAW), per (focal, neighbor, sample)
sids = sort(unique(as.character(adjacencies_and_mapk$sample_id)))
by_patient_combos = CJ(ct_from = focal_glia, ct_to = cell_types_in_dt, sample_id = sids, unique = TRUE)
mapk_adj_results_by_patient = rbindlist(mapply(function(ct1, ct2, sid) {
    tmp = adjacencies_and_mapk[cff == ct1 & sample_id == sid]
    if (nrow(tmp) == 0 || all(tmp[[ct2]] == 0)) return(NULL)
    m = glm(as.formula(sprintf('`%s` ~ MAPK_overlap_score', ct2)),
            data = tmp, family = binomial(link = 'logit'), weights = weights)
    co = as.data.table(summary(m)$coefficients, keep.rownames = 'variables')[variables == 'MAPK_overlap_score']
    # underflow-safe -log10(p): some per-patient panels reach -log10(p) > 800, where Pr
    # floors to 0; compute it in log-space from the z value instead.
    co[, .(ct_from = ct1, ct_to = ct2, sample_id = sid, Estimate, effect = exp(Estimate),
           p_value = `Pr(>|z|)`,
           neg_log10_p_value = -(log(2) + pnorm(-abs(`z value`), log.p = TRUE)) / log(10))]
}, by_patient_combos$ct_from, by_patient_combos$ct_to, by_patient_combos$sample_id, SIMPLIFY = FALSE))
mapk_adj_results_by_patient[, q_value := p.adjust(p_value, method = 'BH'), .(ct_from, sample_id)]
mapk_adj_results_by_patient = merge(mapk_adj_results_by_patient,
                                    unique(mapk_scores[, .(sample_id, alteration)]), by = 'sample_id')
mapk_adj_results_by_patient[, alteration := factor(alteration, levels = alteration_levels)]
mapk_adj_results_by_patient[, `:=`(ct_from_disp = disp(ct_from), ct_to_disp = disp(ct_to))]
fwrite(mapk_adj_results_by_patient, paste0(out_dir, 'mapk_adj_results_by_patient.tsv'), sep = '\t')
```

```{r}
#| fig-width: 13
#| fig-height: 7.5
# per-panel context: focal-cell count + % of the sample, in the strip subtitle so a null
# panel can be read against how much tissue/power backed it (original convention).
sp_composition = mapk_scores[type == 'spatial' & !is.na(cell_type)]
sample_totals = sp_composition[, .(total_n = .N), sample_id]
focal_composition = sp_composition[cell_type %in% focal_glia, .(n = .N), .(sample_id, cell_type)]
focal_composition = merge(focal_composition, sample_totals, by = 'sample_id')
focal_composition[, pct := 100 * n / total_n]
setnames(focal_composition, 'cell_type', 'ct_from')

plot_dt = merge(copy(mapk_adj_results_by_patient),
                focal_composition[, .(sample_id, ct_from, n, pct)], by = c('sample_id', 'ct_from'))
plot_dt[, panel_label := paste0(ct_from_disp, '  ·  ', alteration, '\n',
                                sprintf('n=%s (%.1f%% of sample)', format(n, big.mark = ','), pct))]
panel_levels = unique(plot_dt[order(factor(ct_from, levels = focal_glia), alteration), panel_label])
plot_dt[, panel_label := factor(panel_label, levels = panel_levels)]

# facet_wrap (not grid) for fully independent x AND y per panel -- one outlier panel
# (-log10(p) up to ~800) would otherwise flatten the rest. Reference theme + styled labels.
plot_dt[, sig := ifelse(q_value >= 0.05, 'not significant', ifelse(as.numeric(effect) > 1, 'enriched', 'depleted'))]
ggplot(plot_dt, aes(x = Estimate, y = neg_log10_p_value)) +
    geom_hline(yintercept = -log10(0.05), linetype = 'dotted', color = 'grey55', linewidth = 0.4) +
    geom_vline(xintercept = 0, linetype = 'dashed', color = 'grey40', linewidth = 0.5) +
    geom_point(aes(color = sig), size = 2, alpha = 0.7) +
    geom_text_repel(aes(label = ifelse(q_value < 0.05, ct_to_disp, ''), color = sig),
                    max.overlaps = 100, force = 4, size = 2.8, fontface = 'bold',
                    segment.size = 0.3, segment.color = 'grey60', show.legend = FALSE) +
    facet_wrap(~ panel_label, scales = 'free', ncol = 4) +
    scale_color_manual(values = c('enriched' = '#b2182b', 'depleted' = '#2166ac', 'not significant' = 'grey75'), name = NULL) +
    labs(title = 'Neighbor cell-type composition vs. MAPK activation, per patient',
         subtitle = 'Panel: focal cell type (OPC/Astrocyte) · driver alteration  ·  dotted line: q = 0.05',
         x = 'Effect size (log coefficient)', y = expression(-log[10](p))) +
    ref_theme
```

# Q2 -- Myeloid programs by glial neighborhood (nested 3-group)

For each myeloid cell we classify its glial neighborhood:

- **pa_niche_glia** -- >=1 glial neighbor whose niche is a PA glial niche
- **other_glia** -- has glial neighbor(s), but none in a PA niche
- **no_glial** -- no glial neighbor at all

We fit two models per sample x signature:

- **Model 1 (primary, all myeloid cells):** `signature ~ exposure_group`,
  reference = `other_glia`, so the `pa_niche_glia` coefficient is the direct
  PA-specific effect (does PA-tumor-niche glial exposure shift the program beyond
  generic glial exposure?).
- **Model 2 (refinement, glia-adjacent cells only):**
  `signature ~ (pa_niche_glia vs other_glia) + mean_glial_mapk`. The MAPK dose is
  undefined for no-glial cells, so they are dropped here. This asks whether
  PA-niche proximity adds anything *beyond* the glial cells' own MAPK level.

Signatures (56: 50 MSigDB Hallmark + 6 myeloid/immune literature sets) are
z-scored within each patient before fitting.

**Inference framing (important).** The models are fit *per cell*, so with thousands
of cells per group the per-cell p-values collapse and nearly every signature clears
q < 0.05 within a sample. That per-cell hit count is **not** a cohort claim -- it is
pseudoreplication. The real inferential unit here is the **sample (n = 4)**. We
therefore read Q2 **descriptively / hypothesis-generating**: standardized effect
size + directional consistency across the 4 PA samples, not cohort-level
significance (`project_cohort_granularity`: n confounded, broad strokes only). For
calibration, a 4/4 same-direction sign test is only p = 0.125, and ~7 of 56
signatures would reach 4/4 by chance, so the all-4-consistent set is *suggestive
above chance*, not individually significant.

**Contrast structure.** Because the PA glial niches (3/5) hold most glial cells in
PA, most glia-adjacent myeloid cells land in `pa_niche_glia` (a broad tumor-core
tent) while `other_glia` is a smaller tumor-periphery group. This imbalance is the
intended tumor-core-vs-periphery contrast, by design, not a defect.

## Build myeloid glial-exposure table

```{r}
glial_mapk_lookup  = mapk_scores[type == 'spatial', .(cell_id, glial_mapk = MAPK_overlap_score)]
glial_niche_lookup = cell_meta_pa[cell_type %in% glial_types, .(to_id = cell_id, to_niche = niche)]

# myeloid -> glial directed edges, tagged with each glial neighbor's niche + MAPK
myeloid_glial_edges = adjacencies[cell_type_from == 'myeloid' & cell_type_to %in% glial_types]
myeloid_glial_edges = merge(myeloid_glial_edges, glial_niche_lookup, by = 'to_id')
myeloid_glial_edges = merge(myeloid_glial_edges, glial_mapk_lookup, by.x = 'to_id', by.y = 'cell_id')
myeloid_glial_edges[, to_in_pa_niche := to_niche %in% pa_glial_niches]

# aggregate per myeloid cell: glial-neighbor count, PA-niche-glial count, mean glial MAPK
glial_agg = myeloid_glial_edges[, .(
    n_glial         = .N,
    n_pa_glial      = sum(to_in_pa_niche),
    mean_glial_mapk = mean(glial_mapk)      # over ALL glial neighbors -> Model 2 dose covariate
), .(cell_id = from_id, sample_id)]

# keep every myeloid cell, including those with zero glial neighbors
all_myeloid = unique(adjacencies[cell_type_from == 'myeloid', .(cell_id = from_id, sample_id)])
myeloid_exposure = merge(all_myeloid, glial_agg, by = c('cell_id', 'sample_id'), all.x = TRUE)
myeloid_exposure[is.na(n_glial), `:=`(n_glial = 0L, n_pa_glial = 0L, mean_glial_mapk = NA_real_)]
myeloid_exposure[, exposure_group := fcase(
    n_pa_glial >= 1, 'pa_niche_glia',
    n_glial   >= 1,  'other_glia',
    default = 'no_glial'
)]
# reference = other_glia so the pa_niche_glia coefficient is the PA-specific effect
myeloid_exposure[, exposure_group := factor(exposure_group, levels = c('other_glia', 'pa_niche_glia', 'no_glial'))]
myeloid_exposure[, sample_id := as.character(sample_id)]

# per-sample group counts (feasibility of Model 1 / Model 2 per sample)
group_counts = dcast(myeloid_exposure[, .N, .(sample_id, exposure_group)],
                     sample_id ~ exposure_group, value.var = 'N', fill = 0)
cat('Q2 myeloid exposure-group counts per sample:\n'); print(group_counts)
fwrite(myeloid_exposure, paste0(out_dir, 'myeloid_glial_exposure.tsv'), sep = '\t')
```

## Load myeloid signature scores and z-score per patient

```{r}
# myeloid_signature_scores.csv: myeloid cells x 56 score columns + cell_id + sample_id
# (Phase 1: sc.tl.score_genes on the spatial ENVI imputation, subset to new myeloid label).
sig_scores = fread(paste0(out_dir, 'myeloid_signature_scores.csv'))
score_cols = grep('_score$', names(sig_scores), value = TRUE)
sig_scores[, sample_id := as.character(sample_id)]

myeloid_data = merge(
    myeloid_exposure,
    sig_scores[, c('cell_id', 'sample_id', score_cols), with = FALSE],
    by = c('cell_id', 'sample_id')
)
cat(sprintf('Joined %d / %d myeloid cells to signature scores (%.1f%% match); %d signatures.\n',
            nrow(myeloid_data), nrow(myeloid_exposure),
            100 * nrow(myeloid_data) / nrow(myeloid_exposure), length(score_cols)))

# z-score each signature within each patient (per-sample score_genes baselines)
myeloid_data[
    , (score_cols) := lapply(.SD, function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))
    , .SDcols = score_cols, by = sample_id
]
```

## Fit both models

```{r}
#| results: hold
MIN_N_GRP = 30   # min cells per contrasted group

# Fit lm(sig ~ rhs) and return the single coefficient whose term == term_keep.
# Name-based selection (unlike a single-predictor helper) so it works with the
# 3-level factor and the covariate-adjusted model.
fit_term_row = function(dt, sig_col, rhs, term_keep, weight_col = NULL) {
    fml = as.formula(paste0('`', sig_col, '` ~ ', rhs))
    wts = if (!is.null(weight_col)) dt[[weight_col]] else NULL
    m = tryCatch(lm(fml, data = dt, weights = wts), error = function(e) NULL)
    if (is.null(m)) return(NULL)
    co = as.data.table(summary(m)$coefficients, keep.rownames = 'term')
    row = co[term == term_keep]
    if (nrow(row) != 1) return(NULL)
    data.table(signature = sig_col, Estimate = row$Estimate, SE = row[['Std. Error']],
               t_value = row[['t value']], p_value = row[['Pr(>|t|)']], n_cells = nrow(dt))
}

# empty, correctly-typed results table (so BH/merge don't error if nothing fits)
empty_results = function() data.table(
    signature = character(0), Estimate = numeric(0), SE = numeric(0), t_value = numeric(0),
    p_value = numeric(0), n_cells = integer(0), sample_id = character(0),
    q_value = numeric(0), alteration = factor(character(0), levels = alteration_levels)
)

finalize = function(res, path) {
    if (nrow(res) == 0 || !'sample_id' %in% names(res)) {
        res = empty_results()
    } else {
        res[, q_value := p.adjust(p_value, method = 'BH'), by = sample_id]
        res = merge(res, unique(mapk_scores[, .(sample_id, alteration)]), by = 'sample_id')
        res[, alteration := factor(alteration, levels = alteration_levels)]
    }
    fwrite(res, path, sep = '\t')
    res
}

samples = unique(myeloid_data$sample_id)

# --- Model 1: 3-group, all myeloid cells, PA-niche-glia vs other-glia ---------
combos = CJ(sample_id = samples, signature = score_cols)
nested_results = rbindlist(mapply(function(sid, sig) {
    dt = myeloid_data[sample_id == sid, .SD, .SDcols = c(sig, 'exposure_group')]
    dt = dt[complete.cases(dt)]
    dt[, exposure_group := droplevels(exposure_group)]
    # need both the reference (other_glia) and pa_niche_glia populated
    if (dt[exposure_group == 'pa_niche_glia', .N] < MIN_N_GRP ||
        dt[exposure_group == 'other_glia',   .N] < MIN_N_GRP) return(NULL)
    res = fit_term_row(dt, sig, 'exposure_group', 'exposure_grouppa_niche_glia')
    if (!is.null(res)) res[, sample_id := sid]
    res
}, combos$sample_id, combos$signature, SIMPLIFY = FALSE))
nested_results = finalize(nested_results, paste0(out_dir, 'myeloid_glial_exposure_nested_results.tsv'))
cat('Model 1 (3-group): significant (q<0.05) signature-patient hits:',
    sum(nested_results$q_value < 0.05, na.rm = TRUE), '\n')

# --- Model 2: covariate-adjusted, glia-adjacent cells only --------------------
# drop no_glial (mean_glial_mapk undefined); binary pa vs other + MAPK dose
adj_base = myeloid_data[exposure_group %in% c('pa_niche_glia', 'other_glia')]
adj_base[, exposure_group := droplevels(exposure_group)]   # reference = other_glia
covar_results = rbindlist(mapply(function(sid, sig) {
    dt = adj_base[sample_id == sid, .SD, .SDcols = c(sig, 'exposure_group', 'mean_glial_mapk')]
    dt = dt[complete.cases(dt)]
    if (dt[exposure_group == 'pa_niche_glia', .N] < MIN_N_GRP ||
        dt[exposure_group == 'other_glia',   .N] < MIN_N_GRP) return(NULL)
    res = fit_term_row(dt, sig, 'exposure_group + mean_glial_mapk', 'exposure_grouppa_niche_glia')
    if (!is.null(res)) res[, sample_id := sid]
    res
}, combos$sample_id, combos$signature, SIMPLIFY = FALSE))
covar_results = finalize(covar_results, paste0(out_dir, 'myeloid_glial_exposure_covariate_results.tsv'))
cat('Model 2 (MAPK-adjusted): significant (q<0.05) signature-patient hits:',
    sum(covar_results$q_value < 0.05, na.rm = TRUE), '\n')
```

## Cohort readout: effect size + cross-sample consistency (the Q2 answer)

The per-cell q<0.05 hit counts above are **within-sample descriptive only** and are
dominated by pseudoreplication (thousands of cells per group ⇒ SEs collapse). The
cohort-level Q2 answer is read off the **per-sample standardized effect sizes** and
their **directional consistency across the 4 PA samples**. A signature is
"all-4-consistent" if it was fittable in all 4 samples and every per-sample beta
shares the same sign. Calibration: ~7 of 56 signatures would reach 4/4 by chance,
so this set is *suggestive above chance*, not individually significant at n=4.

```{r}
# Collapse per-sample betas to one row per signature: directional consistency +
# mean standardized effect. Uses ALL fitted per-sample betas (not the per-cell
# q-value), because the sample is the inferential unit here.
q2_consistency = function(res, label) {
    if (nrow(res) == 0) return(data.table(signature = character(0)))
    s = res[!is.na(Estimate), .(
        n_samples = .N,                       # samples where the contrast was fittable (N>=MIN_N_GRP)
        n_pos     = sum(Estimate > 0),
        n_neg     = sum(Estimate < 0),
        mean_beta = mean(Estimate)            # mean standardized effect across samples
    ), by = signature]
    s[, consistent_4 := n_samples == 4L & (n_pos == 4L | n_neg == 4L)]
    s[, model := label]
    s[order(-consistent_4, -abs(mean_beta))]
}
nested_consistency = q2_consistency(nested_results, 'Model1_3group')
covar_consistency  = q2_consistency(covar_results,  'Model2_MAPKadj')

# Q2 headline table: the all-4-consistent signatures under Model 1, most-affected first.
headline = nested_consistency[consistent_4 == TRUE][order(-abs(mean_beta))]
cat('All-4-sample-consistent signatures (Model 1):', nrow(headline),
    '(vs ~7 expected by chance across 56)\n')
kable(headline[, .(signature = sub('_score$', '', signature),
                   direction = ifelse(mean_beta > 0, 'up', 'down'),
                   mean_beta = round(mean_beta, 3))],
      caption = 'Q2 headline: myeloid programs consistently shifted near PA-niche glia across all 4 PA samples')

# Model 2 attenuation: how much does adjusting for neighbor-glial MAPK dose shrink
# the PA-niche effect on this consistent set? Persistence (small attenuation) is the
# evidence that PA-niche proximity is not merely a proxy for the glia's MAPK level.
atten = merge(
    headline[, .(signature, beta_m1 = mean_beta)],
    covar_consistency[, .(signature, beta_m2 = mean_beta)],
    by = 'signature')
atten[, atten_pct := round(100 * (1 - beta_m2 / beta_m1), 1)]   # % shrink of |effect| under MAPK adjustment
cat(sprintf('Model 2 attenuation over the consistent set: median %.0f%% (IQR %.0f-%.0f%%); %d/%d retain sign.\n',
            median(atten$atten_pct, na.rm = TRUE),
            quantile(atten$atten_pct, 0.25, na.rm = TRUE),
            quantile(atten$atten_pct, 0.75, na.rm = TRUE),
            atten[sign(beta_m2) == sign(beta_m1), .N], nrow(atten)))
kable(atten[order(-abs(beta_m1)), .(signature = sub('_score$', '', signature),
            beta_model1 = round(beta_m1, 3), beta_model2 = round(beta_m2, 3),
            attenuation_pct = atten_pct)],
      caption = 'PA-niche effect attenuates modestly but persists after adjusting for glial MAPK dose')
```

**Reading this.** The PA-niche myeloid shift is carried by the all-4-consistent set,
not by the raw per-cell hit count. Under MAPK adjustment (Model 2) those effects
attenuate only modestly and keep their sign — so neighbor-glial MAPK dose explains
part but not most of the signal, and PA-niche proximity is not simply a MAPK proxy.

### Interpretation

The **20 of 56** signatures that move the same direction in all 4 PA samples (~7 expected
by chance ⇒ ~3× enrichment; per-cell q is pseudoreplicated and not the claim) split into
two coherent, biologically consistent themes — myeloid cells near PA-niche glia (the tumor
core), relative to myeloid near other glia (periphery):

- **Immunosuppressive / alternatively-activated TAM shift** — ↓ inflammatory response,
  ↓ IFN-γ response, ↓ allograft rejection, ↓ homeostatic (TIM3-neg) microglia, ↓ FOSL2
  regulon; ↑ complement, ↑ activated-microglia (MG_Act). A pull away from pro-inflammatory/
  antigen-presenting microglia toward a complement-high, immunosuppressive-leaning TAM state.
- **Metabolic reprogramming** — ↓ glycolysis, ↓ hypoxia, ↓ PI3K–AKT–mTOR, ↓ apoptosis;
  ↑ heme metabolism, ↑ adipogenesis (lipid), ↑ xenobiotic metabolism, ↑ coagulation. A switch
  from glycolytic/hypoxic signaling toward lipid/heme scavenging — the metabolic signature of
  scavenger/anti-inflammatory macrophages.

Because these persist under MAPK adjustment (Model 2: ~9% median attenuation, 20/20 keep
sign), the phenotype tracks **PA-niche proximity itself, not the neighboring glia's MAPK
dose**. Caveat: n = 4 (suggestive/hypothesis-generating, not cohort-significant), ENVI-imputed
on a 266-gene panel.

## Heatmaps

Signatures sorted by mean |beta| across patients (most-affected at top). Bold
rows = myeloid/immune-specific signatures; plain = generic Hallmark controls.
Fill = standardized beta where the **within-sample** per-cell q < 0.05 (BH per
patient); grey = n.s. The per-cell q here is descriptive within each sample (it is
inflated by pseudoreplication) — the cohort-level answer is the all-4-consistent
set in the table above, and these heatmaps show the per-sample effect **sizes** and
their sign agreement across the 4 columns.

```{r}
#| fig-width: 9
#| fig-height: 14
immune_sigs = c(
    'HALLMARK_INFLAMMATORY_RESPONSE_score', 'HALLMARK_INTERFERON_GAMMA_RESPONSE_score',
    'HALLMARK_INTERFERON_ALPHA_RESPONSE_score', 'HALLMARK_TNFA_SIGNALING_VIA_NFKB_score',
    'HALLMARK_COMPLEMENT_score', 'HALLMARK_IL6_JAK_STAT3_SIGNALING_score',
    'HALLMARK_IL2_STAT5_SIGNALING_score', 'HALLMARK_ALLOGRAFT_REJECTION_score',
    'MG_Act_Tripathi_score', 'PA_TIM3pos_Microglia_score', 'PA_TIM3neg_Microglia_score',
    'Bernstein_Complement_Immunosuppressive_score', 'Bernstein_Scavenger_Immunosuppressive_score',
    'FOSL2_Regulon_Small_score'
)

plot_exposure_heatmap = function(results_dt, title, subtitle) {
    if (nrow(results_dt) == 0) {
        return(ggplot() +
            annotate('text', x = 0.5, y = 0.5, size = 4.5, color = 'grey30',
                     label = paste0(title, '\n\n(no fitted rows: contrasted groups under-populated in this cohort)')) +
            theme_void())
    }
    dt = copy(results_dt)
    dt[, sig_label := sub('_score$', '', signature)]
    row_order = dt[, .(mean_abs = mean(abs(Estimate), na.rm = TRUE)), sig_label][order(-mean_abs)]$sig_label
    dt[, sig_label := factor(sig_label, levels = row_order)]
    dt[, fill_val := ifelse(q_value < 0.05, Estimate, NA_real_)]
    is_immune = sub('_score$', '', immune_sigs)
    face_vec = ifelse(levels(dt$sig_label) %in% is_immune, 'bold', 'plain')

    ggplot(dt, aes(x = alteration, y = sig_label, fill = fill_val)) +
        geom_tile(color = 'white', linewidth = 0.6) +
        scale_fill_gradient2(low = '#2166ac', mid = '#f7f7f7', high = '#b2182b', midpoint = 0,
                             na.value = 'grey92', limits = c(-1, 1), name = 'Standardized\nbeta (q<0.05)') +
        scale_x_discrete(expand = expansion(add = 0.6)) +
        scale_y_discrete(expand = expansion(add = 0.6)) +
        labs(title = title, subtitle = subtitle) +
        theme_minimal(base_size = 12) +
        theme(axis.title = element_blank(),
              axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
              axis.text.y = element_text(size = 8, face = face_vec),
              panel.grid = element_blank(),
              plot.title = element_text(face = 'bold', size = 14, margin = margin(b = 4)),
              plot.subtitle = element_text(size = 9, color = 'grey40', margin = margin(b = 8)),
              legend.title = element_text(size = 9), legend.key.height = unit(1.4, 'lines'))
}

plot_exposure_heatmap(
    nested_results,
    'Q2 Model 1: PA-niche-glial vs other-glial exposure on myeloid programs',
    'Fill: beta for pa_niche_glia vs other_glia (reference), z-scored signature, all myeloid cells.\nBold = myeloid/immune-specific; plain = generic Hallmark controls.'
)
```

```{r}
#| fig-width: 9
#| fig-height: 14
plot_exposure_heatmap(
    covar_results,
    'Q2 Model 2: PA-niche-glial effect adjusted for glial MAPK dose',
    'Fill: beta for pa_niche_glia vs other_glia, adjusting for mean neighbor-glial MAPK (glia-adjacent cells only).\nSurviving signal = PA-niche proximity effect beyond the glial cells own MAPK level.'
)
```
