---
title: "Modeling of Cell Type Proportions with MAPK Activation"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 8
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---

# Introduction

Here we will be using linear models to test the association between cell types and MAPK activation. We previously generated imputed MAPK overlap geneset score on the xenium data.

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(dbscan)
library(circlize)
library(Seurat)
library(scales)
library(htmlwidgets)
library(ggrepel)
library(rcartocolor)


# local vs remote
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (Sys.getenv("HOME") == "/PHShome/yz762"){
    # in eristwo, the home directory is at /PHShome/yz762
    workdir <- "/data/beroukhim1/"
} else if (Sys.getenv("HOME") == "/home/yz762"){
    # on the new server, the home directory is at /home/yz762
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    # in dipg or uger, the home directory is '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
```


# Preperations

## Loading the data

The objects are here:

RDS object with Leiden Clusters: `/data/beroukhim1/youyun/plgg/data/Xenium_annotations/banksy_clusters_connected_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds`

    -   `/data/beroukhim1/youyun/plgg/data/Xenium_annotations/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds`

Cell type clusters annotation: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv`  


```{r}
# load the banksy embeddings
banksy_embeddings <- readRDS(ifelse(
    !Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng"), paste0(
        workdir, "youyun/plgg/data/Xenium_annotations/banksy_clusters_connected_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds"
    ), "/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds"
))
colData(banksy_embeddings)$sample_id = factor(colData(banksy_embeddings)$sample_id)

# Load the cell type clusters annotation
prelim_annotation_res1 = fread(paste0(
    workdir,'youyun/plgg/data/xenium_cell_types/banksy_proseg_cell_cluster_annotation.tsv'
))

# specify the cluster names 
# this is mainly to keep track of which resolution we are using
cell_res = 1
cell_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k50_res%.1f',cell_res))
niche_res = 1
niche_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.8_k50_res%.1f',niche_res))

# # filter the banksy object to only include the cells with annotated cell types
# annotated_cell_clusters = sort(unique(prelim_annotation_res1$cluster))
# banksy_embeddings = banksy_embeddings[,colData(banksy_embeddings)[,cell_clust_name] %in% annotated_cell_clusters]
```


## Build the color palette for all niches and cell types downstream

```{r}
set.seed(5555)
cluster_classes = levels(colData(banksy_embeddings)[[niche_clust_name]])
niche_color_manual = structure(
    pals::polychrome()[c(1:length(cluster_classes))],
    names = sample(cluster_classes)
)
# for niche resolution 0.5
# niches_to_show = c(1:11, 13, 14, 16, 17)
# for niche resolution 1
niches_to_show = c(1:20, 22:25)

cell_type_to_test = 'Coarse Cell Type'
cell_type_to_test = 'Fine Cell Type'
cell_type_to_test = 'Fine Cell Type UMAP'

cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))
cell_type_color_manual = structure(
    pals::alphabet()[c(1:length(cell_type_class))],
    names = cell_type_class
)
```

## MAPK score

```{r}
mapk_scores = fread(paste0(workdir,'youyun/plgg/code/sc_integration/ENVI/2_spatial_imputation/PA/MAPK_overlap_scores.csv'))
```

# Visualize MAPK by niche

```{r}
cell_niches = as.data.table(colData(banksy_embeddings[
    , colData(banksy_embeddings)$histology == 'pilocytic astrocytoma'
])[, c('alteration',niche_clust_name)], keep.rownames = 'cell_id')[
    , alteration := fcase(
        grepl('188570', cell_id), 'KIAA1549::BRAF',
        grepl('258482', cell_id), 'FGFR1 ITD',
        grepl('320784', cell_id), 'FAM131B::BRAF',
        grepl('328368', cell_id), 'RAF1::TOP2B'
    )
]
setnames(cell_niches, c(niche_clust_name), c('niche_type'))
mapk_scores_w_niches = merge(
    mapk_scores, cell_niches,
    by.x = 'cell_id', by.y = 'cell_id', all.y = TRUE
)[
    , sample_id := gsub('.*__','',cell_id)
][
    , niche_type := ifelse(
        # niche_type %in% c(1,2,3,5,7,9,19,11,14,16,18,19,22,27),
        niche_type %in% c(7, 18, 22),
        niche_type, 'Other'
    )
]

ggplot(
    mapk_scores_w_niches[
        grepl('Astro|OPC', cell_type) & niche_type != 'Other', 
        .(
            MAPK_overlap_score_mean = mean(MAPK_overlap_score),
            MAPK_overlap_score_se = sd(MAPK_overlap_score)/sqrt(.N)
        ), .(cell_type, niche_type = factor(as.numeric(niche_type)))
    ], aes(
        x = factor(cell_type), y = MAPK_overlap_score_mean
    )
) +
    geom_point(
        aes(group = niche_type, color = niche_type),
        position = position_dodge(width = 0.8)
    ) +
    geom_errorbar(
        aes(
            y = MAPK_overlap_score_mean,
            ymin = MAPK_overlap_score_mean - MAPK_overlap_score_se,
            ymax = MAPK_overlap_score_mean + MAPK_overlap_score_se,
            color = niche_type,
            group = interaction(cell_type, niche_type)
        ),
        position = position_dodge(width = 0.8),
        width = 0.2, size = 1
    ) +
    theme_bw(base_size = 20) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'bottom'
    ) +
    labs(
        title = "MAPK Activation by Niche Type",
        x = "Cell Type",
        y = "MAPK Activation Score"
    ) +
    scale_color_carto_d(name = "Niche Type", palette = "Vivid") +
    guides(
        fill = guide_legend(title = "Niche Type", nrow = 2, byrow = TRUE)
    )
```

# Neighborhood analysis

## Setting up Neighborhood Analysis

Here we will look at the niches in the data and annotate them using the cell types of self and neighbors.

```{r}
k_geom = 15
locs = spatialCoords(banksy_embeddings[, colData(banksy_embeddings)$histology == 'pilocytic astrocytoma'])
knn = dbscan::kNN(x = locs, k = k_geom)
knnDT <- data.table(
    from = rep(seq_len(nrow(knn$id)), k_geom),
    from_id = rep(rownames(knn$id), k_geom),
    to = as.vector(knn$id),
    to_id = rownames(knn$id)[as.vector(knn$id)],
    distance = as.vector(knn$dist)
)
cell_cell_type_dt = data.table(
    cell_id = rownames(colData(banksy_embeddings[])),
    cell_type = as.character(colData(banksy_embeddings)[,cell_clust_name]),
    sample_id = colData(banksy_embeddings)$sample_id,
    histology = colData(banksy_embeddings)$histology
)[histology == 'pilocytic astrocytoma']

sparsity_threshold = 0.01
cell_cell_type_dt[
    , num_cells_per_sample := .N, .(sample_id)
][
    , cell_type := ifelse(
        .N/num_cells_per_sample < sparsity_threshold |
            is.na(cell_type), 'Other', cell_type
    ),.(sample_id, cell_type)
]
```

# Cell Cell Adjacency Analysis

To do this, we compare each niche's cell cell adjacency to all other niches' cell cell adjacency. 

```{r}
cell_type_assignment_vector = structure(
    c(prelim_annotation_res1[[cell_type_to_test]], 'Other'),
    names = c(prelim_annotation_res1$cluster,'Other')
)
adjacencies = knnDT[,.(from_id,to_id,distance)][
    cell_cell_type_dt[,.(cell_id,cell_type_to = cell_type)], 
    on = c('to_id' = 'cell_id')
][
    cell_cell_type_dt[,.(
        cell_id, cell_type_from = cell_type, sample_id, histology
    )], on = c('from_id' = 'cell_id')
][
    , c('cell_type_test_from','cell_type_test_to') := list(
        cell_type_assignment_vector[cell_type_from],
        cell_type_assignment_vector[cell_type_to]
    )
]
```

# Modeling cell type proportions with MAPK activation

```{r}
adjacencies_and_mapk = merge(
    dcast(
        adjacencies[,.(neighbor_prop = .N/k_geom),.(from_id, cell_type_test_from, cell_type_test_to, sample_id)],
        from_id + cell_type_test_from + sample_id ~ cell_type_test_to,
        value.var = 'neighbor_prop', fill = 0
    ),
    mapk_scores[,.(cell_id, MAPK_overlap_score)], by.x = 'from_id', by.y = 'cell_id'
)[, weights := k_geom]
cell_types_in_dt = setdiff(colnames(adjacencies_and_mapk), c('from_id', 'cell_type_test_from', 'sample_id', 'MAPK_overlap_score', 'weights'))
cell_type_pairs_to_test = expand.grid(cell_type_from = cell_types_in_dt, cell_type_to = cell_types_in_dt)

mapk_adj_results = rbindlist(mapply(
    function(ct1, ct2) {
        print(paste0("Testing ", ct1, " -> ", ct2))
        if (nrow(adjacencies_and_mapk[cell_type_test_from == ct1]) == 0) {
            message(sprintf("Cell type %s not found in the data!", ct1))
            return(NULL)
        }
        if (all(adjacencies_and_mapk[[ct2]] == 0)) {
            message(sprintf("Cell type %s has all zeros!", ct))
            return(NULL)
        }
        
        tmp_dt = adjacencies_and_mapk[cell_type_test_from == ct1]

        if (length(unique(tmp_dt$sample_id)) == 1){
            model = glm(
                as.formula(paste0('`',ct2, "` ~ MAPK_overlap_score")),
                data = tmp_dt, family = binomial(link = 'logit'), weights = weights
            )   
        } else {
            model = glm(
                as.formula(paste0('`',ct2, "` ~ MAPK_overlap_score + sample_id")),
                data = tmp_dt, family = binomial(link = 'logit'), weights = weights
            )
        }
        model_coef = as.data.frame(summary(model)$coefficients)
        model_coef$variables = rownames(model_coef)
        model_coef = as.data.table(model_coef)[variables == 'MAPK_overlap_score'][
            ,c('neg_log10_p_value', 'effect','ct_from','ct_to') := list(
                -log10(`Pr(>|z|)`), exp(Estimate),
                ct1, ct2
            )
        ]
        return(model_coef)
    }, 
    cell_type_pairs_to_test$cell_type_from, 
    cell_type_pairs_to_test$cell_type_to, 
    SIMPLIFY = FALSE
))[
    , q_value := p.adjust(`Pr(>|z|)`, method = 'BH')
]
```

# Results

```{r}
ggplot(mapk_adj_results[grepl('Astro|OPC', ct_from)]) +
    geom_point(aes(
        x = Estimate, y = neg_log10_p_value,
        color = ifelse(
            q_value >= 0.05,'not significant',
            ifelse(
                as.numeric(effect) > 1, 'enriched','depleted'
            )
        ), alpha = 0.5
    )) +
    geom_text_repel(aes(
        x = Estimate, y = neg_log10_p_value,
        label = ifelse(q_value < 0.05, gsub('`','',ct_to), '')
    ), max.overlaps = 100, force = 3) +
    theme_linedraw() +
    theme(
        text = element_text(size = 15), 
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'right', 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()
    ) + scale_color_manual(values = c(
        'enriched' = 'red', 'depleted' = 'blue',
        'not significant' = 'grey'
    )) +
    geom_vline(xintercept = 0, linetype = 'dashed', color = 'black') +
    xlim(-1, 0.5) +
    labs(
        title = "Effect of Imputed MAPK on Neighbor Cell Type Proportions",
        x = "Effect size (log coefficient)",
        y = "-log10(p-value)",
        color = "Significance"
    ) + facet_wrap(~ ct_from, nrow = 1)
```