---
title: "PLGG Cohort Description"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 12
        toc: true
        toc-depth: 3
        embed-resources: true
        results: hold
engine: knitr
knitr:
    opts_chunk: 
        R.options:
            width: 120
---

```{r}
library(data.table)
library(knitr)
library(readxl)
# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}
```

# Introduction

This document will look at the cohort of spatial PLGG dataset we are receiving from DKFZ. The metadata is here *`r workdir`coja/Spatial_PLGG/data/metadata/GNT.xlsx* and *`r workdir`coja/Spatial_PLGG/data/metadata/Xenium_PS.xlsx*. 

```{r}
metadata <- data.table(read_xlsx(paste0(
    workdir, "coja/Spatial_PLGG/data/metadata/GNT.xlsx"
)))
kable(metadata[
    xenium == 'yes',.(
    `sample #`, id, np_nr, protocol,max_super_family_class,
    max_family_class, max_class
)])
```

# Single Cell Data

Single cell data are either here `r workdir`coja/Spatial_PLGG/data/SN0304052 or here `r workdir`coja/Spatial_PLGG/data/snRNA_seq

We will check which of these samples are received and where

```{r}
library(ggalluvial)
library(ggrepel)
ggplot(metadata[,.N, .(
    max_super_family_class, max_family_class, 
    max_class, xenium
)], aes(
    axis1 = max_super_family_class, 
    axis2 = max_family_class,
    axis3 = max_class,
    y = N
)) + geom_alluvium(
    aes(fill = xenium), width = 1/12, color = "black"
) + geom_stratum(width = 1/12, fill = "black", color = "grey") + 
    geom_label_repel(stat = "stratum", aes(label = after_stat(
        stringr::str_wrap(stratum, 20)
    )),alpha = 0.7) + 
    theme_minimal() + theme(
        legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.margin = unit(c(1,1,1,1), "cm")
    ) + 
    ggtitle("Single Cell Sample Histologies") +
    scale_fill_brewer(palette = "Paired") 
```

# Spatial Data
