---
title: ""
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 6
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk:
        results: hold
        R.options:
            width: 120
---

# Introduction

In this notebook we systematically analyze the single nucleus RNA-seq data in terms of cell type composition.

The data is here on terra google bucket:

```
gcloud storage cp 'gs://fc-secure-351a0c20-e431-4bd2-aa5f-7cfeb0216d9a/data/sc_sn/Extended_sn/harmony_lambda2_t1.rds' /mnt/storage/dept/medonc/beroukhim/youyun/plgg/data/single_cell/harmony_lambda2_t1.rds
```

The single cell object is here -- `/data/beroukhim1/youyun/plgg/data/single_cell/harmony_lambda2_t1.rds`

Metadata is extracted from the Seurat object and saved as a tsv file here -- `/data/beroukhim1/youyun/plgg/data/single_cell/sn_cohort_cell_metadata.tsv`

```{r}
#| echo: FALSE
library(data.table)
library(emmeans)
library(ggalluvial)
library(ggplot2)
library(glmmTMB)
library(knitr)
library(rcartocolor)

# local vs remote
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usually at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (Sys.getenv("HOME") == "/PHShome/yz762") {
    # in eristwo, the home directory is at /PHShome/yz762
    workdir <- "/data/beroukhim1/"
} else if (Sys.getenv("HOME") == "/home/yz762") {
    # on the new server, the home directory is at /home/yz762
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    # in dipg or uger, the home directory is '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}
```

# Load Data

```{r}
# sn <- readRDS(paste0(workdir, "youyun/plgg/data/single_cell/harmony_lambda2_t1.rds"))
# metadata = sn[[c(
#     'orig.ident', 'age', 'Sex', 'IntDx', 'type',
#     'Location_standard', 'Location_specific',
#     'annot_v2', 'annot_v1', 'annot_v0'
# )]]
# setnames(metadata, c('Location_standard'), c('Location'))
# write.table(
#     metadata, paste0(workdir, "youyun/plgg/data/single_cell/sn_cohort_cell_metadata.tsv"),
#     sep = "\t", quote = FALSE, row.names = TRUE, col.names = TRUE
# )
```

```{r}
# The TSV was written with row names, so the first column is the cell barcode
# and the true sample ID is the second column.
metadata <- fread(
    paste0(workdir, "youyun/plgg/data/single_cell/sn_cohort_cell_metadata.tsv"),
    skip = 1,
    header = FALSE,
    col.names = c(
        "barcode", "orig.ident", "age", "Sex", "IntDx", "type", "Location",
        "Location_specific", "annot_v2", "annot_v1", "annot_v0"
    )
)
metadata[, age := suppressWarnings(as.numeric(age))]
metadata[Location == "NA", Location := ""]
```

```{r}
#| layout-nrow: 1
kable(table(metadata$annot_v0))
kable(table(metadata$annot_v1))
kable(table(metadata$annot_v2))
```

# Visualize the cohort metadata

```{r}
cohort_dt <- unique(metadata[, .(orig.ident, age, Sex, Location, IntDx, type)])
cohort_dt[is.na(Location) | !nzchar(Location), Location := "Missing"]

ggplot(
    cohort_dt,
    aes(axis1 = Location, axis2 = IntDx, axis3 = Sex, axis4 = type, axis5 = as.character(age))
) +
    geom_alluvium(aes(fill = IntDx)) +
    geom_stratum() +
    geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
    theme_minimal() +
    theme(panel.grid = element_blank(), axis.text.x = element_blank()) +
    scale_fill_carto_d(palette = "Safe") +
    scale_y_continuous(breaks = seq(0, nrow(cohort_dt), by = 5))
```

# Modeling cell type proportions

We fit sample-level beta-binomial models. Sex is the only adjustment term in the two primary models because finer adjustment is not supportable in this cohort.

## Data cleanup

Options for analysis

```{r}
cell_type_column <- "annot_v0"   # one of: annot_v0, annot_v1, annot_v2
min_total_cells_per_type <- 50
min_samples_with_cells <- 3
rare_fraction_threshold <- 0.005
min_group_samples <- 3L
min_samples_per_sex <- 2L
alpha <- 0.25
```

```{r}
setDT(metadata)
metadata <- metadata[
    !is.na(orig.ident) & !is.na(IntDx) & !is.na(Sex) & Sex %in% c("Female", "Male")
]

metadata[, cell_type := as.character(get(cell_type_column))]
metadata <- metadata[!is.na(cell_type) & nzchar(cell_type)]

global_ct <- metadata[, .N, by = cell_type][order(-N)]
global_ct[, frac := N / sum(N)]
rare_ct <- unique(c(
    global_ct[N < min_total_cells_per_type, cell_type],
    global_ct[frac < rare_fraction_threshold, cell_type]
))

metadata_f <- copy(metadata)
metadata_f[cell_type %in% rare_ct, cell_type := "Other"]

# Build the sample x cell-type table once so every model uses the same counts.
sample_dt <- unique(metadata_f[, .(orig.ident, IntDx, type, Location, Sex)])
total_dt <- metadata_f[, .(total = .N), by = orig.ident]
counts_dt <- metadata_f[, .(n_type = .N), by = .(orig.ident, cell_type)]
all_cell_types <- sort(unique(metadata_f$cell_type))

dat <- CJ(orig.ident = sample_dt$orig.ident, cell_type = all_cell_types, unique = TRUE)
dat <- dat[sample_dt, on = "orig.ident"]
dat <- counts_dt[dat, on = .(orig.ident, cell_type)]
dat[is.na(n_type), n_type := 0L]
dat <- total_dt[dat, on = "orig.ident"]

ct_presence <- dat[, .(
    samples_present = sum(n_type > 0L),
    total_cells = sum(n_type)
), by = cell_type]
keep_ct <- ct_presence[
    samples_present >= min_samples_with_cells & total_cells >= min_total_cells_per_type,
    cell_type
]
keep_ct <- union(setdiff(keep_ct, "Other"), "Other")
dat <- dat[cell_type %in% keep_ct]
dat[, prop := n_type / pmax(total, 1L)]

cat("Unique samples:", sample_dt[, uniqueN(orig.ident)], "\n")
cat("Cell types kept for modeling:", length(setdiff(keep_ct, "Other")), "\n")
```

## Analysis support

```{r}
make_support_table <- function(sample_dt, group_var) {
    out <- copy(sample_dt)
    out[, level := as.character(get(group_var))]
    out[is.na(level) | !nzchar(level), level := "(blank)"]
    out <- out[, .(
        n_samples = .N,
        n_female = sum(Sex == "Female"),
        n_male = sum(Sex == "Male")
    ), by = level][order(level)]
    out[, included := n_samples >= min_group_samples &
        n_female >= min_samples_per_sex &
        n_male >= min_samples_per_sex]
    out[, reason := fifelse(
        included,
        sprintf("included: >=%d samples and >=%d per sex", min_group_samples, min_samples_per_sex),
        fifelse(
            n_samples < min_group_samples,
            sprintf("excluded: fewer than %d samples", min_group_samples),
            sprintf("excluded: fewer than %d female and %d male samples", min_samples_per_sex, min_samples_per_sex)
        )
    )]
    out
}

support_intdx <- make_support_table(sample_dt, "IntDx")
keep_intdx <- support_intdx[included == TRUE, level]

support_location <- make_support_table(sample_dt, "Location")
keep_location <- support_location[included == TRUE, level]
```

Model 1 keeps only diagnosis levels with at least 3 samples and at least 2 female plus 2 male samples.

```{r}
kable(setnames(copy(support_intdx), "level", "IntDx"))
```

Model 2 uses the same sex-support rule for location levels.

```{r}
kable(setnames(copy(support_location), "level", "Location"))
```

## Model details

For cell type $j$ in sample $i$, we model the count $n_{ij}$ out of $N_i$ total cells with a beta-binomial model:

$$
    n_{ij} \sim \mathrm{BetaBinomial}(N_i, p_{ij}, \phi_j), \qquad
    \mathrm{logit}(p_{ij}) = \beta_{0j} + x_i^\top \beta_j.
$$

The primary models adjust only for sex. Model 1 is interpreted as diagnosis-associated composition. Model 2 is interpreted as location-associated composition. The omnibus row is a global likelihood-ratio test for whether the focal predictor matters for a cell type at all after sex adjustment. The `emmeans` estimated proportions are the level-specific quantities to interpret biologically. In the estimate plots, the y-axis is the adjusted marginal proportion, and the point color indicates whether that level is higher or lower than the equal-weight model-wide mean for that cell type on the response scale. For Models 1 and 2, only cell types with omnibus BH-adjusted `p < 0.25` are highlighted in the main `emmeans` plots; the rest are treated as weaker or noisier evidence. The DLGNT subtype comparison is exploratory because subtype and location are strongly confounded in this cohort.

```{r}
# Keep the result-shaping logic in one place because emmeans uses slightly
# different column names across model families.
pick_col <- function(x, candidates) {
    hit <- intersect(candidates, names(x))
    if (length(hit)) hit[[1]] else NA_character_
}

bind_rows <- function(x) {
    x <- Filter(function(y) !is.null(y) && nrow(y), x)
    if (!length(x)) return(data.table())
    rbindlist(x, use.names = TRUE, fill = TRUE)
}

tidy_emmeans <- function(emm_sum, predictor, model_name, cell_type, n_samples) {
    level_col <- names(emm_sum)[1]
    est_col <- pick_col(emm_sum, c("prob", "response", "rate", "emmean"))
    lower_col <- pick_col(emm_sum, c("lower.CL", "asymp.LCL"))
    upper_col <- pick_col(emm_sum, c("upper.CL", "asymp.UCL"))
    data.table(
        model = model_name,
        result_type = "emmeans",
        cell_type = cell_type,
        predictor = predictor,
        term_or_contrast = as.character(emm_sum[[level_col]]),
        estimate = if (!is.na(est_col)) as.numeric(emm_sum[[est_col]]) else NA_real_,
        std.error = if ("SE" %in% names(emm_sum)) as.numeric(emm_sum[["SE"]]) else NA_real_,
        lower.CL = if (!is.na(lower_col)) as.numeric(emm_sum[[lower_col]]) else NA_real_,
        upper.CL = if (!is.na(upper_col)) as.numeric(emm_sum[[upper_col]]) else NA_real_,
        p.value = NA_real_,
        p_adj = NA_real_,
        n_samples = n_samples
    )
}

clean_effect_label <- function(x) {
    sub(" effect$", "", as.character(x))
}

tidy_effects <- function(eff_raw_sum, eff_bh_sum, predictor, model_name, cell_type, n_samples) {
    label_col <- names(eff_raw_sum)[1]
    lower_col <- pick_col(eff_raw_sum, c("lower.CL", "asymp.LCL"))
    upper_col <- pick_col(eff_raw_sum, c("upper.CL", "asymp.UCL"))

    out <- data.table(
        model = model_name,
        result_type = "eff_vs_mean",
        cell_type = cell_type,
        predictor = predictor,
        term_or_contrast = clean_effect_label(eff_raw_sum[[label_col]]),
        estimate = if ("estimate" %in% names(eff_raw_sum)) as.numeric(eff_raw_sum[["estimate"]]) else NA_real_,
        std.error = if ("SE" %in% names(eff_raw_sum)) as.numeric(eff_raw_sum[["SE"]]) else NA_real_,
        lower.CL = if (!is.na(lower_col)) as.numeric(eff_raw_sum[[lower_col]]) else NA_real_,
        upper.CL = if (!is.na(upper_col)) as.numeric(eff_raw_sum[[upper_col]]) else NA_real_,
        p.value = if ("p.value" %in% names(eff_raw_sum)) as.numeric(eff_raw_sum[["p.value"]]) else NA_real_,
        p_adj = NA_real_,
        n_samples = n_samples
    )

    bh_dt <- data.table(
        term_or_contrast = clean_effect_label(eff_bh_sum[[label_col]]),
        p_adj = if ("p.value" %in% names(eff_bh_sum)) as.numeric(eff_bh_sum[["p.value"]]) else NA_real_
    )
    out[bh_dt, on = "term_or_contrast", p_adj := i.p_adj]
    out
}

fit_beta_model <- function(dd, predictor, model_name, adjust_sex = TRUE, do_omnibus = TRUE) {
    cell_id <- dd$cell_type[[1]]
    dd <- copy(dd[total > 0])
    dd[, (predictor) := droplevels(factor(get(predictor)))]
    if (adjust_sex) dd[, Sex := droplevels(factor(Sex))]

    if (dd[, uniqueN(get(predictor))] < 2L) {
        return(list(
            log = data.table(model = model_name, cell_type = cell_id, status = "skipped: fewer than 2 predictor levels"),
            omnibus = NULL, emmeans = NULL, effects = NULL
        ))
    }

    full_terms <- c(predictor, if (adjust_sex) "Sex")
    full_form <- as.formula(paste0("cbind(n_type, total - n_type) ~ ", paste(full_terms, collapse = " + ")))
    reduced_form <- as.formula(paste0("cbind(n_type, total - n_type) ~ ", if (adjust_sex) "Sex" else "1"))

    full_fit <- tryCatch(
        glmmTMB(full_form, data = dd, family = betabinomial(link = "logit")),
        error = function(e) e
    )
    if (inherits(full_fit, "error")) {
        return(list(
            log = data.table(model = model_name, cell_type = cell_id, status = paste("fit failed:", full_fit$message)),
            omnibus = NULL, emmeans = NULL, effects = NULL
        ))
    }

    omnibus_dt <- NULL
    if (do_omnibus) {
        reduced_fit <- tryCatch(
            glmmTMB(reduced_form, data = dd, family = betabinomial(link = "logit")),
            error = function(e) e
        )
        if (inherits(reduced_fit, "error")) {
            return(list(
                log = data.table(model = model_name, cell_type = cell_id, status = paste("reduced fit failed:", reduced_fit$message)),
                omnibus = NULL, emmeans = NULL, effects = NULL
            ))
        }
        lr_tab <- tryCatch(as.data.table(anova(reduced_fit, full_fit)), error = function(e) NULL)
        p_col <- if (!is.null(lr_tab)) grep("^Pr\\(>Chisq\\)$", names(lr_tab), value = TRUE) else character()
        chisq_col <- if (!is.null(lr_tab)) grep("Chisq", names(lr_tab), value = TRUE) else character()
        omnibus_dt <- data.table(
            model = model_name,
            result_type = "omnibus",
            cell_type = cell_id,
            predictor = predictor,
            term_or_contrast = predictor,
            estimate = if (length(chisq_col)) as.numeric(lr_tab[[chisq_col[[1]]]][nrow(lr_tab)]) else NA_real_,
            std.error = NA_real_,
            lower.CL = NA_real_,
            upper.CL = NA_real_,
            p.value = if (length(p_col)) as.numeric(lr_tab[[p_col[[1]]]][nrow(lr_tab)]) else NA_real_,
            p_adj = NA_real_,
            n_samples = uniqueN(dd$orig.ident)
        )
    }

    emm <- tryCatch(emmeans(full_fit, specs = as.formula(paste0("~ ", predictor))), error = function(e) NULL)
    if (is.null(emm)) {
        return(list(
            log = data.table(model = model_name, cell_type = cell_id, status = "fit complete; emmeans failed"),
            omnibus = omnibus_dt, emmeans = NULL, effects = NULL
        ))
    }

    emm_sum <- tryCatch(as.data.table(summary(emm, type = "response", infer = TRUE)), error = function(e) NULL)
    if (is.null(emm_sum)) {
        return(list(
            log = data.table(model = model_name, cell_type = cell_id, status = "fit complete; emmeans summary failed"),
            omnibus = omnibus_dt, emmeans = NULL, effects = NULL
        ))
    }

    emm_dt <- tidy_emmeans(emm_sum, predictor, model_name, cell_id, uniqueN(dd$orig.ident))
    eff_dt <- NULL
    eff_obj <- tryCatch(contrast(regrid(emm, transform = "response"), "eff", by = NULL), error = function(e) NULL)
    if (!is.null(eff_obj)) {
        eff_raw_sum <- tryCatch(as.data.table(summary(eff_obj, infer = TRUE, adjust = "none")), error = function(e) NULL)
        eff_bh_sum <- tryCatch(as.data.table(summary(eff_obj, infer = TRUE, adjust = "BH")), error = function(e) NULL)
        if (!is.null(eff_raw_sum) && !is.null(eff_bh_sum)) {
            eff_dt <- tidy_effects(eff_raw_sum, eff_bh_sum, predictor, model_name, cell_id, uniqueN(dd$orig.ident))
        }
    }

    list(
        log = data.table(model = model_name, cell_type = cell_id, status = "fit"),
        omnibus = omnibus_dt,
        emmeans = emm_dt,
        effects = eff_dt
    )
}

run_model <- function(dat, predictor, keep_levels, model_name, adjust_sex = TRUE, do_omnibus = TRUE) {
    model_dt <- copy(dat[cell_type != "Other" & get(predictor) %in% keep_levels])
    fits <- lapply(
        split(model_dt, by = "cell_type", keep.by = TRUE),
        fit_beta_model,
        predictor = predictor,
        model_name = model_name,
        adjust_sex = adjust_sex,
        do_omnibus = do_omnibus
    )

    out <- list(
        log = bind_rows(lapply(fits, `[[`, "log")),
        omnibus = bind_rows(lapply(fits, `[[`, "omnibus")),
        emmeans = bind_rows(lapply(fits, `[[`, "emmeans")),
        effects = bind_rows(lapply(fits, `[[`, "effects"))
    )

    if (nrow(out$omnibus)) out$omnibus[, p_adj := p.adjust(p.value, method = "BH")]
    out
}

plot_emmeans_hits <- function(res, predictor_levels, predictor_name, plot_title) {
    sig_ct <- character()
    if (nrow(res$omnibus) && "cell_type" %in% names(res$omnibus)) {
        sig_ct <- res$omnibus[!is.na(p_adj) & p_adj < alpha, unique(cell_type)]
    }

    if (!length(sig_ct)) {
        cat("No omnibus-significant", predictor_name, "hits at BH <", alpha, "\n")
        return(invisible(NULL))
    }

    plot_dt <- copy(res$emmeans[cell_type %in% sig_ct])
    eff_dt <- copy(res$effects[cell_type %in% sig_ct, .(
        cell_type,
        term_or_contrast,
        eff_estimate = estimate,
        eff_p_adj = p_adj
    )])
    plot_dt[eff_dt, on = .(cell_type, term_or_contrast), `:=`(
        eff_estimate = i.eff_estimate,
        eff_p_adj = i.eff_p_adj
    )]
    plot_dt[, effect_class := fifelse(
        !is.na(eff_p_adj) & eff_p_adj < alpha & eff_estimate > 0,
        "higher",
        fifelse(
            !is.na(eff_p_adj) & eff_p_adj < alpha & eff_estimate < 0,
            "lower",
            "not_significant"
        )
    )]
    plot_dt[, effect_class := factor(effect_class, levels = c("higher", "lower", "not_significant"))]
    plot_dt[, term_or_contrast := factor(term_or_contrast, levels = predictor_levels)]

    ggplot(
        plot_dt,
        aes(
            x = term_or_contrast,
            y = estimate,
            ymin = lower.CL,
            ymax = upper.CL,
            color = effect_class
        )
    ) +
        geom_pointrange() +
        facet_wrap(~ cell_type, scales = "free_y") +
        theme_minimal(base_size = 18) +
        scale_color_manual(
            values = c(higher = "firebrick4", lower = "steelblue4", not_significant = "grey70"),
            labels = c(higher = "Higher than mean", lower = "Lower than mean", not_significant = "Not significant"),
            name = "Effect vs mean"
        ) +
        labs(
            x = predictor_name,
            y = "Estimated proportion",
            title = plot_title,
            subtitle = paste0(
                "Only cell types with omnibus BH-adjusted p < ", alpha,
                " are shown; color marks response-scale deviation from the equal-weight model mean"
            )
        ) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Model 1: `IntDx + Sex`

Included IntDx levels: `r paste(keep_intdx, collapse = ", ")`

Model formula: `cbind(n_type, total - n_type) ~ IntDx + Sex`

```{r}
intdx_results <- run_model(dat, predictor = "IntDx", keep_levels = keep_intdx, model_name = "intdx")
kable(intdx_results$log[, .N, by = status][order(-N)])
if (nrow(intdx_results$omnibus)) {
    kable(intdx_results$omnibus[order(p_adj)])
} else {
    cat("No diagnosis models completed.\n")
}
```

```{r}
#| fig-height: 10
plot_emmeans_hits(
    intdx_results,
    predictor_levels = keep_intdx,
    predictor_name = "Diagnosis",
    plot_title = "Model 1: estimated cell-type proportions by diagnosis"
)
```

## Model 2: `Location + Sex`

Included location levels: `r paste(keep_location, collapse = ", ")`

Model formula: `cbind(n_type, total - n_type) ~ Location + Sex`

```{r}
location_results <- run_model(dat, predictor = "Location", keep_levels = keep_location, model_name = "location")
kable(location_results$log[, .N, by = status][order(-N)])
if (nrow(location_results$omnibus)) {
    kable(location_results$omnibus[order(p_adj)])
} else {
    cat("No location models completed.\n")
}
```

```{r}
#| fig-height: 10
plot_emmeans_hits(
    location_results,
    predictor_levels = keep_location,
    predictor_name = "Location",
    plot_title = "Model 2: estimated cell-type proportions by location"
)
```

## Model 3: Exploratory `DLGNT_1` vs `DLGNT_2`

This comparison is reported separately because subtype and location are strongly confounded in this cohort, so it should be interpreted as exploratory rather than adjusted.

Model formula: `cbind(n_type, total - n_type) ~ IntDx`

```{r}
dlgnt_levels <- c("DLGNT_1", "DLGNT_2")
dlgnt_results <- run_model(
    dat,
    predictor = "IntDx",
    keep_levels = dlgnt_levels,
    model_name = "dlgnt12",
    adjust_sex = FALSE,
    do_omnibus = TRUE
)
kable(dlgnt_results$log[, .N, by = status][order(-N)])
```

### `DLGNT_1` vs `DLGNT_2` omnibus results

For this two-level exploratory model, the omnibus likelihood-ratio test asks whether the cell-type proportion differs at all between `DLGNT_1` and `DLGNT_2`.

```{r}
if (nrow(dlgnt_results$omnibus)) {
    kable(dlgnt_results$omnibus[order(p_adj)])
    dlgnt_sig <- dlgnt_results$omnibus[!is.na(p_adj) & p_adj < alpha, cell_type]
    if (length(dlgnt_sig)) {
        cat("BH-adjusted p <", alpha, ":", paste(dlgnt_sig, collapse = ", "), "\n")
    } else {
        cat("No DLGNT cell-type models were BH-significant at", alpha, ".\n")
    }
} else {
    cat("No DLGNT subtype models completed.\n")
}
```

### Estimated DLGNT subtype proportions

```{r}
if (nrow(dlgnt_results$emmeans)) {
    kable(dlgnt_results$emmeans[order(cell_type, term_or_contrast)])
} else {
    cat("No DLGNT subtype estimates completed.\n")
}
```

```{r}
#| fig-height: 10
plot_emmeans_hits(
    dlgnt_results,
    predictor_levels = dlgnt_levels,
    predictor_name = "DLGNT subtype",
    plot_title = "Model 3: estimated cell-type proportions by DLGNT subtype"
)
```

### Observed sample proportions

```{r}
#| fig-height: 10
if (nrow(dlgnt_results$emmeans)) {
    plot_ct <- unique(dlgnt_results$emmeans$cell_type)
    plot_dt <- dat[
        cell_type %in% plot_ct & IntDx %in% dlgnt_levels,
        .(orig.ident, cell_type, IntDx, Location, Sex, prop)
    ]
    plot_dt[is.na(Location) | !nzchar(Location), Location := "(blank)"]
    plot_dt[, IntDx := factor(IntDx, levels = dlgnt_levels)]

    ggplot(
        plot_dt,
        aes(
            x = IntDx,
            y = prop,
            color = Sex,
            shape = Location
        )
    ) +
        geom_point(
            position = position_jitter(width = 0.08, height = 0),
            size = 2.8,
            alpha = 0.9
        ) +
        facet_wrap(~ cell_type, scales = "free_y") +
        theme_minimal(base_size = 18) +
        labs(
            x = "DLGNT subtype",
            y = "Observed sample proportion",
            title = "Exploratory DLGNT_1 vs DLGNT_2 observed proportions",
            subtitle = "Points are individual samples; color shows sex and shape shows location"
        )
} else {
    cat("No DLGNT subtype estimates to plot.\n")
}
```

## Save results

```{r}
combine_results <- function(res) {
    bind_rows(list(res$omnibus, res$emmeans, res$effects))
}

outdir <- paste0(workdir, "youyun/plgg/code/sample_comparisons/")
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

intdx_out <- file.path(outdir, paste0("betabinom_intdx_", cell_type_column, "_results.tsv"))
location_out <- file.path(outdir, paste0("betabinom_location_", cell_type_column, "_results.tsv"))
dlgnt_out <- file.path(outdir, paste0("betabinom_dlgnt12_", cell_type_column, "_results.tsv"))

fwrite(combine_results(intdx_results), intdx_out, sep = "\t")
fwrite(combine_results(location_results), location_out, sep = "\t")
fwrite(combine_results(dlgnt_results), dlgnt_out, sep = "\t")

cat("Wrote:", intdx_out, "\n")
cat("Wrote:", location_out, "\n")
cat("Wrote:", dlgnt_out, "\n")
```
