---
title: "CRAWDAD niches analysis"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 6
        toc: true
        toc-depth: 3
        embed-resources: true
engine: knitr
knitr:
    opts_chunk: 
        results: hold
        R.options:
            width: 120
---


# Introduction

Testing out crawdad's multiscale niches analysis on our spatial transcriptomics data using previously assigned cell types. 

```{r}
#| echo: FALSE
#| output: false
library(data.table)
library(ggplot2)
library(knitr)

library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
library(harmony)
library(scran)
library(Seurat)
library(dbscan)
library(circlize)
library(Seurat)
library(plotly)
library(scales)
library(htmlwidgets)
library(metap)
library(crawdad)

# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}

source(paste0(workdir,'/youyun/plgg/code/helpers/test_enrichment.R'))
```

# Load data

# Load the data

The objects are here:

RDS object with Leiden Clusters: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds`

    -   `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_embeddings/total_spatial_banksy_clusters_subset_20240719_092550.rds`


Cell type clusters annotation: `/xchip/beroukhimlab/coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv`  


```{r}
# load the banksy embeddings
banksy_embeddings = readRDS(ifelse(
    workdir == '/xchip/beroukhimlab/',
    paste0(workdir,'coja/Spatial_PLGG/data/Xenium/Xenium_Objects/total_spatial_banksy_clusters_20240719_092550.rds'),
    '/Users/youyun/Documents/HMS/PhD/beroukhimlab/plgg/data/total_spatial_banksy_clusters_20240719_092550.rds'
))
colData(banksy_embeddings)$sample_id = factor(colData(banksy_embeddings)$sample_id)

# Load the cell type clusters annotation
prelim_annotation_res1 = fread(paste0(
    workdir,'coja/Spatial_PLGG/data/Xenium/banksy/banksy_cell_cluster_annotation.tsv'
))

# specify the cluster names 
# this is mainly to keep track of which resolution we are using
cell_res = 1
cell_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.2_k50_res%.1f',cell_res))
niche_res = 1
niche_clust_name = gsub('.0$','',sprintf('clust_Harmony_BANKSY_lam0.8_k50_res%.1f',niche_res))
```

Build the color palette for all niches and cell types downstream

```{r}
set.seed(5555)
cluster_classes = levels(colData(banksy_embeddings)[[niche_clust_name]])
niche_color_manual = structure(
    pals::polychrome()[c(1:length(cluster_classes))],
    names = sample(cluster_classes)
)
# for niche resolution 0.5
niches_to_show = c(1:11, 13, 14, 16, 17)
# for niche resolution 1
# niches_to_show = c(1:20, 22:25)

cell_type_to_test = 'Coarse Cell Type'
cell_type_to_test = 'Fine Cell Type'
cell_type_to_test = 'Fine Cell Type UMAP'

cell_type_class = sort(unique(prelim_annotation_res1[[cell_type_to_test]]))
cell_type_color_manual = structure(
    pals::alphabet()[c(1:length(cell_type_class))],
    names = cell_type_class
)

```

```{r}
# get spatial coordinates 
spat_df = data.frame(spatialCoords(banksy_embeddings))
# get the cell type labels
label_df = data.frame(colData(banksy_embeddings)[,c('sample_id',cell_clust_name)])
remove(banksy_embeddings)
# explicitly merge by cell barcode
spat_df$cell_barcode = rownames(spat_df)
label_df$cell_barcode = rownames(label_df)
spat_df = data.table(spat_df)
label_df = data.table(label_df)
total_df = spat_df[label_df,on = 'cell_barcode']

# convert cell type labels
cell_type_map_vector = structure(
    prelim_annotation_res1[[cell_type_to_test]],
    names = prelim_annotation_res1$cluster
)
total_df[
    , cell_type := ifelse(
        is.na(cell_type_map_vector[as.character(get(cell_clust_name))]),
        'NA', cell_type_map_vector[as.character(get(cell_clust_name))]
    ),.(cell_barcode)
]

```


```{r}
data(pkhl)
data(slide)
data(seq)
rbind(
    total_df[,.(
        .N, span_x = max(sdimx) - min(sdimx), span_y = max(sdimy) - min(sdimy),
        num_cell_type = length(unique(cell_type))
    ), .(sample_id)][order(-N),.(
        sample_id, N, span_x, span_y, num_cell_type, 
        cell_density = N/ (span_x * span_y)
    )],
    data.table(
        sample_id = c('pkhl','slide','seq'),
        N = c(nrow(pkhl), nrow(slide), nrow(seq)),
        span_x = c(max(pkhl$x) - min(pkhl$x), max(slide$x) - min(slide$x), max(seq$x) - min(seq$x)),
        span_y = c(max(pkhl$y) - min(pkhl$y), max(slide$y) - min(slide$y), max(seq$y) - min(seq$y)),
        num_cell_type = c(length(unique(pkhl$celltypes)), length(unique(slide$celltypes)), length(unique(seq$celltypes)))
    )[
        , cell_density := N / (span_x * span_y)
    ]
)
```

```{r}
cell_type_dt = rbind(
    total_df[
        , .N, .(sample_id, cell_type)
    ][
        , .(cell_type_proportion = N / sum(N)), .(sample_id)
    ],
    rbind(
        data.table(pkhl)[,sample_id := 'pkhl'], data.table(slide)[,sample_id := 'slide'], 
        data.table(seq)[,sample_id := 'seq']
    )[
        , .N, .(sample_id, celltypes)
    ][
        , .(cell_type_proportion = N / sum(N)), .(sample_id)
    ]
)
ggplot(
    cell_type_dt, aes(x = cell_type_proportion, fill = sample_id)
) + geom_histogram() + theme_linedraw() + theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 15)
) + facet_wrap(~sample_id, nrow = 3) +
    scale_x_log10() +
    labs(
        title = 'Cell type proportion distribution'
    )
remove(cell_type_dt, pkhl, slide, seq)
```

1.  It looks like, compared to the reference dataset used in demo of crawdad, our dataset is in general larger spatially and by cell count. 
2.  On the other hand, our cell density is on the lower end. This could be technical, but this could also because we have large white spaces on the FOV due to the irregularity of the biopsy shape
3.  It looks like we have quite a few cell types that are ultra rare. This may cause us to do too many useless tests and get no statistical significance

# CRAWDAD niches analysis 

```{r}
min_threshold = 0.005
verbosity = FALSE
scales <- c(100, 500, 1000)
ncores = 7
```

## CLN sample 87352

### Prepare the data

```{r}
CLN_sample = total_df[sample_id == '87352',.(
    x = sdimx - min(sdimx), y = sdimy - min(sdimy), cell_type
), .(sample_id)][x >= 7000 & y <= 2000]
# get rid of the rare cell types
cell_type_dt = CLN_sample[, .N, .(sample_id, cell_type)][, .(cell_type_proportion = N/sum(N), cell_type), .(sample_id)]
rare_cell_types = cell_type_dt[cell_type_proportion < min_threshold, cell_type]
CLN_cells = crawdad::toSF(
    pos = data.frame(CLN_sample[,.(x,y)]),
    cellTypes = as.factor(as.character(
        ifelse(CLN_sample$cell_type %in% rare_cell_types, 'Rare', CLN_sample$cell_type)
    ))
)
```

```{r}
#| fig-height: 12
vizEachCluster(cells = CLN_cells, coms = CLN_cells$celltypes, s = 5)
```

### Visualize different size reference neighborhoods

```{r}
#| fig-width: 6
#| fig-height: 3
## histogram for neighDist = 10
vizCelltypeProportions(CLN_cells, neighDist = 10) + 
    labs(title = 'Cell-type proportions for a neighborhood distance of 10')
## histogram for neighDist = 25
vizCelltypeProportions(CLN_cells, neighDist = 25) + 
    labs(title = 'Cell-type proportions for a neighborhood distance of 25')
## histogram for neighDist = 50
vizCelltypeProportions(CLN_cells, neighDist = 50) + 
    labs(title = 'Cell-type proportions for a neighborhood distance of 50')
## histogram for neighDist = 100
vizCelltypeProportions(CLN_cells, neighDist = 100) + 
    labs(title = 'Cell-type proportions for a neighborhood distance of 100')
```

### crawdad

```{r}
## shuffle cells to create null background
CLN_shuffle_list <- crawdad:::makeShuffledCells(
    CLN_cells,scales = scales, perms = 3, ncore = ncores,
    seed = 1,verbose = verbosity
)
## calculate the zscore for the cell-type pairs at different scales
CLN_results <- crawdad::findTrends(
    CLN_cells, neighDist = 25, shuffleList = CLN_shuffle_list,
    ncore = ncores, verbose = verbosity,
    returnMeans = FALSE
)
CLN_dat <- crawdad::meltResultsList(CLN_results, withPerms = TRUE)
## calculate the zscore for the multiple-test correction
CLN_zsig <- correctZBonferroni(CLN_dat)
## summary visualization
vizColocDotplot(
    CLN_dat, reorder = TRUE,
    zSigThresh = CLN_zsig, zScoreLimit = 2*CLN_zsig, 
    dotSizes = c(3,15)
) + theme(axis.text.x = element_text(angle = 35, h = 0, size = 15))
saveRDS(
    CLN_dat,
    paste0(workdir,'youyun/plgg/code/niche/crawdad/crawdad_niches_CLN_sample_87352.rds')
)
```

## PA sample 258482

### Prepare the data

```{r}
PA_sample = total_df[sample_id == '320784',.(
    x = sdimx - min(sdimx), y = sdimy - min(sdimy), cell_type
), .(sample_id)]#[x <= 3000 & x >= 1000 & y <= 5000 & y >= 3000]
# get rid of the rare cell types
cell_type_dt = PA_sample[, .N, .(sample_id, cell_type)][, .(cell_type_proportion = N/sum(N), cell_type), .(sample_id)]
rare_cell_types = cell_type_dt[cell_type_proportion < min_threshold, cell_type]
PA_cells = crawdad::toSF(
    pos = data.frame(PA_sample[,.(x,y)]),
    cellTypes = as.factor(as.character(
        ifelse(PA_sample$cell_type %in% rare_cell_types, 'Rare', PA_sample$cell_type)
    ))
)
```

```{r}
#| fig-height: 12
vizEachCluster(cells = PA_cells, coms = PA_cells$celltypes, s = 2)
```

### Visualize different size reference neighborhoods

```{r}
#| fig-width: 6
#| fig-height: 3
## histogram for neighDist = 10
vizCelltypeProportions(PA_cells, neighDist = 10) + 
    labs(title = 'Cell-type proportions for a neighborhood distance of 10')
## histogram for neighDist = 25
vizCelltypeProportions(PA_cells, neighDist = 25) + 
    labs(title = 'Cell-type proportions for a neighborhood distance of 25')
## histogram for neighDist = 50
vizCelltypeProportions(PA_cells, neighDist = 50) + 
    labs(title = 'Cell-type proportions for a neighborhood distance of 50')
## histogram for neighDist = 100
vizCelltypeProportions(PA_cells, neighDist = 100) + 
    labs(title = 'Cell-type proportions for a neighborhood distance of 100')
```

### crawdad

```{r}
## shuffle cells to create null background
PA_shuffle_list <- crawdad:::makeShuffledCells(
    PA_cells,scales = scales, perms = 3, ncore = ncores,
    seed = 1,verbose = verbosity
)
## calculate the zscore for the cell-type pairs at different scales
PA_results <- crawdad::findTrends(
    PA_cells, neighDist = 25, shuffleList = PA_shuffle_list,
    ncore = ncores, verbose = verbosity,
    returnMeans = FALSE
)
PA_dat <- crawdad::meltResultsList(PA_results, withPerms = TRUE)
## calculate the zscore for the multiple-test correction
PA_zsig <- correctZBonferroni(PA_dat)
## summary visualization
vizColocDotplot(
    PA_dat, reorder = TRUE,
    zSigThresh = PA_zsig, zScoreLimit = 2*PA_zsig, 
    dotSizes = c(3,15)
) + theme(axis.text.x = element_text(angle = 35, h = 0, size = 15))
saveRDS(
    PA_dat,
    paste0(workdir,'youyun/plgg/code/niche/crawdad/crawdad_niches_PA_sample_258482.rds')
)
```


