---
title: "Unassigned Transcripts in Spatial Data"
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 6
        toc: true
        toc-depth: 3
        embed-resources: true
---

# Introduction

This notebook quantifies unassigned transcripts in the ProSeg-segmented Xenium
cohort. The first analysis asks whether any sample has an unusually high
fraction of gene transcripts assigned to `UNASSIGNED`. The second analysis is
cell-centric: for a reproducible 10% sample of annotated cells, it calculates
the percentage of nearby gene transcripts within 50 um of the cell centroid that
are unassigned. The downstream source-profiling analysis adds a region-centric
grid of transcript-supported candidate locations, keeps only locations supported
by nearby segmented cells, and then compares high-unassigned patches against
their own local assigned transcript milieu.

The primary focus is the cerebellar liponeurocytoma samples (`dx == "LIPN"`),
but the abnormality ranking uses all Xenium samples as the background.

```{r}
#| echo: FALSE
suppressPackageStartupMessages({
    library(data.table)
    library(ggplot2)
    library(knitr)
    library(arrow)
    library(readxl)
    library(rcartocolor)
    library(SummarizedExperiment)
    library(SpatialExperiment)
})

set.seed(5555)

# local vs remote
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # Local mac mount.
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/dfci_mount/"
} else if (Sys.getenv("HOME") == "/PHShome/yz762"){
    # ERIS2 mount.
    workdir <- "/data/beroukhim1/"
} else if (Sys.getenv("HOME") == "/home/yz762"){
    # New server mount.
    workdir <- "/mnt/storage/dept/medonc/beroukhim/"
} else {
    # DIPG/UGER fallback.
    workdir <- "/xchip/beroukhimlab/"
}
workdir = normalizePath(path.expand(workdir), mustWork = FALSE)

data_root = file.path(workdir, "youyun/plgg/data")
proseg_root = file.path(data_root, "segmentation/proseg_run_121024")
cache_dir = file.path(data_root, "segmentation/unassigned_transcripts_50um_gene_only")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

radius_um = as.numeric(Sys.getenv("UNASSIGNED_RADIUS_UM", "50"))
cell_sample_fraction = as.numeric(Sys.getenv("UNASSIGNED_CELL_SAMPLE_FRACTION", "0.1"))
local_seed = as.integer(Sys.getenv("UNASSIGNED_CELL_SAMPLE_SEED", "5555"))
tail_fraction = as.numeric(Sys.getenv("UNASSIGNED_TAIL_FRACTION", "0.20"))
if (is.na(tail_fraction) || tail_fraction <= 0 || tail_fraction >= 0.5) {
    stop("UNASSIGNED_TAIL_FRACTION must be between 0 and 0.5.")
}
tail_high_label = "high_unassigned_neighborhood"
tail_low_label = "low_unassigned_neighborhood"
tail_enrichment_q_threshold = as.numeric(Sys.getenv("UNASSIGNED_TAIL_ENRICHMENT_Q", "0.10"))
if (
    is.na(tail_enrichment_q_threshold) ||
        tail_enrichment_q_threshold <= 0 ||
        tail_enrichment_q_threshold > 1
) {
    stop("UNASSIGNED_TAIL_ENRICHMENT_Q must be between 0 and 1.")
}
region_gene_q_threshold = as.numeric(Sys.getenv("UNASSIGNED_REGION_GENE_Q_THRESHOLD", "0.05"))
if (
    is.na(region_gene_q_threshold) ||
        region_gene_q_threshold <= 0 ||
        region_gene_q_threshold > 1
) {
    stop("UNASSIGNED_REGION_GENE_Q_THRESHOLD must be between 0 and 1.")
}
region_gene_min_log2_fc = as.numeric(Sys.getenv("UNASSIGNED_REGION_GENE_MIN_LOG2_FC", "1"))
if (is.na(region_gene_min_log2_fc) || region_gene_min_log2_fc < 0) {
    stop("UNASSIGNED_REGION_GENE_MIN_LOG2_FC must be non-negative.")
}
region_gene_min_detected_regions = as.integer(
    Sys.getenv("UNASSIGNED_REGION_GENE_MIN_DETECTED_REGIONS", "10")
)
if (is.na(region_gene_min_detected_regions) || region_gene_min_detected_regions < 1) {
    stop("UNASSIGNED_REGION_GENE_MIN_DETECTED_REGIONS must be a positive integer.")
}
region_grid_spacing_um = as.numeric(Sys.getenv("UNASSIGNED_REGION_GRID_SPACING_UM", radius_um))
if (is.na(region_grid_spacing_um) || region_grid_spacing_um <= 0) {
    stop("UNASSIGNED_REGION_GRID_SPACING_UM must be positive.")
}
min_region_gene_transcripts = as.integer(Sys.getenv("UNASSIGNED_REGION_MIN_GENE_TRANSCRIPTS", "25"))
if (is.na(min_region_gene_transcripts) || min_region_gene_transcripts < 1) {
    stop("UNASSIGNED_REGION_MIN_GENE_TRANSCRIPTS must be a positive integer.")
}
min_region_assigned_transcripts = as.integer(Sys.getenv("UNASSIGNED_REGION_MIN_ASSIGNED_TRANSCRIPTS", "10"))
if (is.na(min_region_assigned_transcripts) || min_region_assigned_transcripts < 1) {
    stop("UNASSIGNED_REGION_MIN_ASSIGNED_TRANSCRIPTS must be a positive integer.")
}
min_region_cells_env = Sys.getenv("UNASSIGNED_REGION_MIN_CELLS", "3")
min_region_cells = suppressWarnings(as.integer(min_region_cells_env))
if (
    is.na(min_region_cells) ||
        min_region_cells < 1 ||
        !grepl("^[0-9]+$", min_region_cells_env)
) {
    stop("UNASSIGNED_REGION_MIN_CELLS must be a positive integer.")
}
region_location_sample_n = as.integer(Sys.getenv("UNASSIGNED_REGION_SAMPLE_N", "5000"))
if (is.na(region_location_sample_n) || region_location_sample_n < 1) {
    stop("UNASSIGNED_REGION_SAMPLE_N must be a positive integer.")
}
tail_fraction_label = gsub(
    "\\.",
    "p",
    format(tail_fraction * 100, trim = TRUE, scientific = FALSE)
)
region_grid_spacing_label = gsub(
    "\\.",
    "p",
    format(region_grid_spacing_um, trim = TRUE, scientific = FALSE)
)

# These environment variables make quick smoke tests possible without editing
# the notebook or overwriting the full-run cache files.
sample_id_filter_env = Sys.getenv("UNASSIGNED_SAMPLE_IDS", "")
sample_id_filter = if (nzchar(sample_id_filter_env)) {
    strsplit(sample_id_filter_env, ",", fixed = TRUE)[[1]]
} else {
    character()
}
max_row_groups_env = Sys.getenv("UNASSIGNED_MAX_ROW_GROUPS", "")
max_row_groups = if (nzchar(max_row_groups_env)) as.integer(max_row_groups_env) else NA_integer_
force_recompute = Sys.getenv("UNASSIGNED_RECOMPUTE", "0") == "1"
is_smoke_run = length(sample_id_filter) > 0 || !is.na(max_row_groups)

cache_suffix = if (is_smoke_run) {
    paste0(
        "_smoke_",
        if (length(sample_id_filter) > 0) paste(sample_id_filter, collapse = "-") else "all",
        "_rg", ifelse(is.na(max_row_groups), "all", max_row_groups)
    )
} else {
    ""
}

sample_summary_cache = file.path(cache_dir, paste0("sample_gene_unassigned_summary", cache_suffix, ".tsv"))
cell_metadata_cache = file.path(cache_dir, "banksy_cell_metadata_proseg_centroids.rds")
sampled_cell_cache = file.path(
    cache_dir,
    paste0(
        "sampled_cells_proseg_centroids_",
        round(cell_sample_fraction * 100), "pct_seed", local_seed, cache_suffix, ".rds"
    )
)
local_summary_cache = file.path(
    cache_dir,
    paste0("cell_local_unassigned_", radius_um, "um_proseg_centroids", cache_suffix, ".tsv")
)
region_location_cache = file.path(
    cache_dir,
    paste0(
        "sampled_locations_",
        radius_um, "um_grid", region_grid_spacing_label, "um_gene_only",
        "_mincells", min_region_cells,
        "_n", region_location_sample_n, "_seed", local_seed,
        cache_suffix, ".tsv"
    )
)
region_local_profile_cache = file.path(
    cache_dir,
    paste0(
        "location_local_gene_profiles_",
        radius_um, "um_tail", tail_fraction_label, "pct_grid",
        region_grid_spacing_label, "um_n", region_location_sample_n,
        "_mincells", min_region_cells, "_seed", local_seed, cache_suffix, ".tsv"
    )
)
region_local_summary_cache = file.path(
    cache_dir,
    paste0(
        "location_local_unassigned_",
        radius_um, "um_tail", tail_fraction_label, "pct_grid",
        region_grid_spacing_label, "um_n", region_location_sample_n,
        "_mincells", min_region_cells, "_seed", local_seed, cache_suffix, ".tsv"
    )
)
```

```{r}
theme_set(theme_minimal(base_size = 25))
```

# Inputs and helper functions

```{r}
xenium_metadata = fread(file.path(data_root, "metadata/Xenium_metadata.csv"))
setnames(xenium_metadata, "tumor location", "tumor_location")
xenium_metadata[, `:=`(
    id = as.character(id),
    file = as.character(file)
)]

proseg_dt = data.table(
    file = list.files(proseg_root, full.names = FALSE, recursive = FALSE)
)[
    , parquet_file := file.path(
        proseg_root, file, paste0(file, "_proseg"), "outs", "transcripts.parquet"
    )
][
    , cells_file := file.path(
        proseg_root, file, paste0(file, "_proseg"), "outs", "cells.parquet"
    )
][
    xenium_metadata,
    on = "file"
][
    !is.na(id)
]

if (length(sample_id_filter) > 0) {
    proseg_dt = proseg_dt[id %in% sample_id_filter]
}

if (nrow(proseg_dt) == 0) {
    stop("No ProSeg transcript parquet files matched the metadata/sample filter.")
}

missing_parquet = proseg_dt[!file.exists(parquet_file)]
if (nrow(missing_parquet) > 0) {
    stop(
        "Missing transcript parquet files for: ",
        paste(missing_parquet$file, collapse = ", ")
    )
}
missing_cells = proseg_dt[!file.exists(cells_file)]
if (nrow(missing_cells) > 0) {
    stop(
        "Missing ProSeg/Xenium cell centroid files for: ",
        paste(missing_cells$file, collapse = ", ")
    )
}

lipn_sample_ids = xenium_metadata[dx == "LIPN", id]
proseg_dt[, focus_group := fifelse(id %in% lipn_sample_ids, "LIPN focus", "Comparator")]

kable(proseg_dt[, .(
    sample_id = id, file, dx, histology = mc, alteration = alt, tumor_location,
    focus_group
)][order(dx, sample_id)])
```

```{r}
schema_names = function(parquet_file) {
    names(ParquetFileReader$create(parquet_file)$GetSchema())
}

column_indices = function(parquet_file, cols) {
    idx = match(cols, schema_names(parquet_file))
    if (any(is.na(idx))) {
        stop(
            "Missing required parquet columns in ", parquet_file, ": ",
            paste(cols[is.na(idx)], collapse = ", ")
        )
    }
    idx - 1L
}

iter_row_groups = function(parquet_file, cols, fun, max_row_groups = NA_integer_) {
    reader = ParquetFileReader$create(parquet_file)
    row_groups = seq.int(0L, reader$num_row_groups - 1L)
    if (!is.na(max_row_groups)) {
        row_groups = head(row_groups, max_row_groups)
    }
    col_idx = column_indices(parquet_file, cols)

    for (row_group in row_groups) {
        batch = as.data.table(as.data.frame(reader$ReadRowGroup(
            row_group,
            column_indices = col_idx
        )))
        fun(batch, row_group)
        rm(batch)
        gc(verbose = FALSE)
    }
    invisible(NULL)
}

add_sample_outlier_scores = function(sample_summary) {
    sample_summary = copy(sample_summary)
    cohort_median = median(sample_summary$prop_gene_unassigned, na.rm = TRUE)
    cohort_mad = mad(sample_summary$prop_gene_unassigned, constant = 1.4826, na.rm = TRUE)
    sample_summary[, robust_z := if (cohort_mad > 0) {
        (prop_gene_unassigned - cohort_median) / cohort_mad
    } else {
        NA_real_
    }]
    sample_summary[, percentile_rank := frank(
        prop_gene_unassigned,
        ties.method = "average"
    ) / .N]
    sample_summary[, outlier_tier := fcase(
        robust_z >= 3, "Strong high outlier",
        robust_z >= 2, "Moderate high outlier",
        default = "Not robust high outlier"
    )]
    sample_summary[]
}
```

# Sample-level unassigned transcript burden

```{r}
summarize_one_sample = function(sample_row) {
    sample_row = as.list(sample_row)
    message("Summarizing gene transcript assignment for sample ", sample_row$id)

    n_gene_transcripts = 0
    n_gene_unassigned = 0

    iter_row_groups(
        parquet_file = sample_row$parquet_file,
        cols = c("cell_id", "is_gene"),
        max_row_groups = max_row_groups,
        fun = function(batch, row_group) {
            gene_idx = !is.na(batch$is_gene) & batch$is_gene
            n_gene_transcripts <<- n_gene_transcripts + sum(gene_idx)
            n_gene_unassigned <<- n_gene_unassigned + sum(
                gene_idx & batch$cell_id == "UNASSIGNED",
                na.rm = TRUE
            )
        }
    )

    data.table(
        sample_id = sample_row$id,
        file = sample_row$file,
        dx = sample_row$dx,
        histology = sample_row$mc,
        alteration = sample_row$alt,
        tumor_location = sample_row$tumor_location,
        focus_group = sample_row$focus_group,
        n_gene_transcripts = n_gene_transcripts,
        n_gene_unassigned = n_gene_unassigned,
        n_gene_assigned = n_gene_transcripts - n_gene_unassigned,
        prop_gene_unassigned = n_gene_unassigned / n_gene_transcripts
    )
}

if (!file.exists(sample_summary_cache) || force_recompute) {
    sample_summary = rbindlist(lapply(seq_len(nrow(proseg_dt)), function(i) {
        summarize_one_sample(proseg_dt[i])
    }))
    sample_summary = add_sample_outlier_scores(sample_summary)
    fwrite(sample_summary, sample_summary_cache, sep = "\t")
} else {
    sample_summary = fread(sample_summary_cache)
}

sample_summary[, sample_id := as.character(sample_id)]
sample_summary = sample_summary[order(-prop_gene_unassigned)]

kable(sample_summary[, .(
    sample_id, dx, histology, tumor_location,
    n_gene_transcripts,
    n_gene_unassigned,
    prop_gene_unassigned = round(prop_gene_unassigned, 4),
    robust_z = round(robust_z, 2),
    percentile_rank = round(percentile_rank, 2),
    outlier_tier
)])
```

```{r}
sample_summary[, sample_id_plot := factor(sample_id, levels = sample_id)]

ggplot(sample_summary, aes(
    x = sample_id_plot,
    y = prop_gene_unassigned,
    fill = dx,
    alpha = focus_group
)) +
    geom_col(color = "grey25", linewidth = 0.2) +
    geom_text(
        aes(label = sprintf("%.1f%%", 100 * prop_gene_unassigned)),
        hjust = -0.1,
        size = 3.2
    ) +
    coord_flip(clip = "off") +
    scale_alpha_manual(values = c("LIPN focus" = 1, "Comparator" = 0.55)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.15))) +
    labs(
        title = "Sample-level gene transcript unassigned fraction",
        x = "Sample",
        y = "Gene transcripts assigned to UNASSIGNED",
        fill = "Diagnosis",
        alpha = ""
    ) +
    theme(legend.position = "bottom")
```

```{r}
ggplot(sample_summary, aes(
    x = dx,
    y = prop_gene_unassigned,
    color = focus_group
)) +
    geom_boxplot(aes(group = dx), color = "grey45", outlier.shape = NA) +
    geom_point(aes(size = n_gene_transcripts), alpha = 0.85) +
    geom_text(aes(label = sample_id), nudge_y = 0.01, size = 3, show.legend = FALSE) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_color_manual(values = c("LIPN focus" = "#D55E00", "Comparator" = "#0072B2")) +
    labs(
        title = "Unassigned gene transcript burden by diagnosis",
        x = "Diagnosis",
        y = "Gene transcripts assigned to UNASSIGNED",
        color = "",
        size = "Gene transcript count"
    ) +
    theme(legend.position = "bottom")
```

# Cell-centric local unassigned transcript burden

```{r}
load_banksy_cell_metadata = function() {
    if (file.exists(cell_metadata_cache) && !force_recompute) {
        return(readRDS(cell_metadata_cache))
    }

    banksy_subset_rds_path = file.path(
        data_root, "Xenium_annotations",
        "banksy_clusters_connected_subset_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds"
    )
    banksy_full_rds_path = file.path(
        data_root, "Xenium_annotations",
        "banksy_clusters_connected_k_geom_15_30_pc_20_lam_0.2_0.8_k_leiden_30_50_res_0.75_1_20241230_232215_annotated.rds"
    )
    banksy_rds_path = if (file.exists(banksy_subset_rds_path)) {
        banksy_subset_rds_path
    } else {
        banksy_full_rds_path
    }
    message("Using BANKSY annotated object: ", banksy_rds_path)

    banksy_embeddings = readRDS(banksy_rds_path)
    cell_res = 1
    cell_clust_name = gsub(
        ".0$", "",
        sprintf("clust_Harmony_BANKSY_lam0.2_k50_res%.1f", cell_res)
    )
    niche_res = 1
    niche_clust_name = gsub(
        ".0$", "",
        sprintf("clust_Harmony_BANKSY_lam0.8_k50_res%.1f", niche_res)
    )

    required_cols = c(
        "cell_id", "sample_id", "histology", "alteration",
        cell_clust_name, niche_clust_name
    )
    missing_cols = setdiff(required_cols, colnames(colData(banksy_embeddings)))
    if (length(missing_cols) > 0) {
        stop("Missing required BANKSY metadata columns: ", paste(missing_cols, collapse = ", "))
    }

    prelim_annotation_res1 = fread(file.path(
        data_root, "xenium_cell_types/banksy_proseg_cell_cluster_annotation.tsv"
    ))
    setnames(prelim_annotation_res1, c(
        "Coarse Cell Type", "Fine Cell Type",
        "Fine Cell Type UMAP", "Fine Cell Type Annotation"
    ), c(
        "coarse_cell_type", "fine_cell_type",
        "fine_cell_type_umap", "fine_cell_type_annotation"
    ))

    coords = as.data.table(spatialCoords(banksy_embeddings))
    setnames(coords, names(coords)[1:2], c("banksy_x", "banksy_y"))
    cd = as.data.table(as.data.frame(colData(banksy_embeddings)))
    cell_id_from_coldata = as.character(cd$cell_id)
    sample_id_from_coldata = as.character(cd$sample_id)
    sample_suffix = paste0("__", sample_id_from_coldata)
    banksy_cell_key_from_coldata = ifelse(
        endsWith(cell_id_from_coldata, sample_suffix),
        cell_id_from_coldata,
        paste0(cell_id_from_coldata, sample_suffix)
    )
    banksy_cell_key_from_rownames = rownames(colData(banksy_embeddings))
    banksy_cell_key = ifelse(
        !is.na(banksy_cell_key_from_rownames) & endsWith(
            banksy_cell_key_from_rownames,
            sample_suffix
        ),
        banksy_cell_key_from_rownames,
        banksy_cell_key_from_coldata
    )

    cell_dt = data.table(
        cell_key = banksy_cell_key,
        cell_id = as.character(cd$cell_id),
        sample_id = as.character(cd$sample_id),
        histology = as.character(cd$histology),
        alteration = as.character(cd$alteration),
        cell_cluster = as.integer(as.character(cd[[cell_clust_name]])),
        niche_type = as.character(cd[[niche_clust_name]]),
        pathology_annotation = if ("pathology_annotation" %in% names(cd)) {
            as.character(cd$pathology_annotation)
        } else {
            NA_character_
        },
        banksy_x = coords$banksy_x,
        banksy_y = coords$banksy_y
    )
    duplicated_banksy_keys = cell_dt[duplicated(cell_key), unique(cell_key)]
    if (length(duplicated_banksy_keys) > 0) {
        stop(
            "Duplicated BANKSY cell_key values. Examples: ",
            paste(head(duplicated_banksy_keys, 10), collapse = ", ")
        )
    }
    cell_dt = merge(
        cell_dt,
        prelim_annotation_res1,
        by.x = "cell_cluster",
        by.y = "cluster",
        all.x = TRUE,
        sort = FALSE
    )
    cell_dt[is.na(coarse_cell_type), coarse_cell_type := "Other"]
    cell_dt[is.na(fine_cell_type), fine_cell_type := "Other"]

    # BANKSY spatial coordinates are offset for cohort-level plotting. The
    # transcript parquet files use per-sample Xenium coordinates, so local
    # transcript counting must use the per-sample cell centroids from the
    # ProSeg/Xenium output.
    proseg_centroids = rbindlist(lapply(seq_len(nrow(proseg_dt)), function(i) {
        centroid_dt = as.data.table(read_parquet(
            proseg_dt$cells_file[i],
            col_select = c("cell_id", "x_centroid", "y_centroid")
        ))
        centroid_dt[, `:=`(
            sample_id = as.character(proseg_dt$id[i]),
            cell_id = as.character(cell_id)
        )]
        centroid_dt[, cell_key := paste0(cell_id, "__", sample_id)]
        centroid_dt
    }), use.names = TRUE, fill = TRUE)
    duplicated_centroid_keys = proseg_centroids[duplicated(cell_key), unique(cell_key)]
    if (length(duplicated_centroid_keys) > 0) {
        stop(
            "Duplicated ProSeg centroid cell_key values. Examples: ",
            paste(head(duplicated_centroid_keys, 10), collapse = ", ")
        )
    }

    cell_dt = merge(
        cell_dt,
        proseg_centroids[, .(
            cell_key,
            x = x_centroid,
            y = y_centroid
        )],
        by = "cell_key",
        all.x = TRUE,
        sort = FALSE
    )
    missing_centroids = cell_dt[is.na(x) | is.na(y), .N]
    if (missing_centroids > 0) {
        missing_examples = cell_dt[is.na(x) | is.na(y), head(cell_key, 10)]
        stop(
            "Missing ProSeg/Xenium centroids for ", missing_centroids,
            " BANKSY cells. Examples: ",
            paste(missing_examples, collapse = ", "),
            ". Cannot compare cells and transcripts in the same coordinate frame."
        )
    }

    saveRDS(cell_dt, cell_metadata_cache)
    cell_dt
}

cell_metadata = load_banksy_cell_metadata()
if (length(sample_id_filter) > 0) {
    cell_metadata = cell_metadata[sample_id %in% sample_id_filter]
}

kable(cell_metadata[, .(
    n_cells = .N,
    n_coarse_cell_types = uniqueN(coarse_cell_type),
    n_niche_types = uniqueN(niche_type)
), .(sample_id, histology)][order(sample_id)])
```

```{r}
sample_cells_for_local_analysis = function(cell_dt) {
    if (file.exists(sampled_cell_cache) && !force_recompute) {
        return(readRDS(sampled_cell_cache))
    }

    set.seed(local_seed)
    sampled = cell_dt[
        ,
        {
            n_stratum = .N
            n_keep = max(1L, ceiling(n_stratum * cell_sample_fraction))
            keep_idx = sample.int(n_stratum, n_keep)
            .SD[keep_idx][, `:=`(
                stratum_n_cells = n_stratum,
                stratum_sampled_cells = n_keep,
                sampling_weight = n_stratum / n_keep
            )]
        },
        by = .(sample_id, coarse_cell_type)
    ]
    sampled[, local_cell_index := seq_len(.N)]
    saveRDS(sampled, sampled_cell_cache)
    sampled
}

sampled_cells = sample_cells_for_local_analysis(cell_metadata)

kable(sampled_cells[, .(
    sampled_cells = .N,
    represented_cells = sum(sampling_weight)
), .(sample_id, coarse_cell_type)][order(sample_id, coarse_cell_type)])
```

```{r}
count_local_unassigned_for_sample = function(sample_row, sampled_cells, radius_um) {
    sample_row = as.list(sample_row)
    sample_cells = copy(sampled_cells[sample_id == sample_row$id])
    if (nrow(sample_cells) == 0) {
        return(data.table())
    }

    message(
        "Counting local gene transcript assignment for sample ",
        sample_row$id, " with ", nrow(sample_cells), " sampled cells"
    )

    sample_cells[, cell_index := seq_len(.N)]
    sample_cells[, `:=`(
        cell_x = x,
        cell_y = y,
        gx = floor(x / radius_um),
        gy = floor(y / radius_um)
    )]
    cell_grid = sample_cells[, .(gx, gy, cell_index, cell_x, cell_y)]

    n_total = integer(nrow(sample_cells))
    n_unassigned = integer(nrow(sample_cells))
    radius_sq = radius_um^2
    bbox = list(
        xmin = min(sample_cells$cell_x) - radius_um,
        xmax = max(sample_cells$cell_x) + radius_um,
        ymin = min(sample_cells$cell_y) - radius_um,
        ymax = max(sample_cells$cell_y) + radius_um
    )
    offsets = CJ(dx = -1L:1L, dy = -1L:1L)

    iter_row_groups(
        parquet_file = sample_row$parquet_file,
        cols = c("cell_id", "x_location", "y_location", "is_gene"),
        max_row_groups = max_row_groups,
        fun = function(batch, row_group) {
            # Keep only gene transcript coordinates that could fall in any
            # sampled-cell radius; this avoids building unnecessary grid joins.
            batch = batch[
                !is.na(is_gene) & is_gene &
                    x_location >= bbox$xmin & x_location <= bbox$xmax &
                    y_location >= bbox$ymin & y_location <= bbox$ymax,
                .(
                    transcript_x = x_location,
                    transcript_y = y_location,
                    unassigned = cell_id == "UNASSIGNED"
                )
            ]
            if (nrow(batch) == 0) {
                return(invisible(NULL))
            }

            batch[, transcript_index := seq_len(.N)]
            batch[, `:=`(
                gx = floor(transcript_x / radius_um),
                gy = floor(transcript_y / radius_um)
            )]

            neighbor_bins = batch[rep(seq_len(.N), each = nrow(offsets))]
            neighbor_bins[, `:=`(
                gx = gx + rep(offsets$dx, times = nrow(batch)),
                gy = gy + rep(offsets$dy, times = nrow(batch))
            )]

            candidates = cell_grid[
                neighbor_bins,
                on = .(gx, gy),
                allow.cartesian = TRUE,
                nomatch = 0L
            ]
            if (nrow(candidates) == 0) {
                return(invisible(NULL))
            }

            candidates = candidates[
                (transcript_x - cell_x)^2 + (transcript_y - cell_y)^2 <= radius_sq
            ]
            if (nrow(candidates) == 0) {
                return(invisible(NULL))
            }

            counts = candidates[, .(
                n_gene_transcripts_50um = .N,
                n_gene_unassigned_50um = sum(unassigned, na.rm = TRUE)
            ), by = cell_index]

            n_total[counts$cell_index] <<- n_total[counts$cell_index] +
                counts$n_gene_transcripts_50um
            n_unassigned[counts$cell_index] <<- n_unassigned[counts$cell_index] +
                counts$n_gene_unassigned_50um
            invisible(NULL)
        }
    )

    sample_cells[, `:=`(
        n_gene_transcripts_50um = n_total,
        n_gene_unassigned_50um = n_unassigned,
        prop_gene_unassigned_50um = fifelse(n_total > 0, n_unassigned / n_total, NA_real_)
    )]

    sample_cells[, .(
        sample_id, cell_key, cell_id, x, y,
        histology, alteration, coarse_cell_type, fine_cell_type,
        fine_cell_type_umap, niche_type, pathology_annotation,
        stratum_n_cells, stratum_sampled_cells, sampling_weight,
        n_gene_transcripts_50um, n_gene_unassigned_50um,
        prop_gene_unassigned_50um
    )]
}

if (!file.exists(local_summary_cache) || force_recompute) {
    local_unassigned = rbindlist(lapply(seq_len(nrow(proseg_dt)), function(i) {
        count_local_unassigned_for_sample(proseg_dt[i], sampled_cells, radius_um)
    }), use.names = TRUE, fill = TRUE)
    fwrite(local_unassigned, local_summary_cache, sep = "\t")
} else {
    local_unassigned = fread(local_summary_cache)
}

local_unassigned[, sample_id := as.character(sample_id)]
local_unassigned = local_unassigned[
    !is.na(prop_gene_unassigned_50um) & n_gene_transcripts_50um > 0
]

if (nrow(local_unassigned) == 0) {
    message(
        "No sampled cells had gene transcripts within ", radius_um,
        " um after filtering. Local milieu summaries and plots will be skipped."
    )
}

kable(local_unassigned[, .(
    sampled_cells_with_transcripts = .N,
    median_local_unassigned = round(median(prop_gene_unassigned_50um, na.rm = TRUE), 4),
    aggregate_local_unassigned = round(
        sum(n_gene_unassigned_50um) / sum(n_gene_transcripts_50um),
        4
    )
), .(sample_id, histology)][order(-aggregate_local_unassigned)])
```

```{r}
if (nrow(local_unassigned) > 0) {
    cell_type_local_summary = local_unassigned[, .(
        sampled_cells = .N,
        represented_cells = sum(sampling_weight),
        median_local_unassigned = median(prop_gene_unassigned_50um, na.rm = TRUE),
        iqr_local_unassigned = IQR(prop_gene_unassigned_50um, na.rm = TRUE),
        aggregate_gene_transcripts_50um = sum(n_gene_transcripts_50um),
        aggregate_gene_unassigned_50um = sum(n_gene_unassigned_50um),
        aggregate_local_unassigned = sum(n_gene_unassigned_50um) / sum(n_gene_transcripts_50um)
    ), .(
        focus_group = fifelse(sample_id %in% lipn_sample_ids, "LIPN focus", "Comparator"),
        coarse_cell_type
    )]

    cell_type_local_summary = cell_type_local_summary[order(
        focus_group,
        -aggregate_local_unassigned
    )]
} else {
    cell_type_local_summary = data.table(
        focus_group = character(),
        coarse_cell_type = character(),
        sampled_cells = integer(),
        represented_cells = numeric(),
        median_local_unassigned = numeric(),
        iqr_local_unassigned = numeric(),
        aggregate_local_unassigned = numeric(),
        aggregate_gene_transcripts_50um = integer(),
        aggregate_gene_unassigned_50um = integer()
    )
}

kable(cell_type_local_summary[, .(
    focus_group, coarse_cell_type, sampled_cells,
    represented_cells = round(represented_cells),
    median_local_unassigned = round(median_local_unassigned, 4),
    iqr_local_unassigned = round(iqr_local_unassigned, 4),
    aggregate_local_unassigned = round(aggregate_local_unassigned, 4),
    aggregate_gene_transcripts_50um,
    aggregate_gene_unassigned_50um
)])
```

```{r}
cell_type_levels = sort(unique(local_unassigned$coarse_cell_type))

if (length(cell_type_levels) > 0) {
    safe_native = rcartocolor::cartocolors$Safe
    safe_colors = rcartocolor::carto_pal(
        n = min(length(cell_type_levels), length(safe_native)),
        name = "Safe"
    )

    if (length(cell_type_levels) > length(safe_colors)) {
        safe_colors = grDevices::colorRampPalette(safe_colors)(length(cell_type_levels))
    }

    cell_type_palette = setNames(safe_colors[seq_along(cell_type_levels)], cell_type_levels)
} else {
    cell_type_palette = character()
}
```

```{r}
#| fig-height: 12
if (nrow(local_unassigned) > 0) {
    ggplot(local_unassigned, aes(
        x = reorder(coarse_cell_type, prop_gene_unassigned_50um, FUN = median),
        y = prop_gene_unassigned_50um,
        fill = coarse_cell_type
    )) +
        geom_boxplot(outlier.alpha = 0.15, linewidth = 0.2) +
        facet_wrap(
            ~ histology,
            nrow = 2,
            scales = "free_x"
        ) +
        scale_fill_manual(values = cell_type_palette, drop = FALSE) +
        scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
        labs(
            title = paste0("Local unassigned gene transcript burden within ", radius_um, " um"),
            x = "Coarse cell type",
            y = "Nearby gene transcripts assigned to UNASSIGNED",
            fill = "Coarse cell type"
        ) +
        theme(
            axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "none"
        )
} else {
    message("Skipping cell-type local unassigned plot because local_unassigned has no rows.")
}
```

```{r}
#| fig-height: 10
if (nrow(local_unassigned) > 0) {
    ggplot(local_unassigned, aes(
        x = sample_id,
        y = prop_gene_unassigned_50um,
        fill = coarse_cell_type
    )) +
        geom_boxplot(outlier.alpha = 0.1, linewidth = 0.2) +
        facet_wrap(~ histology, scales = "free_x") +
        scale_fill_manual(values = cell_type_palette, drop = FALSE) +
        scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
        guides(fill = guide_legend(nrow = 3, byrow = TRUE)) +
        labs(
            title = paste0("Sample-level cell milieu burden within ", radius_um, " um"),
            x = "Sample",
            y = "Nearby gene transcripts assigned to UNASSIGNED",
            fill = "Coarse cell type"
        ) +
        theme(
            axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "bottom"
        )
} else {
    message("Skipping sample-level local unassigned plot because local_unassigned has no rows.")
}
```

```{r}
#| fig-height: 10
#| fig-width: 12

spatial_local = copy(local_unassigned)

if (nrow(spatial_local) > 0) {
    ggplot(spatial_local, aes(
        x = x,
        y = y,
        color = prop_gene_unassigned_50um
    )) +
        geom_point(size = 0.25, alpha = 0.9) +
        facet_wrap(~ histology + sample_id, ncol = 4) +
        scale_color_viridis_c(
            labels = scales::percent_format(accuracy = 1),
            option = "magma"
        ) +
        coord_equal() +
        labs(
            title = paste0("All sampled cells colored by local ", radius_um, " um unassigned burden"),
            x = "X location",
            y = "Y location",
            color = "Unassigned"
        ) +
        theme(
            axis.text = element_blank(),
            axis.ticks = element_blank(),
            panel.grid = element_blank()
        )
}
```

# Region-centered high-unassigned neighborhoods

The cell-centered plots above use sampled cell centroids as neighborhood anchors.
That asks which segmented cells sit near unassigned transcripts. The region-centered
analysis below instead starts from a deterministic grid of transcript-containing
candidate locations, then keeps only grid locations with nearby segmented-cell
support before counting local transcripts. This makes the tissue-supported patch
the unit of analysis, then asks whether the unassigned transcripts in that patch
resemble the assigned transcript milieu already observed in the same 50 um disk.

```{r}
sample_region_locations_for_sample = function(
    sample_row, grid_spacing_um, cell_metadata, radius_um, min_region_cells
) {
    sample_row = as.list(sample_row)
    message("Sampling transcript-supported grid locations for sample ", sample_row$id)

    occupied_bins = list()
    iter_row_groups(
        parquet_file = sample_row$parquet_file,
        cols = c("x_location", "y_location", "is_gene"),
        max_row_groups = max_row_groups,
        fun = function(batch, row_group) {
            batch = batch[
                !is.na(is_gene) & is_gene,
                .(
                    gx = floor(x_location / grid_spacing_um),
                    gy = floor(y_location / grid_spacing_um)
                )
            ]
            if (nrow(batch) == 0) {
                return(invisible(NULL))
            }

            occupied_bins[[length(occupied_bins) + 1L]] <<- batch[
                ,
                .(seed_gene_transcripts = .N),
                by = .(gx, gy)
            ]
            invisible(NULL)
        }
    )

    if (length(occupied_bins) == 0) {
        return(data.table())
    }

    candidate_locations = rbindlist(occupied_bins, use.names = TRUE)[
        ,
        .(seed_gene_transcripts = sum(seed_gene_transcripts)),
        by = .(gx, gy)
    ][order(gx, gy)]
    candidate_locations[, candidate_grid_locations := .N]

    # Bound the region-centered analysis so notebook renders remain practical.
    # Sampling from transcript-supported grid bins preserves tissue coverage while
    # avoiding a full all-grid neighborhood count.
    if (nrow(candidate_locations) > region_location_sample_n) {
        set.seed(local_seed + as.integer(sample_row$id))
        keep_idx = sample.int(nrow(candidate_locations), region_location_sample_n)
        region_locations = candidate_locations[keep_idx][order(gx, gy)]
    } else {
        region_locations = candidate_locations
    }

    region_locations[, `:=`(
        sample_id = as.character(sample_row$id),
        histology = as.character(sample_row$mc),
        alteration = as.character(sample_row$alt),
        location_id = sprintf("grid_%s_%s", gx, gy),
        x = (gx + 0.5) * grid_spacing_um,
        y = (gy + 0.5) * grid_spacing_um
    )]
    region_locations[, region_key := paste0(location_id, "__", sample_id)]

    # Tissue support is defined from nearby segmented-cell centroids before any
    # expensive transcript-radius counting. This removes transcript-only,
    # out-of-tissue grid bins from the high-unassigned tail definition.
    sample_cells = copy(cell_metadata[sample_id == sample_row$id])
    if (nrow(sample_cells) == 0) {
        stop("No cell centroids available for sample ", sample_row$id)
    }
    region_locations[, location_index := seq_len(.N)]
    region_locations[, sampled_candidate_grid_locations := .N]
    region_locations[, `:=`(
        local_gx = floor(x / radius_um),
        local_gy = floor(y / radius_um)
    )]
    location_grid = region_locations[, .(
        local_gx, local_gy, location_index, location_x = x, location_y = y
    )]

    sample_cells[, `:=`(
        local_gx = floor(x / radius_um),
        local_gy = floor(y / radius_um),
        cell_x = x,
        cell_y = y
    )]
    offsets = CJ(dx = -1L:1L, dy = -1L:1L)
    neighbor_bins = sample_cells[rep(seq_len(.N), each = nrow(offsets))]
    neighbor_bins[, `:=`(
        local_gx = local_gx + rep(offsets$dx, times = nrow(sample_cells)),
        local_gy = local_gy + rep(offsets$dy, times = nrow(sample_cells))
    )]

    nearby_cells = location_grid[
        neighbor_bins,
        on = .(local_gx, local_gy),
        allow.cartesian = TRUE,
        nomatch = 0L
    ][
        (cell_x - location_x)^2 + (cell_y - location_y)^2 <= radius_um^2
    ]
    if (nrow(nearby_cells) > 0) {
        cell_support = nearby_cells[, .(
            n_cells_50um = .N,
            n_coarse_cell_types_50um = uniqueN(coarse_cell_type)
        ), by = location_index]
        region_locations = merge(
            region_locations,
            cell_support,
            by = "location_index",
            all.x = TRUE,
            sort = FALSE
        )
    } else {
        region_locations[, `:=`(
            n_cells_50um = 0L,
            n_coarse_cell_types_50um = 0L
        )]
    }
    region_locations[is.na(n_cells_50um), n_cells_50um := 0L]
    region_locations[
        is.na(n_coarse_cell_types_50um),
        n_coarse_cell_types_50um := 0L
    ]
    region_locations = region_locations[n_cells_50um >= min_region_cells]
    if (nrow(region_locations) == 0) {
        return(data.table())
    }

    region_locations[, .(
        sample_id, region_key, location_id, gx, gy, x, y,
        histology, alteration, seed_gene_transcripts,
        candidate_grid_locations, sampled_candidate_grid_locations,
        n_cells_50um, n_coarse_cell_types_50um
    )]
}

if (!file.exists(region_location_cache) || force_recompute) {
    region_locations = rbindlist(lapply(seq_len(nrow(proseg_dt)), function(i) {
        sample_region_locations_for_sample(
            proseg_dt[i],
            region_grid_spacing_um,
            cell_metadata,
            radius_um,
            min_region_cells
        )
    }), use.names = TRUE, fill = TRUE)
    fwrite(region_locations, region_location_cache, sep = "\t")
} else {
    region_locations = fread(region_location_cache)
}

region_locations[, sample_id := as.character(sample_id)]

kable(region_locations[, .(
    sampled_candidate_grid_locations = unique(sampled_candidate_grid_locations)[1],
    cell_supported_grid_locations = .N,
    candidate_grid_locations = unique(candidate_grid_locations)[1],
    region_sampling_fraction = unique(sampled_candidate_grid_locations)[1] /
        unique(candidate_grid_locations)[1],
    cell_support_retention_fraction = .N /
        unique(sampled_candidate_grid_locations)[1],
    median_cells_50um = as.numeric(median(n_cells_50um)),
    median_coarse_cell_types_50um = as.numeric(median(n_coarse_cell_types_50um)),
    median_seed_gene_transcripts = as.numeric(median(seed_gene_transcripts)),
    grid_spacing_um = region_grid_spacing_um
), by = .(sample_id, histology)][order(sample_id)])
```

```{r}
count_region_unassigned_for_sample = function(sample_row, region_locations, radius_um) {
    sample_row = as.list(sample_row)
    sample_locations = copy(region_locations[sample_id == sample_row$id])
    if (nrow(sample_locations) == 0) {
        return(data.table())
    }

    message(
        "Counting local gene transcript assignment for sample ",
        sample_row$id, " with ", nrow(sample_locations), " grid locations"
    )

    # These locations have already passed the segmented-cell support threshold;
    # transcript counts are only computed for tissue-supported neighborhoods.
    sample_locations[, location_index := seq_len(.N)]
    sample_locations[, `:=`(
        location_x = x,
        location_y = y,
        local_gx = floor(x / radius_um),
        local_gy = floor(y / radius_um)
    )]
    location_grid = sample_locations[, .(
        local_gx, local_gy, location_index, location_x, location_y
    )]

    n_total = integer(nrow(sample_locations))
    n_unassigned = integer(nrow(sample_locations))
    radius_sq = radius_um^2
    bbox = list(
        xmin = min(sample_locations$location_x) - radius_um,
        xmax = max(sample_locations$location_x) + radius_um,
        ymin = min(sample_locations$location_y) - radius_um,
        ymax = max(sample_locations$location_y) + radius_um
    )
    offsets = CJ(dx = -1L:1L, dy = -1L:1L)

    iter_row_groups(
        parquet_file = sample_row$parquet_file,
        cols = c("cell_id", "x_location", "y_location", "is_gene"),
        max_row_groups = max_row_groups,
        fun = function(batch, row_group) {
            # Keep only gene transcripts that can fall in at least one grid disk.
            batch = batch[
                !is.na(is_gene) & is_gene &
                    x_location >= bbox$xmin & x_location <= bbox$xmax &
                    y_location >= bbox$ymin & y_location <= bbox$ymax,
                .(
                    transcript_x = x_location,
                    transcript_y = y_location,
                    unassigned = cell_id == "UNASSIGNED"
                )
            ]
            if (nrow(batch) == 0) {
                return(invisible(NULL))
            }

            batch[, `:=`(
                local_gx = floor(transcript_x / radius_um),
                local_gy = floor(transcript_y / radius_um)
            )]
            neighbor_bins = batch[rep(seq_len(.N), each = nrow(offsets))]
            neighbor_bins[, `:=`(
                local_gx = local_gx + rep(offsets$dx, times = nrow(batch)),
                local_gy = local_gy + rep(offsets$dy, times = nrow(batch))
            )]

            candidates = location_grid[
                neighbor_bins,
                on = .(local_gx, local_gy),
                allow.cartesian = TRUE,
                nomatch = 0L
            ]
            if (nrow(candidates) == 0) {
                return(invisible(NULL))
            }

            candidates = candidates[
                (transcript_x - location_x)^2 +
                    (transcript_y - location_y)^2 <= radius_sq
            ]
            if (nrow(candidates) == 0) {
                return(invisible(NULL))
            }

            counts = candidates[, .(
                n_gene_transcripts_50um = .N,
                n_gene_unassigned_50um = sum(unassigned, na.rm = TRUE)
            ), by = location_index]
            n_total[counts$location_index] <<- n_total[counts$location_index] +
                counts$n_gene_transcripts_50um
            n_unassigned[counts$location_index] <<- n_unassigned[counts$location_index] +
                counts$n_gene_unassigned_50um
            invisible(NULL)
        }
    )

    sample_locations[, `:=`(
        n_gene_transcripts_50um = n_total,
        n_gene_unassigned_50um = n_unassigned,
        n_gene_assigned_50um = n_total - n_unassigned,
        prop_gene_unassigned_50um = fifelse(n_total > 0, n_unassigned / n_total, NA_real_)
    )]
    sample_locations[, .(
        sample_id, region_key, location_id, x, y, gx, gy,
        histology, alteration, seed_gene_transcripts,
        candidate_grid_locations, sampled_candidate_grid_locations,
        n_cells_50um, n_coarse_cell_types_50um,
        n_gene_transcripts_50um, n_gene_unassigned_50um,
        n_gene_assigned_50um, prop_gene_unassigned_50um
    )]
}

if (!file.exists(region_local_summary_cache) || force_recompute) {
    region_local_unassigned = rbindlist(lapply(seq_len(nrow(proseg_dt)), function(i) {
        count_region_unassigned_for_sample(proseg_dt[i], region_locations, radius_um)
    }), use.names = TRUE, fill = TRUE)

    # Tail assignment uses only tissue-supported regions with enough local
    # total and assigned transcript support. Keeping the assigned-transcript
    # floor here prevents all-unassigned locations from defining the high tail.
    region_local_unassigned = region_local_unassigned[
        !is.na(prop_gene_unassigned_50um) &
            n_cells_50um >= min_region_cells &
            n_gene_transcripts_50um >= min_region_gene_transcripts &
            n_gene_assigned_50um >= min_region_assigned_transcripts
    ]
    region_thresholds = region_local_unassigned[, .(
        low_unassigned_threshold = quantile(
            prop_gene_unassigned_50um,
            probs = tail_fraction,
            na.rm = TRUE
        ),
        high_unassigned_threshold = quantile(
            prop_gene_unassigned_50um,
            probs = 1 - tail_fraction,
            na.rm = TRUE
        ),
        n_supported_locations = .N
    ), by = sample_id]
    region_local_unassigned = merge(
        region_local_unassigned,
        region_thresholds,
        by = "sample_id",
        all.x = TRUE,
        sort = FALSE
    )
    region_local_unassigned[, `:=`(
        local_unassigned_rank = frank(
            prop_gene_unassigned_50um,
            ties.method = "average"
        ),
        tail_label = fcase(
            prop_gene_unassigned_50um >= high_unassigned_threshold,
            tail_high_label,
            prop_gene_unassigned_50um <= low_unassigned_threshold,
            tail_low_label,
            default = "middle"
        )
    ), by = sample_id]
    region_local_unassigned[, local_unassigned_percentile := (
        local_unassigned_rank - 0.5
    ) / n_supported_locations]
    fwrite(region_local_unassigned, region_local_summary_cache, sep = "\t")
} else {
    region_local_unassigned = fread(region_local_summary_cache)
}

region_local_unassigned[, sample_id := as.character(sample_id)]

if (nrow(region_local_unassigned) > 0) {
    region_tail_summary = region_local_unassigned[, .(
        sampled_locations = .N,
        median_local_unassigned = median(prop_gene_unassigned_50um, na.rm = TRUE),
        aggregate_local_unassigned = sum(n_gene_unassigned_50um) /
            sum(n_gene_transcripts_50um),
        median_gene_transcripts_50um = as.numeric(median(n_gene_transcripts_50um)),
        median_gene_assigned_50um = as.numeric(median(n_gene_assigned_50um)),
        median_cells_50um = as.numeric(median(n_cells_50um)),
        median_coarse_cell_types_50um = as.numeric(median(n_coarse_cell_types_50um))
    ), by = .(sample_id, histology, tail_label)][
        order(sample_id, factor(tail_label, levels = c(
            tail_high_label, "middle", tail_low_label
        )))
    ]

    kable(region_tail_summary[, .(
        sample_id,
        histology,
        tail_label,
        sampled_locations,
        median_cells_50um = round(median_cells_50um),
        median_coarse_cell_types_50um = round(median_coarse_cell_types_50um),
        median_gene_transcripts_50um = round(median_gene_transcripts_50um),
        median_gene_assigned_50um = round(median_gene_assigned_50um),
        median_local_unassigned = round(median_local_unassigned, 4),
        aggregate_local_unassigned = round(aggregate_local_unassigned, 4)
    )])
} else {
    message("Skipping region tail labels because no supported grid locations are available.")
}
```

```{r}
#| fig-height: 10
#| fig-width: 12
if (nrow(region_local_unassigned) > 0) {
    ggplot(region_local_unassigned, aes(
        x = x,
        y = y,
        color = prop_gene_unassigned_50um
    )) +
        geom_point(size = 0.2, alpha = 0.85) +
        facet_wrap(~ histology + sample_id, ncol = 4) +
        scale_color_viridis_c(
            labels = scales::percent_format(accuracy = 1),
            option = "magma"
        ) +
        coord_equal() +
        labs(
            title = paste0("Grid locations colored by local ", radius_um, " um unassigned burden"),
            x = "X location",
            y = "Y location",
            color = "Unassigned"
        ) +
        theme(
            axis.text = element_blank(),
            axis.ticks = element_blank(),
            panel.grid = element_blank()
        )
}
```

```{r}
count_region_gene_profiles_for_sample = function(
    sample_row, region_local_unassigned, cell_metadata, radius_um
) {
    sample_row = as.list(sample_row)
    sample_locations = copy(region_local_unassigned[
        sample_id == sample_row$id &
            tail_label %in% c(tail_high_label, tail_low_label)
    ])
    if (nrow(sample_locations) == 0) {
        return(data.table())
    }

    message(
        "Counting region-level gene profiles for sample ",
        sample_row$id, " with ", nrow(sample_locations), " high/low-tail locations"
    )

    sample_locations[, location_index := seq_len(.N)]
    sample_locations[, `:=`(
        location_x = x,
        location_y = y,
        local_gx = floor(x / radius_um),
        local_gy = floor(y / radius_um)
    )]
    location_grid = sample_locations[, .(
        local_gx, local_gy, location_index, location_x, location_y
    )]

    transcript_source_map = unique(cell_metadata[
        sample_id == sample_row$id,
        .(
            transcript_cell_id = as.character(cell_id),
            transcript_coarse_cell_type = as.character(coarse_cell_type)
        )
    ])

    radius_sq = radius_um^2
    bbox = list(
        xmin = min(sample_locations$location_x) - radius_um,
        xmax = max(sample_locations$location_x) + radius_um,
        ymin = min(sample_locations$location_y) - radius_um,
        ymax = max(sample_locations$location_y) + radius_um
    )
    offsets = CJ(dx = -1L:1L, dy = -1L:1L)
    counts_list = list()

    iter_row_groups(
        parquet_file = sample_row$parquet_file,
        cols = c("cell_id", "feature_name", "x_location", "y_location", "is_gene"),
        max_row_groups = max_row_groups,
        fun = function(batch, row_group) {
            batch = batch[
                !is.na(is_gene) & is_gene &
                    x_location >= bbox$xmin & x_location <= bbox$xmax &
                    y_location >= bbox$ymin & y_location <= bbox$ymax,
                .(
                    transcript_cell_id = as.character(cell_id),
                    feature_name = as.character(feature_name),
                    transcript_x = x_location,
                    transcript_y = y_location
                )
            ]
            if (nrow(batch) == 0) {
                return(invisible(NULL))
            }

            batch = merge(
                batch,
                transcript_source_map,
                by = "transcript_cell_id",
                all.x = TRUE,
                sort = FALSE
            )
            batch[, assignment_group := fifelse(
                transcript_cell_id == "UNASSIGNED",
                "UNASSIGNED",
                "ASSIGNED"
            )]
            batch[
                assignment_group == "UNASSIGNED",
                transcript_coarse_cell_type := "UNASSIGNED"
            ]
            batch[
                assignment_group == "ASSIGNED" & is.na(transcript_coarse_cell_type),
                transcript_coarse_cell_type := "Assigned unannotated"
            ]

            batch[, `:=`(
                local_gx = floor(transcript_x / radius_um),
                local_gy = floor(transcript_y / radius_um)
            )]
            neighbor_bins = batch[rep(seq_len(.N), each = nrow(offsets))]
            neighbor_bins[, `:=`(
                local_gx = local_gx + rep(offsets$dx, times = nrow(batch)),
                local_gy = local_gy + rep(offsets$dy, times = nrow(batch))
            )]

            candidates = location_grid[
                neighbor_bins,
                on = .(local_gx, local_gy),
                allow.cartesian = TRUE,
                nomatch = 0L
            ]
            if (nrow(candidates) == 0) {
                return(invisible(NULL))
            }

            candidates = candidates[
                (transcript_x - location_x)^2 +
                    (transcript_y - location_y)^2 <= radius_sq
            ]
            if (nrow(candidates) == 0) {
                return(invisible(NULL))
            }

            counts_list[[length(counts_list) + 1L]] <<- candidates[, .(
                n_transcripts = .N
            ), by = .(
                location_index,
                feature_name,
                assignment_group,
                transcript_coarse_cell_type
            )]
            invisible(NULL)
        }
    )

    if (length(counts_list) == 0) {
        return(data.table())
    }

    profile_counts = rbindlist(counts_list, use.names = TRUE)[
        ,
        .(n_transcripts = sum(n_transcripts)),
        by = .(
            location_index, feature_name, assignment_group,
            transcript_coarse_cell_type
        )
    ]

    merge(
        sample_locations[, .(
            location_index, sample_id, region_key, location_id,
            histology, alteration, x, y, tail_label,
            n_cells_50um, n_coarse_cell_types_50um,
            n_gene_transcripts_50um, n_gene_unassigned_50um,
            n_gene_assigned_50um
        )],
        profile_counts,
        by = "location_index",
        all.y = TRUE,
        sort = FALSE
    )[, location_index := NULL][]
}

if (nrow(region_local_unassigned) > 0 &&
        (!file.exists(region_local_profile_cache) || force_recompute)) {
    region_local_profiles = rbindlist(lapply(seq_len(nrow(proseg_dt)), function(i) {
        count_region_gene_profiles_for_sample(
            proseg_dt[i],
            region_local_unassigned,
            cell_metadata,
            radius_um
        )
    }), use.names = TRUE, fill = TRUE)
    fwrite(region_local_profiles, region_local_profile_cache, sep = "\t")
} else if (file.exists(region_local_profile_cache)) {
    region_local_profiles = fread(region_local_profile_cache)
} else {
    region_local_profiles = data.table()
}

if (nrow(region_local_profiles) > 0) {
    region_local_profiles[, sample_id := as.character(sample_id)]
    kable(region_local_profiles[, .(
        profiled_locations = uniqueN(region_key),
        n_gene_profile_rows = .N,
        n_profile_transcripts = sum(n_transcripts),
        n_unassigned_profile_transcripts = sum(
            n_transcripts[assignment_group == "UNASSIGNED"],
            na.rm = TRUE
        )
    ), by = .(sample_id, histology, tail_label)][order(sample_id, tail_label)])
} else {
    message("Skipping region gene profile summaries because no region profile rows are available.")
}
```

```{r}
compute_region_cell_type_enrichment = function(region_local_profiles, pseudocount = 0.5) {
    if (nrow(region_local_profiles) == 0) {
        return(data.table())
    }

    assigned_dt = region_local_profiles[
        assignment_group == "ASSIGNED" &
            tail_label %in% c(tail_high_label, tail_low_label),
        .(
            n_transcripts = sum(n_transcripts),
            n_locations = uniqueN(region_key)
        ),
        by = .(
            sample_id, histology, tail_label,
            coarse_cell_type = transcript_coarse_cell_type
        )
    ]
    if (nrow(assigned_dt) == 0) {
        return(data.table())
    }

    sample_cell_types = unique(assigned_dt[, .(sample_id, histology, coarse_cell_type)])
    tail_labels = data.table(tail_label = c(tail_high_label, tail_low_label))
    sample_cell_types[, cross_join_key := 1L]
    tail_labels[, cross_join_key := 1L]
    enrichment_grid = merge(
        sample_cell_types,
        tail_labels,
        by = "cross_join_key",
        allow.cartesian = TRUE
    )[, cross_join_key := NULL]

    assigned_dt = merge(
        enrichment_grid,
        assigned_dt,
        by = c("sample_id", "histology", "coarse_cell_type", "tail_label"),
        all.x = TRUE,
        sort = FALSE
    )
    assigned_dt[is.na(n_transcripts), n_transcripts := 0L]
    assigned_dt[is.na(n_locations), n_locations := 0L]

    tail_totals = assigned_dt[, .(
        total_assigned_transcripts = sum(n_transcripts)
    ), by = .(sample_id, tail_label)]
    assigned_dt = merge(
        assigned_dt,
        tail_totals,
        by = c("sample_id", "tail_label"),
        all.x = TRUE,
        sort = FALSE
    )

    enrichment_wide = dcast(
        assigned_dt,
        sample_id + histology + coarse_cell_type ~ tail_label,
        value.var = c("n_transcripts", "total_assigned_transcripts"),
        fill = 0
    )
    high_count_col = paste0("n_transcripts_", tail_high_label)
    low_count_col = paste0("n_transcripts_", tail_low_label)
    high_total_col = paste0("total_assigned_transcripts_", tail_high_label)
    low_total_col = paste0("total_assigned_transcripts_", tail_low_label)
    for (missing_col in setdiff(
        c(high_count_col, low_count_col, high_total_col, low_total_col),
        names(enrichment_wide)
    )) {
        enrichment_wide[, (missing_col) := 0]
    }

    n_types = uniqueN(enrichment_wide$coarse_cell_type)
    enrichment_wide[, `:=`(
        high_count = as.integer(get(high_count_col)),
        low_count = as.integer(get(low_count_col)),
        high_total = as.integer(get(high_total_col)),
        low_total = as.integer(get(low_total_col))
    )]
    enrichment_wide[, `:=`(
        high_fraction = (high_count + pseudocount) /
            (high_total + pseudocount * n_types),
        low_fraction = (low_count + pseudocount) /
            (low_total + pseudocount * n_types)
    )]
    enrichment_wide[, log2_high_low_fc := log2(high_fraction / low_fraction)]
    enrichment_wide[, fisher_p_value := mapply(
        function(high_count, high_total, low_count, low_total) {
            fisher.test(matrix(
                c(
                    high_count,
                    max(high_total - high_count, 0),
                    low_count,
                    max(low_total - low_count, 0)
                ),
                nrow = 2,
                byrow = TRUE
            ))$p.value
        },
        high_count,
        high_total,
        low_count,
        low_total
    )]
    enrichment_wide[, fisher_q_value := p.adjust(fisher_p_value, method = "BH"), by = sample_id]
    enrichment_wide[, .(
        sample_id, histology, coarse_cell_type,
        high_count, low_count, high_total, low_total,
        high_fraction, low_fraction, log2_high_low_fc,
        fisher_p_value, fisher_q_value
    )][order(sample_id, -log2_high_low_fc)]
}

region_cell_type_enrichment = compute_region_cell_type_enrichment(region_local_profiles)

if (nrow(region_cell_type_enrichment) > 0) {
    kable(region_cell_type_enrichment[, .(
        sample_id,
        histology,
        coarse_cell_type,
        high_count,
        low_count,
        high_fraction = round(high_fraction, 3),
        low_fraction = round(low_fraction, 3),
        log2_high_low_fc = round(log2_high_low_fc, 2),
        fisher_p_value = signif(fisher_p_value, 3),
        fisher_q_value = signif(fisher_q_value, 3)
    )])
} else {
    message("Skipping region cell-type enrichment because no assigned profile rows are available.")
}
```

```{r}
#| fig-height: 12
#| fig-width: 14
if (nrow(region_cell_type_enrichment) > 0) {
    region_volcano_dt = copy(region_cell_type_enrichment)
    region_volcano_dt[, `:=`(
        neg_log10_fisher_p = -log10(pmax(fisher_p_value, .Machine$double.xmin)),
        tail_assigned_transcripts = high_count + low_count,
        enrichment_status = fcase(
            fisher_q_value <= tail_enrichment_q_threshold & log2_high_low_fc > 0,
            "Enriched",
            fisher_q_value <= tail_enrichment_q_threshold & log2_high_low_fc < 0,
            "Depleted",
            default = "Not significant"
        )
    )]

    ggplot(region_volcano_dt, aes(
        x = log2_high_low_fc,
        y = neg_log10_fisher_p
    )) +
        geom_vline(xintercept = 0, linewidth = 0.3, color = "grey70") +
        geom_point(aes(
            color = enrichment_status,
            size = tail_assigned_transcripts
        ), alpha = 0.85) +
        ggrepel::geom_text_repel(
            aes(label = coarse_cell_type),
            size = 3,
            min.segment.length = 0,
            max.overlaps = Inf,
            show.legend = FALSE
        ) +
        facet_wrap(~ histology + sample_id, ncol = 4, scales = "free_y") +
        scale_color_manual(
            values = c(
                "Enriched" = "#B2182B",
                "Depleted" = "#2166AC",
                "Not significant" = "grey70"
            ),
            drop = FALSE
        ) +
        scale_size_continuous(range = c(1.5, 6), name = "Tail assigned\ntranscripts") +
        labs(
            title = paste0(
                "Region-level assigned cell-type milieu in top vs bottom ",
                round(tail_fraction * 100), "% unassigned neighborhoods"
            ),
            subtitle = paste0(
                "Grid locations, not anchor cells; color uses exploratory FDR <= ",
                tail_enrichment_q_threshold
            ),
            x = "Assigned transcript composition log2 high/low fold change",
            y = "-log10 exploratory Fisher p-value",
            color = ""
        ) +
        theme(
            legend.position = "bottom",
            panel.grid.minor = element_blank()
        )
} else {
    message("Skipping region cell-type enrichment volcano because no enrichment rows are available.")
}
```

# Region-level assigned-vs-unassigned gene composition

This section compares gene composition between unassigned transcripts and the
assigned transcript mixture within the same retained high-unassigned grid
locations. For each retained region, each gene is normalized to its fraction of
the local assigned or unassigned transcript pool, then summarized as
`log2(unassigned_fraction / assigned_fraction)`. This is not cell-level
differential expression; it is a paired within-region transcript-composition
contrast that helps identify genes preferentially represented in the unassigned
pool after using the local assigned pool as the reference. Because neighboring
grid locations overlap, the p/q values are treated as exploratory and plots
emphasize supported effect sizes and marker identity.

```{r}
marker_path = file.path(data_root, "markers/final_manual_markers.csv")
signature_markers = if (file.exists(marker_path)) {
    fread(marker_path)[
        category == "Cell Type" & !is.na(marker) & !is.na(annotation),
        .(
            feature_name = as.character(marker),
            marker_group = as.character(annotation)
        )
    ]
} else {
    data.table(feature_name = character(), marker_group = character())
}
signature_markers = unique(signature_markers)

gene_marker_groups = if (nrow(signature_markers) > 0) {
    signature_markers[, .(
        marker_group = paste(sort(unique(marker_group)), collapse = "; ")
    ), by = feature_name]
} else {
    data.table(feature_name = character(), marker_group = character())
}

compute_region_assigned_unassigned_gene_contrast = function(
    region_local_profiles, pseudocount = 0.5
) {
    if (nrow(region_local_profiles) == 0) {
        return(data.table())
    }

    high_profiles = region_local_profiles[
        tail_label == tail_high_label &
            assignment_group %in% c("ASSIGNED", "UNASSIGNED")
    ]
    if (nrow(high_profiles) == 0) {
        return(data.table())
    }
    high_profiles[, `:=`(
        x = as.numeric(x),
        y = as.numeric(y),
        n_transcripts = as.numeric(n_transcripts),
        n_cells_50um = as.numeric(n_cells_50um),
        n_coarse_cell_types_50um = as.numeric(n_coarse_cell_types_50um),
        n_gene_transcripts_50um = as.numeric(n_gene_transcripts_50um),
        n_gene_unassigned_50um = as.numeric(n_gene_unassigned_50um),
        n_gene_assigned_50um = as.numeric(n_gene_assigned_50um)
    )]

    region_keys = unique(high_profiles[, .(
        sample_id, histology, region_key, location_id, x, y,
        n_cells_50um, n_coarse_cell_types_50um,
        n_gene_transcripts_50um, n_gene_unassigned_50um,
        n_gene_assigned_50um
    )])
    region_keys = region_keys[
        n_gene_assigned_50um >= min_region_assigned_transcripts &
            n_gene_unassigned_50um > 0
    ]
    if (nrow(region_keys) == 0) {
        return(data.table())
    }

    gene_counts = high_profiles[
        region_keys[, .(sample_id, region_key)],
        on = .(sample_id, region_key),
        nomatch = 0L
    ][
        ,
        .(n_transcripts = as.numeric(sum(n_transcripts, na.rm = TRUE))),
        by = .(
            sample_id, histology, region_key, location_id, x, y,
            feature_name, assignment_group
        )
    ]
    count_wide = dcast(
        gene_counts,
        sample_id + histology + region_key + location_id + x + y + feature_name ~
            assignment_group,
        value.var = "n_transcripts",
        fill = 0
    )
    for (missing_col in setdiff(c("ASSIGNED", "UNASSIGNED"), names(count_wide))) {
        count_wide[, (missing_col) := 0L]
    }

    observed_genes = unique(high_profiles[, .(sample_id, histology, feature_name)])
    region_gene_grid = merge(
        region_keys[, .(
            sample_id, histology, region_key, location_id, x, y,
            n_cells_50um, n_coarse_cell_types_50um,
            n_gene_unassigned_50um, n_gene_assigned_50um
        )],
        observed_genes,
        by = c("sample_id", "histology"),
        allow.cartesian = TRUE
    )
    region_gene_grid = merge(
        region_gene_grid,
        count_wide,
        by = c("sample_id", "histology", "region_key", "location_id", "x", "y", "feature_name"),
        all.x = TRUE,
        sort = FALSE
    )
    region_gene_grid[is.na(ASSIGNED), ASSIGNED := 0L]
    region_gene_grid[is.na(UNASSIGNED), UNASSIGNED := 0L]
    region_gene_grid[, n_observed_genes := uniqueN(feature_name), by = .(sample_id, histology)]
    # Normalize each gene to its region-level assigned and unassigned transcript
    # pools before forming the paired composition contrast.
    region_gene_grid[, `:=`(
        assigned_fraction = (ASSIGNED + pseudocount) /
            (n_gene_assigned_50um + pseudocount * n_observed_genes),
        unassigned_fraction = (UNASSIGNED + pseudocount) /
            (n_gene_unassigned_50um + pseudocount * n_observed_genes)
    )]
    region_gene_grid[, log2_unassigned_assigned_fc := log2(
        unassigned_fraction / assigned_fraction
    )]

    contrast_dt = region_gene_grid[
        ,
        {
            values = log2_unassigned_assigned_fc[
                is.finite(log2_unassigned_assigned_fc)
            ]
            p_value = if (length(values) > 1 && any(values != 0)) {
                tryCatch(
                    wilcox.test(values, mu = 0, exact = FALSE)$p.value,
                    error = function(e) NA_real_
                )
            } else {
                NA_real_
            }

            .(
                n_regions_tested = as.numeric(uniqueN(region_key)),
                n_regions_unassigned_detected = as.numeric(sum(UNASSIGNED > 0, na.rm = TRUE)),
                n_regions_assigned_detected = as.numeric(sum(ASSIGNED > 0, na.rm = TRUE)),
                unassigned_count = as.numeric(sum(UNASSIGNED, na.rm = TRUE)),
                assigned_count = as.numeric(sum(ASSIGNED, na.rm = TRUE)),
                median_region_unassigned_transcripts = median(
                    n_gene_unassigned_50um,
                    na.rm = TRUE
                ),
                median_region_assigned_transcripts = median(
                    n_gene_assigned_50um,
                    na.rm = TRUE
                ),
                median_log2_unassigned_assigned_fc = median(values, na.rm = TRUE),
                mean_log2_unassigned_assigned_fc = mean(values, na.rm = TRUE),
                median_unassigned_fraction = median(unassigned_fraction, na.rm = TRUE),
                median_assigned_fraction = median(assigned_fraction, na.rm = TRUE),
                p_value = p_value
            )
        },
        by = .(sample_id, histology, feature_name)
    ]
    contrast_dt[, q_value := p.adjust(p_value, method = "BH"), by = sample_id]
    contrast_dt[, `:=`(
        # Gene-composition interpretation uses stricter per-sample criteria than
        # the broader tail-enrichment screen because this plot highlights genes,
        # not region or cell-type summaries.
        passes_q_only = !is.na(q_value) & q_value < region_gene_q_threshold,
        passes_unassigned_region_support =
            n_regions_unassigned_detected >= region_gene_min_detected_regions,
        passes_unassigned_effect_threshold =
            median_log2_unassigned_assigned_fc > region_gene_min_log2_fc
    )]
    contrast_dt[, passes_supported_unassigned_enriched := (
        passes_q_only &
            passes_unassigned_effect_threshold &
            passes_unassigned_region_support
    )]
    contrast_dt[, passes_interpretation_threshold := passes_supported_unassigned_enriched]
    contrast_dt[order(sample_id, -median_log2_unassigned_assigned_fc)]
}

region_gene_composition_contrast = compute_region_assigned_unassigned_gene_contrast(
    region_local_profiles
)

if (nrow(region_gene_composition_contrast) > 0) {
    region_gene_contrast_diagnostics = region_gene_composition_contrast[, .(
        n_tested_high_unassigned_regions = unique(n_regions_tested)[1],
        median_high_region_unassigned_transcripts = unique(
            median_region_unassigned_transcripts
        )[1],
        median_high_region_assigned_transcripts = unique(
            median_region_assigned_transcripts
        )[1],
        n_genes_tested = .N,
        n_genes_passing_q_only = sum(passes_q_only, na.rm = TRUE),
        n_genes_passing_q_and_lfc = sum(
            passes_q_only & passes_unassigned_effect_threshold,
            na.rm = TRUE
        ),
        n_genes_supported_unassigned_enriched = sum(
            passes_supported_unassigned_enriched,
            na.rm = TRUE
        )
    ), by = .(sample_id, histology)][order(sample_id)]

    kable(region_gene_contrast_diagnostics[, .(
        sample_id,
        histology,
        n_tested_high_unassigned_regions,
        median_high_region_unassigned_transcripts = round(
            median_high_region_unassigned_transcripts
        ),
        median_high_region_assigned_transcripts = round(
            median_high_region_assigned_transcripts
        ),
        n_genes_tested,
        n_genes_passing_q_only,
        n_genes_passing_q_and_lfc,
        n_genes_supported_unassigned_enriched
    )])

    supported_unassigned_genes = region_gene_composition_contrast[
        passes_supported_unassigned_enriched == TRUE
    ][
        order(sample_id, -median_log2_unassigned_assigned_fc)
    ]
    supported_unassigned_genes = merge(
        supported_unassigned_genes,
        gene_marker_groups,
        by = "feature_name",
        all.x = TRUE,
        sort = FALSE
    )
    supported_unassigned_genes[is.na(marker_group), marker_group := "Other"]

    kable(supported_unassigned_genes[
        ,
        .(
            sample_id,
            histology,
            feature_name,
            marker_group,
            n_regions_tested,
            n_regions_unassigned_detected,
            unassigned_count,
            median_log2_fc = round(median_log2_unassigned_assigned_fc, 2),
            p_value = signif(p_value, 3),
            q_value = signif(q_value, 3)
        )
    ])
} else {
    message("Skipping assigned-vs-unassigned gene composition contrast because no high-tail region profiles are available.")
}
```

```{r}
#| fig-height: 12
#| fig-width: 14
if (nrow(region_gene_composition_contrast) > 0) {
    region_gene_volcano_dt = copy(region_gene_composition_contrast)
    region_gene_volcano_dt[, `:=`(
        neg_log10_p = -log10(pmax(p_value, .Machine$double.xmin)),
        detected_region_support = n_regions_unassigned_detected,
        composition_status = fcase(
            passes_supported_unassigned_enriched == TRUE,
            "Supported unassigned-enriched",
            default = "Exploratory/background"
        )
    )]
    label_dt = region_gene_volcano_dt[
        ,
        {
            supported = .SD[
                passes_supported_unassigned_enriched == TRUE
            ][
                order(-median_log2_unassigned_assigned_fc)
            ]
            n_fallback = max(0L, 8L - min(8L, nrow(supported)))
            fallback = .SD[
                passes_supported_unassigned_enriched != TRUE &
                    is.finite(median_log2_unassigned_assigned_fc) &
                    median_log2_unassigned_assigned_fc > 0
            ][
                order(-median_log2_unassigned_assigned_fc)
            ]
            rbindlist(
                list(head(supported, 8L), head(fallback, n_fallback)),
                use.names = TRUE,
                fill = TRUE
            )
        },
        by = .(sample_id, histology)
    ]

    ggplot(region_gene_volcano_dt, aes(
        x = median_log2_unassigned_assigned_fc,
        y = neg_log10_p
    )) +
        geom_vline(xintercept = 0, linewidth = 0.3, color = "grey70") +
        geom_vline(
            xintercept = region_gene_min_log2_fc,
            linewidth = 0.25,
            color = "grey80",
            linetype = "dashed"
        ) +
        geom_point(aes(
            color = composition_status,
            size = detected_region_support
        ), alpha = 0.85) +
        ggrepel::geom_text_repel(
            data = label_dt,
            aes(label = feature_name),
            size = 3,
            min.segment.length = 0,
            max.overlaps = Inf,
            show.legend = FALSE
        ) +
        facet_wrap(~ histology + sample_id, ncol = 4, scales = "free_y") +
        scale_color_manual(
            values = c(
                "Supported unassigned-enriched" = "#B2182B",
                "Exploratory/background" = "grey70"
            ),
            drop = FALSE
        ) +
        scale_y_continuous(
            limits = c(0, NA),
            expand = expansion(mult = c(0, 0.05))
        ) +
        scale_size_continuous(range = c(1.5, 6), name = "Unassigned\nregion\nsupport") +
        labs(
            title = "Gene-composition contrast in high-unassigned regions",
            subtitle = paste0(
                "Color requires q < ", region_gene_q_threshold,
                ", median log2 FC > ", region_gene_min_log2_fc,
                ", and >= ", region_gene_min_detected_regions,
                " high-unassigned regions with that gene detected in the unassigned pool"
            ),
            x = "Median log2(unassigned_fraction / assigned_fraction)",
            y = "-log10 exploratory paired Wilcoxon p-value",
            color = ""
        ) +
        theme(
            legend.position = "bottom",
            panel.grid.minor = element_blank()
        )
} else {
    message("Skipping assigned-vs-unassigned gene volcano because no contrast rows are available.")
}
```

```{r}
#| fig-height: 18
#| fig-width: 12
if (nrow(region_gene_composition_contrast) > 0) {
    histology_sample_counts = unique(
        region_gene_composition_contrast[, .(histology, sample_id)]
    )[
        ,
        .(n_histology_samples = uniqueN(sample_id)),
        by = histology
    ]

    supported_region_gene_dt = region_gene_composition_contrast[
        passes_supported_unassigned_enriched == TRUE
    ]
    histology_gene_meta_dt = if (nrow(supported_region_gene_dt) > 0) {
        supported_region_gene_dt[
            ,
            .(
                n_samples = uniqueN(sample_id),
                min_log2_fc = min(median_log2_unassigned_assigned_fc, na.rm = TRUE),
                median_log2_fc = median(median_log2_unassigned_assigned_fc, na.rm = TRUE),
                mean_log2_fc = mean(median_log2_unassigned_assigned_fc, na.rm = TRUE),
                min_unassigned_detected_regions = min(
                    n_regions_unassigned_detected,
                    na.rm = TRUE
                ),
                combined_p_value = {
                    fisher_stat = -2 * sum(log(pmax(p_value, .Machine$double.xmin)))
                    pchisq(fisher_stat, df = 2 * .N, lower.tail = FALSE)
                }
            ),
            by = .(histology, feature_name)
        ]
    } else {
        data.table(
            histology = character(),
            feature_name = character(),
            n_samples = integer(),
            min_log2_fc = numeric(),
            median_log2_fc = numeric(),
            mean_log2_fc = numeric(),
            min_unassigned_detected_regions = numeric(),
            combined_p_value = numeric()
        )
    }
    histology_gene_meta_dt = merge(
        histology_gene_meta_dt,
        histology_sample_counts,
        by = "histology",
        all.x = TRUE,
        sort = FALSE
    )[
        n_samples == n_histology_samples
    ]
    histology_gene_meta_dt[, histology_meta_q_value := p.adjust(
        combined_p_value,
        method = "BH"
    ), by = histology]

    if (nrow(histology_gene_meta_dt) > 0) {
        histology_gene_meta_dt = merge(
            histology_gene_meta_dt,
            gene_marker_groups,
            by = "feature_name",
            all.x = TRUE,
            sort = FALSE
        )
        histology_gene_meta_dt[is.na(marker_group), marker_group := "Other"]
        marker_group_base_order = unique(c(cell_type_levels, "Neuron", "RGC", "Other"))
        marker_group_rank = function(x) {
            ranks = match(strsplit(x, "; ", fixed = TRUE)[[1]], marker_group_base_order)
            ranks = ranks[!is.na(ranks)]
            if (length(ranks) == 0) {
                length(marker_group_base_order) + 1L
            } else {
                min(ranks)
            }
        }
        marker_group_order_dt = unique(histology_gene_meta_dt[, .(marker_group)])
        marker_group_order_dt[, marker_group_order := vapply(
            marker_group,
            marker_group_rank,
            integer(1)
        )]
        marker_group_order = marker_group_order_dt[
            order(marker_group_order, marker_group),
            marker_group
        ]
        histology_gene_meta_dt[, marker_group := factor(
            marker_group,
            levels = marker_group_order
        )]
        gene_order = histology_gene_meta_dt[, .(
            marker_group = as.character(marker_group[1]),
            marker_group_order = match(as.character(marker_group[1]), marker_group_order),
            median_effect = median(median_log2_fc, na.rm = TRUE)
        ), by = feature_name][
            order(marker_group_order, median_effect, feature_name),
            feature_name
        ]
        histology_order = unique(
            region_gene_composition_contrast[order(histology), histology]
        )

        kable(histology_gene_meta_dt[
            order(histology, marker_group, -median_log2_fc),
            .(
                histology,
                feature_name,
                marker_group,
                n_samples,
                min_log2_fc = round(min_log2_fc, 2),
                median_log2_fc = round(median_log2_fc, 2),
                mean_log2_fc = round(mean_log2_fc, 2),
                min_unassigned_detected_regions,
                combined_p_value = signif(combined_p_value, 3),
                histology_meta_q_value = signif(histology_meta_q_value, 3)
            )
        ])

        ggplot(histology_gene_meta_dt, aes(
            x = factor(histology, levels = histology_order),
            y = factor(feature_name, levels = gene_order),
            fill = median_log2_fc,
            size = min_unassigned_detected_regions
        )) +
            geom_point(shape = 21, color = "grey20", stroke = 0.25, alpha = 0.9) +
            scale_fill_gradient(
                low = "#FDDDBC",
                high = "#B2182B",
                name = "Histology\nmedian\nlog2 FC"
            ) +
            scale_size_continuous(
                range = c(2, 7),
                name = "Minimum\nunassigned\nregions"
            ) +
            facet_grid(marker_group ~ ., scales = "free_y", space = "free_y") +
            labs(
                title = "Histology-level supported unassigned-enriched gene-composition effects",
                subtitle = paste0(
                    "Shown genes pass q < ", region_gene_q_threshold,
                    ", median log2 FC > ", region_gene_min_log2_fc,
                    ", and >= ", region_gene_min_detected_regions,
                    " unassigned-detected high regions in every sample from that histology"
                ),
                x = "Histology",
                y = "Gene"
            ) +
            theme(
                axis.text.x = element_text(angle = 45, hjust = 1),
                legend.position = "bottom",
                panel.grid = element_blank()
            )
    } else {
        message("Skipping histology-level gene meta plot because no gene passes the strict criteria in every sample from a histology.")
    }
} else {
    message("Skipping histology-level gene meta plot because no contrast rows are available.")
}
```

# Cached outputs

```{r}
output_manifest = data.table(
    output = c(
        "Sample-level summary",
        "BANKSY cell metadata cache",
        "Sampled cell cache",
        "Cell-local unassigned summary",
        "Region grid-location cache",
        "Region-local unassigned summary",
        "Region-local gene source profiles"
    ),
    path = c(
        sample_summary_cache,
        cell_metadata_cache,
        sampled_cell_cache,
        local_summary_cache,
        region_location_cache,
        region_local_summary_cache,
        region_local_profile_cache
    ),
    exists = file.exists(c(
        sample_summary_cache,
        cell_metadata_cache,
        sampled_cell_cache,
        local_summary_cache,
        region_location_cache,
        region_local_summary_cache,
        region_local_profile_cache
    ))
)

kable(output_manifest)
```
