---
title: ""
author: "Peter Youyun Zheng"
date: today
date-format: long
format:
    html:
        code-fold: true
        page-layout: full
        fig-width: 12
        fig-height: 6
        toc: true
        toc-depth: 3
        embed-resources: true
---

# Introduction

Using this notebook to run a few runs of QC and determine which cutoff is best. 

```{r}
#| echo: FALSE
library(data.table)
library(ggplot2)
library(knitr)
# local vs UGER
if (Sys.getenv("HOME") %in% c("/Users/youyun", "/Users/youyunzheng")) {
    # in a local mac, the home directory is usuaully at '/Users/[username]'
    workdir <- "~/Documents/HMS/PhD/beroukhimlab/broad_mount/"
} else {
    # in dipg or uger, the home directory is usuaully at '/home/unix/[username]'
    workdir <- "/xchip/beroukhimlab/"
}
```

# Loading Data

```{r}
source(paste0(workdir, "youyun/plgg/code/helpers/spatial_helper.R"))
current_timestamp = format(Sys.time(), "%Y%m%d_%H%M%S")

# LOAD DATA ===========================================================
print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Loading data...'))
total_se = loading_data()

# STAGGER SPATIAL COORDINATES ==========================================
print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Staggering spatial coordinates...'))
total_se_staggered = stagger_spatial_coords(total_se)

# SUBSET to only gene expression and not control probes ===============
print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Subsetting to gene expression...'))
total_se_staggered = total_se_staggered[rowData(total_se_staggered)$feature_type == 'Gene Expression',]

```

# Testing Different Parameters

We are filtering cells on two things:

1.  total transcript count per cell

2.  number of genes detected per cell

We are doing a median absolute deviation inspired by [single cell best practice](https://www.sc-best-practices.org/preprocessing_visualization/quality_control.html). The question is -- how may MADs do we use. 

We will test a couple versions here differing by nmad x threshold x metric. We will make a table here:

| nmad (total) | threshold (total) |  nmad (detected) | threshold (detected) |
|--------------|-------------------|------------------|----------------------|
| 3            | both              | 3                | lower                |
| 3            | both              | 3                | both                 |
| 3.5          | both              | 3.5              | lower                |
| 3.5          | both              | 3.5              | both                 |
| 4            | both              | 4                | lower                |
| 4            | both              | 4                | both                 |
| 3.5          | both              | 3.5              | lower                |
| 3.5          | both              | 3.5              | both                 |
| 5            | both              | 5                | lower                |
| 5            | both              | 5                | both                 |


We will then run UMAP to evaluate whether or not the QC was sufficient.

```{r}
QC_and_normalize = function(spe, nmad_total, threshold_total, nmad_detected, threshold_detected) {
    qcstats <- perCellQCMetrics(spe)
    
    outliers = is_outlier( # outliers for log1p total counts -- upper lower
        log(qcstats$total + 1), nmad = nmad_total, 
        threshold = threshold_total
    ) | is_outlier( # outliers for log1p genes by count -- lower
        log(qcstats$detected + 1), nmad = nmad_detected, 
        threshold = threshold_detected
    )

    print(paste0(
        'Removing ', sum(outliers)/length(outliers)*100,
        '% of cells due to low quality'
    ))
    keep <- !outliers
    spe <- spe[, keep]

    # Normalization to mean library size
    spe <- computeLibraryFactors(spe)
    aname <- "normcounts"
    # log-normalize the counts
    assay(spe, aname) <- normalizeCounts(spe, log = FALSE)
    
    return(spe)
}
```

## No filtering

```{r}
# UMAP
spe = total_se_staggered
spe = computeLibraryFactors(spe)
aname <- "normcounts"
spe = spe[, colData(spe)$sizeFactor > 0]
# log-normalize the counts
assay(spe, aname) <- normalizeCounts(spe)

print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Running UMAP...'))
umap = runUMAP(
    spe, exprs_values = "normcounts", 
    pca = 20, name = 'UMAP 0'
)
plotReducedDim(umap, dimred = 'UMAP 0',point_size = 0.01)
```

## total nmad = 3, threshold = both, detected nmad = 3, threshold = lower

```{r}
spe = QC_and_normalize(
    spe = total_se_staggered, 
    nmad_total = 3, threshold_total = "both", 
    nmad_detected = 3, threshold_detected = "lower"
)

# UMAP
print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Running UMAP...'))
umap = runUMAP(
    spe, exprs_values = "normcounts", 
    pca = 20, name = 'UMAP 1'
)
plotReducedDim(umap, dimred = 'UMAP 1',point_size = 0.01)
```

## total nmad = 3, threshold = both, detected nmad = 3, threshold = both

```{r}
spe = QC_and_normalize(
    spe = total_se_staggered, 
    nmad_total = 3, threshold_total = "both", 
    nmad_detected = 3, threshold_detected = "both"
)

# UMAP
print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Running UMAP...'))
umap = runUMAP(
    spe, exprs_values = "normcounts", 
    pca = 20, name = 'UMAP 2'
)
plotReducedDim(umap, dimred = 'UMAP 2',point_size = 0.01)
```

## total nmad = 4, threshold = both, detected nmad = 4, threshold = lower

```{r}
spe = QC_and_normalize(
    spe = total_se_staggered, 
    nmad_total = 4, threshold_total = "both", 
    nmad_detected = 4, threshold_detected = "lower"
)

# UMAP
print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Running UMAP...'))
umap = runUMAP(
    spe, exprs_values = "normcounts", 
    pca = 20, name = 'UMAP 3'
)
plotReducedDim(umap, dimred = 'UMAP 3',point_size = 0.01)
```

## total nmad = 4, threshold = both, detected nmad = 4, threshold = both

```{r}
spe = QC_and_normalize(
    spe = total_se_staggered, 
    nmad_total = 4, threshold_total = "both", 
    nmad_detected = 4, threshold_detected = "both"
)

# UMAP
print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Running UMAP...'))
umap = runUMAP(
    spe, exprs_values = "normcounts", 
    pca = 20, name = 'UMAP 4'
)
plotReducedDim(umap, dimred = 'UMAP 4',point_size = 0.01)
```

## total nmad = 3.5, threshold = both, detected nmad = 3.5, threshold = lower

```{r}
spe = QC_and_normalize(
    spe = total_se_staggered, 
    nmad_total = 3.5, threshold_total = "both", 
    nmad_detected = 3.5, threshold_detected = "lower"
)

# UMAP
print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Running UMAP...'))
umap = runUMAP(
    spe, exprs_values = "normcounts", 
    pca = 20, name = 'UMAP 5'
)
plotReducedDim(umap, dimred = 'UMAP 5',point_size = 0.01)
```

## total nmad = 3.5, threshold = both, detected nmad = 3.5, threshold = both

```{r}
spe = QC_and_normalize(
    spe = total_se_staggered, 
    nmad_total = 3.5, threshold_total = "both", 
    nmad_detected = 3.5, threshold_detected = "both"
)

# UMAP
print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Running UMAP...'))
umap = runUMAP(
    spe, exprs_values = "normcounts", 
    pca = 20, name = 'UMAP 6'
)
plotReducedDim(umap, dimred = 'UMAP 6',point_size = 0.01)
```

## total nmad = 5, threshold = both, detected nmad = 5, threshold = lower

```{r}
spe = QC_and_normalize(
    spe = total_se_staggered, 
    nmad_total = 5, threshold_total = "both", 
    nmad_detected = 5, threshold_detected = "lower"
)

# UMAP
print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Running UMAP...'))
umap = runUMAP(
    spe, exprs_values = "normcounts", 
    pca = 20, name = 'UMAP 7'
)
plotReducedDim(umap, dimred = 'UMAP 7',point_size = 0.01)
```

## total nmad = 5, threshold = both, detected nmad = 5, threshold = both

```{r}
spe = QC_and_normalize(
    spe = total_se_staggered, 
    nmad_total = 5, threshold_total = "both", 
    nmad_detected = 5, threshold_detected = "both"
)

# UMAP
print(paste0('[',format(Sys.time(), "%Y/%m/%d-%H:%M:%S"),'] | ','Running UMAP...'))
umap = runUMAP(
    spe, exprs_values = "normcounts", 
    pca = 20, name = 'UMAP 8'
)
plotReducedDim(umap, dimred = 'UMAP 8',point_size = 0.01)
```
